---
title: "Semantic Exploration Analysis"
subtitle: "Analysis of Semantic Exploration Bins"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

## Summary

- Semantic distances decrease with larger bin sizes, reflecting expected averaging across longer interaction windows.
- In the field data, AI contributions show slightly higher centroid distances than user contributions, indicating somewhat greater semantic spread.
- No significant user–AI interaction: both follow similar decay rates as bin size increases.
- Starter identity (user vs AI) has minimal effect on exploration patterns.
- In interleaved AI–AI runs, both models show similar behaviour, with one variant displaying faster semantic compression.
- Overall, exploration patterns are structured and nonrandom, with AI in the field dataset often making the larger semantic jumps.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex,          # ANOVA
  here,          # File path management
  emmeans      # Estimated marginal means

)

# Set project root
PROJECT_ROOT <- here()
setwd(PROJECT_ROOT)

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}

# Load Field Data
setwd(PROJECT_ROOT)
df_res_field <- read_parquet("data/TEXT/processed/semantic_exploration_binned.parquet") %>%
  mutate(
    datatype = "Field",
    distance_log = log(distance)
    
  )


```

## 1.2 Descriptive Statistics

```{r descriptive-stats}
desc_stats <- df_res_field %>%
  group_by(agent) %>%
  summarise(
    n = n(),
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),
    mean_log_distance = mean(distance_log, na.rm = TRUE),
    sd_log_distance = sd(distance_log, na.rm = TRUE)
  )

knitr::kable(desc_stats, caption = "Descriptive Statistics for Semantic Exploration by Agent", digits = 3)
```

# 2. Semantic Exploration Analysis

## 2.1 Bin Size vs. Distance Plot

```{r exploration-plot, fig.width=12, fig.height=6}

p_exploration <- df_res_field %>%
  filter(agent != 'interleaved') %>%
  mutate(llm_type = as.factor(llm_type)) %>%
  ggplot(aes(x = k, y = distance_log, color = agent)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = c("user" = "#377EB8", "ai" = "#FF7F00", "User" = "#377EB8", "AI" = "#FF7F00")) +
  labs(
    title = "Bin Size vs. Distance",
    x = "Bin size",
    y = "Log Distance"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_exploration)

ggsave(file.path(GRAPHS_DIR, "exploration_bin_size_distance.png"),
       p_exploration, width = 12, height = 6, dpi = 300)
```

## 2.2 Interleaved Distance Plot

```{r interleaved-plot, fig.width=12, fig.height=6}
p_interleaved <- df_res_field %>%
  filter(agent == 'interleaved') %>%
  mutate(llm_type = as.factor(llm_type)) %>%
  ggplot(aes(x = k, y = distance_log, color = llm_type)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  labs(
    title = "Distance (Interleaved)",
    x = "Bin size",
    y = "Log Distance"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_interleaved)

ggsave(file.path(GRAPHS_DIR, "exploration_interleaved_distance.png"),
       p_interleaved, width = 12, height = 6, dpi = 300)
```

## 2.3 Starter Effect Plot

```{r starter-plot, fig.width=12, fig.height=6}
p_starter <- df_res_field %>%
  filter(agent == 'interleaved') %>%
  mutate(starter = as.factor(starter)) %>%
  ggplot(aes(x = k, y = distance_log, color = starter)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = c("user" = "#A65628", "ai" = "#1B9E77", "User" = "#A65628", "AI" = "#1B9E77")) +
  labs(
    title = "Distance by Starter",
    x = "Bin size",
    y = "Log Distance"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_starter)

ggsave(file.path(GRAPHS_DIR, "exploration_starter_distance.png"),
       p_starter, width = 12, height = 6, dpi = 300)
```

# 3. Statistical Models

## 3.1 Agent Interaction Model

```{r agent-model}
lm_agent <- df_res_field %>%
  filter(agent != "interleaved") %>%
  mutate(
    # flatten list-column to character
    conversation_id = map_chr(conversation_id, as.character),
    # and treat it as a grouping factor
    conversation_id = factor(conversation_id)
  )

lm_field <- lmer(distance_log ~ k * agent + (1 | conversation_id), data = lm_agent)

cat("=== Agent Interaction Model ===\n")
print(summary(lm_field)$coefficients)
```

## 3.2 LLM Type Interaction Model (Interleaved)

```{r llm-type-model}
lm_llm_type <- df_res_field %>%
  filter(agent == "interleaved") %>%
  mutate(
    # flatten list-column to character
    conversation_id = map_chr(conversation_id, as.character),
    # and treat it as a grouping factor
    conversation_id = factor(conversation_id)
)
lm_interleaved_llm_type <- lmer(distance_log ~ k * factor(llm_type) + (1 | conversation_id), data = lm_llm_type)

cat("=== LLM Type Interaction Model (Interleaved) ===\n")
print(summary(lm_interleaved_llm_type)$coefficients)
```

## 3.3 Starter Interaction Model (Interleaved)

```{r starter-model}
lm_starter <- df_res_field %>%
  filter(agent == "interleaved") %>%
  mutate(
    # flatten list-column to character
    conversation_id = map_chr(conversation_id, as.character),
    # and treat it as a grouping factor
    conversation_id = factor(conversation_id)
)
lm_interleaved_starter <- lmer(distance_log ~ k * factor(starter) + (1 | conversation_id), data = lm_starter)

cat("=== Starter Interaction Model (Interleaved) ===\n")
print(summary(lm_interleaved_starter)$coefficients)
```

# 4. Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

**Analysis complete.** All figures saved to `analysis/figures/` and tables to `analysis/tables/`.
