---
title: "Surface Metrics Analysis"
subtitle: "Analysis of TextDescriptives Metrics (Readability, Complexity, POS)"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---
## Summary

- AI text is more syntactically complex and less readable: longer dependency distances, higher grade-level estimates, more nouns/adjectives/determiners.
- Human text is more variable and conversational, using more pronouns, auxiliaries, adverbs, interjections, and shorter dependency structures.
- Several linguistic features change over turn count (e.g., punctuation, entropy, document length), with differing slopes for user vs AI.
- The two agents maintain distinct stylistic signatures across the interaction, not converging linguistically.
- Overall, AI emphasises structural complexity, while humans contribute stylistic diversity and natural conversational markers.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  here,         # File path management
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  broom.mixed   # Tidy model output
)

# Set project root
PROJECT_ROOT <- here()
setwd(PROJECT_ROOT)

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)

# Define colors
colors <- c("User" = "#377EB8", "AI" = "#FF7F00", "user" = "#377EB8", "ai" = "#FF7F00")
```

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load Full Story Data
df_full <- read_parquet(file.path(PROJECT_ROOT, "data/TEXT/processed/full_story_surface_metrics.parquet")) %>%
  mutate(type = factor(type, levels = c("user", "ai"), labels = c("User", "AI")))

df_full <- df_full %>% select(!c('pos_prop_SYM', 'pos_prop_X'))

# Load Interaction Level Data
df_inter <- read_parquet(file.path(PROJECT_ROOT, "data/TEXT/processed/interaction_level_surface_metrics.parquet")) %>%
  mutate(type = factor(type, levels = c("user", "ai"), labels = c("User", "AI")))

df_inter <- df_inter %>% select(!c('pos_prop_SYM', 'pos_prop_X')) %>% filter(interaction_count > 1)


cat("Full Story Data Dimensions:", dim(df_full), "\n")
cat("Interaction Data Dimensions:", dim(df_inter), "\n")
```

# 2. Full Story Analysis

## 2.1 Readability Metrics

```{r readability-plot, fig.width=12, fig.height=6}
readability_metrics <- c('flesch_kincaid_grade', 'smog', 'gunning_fog')

df_readability <- df_full %>%
  select(type, all_of(readability_metrics)) %>%
  pivot_longer(cols = -type, names_to = "metric", values_to = "value")

p_readability <- ggplot(df_readability, aes(x = value, y = metric, fill = type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = colors) +
  labs(title = "Readability Metrics by Type", x = "Score", y = "") +
  theme(legend.position = "top")

print(p_readability)
ggsave(file.path(GRAPHS_DIR, "surface_readability.png"), p_readability, width = 12, height = 6)
```

### Statistical Models (Readability)

```{r readability-stats}
for (m in readability_metrics) {
  cat(paste0("\n=== Model: ", m, " ===\n"))
  model <- lmer(as.formula(paste(m, "~ type + (1 | conversation_id)")), data = df_full)
  print(summary(model)$coefficients)
}
```

## 2.2 Complexity & Coherence

```{r complexity-coherence-plot, fig.width=12, fig.height=8}
complexity_metrics <- c('first_order_coherence', 'second_order_coherence')

df_comp <- df_full %>%
  select(type, all_of(complexity_metrics)) %>%
  pivot_longer(cols = -type, names_to = "metric", values_to = "value")

p_comp <- ggplot(df_comp, aes(x = value, y = metric, fill = type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = colors) +
  labs(title = "Complexity & Coherence Metrics", x = "Value", y = "") +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  theme(legend.position = "top")

print(p_comp)
ggsave(file.path(GRAPHS_DIR, "surface_complexity_coherence.png"), p_comp, width = 12, height = 8)
```

### Statistical Models (Complexity & Coherence)

```{r complexity-stats}
for (m in complexity_metrics) {
  cat(paste0("\n=== Model: ", m, " ===\n"))
  model <- lmer(as.formula(paste(m, "~ type + (1 | conversation_id)")), data = df_full)
  print(summary(model)$coefficients)
}
```

## 2.3 Dependency Metrics

```{r dependency-plot, fig.width=12, fig.height=6}
dep_metrics <- c('dependency_distance_mean', 'prop_adjacent_dependency_relation_mean')

df_dep <- df_full %>%
  select(type, all_of(dep_metrics)) %>%
  pivot_longer(cols = -type, names_to = "metric", values_to = "value")

p_dep <- ggplot(df_dep, aes(x = value, y = metric, fill = type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = colors) +
  labs(title = "Dependency Metrics", x = "Value", y = "") +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  theme(legend.position = "top")

print(p_dep)
ggsave(file.path(GRAPHS_DIR, "surface_dependency.png"), p_dep, width = 12, height = 6)
```

### Statistical Models (Dependency)

```{r dependency-stats}
for (m in dep_metrics) {
  cat(paste0("\n=== Model: ", m, " ===\n"))
  model <- lmer(as.formula(paste(m, "~ type + (1 | conversation_id)")), data = df_full)
  print(summary(model)$coefficients)
}
```

## 2.4 Part-of-Speech (POS) Analysis

```{r pos-prep}
# POS Mapping
pos_names <- c(
  'ADJ' = 'adjective', 'ADP' = 'adposition', 'ADV' = 'adverb', 'AUX' = 'auxiliary',
  'CCONJ' = 'coordinating conjunction', 'DET' = 'determiner', 'INTJ' = 'interjection',
  'NOUN' = 'noun', 'NUM' = 'numeral', 'PART' = 'particle', 'PRON' = 'pronoun',
  'PROPN' = 'proper noun', 'PUNCT' = 'punctuation', 'SCONJ' = 'subordinating conjunction',
  'SYM' = 'symbol', 'VERB' = 'verb', 'X' = 'other'
)

# Select POS columns
pos_cols <- names(df_full)[grepl("^pos_", names(df_full))]
df_pos <- df_full %>%
  select(type, conversation_id, all_of(pos_cols)) %>%
  pivot_longer(cols = starts_with("pos_"), names_to = "pos_tag", values_to = "prop") %>%
  mutate(
    pos_clean = str_remove(pos_tag, "pos_prop_"),
    pos_name = recode(pos_clean, !!!pos_names),
    pos_name = str_to_title(pos_name)
  )
```

### AI-User Mean Differences

```{r pos-diff-plot, fig.width=10, fig.height=8}
# Calculate stats
pos_stats <- df_pos %>%
  group_by(type, pos_name) %>%
  summarise(
    mean = mean(prop, na.rm = TRUE),
    sd = sd(prop, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  ) %>%
  pivot_wider(names_from = type, values_from = c(mean, sd, n)) %>%
  mutate(
    diff = mean_AI - mean_User,
    se_diff = sqrt((sd_AI^2 / n_AI) + (sd_User^2 / n_User))
  ) %>%
  arrange(desc(diff))

# Plot
p_pos_diff <- ggplot(pos_stats, aes(x = diff, y = reorder(pos_name, diff), fill = diff > 0)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_errorbar(aes(xmin = diff - se_diff, xmax = diff + se_diff), width = 0.2) +
  scale_fill_manual(values = c("TRUE" = "#FF7F00", "FALSE" = "#377EB8"), 
                    labels = c("User Higher", "AI Higher"), name = "Direction") +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  labs(
    title = "AI–User Mean Differences in POS Proportions",
    x = "Difference (AI - User) ± SE",
    y = "POS Tag"
  ) +
  theme(legend.position = "top")

print(p_pos_diff)
ggsave(file.path(GRAPHS_DIR, "surface_pos_differences.png"), p_pos_diff, width = 10, height = 8)
```

### Statistical Models (POS)

```{r pos-stats}
# Run LMER for each POS tag
pos_models <- df_pos %>%
  group_by(pos_name) %>%
  nest() %>%
  mutate(
    model = map(data, ~lmer(prop ~ type + (1 | conversation_id), data = .x)),
    coefs = map(model, ~summary(.x)$coefficients)
  )

# Print coefficients
walk2(pos_models$pos_name, pos_models$coefs, ~{
  cat(paste0("\n=== POS Model: ", .x, " ===\n"))
  print(.y)
})
```

# 3. Interaction Level Analysis

## 3.1 POS Proportions Over Time

```{r pos-interaction-plot, fig.width=16, fig.height=12}
# Filter and reshape interaction data
df_inter_pos <- df_inter %>%
  filter(interaction_count > 1) %>%
  select(type, interaction_count, conversation_id, all_of(pos_cols)) %>%
  pivot_longer(cols = starts_with("pos_"), names_to = "pos_tag", values_to = "prop") %>%
  mutate(
    pos_clean = str_remove(pos_tag, "pos_prop_"),
    pos_name = recode(pos_clean, !!!pos_names),
    pos_name = str_to_title(pos_name)
  )

p_inter_pos <- ggplot(df_inter_pos, aes(x = interaction_count, y = prop, color = type)) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~pos_name, scales = "free_y", ncol = 4) +
  scale_color_manual(values = colors) +
  labs(title = "POS Proportions Across Interactions", x = "Interaction Count", y = "Proportion") +
  theme(legend.position = "top")

print(p_inter_pos)
ggsave(file.path(GRAPHS_DIR, "surface_pos_interaction.png"), p_inter_pos, width = 16, height = 12)
```

### Statistical Models (POS Interaction)

```{r pos-interaction-stats}
# Run LMER with interaction term
pos_inter_models <- df_inter_pos %>%
  group_by(pos_name) %>%
  nest() %>%
  mutate(
    model = map(data, ~lmer(prop ~ interaction_count * type + (1 | conversation_id), data = .x)),
    coefs = map(model, ~summary(.x)$coefficients)
  )

walk2(pos_inter_models$pos_name, pos_inter_models$coefs, ~{
  cat(paste0("\n=== POS Interaction Model: ", .x, " ===\n"))
  print(.y)
})
```

## 3.2 Other Metrics Over Time

```{r other-interaction-plot, fig.width=16, fig.height=8}
other_metrics <- c('flesch_kincaid_grade', 'doc_length', 'entropy', 'n_characters', 'oov_ratio', 'proportion_unique_tokens', 'token_length_std')

df_inter_other <- df_inter %>%
  filter(interaction_count > 1) %>%
  select(type, interaction_count, conversation_id, all_of(other_metrics)) %>%
  pivot_longer(cols = all_of(other_metrics), names_to = "metric", values_to = "value")

p_inter_other <- ggplot(df_inter_other, aes(x = interaction_count, y = value, color = type)) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  scale_color_manual(values = colors) +
  labs(title = "Surface Metrics Across Interactions", x = "Interaction Count", y = "Value") +
  theme(legend.position = "top")

print(p_inter_other)
ggsave(file.path(GRAPHS_DIR, "surface_metrics_interaction.png"), p_inter_other, width = 16, height = 8)
```

### Statistical Models (Other Metrics Interaction)

```{r other-interaction-stats}
other_inter_models <- df_inter_other %>%
  group_by(metric) %>%
  nest() %>%
  mutate(
    model = map(data, ~lmer(value ~ interaction_count * type + (1 | conversation_id), data = .x)),
    coefs = map(model, ~summary(.x)$coefficients)
  )

walk2(other_inter_models$metric, other_inter_models$coefs, ~{
  cat(paste0("\n=== Metric Interaction Model: ", .x, " ===\n"))
  print(.y)
})
```

# 4. Session Information

```{r session-info}
sessionInfo()
```
