---
title: "Valence Alignment in Human-AI Story Co-Creation"
subtitle: "Analysis of Field Data and AI-AI Baseline Comparison"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex,          # ANOVA
  here
)

# Set project root
PROJECT_ROOT <- here()

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

## Summary

- Emotional alignment is asymmetric: the model reliably aligns to the user (User → AI), while humans do not mirror the preceding AI tone.
- Field data shows greater variance in valence and weaker overall coordination compared to the simulated baseline.
- Overall, affective dynamics in the field are primarily model-driven rather than mutually convergent.

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load dyadic sentiment scores

df_field <- read_parquet(here("data/TEXT/processed/dyadic_sentiment_scores.parquet"))
df_simulated <- read_parquet(here("data/Berlin/processed/dyadic_sentiment_scores_simulated.parquet"))

# Display structure
cat("Field data dimensions:", dim(df_field), "\n")
cat("Simulated data dimensions:", dim(df_simulated), "\n")

# Show column names
cat("\nField data columns:\n")
print(names(df_field))

cat("\nSimulated data columns:\n")
print(names(df_simulated))
```

## 1.2 Data Preprocessing

```{r preprocess}
# Preprocess field data
df_field <- df_field %>%
  rename(
    ai_valence = ai_sentiment_projection,
    user_valence = user_sentiment_projection
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing
    timestamp = ymd_hms(timestamp),
    
    # Data source label
    data_source = "Field"
  ) %>%
  # Ensure conversation_id is a character/factor
  mutate(conversation_id = as.character(conversation_id)) %>%
  filter(turn > 1, turn < 10)


# Preprocess simulated data
df_simulated <- df_simulated %>%
  rename(
    ai_valence = ai_sentiment_score,
    user_valence = user_sentiment_score
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing (if available)
    timestamp = if_else(!is.na(timestamp), ymd_hms(timestamp), as.POSIXct(NA)),
    
    # Data source label
    data_source = "Simulated"
  ) %>%
  mutate(conversation_id = as.character(conversation_id))

# Summary statistics
cat("\n=== Field Data Summary ===\n")
df_field %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()

cat("\n=== Simulated Data Summary ===\n")
df_simulated %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()
```

## 1.3 Compute Lagged Variables and Cosine

```{r compute-lags}
# Add lagged AI valence for field data
# Note: lag() within group_by ensures no cross-story contamination
# Turn 1 of each story will have NA for lagged values (correct behavior)

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1),
  ) %>%
  ungroup()


cosine_sim <- function(x, y) {
  if (is.null(x) || is.null(y) || any(is.na(x)) || any(is.na(y))) {
    return(NA_real_)
  }
  return(sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2))))
}

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_embedding_prev = lag(ai_embedding),
    user_embedding_prev = lag(user_embedding)
  ) %>%
  rowwise() %>%
  mutate(
    cosine_sim_same = cosine_sim(user_embedding, ai_embedding),
    cosine_sim_prev = cosine_sim(user_embedding, ai_embedding_prev),
    cosine_sim_user_self = cosine_sim(user_embedding, user_embedding_prev),
    cosine_sim_ai_self = cosine_sim(ai_embedding, ai_embedding_prev)
  ) %>%
  ungroup()

# Add lagged AI valence for simulated data
df_simulated <- df_simulated %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1)
  ) %>%
  ungroup()

# Verify no cross-story contamination
cat("=== Lag Variable Verification ===\n")
cat("Field data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_field$turn == 1 & !is.na(df_field$ai_valence_lag1)), "\n")
cat("Simulated data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_simulated$turn == 1 & !is.na(df_simulated$ai_valence_lag1)), "\n")
```

# 2. Descriptive Statistics

## 2.1 Token Count Statistics

```{r token-stats}
# Field data token statistics
token_stats_field <- df_field %>%
  summarise(
    mean_user_tokens = mean(user_token_count, na.rm = TRUE),
    sd_user_tokens = sd(user_token_count, na.rm = TRUE),
    mean_ai_tokens = mean(ai_token_count, na.rm = TRUE),
    sd_ai_tokens = sd(ai_token_count, na.rm = TRUE),
    mean_token_ratio = mean(token_ratio, na.rm = TRUE)
  )

cat("=== Field Data Token Statistics ===\n")
print(token_stats_field)

# Simulated data token statistics
token_stats_sim <- df_simulated %>%
  summarise(
    mean_user_tokens = mean(user_token_count, na.rm = TRUE),
    sd_user_tokens = sd(user_token_count, na.rm = TRUE),
    mean_ai_tokens = mean(ai_token_count, na.rm = TRUE),
    sd_ai_tokens = sd(ai_token_count, na.rm = TRUE),
    mean_token_ratio = mean(token_ratio, na.rm = TRUE)
  )

cat("\n=== Simulated Data Token Statistics ===\n")
print(token_stats_sim)
```

## 2.2 Valence Distribution Statistics

```{r valence-stats}
# Combined statistics
valence_stats <- bind_rows(
  df_field %>% 
    pivot_longer(cols = c(user_valence, ai_valence),
                 names_to = "agent", values_to = "valence") %>%
    mutate(agent = str_remove(agent, "_valence")) %>%
    group_by(data_source, agent) %>%
    summarise(
      mean_valence = mean(valence, na.rm = TRUE),
      sd_valence = sd(valence, na.rm = TRUE),
      median_valence = median(valence, na.rm = TRUE),
      .groups = "drop"
    ),
  
  df_simulated %>%
    pivot_longer(cols = c(user_valence, ai_valence),
                 names_to = "agent", values_to = "valence") %>%
    mutate(agent = str_remove(agent, "_valence")) %>%
    group_by(data_source, agent) %>%
    summarise(
      mean_valence = mean(valence, na.rm = TRUE),
      sd_valence = sd(valence, na.rm = TRUE),
      median_valence = median(valence, na.rm = TRUE),
      .groups = "drop"
    )
)

print(valence_stats)

# Save to CSV
write_csv(valence_stats, file.path(TABLES_DIR, "valence_descriptive_stats.csv"))
```

# 3. Valence Distribution Analysis

## 3.1 Valence Distribution Difference Test
```{r test-valence-distribution}
df_field_long <- df_field %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  mutate(agent = recode(agent, 
                       user_valence = "User", 
                       ai_valence = "LLM"),
         agent = factor(agent, levels = c("User", "LLM"))
         )



df_field_cosine_long <- df_field %>%
  pivot_longer(cols = c(cosine_sim_same, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_same = "Same",
                            cosine_sim_prev = "Prev"),
         direction = factor(direction, levels = c("Same", "Prev"))
         )

df_sim_long <- df_simulated %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  mutate(agent = recode(agent, 
                       user_valence = "AI 1", 
                       ai_valence = "AI 2"))


# Calculate means for vertical lines
means_field <- df_field_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")

means_sim <- df_sim_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")


valence_difference_mixed_model <- lmer(valence ~ agent + (agent | conversation_id), data=df_field_long)
cat("=== Valence Difference Model Coefficients ===\n")
print(round(summary(valence_difference_mixed_model)$coefficients, 4))

```


## 3.2 Combined Valence Distribution Plot

```{r valence-distribution, fig.width=12, fig.height=6}
# Prepare data for plotting

p_distribution_violin <- df_field_long %>%
  ggplot(aes(x = agent, y = valence, fill = agent)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00"), name = "Agent") +
  scale_color_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00"), name = "Agent") +
  labs(
    title = "Valence Distribution by Agent",
    x = "Agent",
    y = "Valence Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

print(p_distribution_violin)

ggsave(file.path(GRAPHS_DIR, "valence_distribution_comparison_violin.png"), 
       p_distribution_violin, width = 12, height = 6, dpi = 300)

# Field data plot
p_field <- ggplot(df_field_long, aes(x = valence, fill = agent)) +
  geom_density(alpha = 0.4, color = NA) +
  scale_fill_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00"), name = "") +
  scale_color_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00")) +
  labs(title = "Field Data", x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(p_field)

# Simulated data plot
#p_sim <- ggplot(df_sim_long, aes(x = valence, fill = agent)) +
#  geom_density(alpha = 0.4, color = NA) +
#  geom_vline(data = means_sim,
#             aes(xintercept = mean_val, color = agent),
#             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
#  scale_fill_brewer(palette = "Set2", name = "") +
#  scale_color_brewer(palette = "Set2") +
#  labs(title = "Simulated Data", x = NULL, y = NULL) +
#  theme(
#    legend.position = "top",
#    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
#    axis.text = element_text(size = 12)
#  )

# Combine plots
#combined_plot <- (p_field) +
#  plot_annotation(
#    title = "Distribution of Valence Scores: Field vs Simulated Data",
#    theme = theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"))
#  )
#
# Add axis labels using cowplot
#final_valence_dist <- ggdraw() +
#  draw_plot(combined_plot, 0, 0, 1, 1) +
#  draw_label("Valence Score", x = 0.5, y = 0.02, size = 16) +
#  draw_label("Density", x = 0.02, y = 0.5, angle = 90, size = 16)

#print(final_valence_dist)

#ggsave(file.path(GRAPHS_DIR, "valence_distribution_comparison.png"), 
#       final_valence_dist, width = 12, height = 6, dpi = 300)
```

# 4. Valence Alignment Analysis

## 4.1 Per-Story Correlation (Same Turn)

```{r alignment-same-turn}
# Field data: same-turn correlation
cor_field_same <- df_field %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%  # Require at least 3 turns
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )

# Simulated data: same-turn correlation
cor_sim_same <- df_simulated %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated",
    .groups = "drop"
  )

# Combine
cor_same_combined <- bind_rows(cor_field_same, cor_sim_same) %>%
  mutate(alignment_type = "Same Turn")

# Summary
cat("=== Same-Turn Alignment Summary ===\n")
cor_same_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

```

## 4.2 Per-Story Correlation (Previous Turn)

```{r alignment-prev-turn}
# Field data: previous-turn correlation (User response to previous AI)
cor_field_prev <- df_field %>%
  group_by(conversation_id) %>%
  filter(!is.na(ai_valence_lag1)) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )

# Simulated data: previous-turn correlation
cor_sim_prev <- df_simulated %>%
  group_by(conversation_id) %>%
  filter(!is.na(ai_valence_lag1)) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated",
    .groups = "drop"
  )

# Combine
cor_prev_combined <- bind_rows(cor_field_prev, cor_sim_prev) %>%
  mutate(alignment_type = "Previous Turn")

# Summary
cat("=== Previous-Turn Alignment Summary ===\n")
cor_prev_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```

## 4.3.1 All turns story correlation (same turn)
```{r alignment across all turns}
#Violates assumptions since turns are not independent (nested in stories)
cor_field_same_all_turns <- df_field %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field"
  )

cor_sim_same_all_turns <- df_simulated %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated"
  )

cor_same_combined_all_turns <- bind_rows(cor_field_same_all_turns, cor_sim_same_all_turns) %>%
  mutate(alignment_tupe = "Same Turn All Turns")

cat("=== Same-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```
## 4.3.2 All turns story correlation (Previous Turn)

```{r previous-turn-alignment-all-turns}
# Violates independence assumption (turns nested in stories)
# Note: use="complete.obs" automatically excludes turn 1 (where ai_valence_lag1 is NA)
# This ensures user_t1 is never paired with ai from previous story

# Verify exclusion of turn 1
cat("=== Previous-Turn Aggregate Correlation ===\n")
cat("Field: Turn 1 observations (excluded):", sum(df_field$turn == 1), "\n")
cat("Field: Valid lag observations:", sum(!is.na(df_field$ai_valence_lag1)), "\n")
cat("Simulated: Turn 1 observations (excluded):", sum(df_simulated$turn == 1), "\n")
cat("Simulated: Valid lag observations:", sum(!is.na(df_simulated$ai_valence_lag1)), "\n\n")

cor_field_same_all_turns_prev <- df_field %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    n_valid = sum(!is.na(ai_valence_lag1)),
    data_source = "Field"
  )

cor_sim_same_all_turns_prev <- df_simulated %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    n_valid = sum(!is.na(ai_valence_lag1)),
    data_source = "Simulated"
  )

cor_same_combined_all_turns_prev <- bind_rows(cor_field_same_all_turns_prev, cor_sim_same_all_turns_prev) %>%
  mutate(alignment_type = "Previous Turn All Turns")

cat("=== Previous-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns_prev %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_total = sum(n_turns),
    n_valid = sum(n_valid)
  ) %>%
  print()



```
## 4.3.3 Probe Sentiment Narrative Development

```{r alignment-probe-example, fig.width=12, fig.height=6}


p_probe_sentiment <- df_field %>%
  mutate(time = ymd_hms(timestamp)) %>%
  
  group_by(conversation_id) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(turn_index = row_number()) %>%
  ungroup() %>%
  filter(conversation_id %in% unique(conversation_id)[22:24]) %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "source", values_to = "valence") %>%
mutate(
  source = recode(source, "user_valence" = "User", "ai_valence" = "LLM")
) %>%
group_by(conversation_id, source) %>%            # NEW: smooth separately per convo + source
arrange(turn_index, .by_group = TRUE) %>%        # NEW: make sure order is correct
mutate(
  valence_smoothed = rollmean(valence, k = 1,    # CHANGE: k = 3 (or 5 if you want very smooth)
                              fill = NA,
                              align = "center")
) %>%
ungroup() %>%
ggplot(aes(x = turn_index, y = valence_smoothed, color = source)) +
  geom_line(size = 1.2) +                                # thicker lines
  geom_point(aes(y = valence), size = 1.2, alpha = 0.5) +# optional: show raw points faintly
  scale_color_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00")) +
  facet_wrap(~ conversation_id, scales = "free_x") +
  labs(
    title = "Smoothed valence over turns (rolling mean, k = 2)",  # keep this in sync with k
    x = "Turn index",
    y = "Valence",
    color = NULL
  ) +
  theme_minimal(base_size = 14) +                        # nicer base size
  theme(
    legend.position = "bottom",                          # cleaner legend
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0)
  )

print(p_probe_sentiment)
ggsave(file.path(GRAPHS_DIR, "valence_probe_temporal.png"),
       p_probe_sentiment, width = 12, height = 6, dpi = 300)

```

## 4.4.1 Alignment Comparison Plot

```{r alignment-comparison-plot, fig.width=12, fig.height=6}
# Combine all correlations
cor_all <- bind_rows(cor_same_combined, cor_prev_combined) %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("Same Turn", "Previous Turn"))
  )

# Boxplot with jittered points
p_alignment <- cor_all %>%
  filter(data_source == "Field") %>%
  ggplot(aes(x = alignment_type, y = rho, fill = alignment_type)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.4) +
  scale_fill_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00")) +
  scale_color_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00", name = "Alignment Type")) +
  labs(
    title = "Valence Alignment: Field vs Simulated Data",
    x = "Alignment Type",
    y = "Pearson ρ (per story)"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_alignment)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison.png"),
       p_alignment, width = 12, height = 6, dpi = 300)
```

## 4.4.2 Comparison: Per-Story vs Aggregate Correlations

```{r alignment-comparison-aggregate, fig.width=12, fig.height=8}
# Prepare aggregate correlation data for overlay
agg_same <- cor_same_combined_all_turns %>%
  mutate(alignment_type = "Same Turn")

agg_prev <- cor_same_combined_all_turns_prev %>%
  mutate(alignment_type = "Previous Turn") %>%
  select(data_source, rho, alignment_type)

agg_data <- bind_rows(agg_same, agg_prev) %>%
  filter(data_source == "Field") %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("Same Turn", "Previous Turn"))
  )

# Create plot with per-story boxplots and aggregate overlay
p_comparison <- cor_all %>%
  filter(data_source == "Field") %>%
  ggplot(aes(x = alignment_type, y = rho, fill = alignment_type)) +
  # Per-story boxplots
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  # Aggregate correlations as large points
  geom_point(data = agg_data, 
             aes(x = alignment_type, y = rho, fill = alignment_type),
             position = position_dodge(width = 0.75),
             size = 6, shape = 23, color = "black", stroke = 1.5) +
  scale_fill_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00")) +
  scale_color_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00", name = "Alignment Type")) +
  labs(
    title = "Valence Alignment: Per-Story vs Aggregate Correlations",
    subtitle = "Boxplots = per-story correlations; Diamonds = aggregate (all turns pooled)",
    x = "Alignment Type",
    y = "Pearson ρ"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_comparison)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison_with_aggregate.png"),
       p_comparison, width = 12, height = 8, dpi = 300)

# Print summary comparison
cat("\n=== Per-Story vs Aggregate Correlation Comparison ===\n")
comparison_summary <- bind_rows(
  cor_all %>%
    group_by(data_source, alignment_type) %>%
    summarise(
      mean_rho = mean(rho, na.rm = TRUE),
      method = "Per-Story Mean",
      .groups = "drop"
    ),
  agg_data %>%
    mutate(mean_rho = rho, method = "Aggregate All Turns") %>%
    select(data_source, alignment_type, mean_rho, method)
) %>%
  pivot_wider(names_from = method, values_from = mean_rho)

print(comparison_summary)

```



```{r}
# We assume df_field already has:


# 1) Build long-format dataset with both directions in a common structure

# Same-turn: User -> AI
df_same_dir <- df_field %>%
  transmute(
    conversation_id,
    turn,
    alignment_direction = "User_to_AI",
    predictor_valence = user_valence,
    response_valence = ai_valence
  )

# Previous-turn: AI -> User (response to previous AI valence)
df_prev_dir <- df_field %>%
  filter(!is.na(ai_valence_lag1)) %>%   # ensure valid lag values
  transmute(
    conversation_id,
    turn,
    alignment_direction = "AI_to_User",
    predictor_valence = ai_valence_lag1,
    response_valence = user_valence
  )

# Combine into one long dataset
df_alignment_long <- bind_rows(df_same_dir, df_prev_dir) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("User_to_AI", "AI_to_User")
    )
  )

cat("=== Directional Alignment Long-Format Data ===\n")
df_alignment_long %>%
  group_by(alignment_direction) %>%
  summarise(
    n_rows = n(),
    mean_predictor = mean(predictor_valence, na.rm = TRUE),
    mean_response  = mean(response_valence,  na.rm = TRUE),
    n_conversations = n_distinct(conversation_id),
    .groups = "drop"
  ) %>%
  print()
```
```{r}

# standardize within the direction groups to ensure beta = correlation
df_alignment_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z = as.numeric(scale(response_valence))
  ) %>%
  ungroup()



model_direction <- lmer(
  response_z ~ predictor_z * alignment_direction + (1 | conversation_id),
  data = df_alignment_std
)
cat("=== Mixed Effects Model (Standardized): Directional Alignment (User -> AI vs AI -> User) ===\n")
## === Mixed Effects Model: Directional Alignment (User -> AI vs AI -> User) ===
direction_summary <- summary(model_direction)
print(round(direction_summary$coefficients, 4))

```

```{r}
# 3) Visualise directional alignment with separate regression lines

p_direction_alignment <- df_alignment_long %>%
  ggplot(aes(
    x = predictor_valence,
    y = response_valence,
    color = alignment_direction
  )) +
  geom_point(alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    fullrange = TRUE,
    size = 1.2
  ) +
  scale_color_manual(
    values = c("User_to_AI" = "#377EB8", "AI_to_User" = "#FF7F00"),
    name = "Alignment Direction",
    labels = c("User → LLM (same turn)", "LLM → User (previous turn)")
  ) +
  labs(
    title = "Directional Valence Alignment in Field Data",
    x = "Predictor Valence",
    y = "Response Valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_direction_alignment)

ggsave(
  file.path(GRAPHS_DIR, "directional_valence_alignment_field.png"),
  p_direction_alignment,
  width = 10,
  height = 6,
  dpi = 300
)
```


## 4.6 Statistical Tests: Alignment Differences

```{r alignment-stats}
# Test 1: Same-turn alignment difference
t_test_same <- t.test(rho ~ data_source, data = cor_same_combined)
#cat("=== T-test: Same-Turn Alignment ===\n")
#print(t_test_same)

# Test 2: Previous-turn alignment difference
t_test_prev <- t.test(rho ~ data_source, data = cor_prev_combined)
#cat("\n=== T-test: Previous-Turn Alignment ===\n")
#print(t_test_prev)

# Test 3: One-sample t-tests against zero (is alignment significant?)
cat("\n=== One-Sample Tests Against Zero ===\n")
cor_all %>%
  group_by(data_source, alignment_type) %>%
  summarise(
    t_stat = t.test(rho, mu = 0)$statistic,
    p_value = t.test(rho, mu = 0)$p.value,
    mean_rho = mean(rho, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  print()

# Test 4: Paired t-test (Directional Asymmetry within Story)
# We join same and prev correlations by conversation_id to create pairs
cat("\n=== Paired T-Test: Directional Asymmetry (Same vs Prev) ===\n")

# Prepare Field pairs
pairs_field <- inner_join(
  cor_field_same %>% select(conversation_id, rho_same = rho),
  cor_field_prev %>% select(conversation_id, rho_prev = rho),
  by = "conversation_id"
)

# Apply Fisher z-transformation before t-test
t_test_paired_field <- t.test(atanh(pairs_field$rho_same), atanh(pairs_field$rho_prev), paired = TRUE)
cat("Field Data: Paired t-test (Fisher z-transformed Same Turn - Previous Turn)\n")
print(t_test_paired_field)

# Calculate back-transformed means for reporting
mean_z_same <- mean(atanh(pairs_field$rho_same), na.rm = TRUE)
mean_z_prev <- mean(atanh(pairs_field$rho_prev), na.rm = TRUE)

cat("\n--- Interpretation (Back-transformed means) ---\n")
cat("Mean Same-Turn r (User->AI):", round(tanh(mean_z_same), 3), "\n")
cat("Mean Prev-Turn r (AI->User):", round(tanh(mean_z_prev), 3), "\n")

# Prepare Simulated pairs
pairs_sim <- inner_join(
  cor_sim_same %>% select(conversation_id, rho_same = rho),
  cor_sim_prev %>% select(conversation_id, rho_prev = rho),
  by = "conversation_id"
)

# Apply Fisher z-transformation before t-test
t_test_paired_sim <- t.test(atanh(pairs_sim$rho_same), atanh(pairs_sim$rho_prev), paired = TRUE)
cat("\nSimulated Data: Paired t-test (Fisher z-transformed Same Turn - Previous Turn)\n")
print(t_test_paired_sim)

# Save results
alignment_test_results <- tibble(
  test = c("Same-turn: Field vs Simulated", "Previous-turn: Field vs Simulated"),
  t_statistic = c(t_test_same$statistic, t_test_prev$statistic),
  p_value = c(t_test_same$p.value, t_test_prev$p.value),
  mean_field = c(mean(cor_field_same$rho, na.rm = TRUE), 
                 mean(cor_field_prev$rho, na.rm = TRUE)),
  mean_sim = c(mean(cor_sim_same$rho, na.rm = TRUE), 
               mean(cor_sim_prev$rho, na.rm = TRUE))
)

write_csv(alignment_test_results, 
          file.path(TABLES_DIR, "alignment_statistical_tests.csv"))
```


## Cosine Distribution Difference Test

```{r cosine-sim-distribution}


df_field_cosine_long <- df_field %>%
  pivot_longer(cols = c(cosine_sim_same, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_same = "Same Turn",
                            cosine_sim_prev = "Previous Turn"),
         direction = factor(direction, levels = c("Same Turn", "Previous Turn"))
         )
df_field_cosine_long_prev_self <- df_field %>%
  pivot_longer(cols = c(cosine_sim_user_self, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_user_self = "Self",
                            cosine_sim_prev = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor"))
         )
df_field_cosine_long_prev_self_ai <- df_field %>%
  pivot_longer(cols = c(cosine_sim_ai_self, cosine_sim_same),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_ai_self = "Self",
                            cosine_sim_same = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor"))
         )


# Calculate means for vertical lines
means_field <- df_field_cosine_long %>%
  group_by(direction) %>%
  summarise(mean_val = mean(cosine_sim, na.rm = TRUE), .groups = "drop")




cosine_difference_mixed_model_self_user <- lmer(cosine_sim ~ direction + (direction | conversation_id), data=df_field_cosine_long_prev_self)
cat("\n=== Cosine Sim: User (Self vs Interlocutor) ===\n")
print(round(summary(cosine_difference_mixed_model_self_user)$coefficients, 4))

cosine_difference_mixed_model_self_ai <- lmer(cosine_sim ~ direction + (direction | conversation_id), data=df_field_cosine_long_prev_self_ai)
cat("\n=== Cosine Sim: AI (Self vs Interlocutor) ===\n")
print(round(summary(cosine_difference_mixed_model_self_ai)$coefficients, 4))

cosine_difference_mixed_model <- lmer(cosine_sim ~ direction + (direction | conversation_id), data=df_field_cosine_long)
cat("\n=== Cosine Sim: General Direction Comparison ===\n")
print(round(summary(cosine_difference_mixed_model)$coefficients, 4))
```


```{r}
# Create plot with per-story boxplots and aggregate overlay
p_comparison_cosine_violin <- df_field_cosine_long %>%
  ggplot(aes(x = direction, y = cosine_sim, fill = direction)) +
  # Per-story boxplots
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00")) +
  scale_color_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00", name = "direction")) +
  labs(
    title = "Semantic Alignment: Cosine Similarity Between Agents",
    x = "Alignment Type",
    y = "Cosine Similarity"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_comparison_cosine_violin)

ggsave(file.path(GRAPHS_DIR, "cosine_sim_violin.png"),
       p_comparison_cosine_violin, width = 12, height = 8, dpi = 300)

p_field <- ggplot(df_field_cosine_long, aes(x = cosine_sim, fill = direction)) +
  geom_density(alpha = 0.4, color = NA) +
  scale_fill_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00"), name = "") +
  scale_color_manual(values = c("Same Turn" = "#377EB8", "Previous Turn" = "#FF7F00")) +
  labs(title = "Field Data", x = "Cosine Similarity", y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(p_field)
ggsave(
  file.path(GRAPHS_DIR, "directional_cosine_alignment_field.png"),
  p_field,
  width = 10,
  height = 6,
  dpi = 300
)
```
## Combined Alignment Plot

```{r combined-alignment-plot, fig.width=12, fig.height=12}
# Combine Valence Alignment (p_comparison) and Semantic Alignment (p_comparison_cosine_violin)
# Note: p_comparison is defined in chunk 'alignment-comparison-aggregate'
#       p_comparison_cosine_violin is defined in the chunk above

# Remove x-axis title and text from the top plot to avoid redundancy
p_comparison_clean <- p_comparison + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

combined_alignment_plot <- (p_comparison_clean / p_comparison_cosine_violin) +
  plot_annotation(
    title = "Comparison of Valence and Semantic Alignment",
    tag_levels = 'A'
  )

print(combined_alignment_plot)

ggsave(file.path(GRAPHS_DIR, "combined_valence_semantic_alignment.png"),
       combined_alignment_plot, width = 12, height = 12, dpi = 300)
```


```{r cosine-self-vs-interlocutor-plot, fig.width=10, fig.height=6}
# Plot for User: Self vs Interlocutor Cosine Similarity
p_cosine_user_self_vs_interlocutor <- df_field_cosine_long_prev_self %>%
  ggplot(aes(x = direction, y = cosine_sim, fill = direction)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("Self" = "#377EB8", "Interlocutor" = "#FF7F00")) +
  labs(
    title = "User Semantic Alignment: Self vs. Interlocutor",
    subtitle = "Cosine Similarity: User(t) ↔ User(t-1) vs. User(t) ↔ AI(t-1)",
    x = "Alignment Target",
    y = "Cosine Similarity"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40")
  )

print(p_cosine_user_self_vs_interlocutor)

ggsave(file.path(GRAPHS_DIR, "cosine_similarity_user_self_vs_interlocutor.png"),
       p_cosine_user_self_vs_interlocutor, width = 10, height = 6, dpi = 300)
```

## Valence Distance Difference Test

```{r valence-distance-distribution}
# Calculate valence distances
df_field_val <- df_field %>%
  mutate(
    val_dist_same = abs(user_valence - ai_valence),
    val_dist_prev = abs(user_valence - ai_valence_lag1),
    val_dist_user_self = abs(user_valence - user_valence_lag1),
    val_dist_ai_self = abs(ai_valence - ai_valence_lag1)
  )

# User: Self vs Partner (Prev)
df_field_val_long_user <- df_field_val %>%
  pivot_longer(cols = c(val_dist_user_self, val_dist_prev),
               names_to = "direction", values_to = "val_dist") %>%
  mutate(direction = recode(direction,
                            val_dist_user_self = "Self",
                            val_dist_prev = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor")))

# AI: Self vs Partner (Same)
df_field_val_long_ai <- df_field_val %>%
  pivot_longer(cols = c(val_dist_ai_self, val_dist_same),
               names_to = "direction", values_to = "val_dist") %>%
  mutate(direction = recode(direction,
                            val_dist_ai_self = "Self",
                            val_dist_same = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor")))

cat("\n=== Valence Distance Model: User (Self vs Interlocutor) ===\n")
val_model_user <- lmer(val_dist ~ direction + (direction | conversation_id), 
                       data = df_field_val_long_user)
print(round(summary(val_model_user)$coefficients, 4))

cat("\n=== Valence Distance Model: AI (Self vs Interlocutor) ===\n")
val_model_ai <- lmer(val_dist ~ direction + (direction | conversation_id), 
                     data = df_field_val_long_ai)
print(round(summary(val_model_ai)$coefficients, 4))
```

```{r valence-distance-plot, fig.width=12, fig.height=6}
# Combine for plotting
df_val_plot <- bind_rows(
  df_field_val_long_user %>% mutate(agent = "User"),
  df_field_val_long_ai %>% mutate(agent = "LLM")
)

p_val_dist <- df_val_plot %>%
  ggplot(aes(x = direction, y = val_dist, fill = direction)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  facet_wrap(~ agent) +
  scale_fill_manual(values = c("Self" = "#377EB8", "Interlocutor" = "#FF7F00")) +
  labs(
    title = "Valence Distance: Self vs Interlocutor Alignment",
    subtitle = "Lower distance = Higher alignment",
    x = "Alignment Target",
    y = "Absolute Valence Distance"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    strip.text = element_text(size = 14, face = "bold")
  )

print(p_val_dist)

ggsave(file.path(GRAPHS_DIR, "valence_distance_comparison.png"),
       p_val_dist, width = 12, height = 6, dpi = 300)
```


# 5. Summary and Conclusions

## 5.1 Key Findings Summary

```{r summary-table}
# Create summary table
summary_results <- tribble(
  ~Metric, ~Field_Data, ~Simulated_Data,
  "N Stories", n_distinct(df_field$conversation_id), n_distinct(df_simulated$conversation_id),
  "Mean User Valence", mean(df_field$user_valence, na.rm = TRUE), mean(df_simulated$user_valence, na.rm = TRUE),
  "Mean AI Valence", mean(df_field$ai_valence, na.rm = TRUE), mean(df_simulated$ai_valence, na.rm = TRUE),
  "Same-Turn Alignment (ρ)", mean(cor_field_same$rho, na.rm = TRUE), mean(cor_sim_same$rho, na.rm = TRUE),
  "Previous-Turn Alignment (ρ)", mean(cor_field_prev$rho, na.rm = TRUE), mean(cor_sim_prev$rho, na.rm = TRUE)
) %>%
  mutate(
    Field_Data = round(Field_Data, 3),
    Simulated_Data = round(Simulated_Data, 3)
  )

print(summary_results)

write_csv(summary_results, file.path(TABLES_DIR, "analysis_summary.csv"))
```


# 6. Supplementary: Interaction Effects

## 6.1 Starter Interaction Effects

### 6.1.1 Same-Turn Alignment by Starter (Removed for now)
```{r starter-same-turn}

lmer_valence_time <- lmer(valence ~ turn + (1 | conversation_id), data = df_field_long)
print(round(summary(lmer_valence_time)$coefficients, 4))

# Mixed model with starter interaction
#model_starter_same <- lmer(ai_valence ~ user_valence * starter + (1 | conversation_id), 
                           #data = df_field)

#cat("=== Same-Turn Alignment: Starter Interaction ===\n")
#print(round(summary(model_starter_same)$coefficients, 4))
```

### 6.1.2 Previous-Turn Alignment by Starter (Removed for now)
```{r starter-prev-turn}
# Mixed model with starter interaction
#df_field_lag <- df_field %>% filter(!is.na(ai_valence_lag1))

#model_starter_prev <- lmer(user_valence ~ ai_valence_lag1 * starter + (1 | conversation_id), 
#                           data = df_field_lag)

#cat("=== Previous-Turn Alignment: Starter Interaction ===\n")
#print(round(summary(model_starter_prev)$coefficients, 4))
```

### 6.1.3 Per-Story Correlations by Starter (Removed for now)
```{r starter-per-story}
# Same-turn correlations stratified by starter
#cor_starter_same <- df_field %>%
#  filter(!is.na(starter)) %>%
#  group_by(conversation_id, starter) %>%
#  filter(n() >= 3) %>%
#  summarise(
#    rho = cor(user_valence, ai_valence, use = "complete.obs"),
#    n_turns = n(),
#    .groups = "drop"
#  )

#cat("=== Same-Turn Correlations by Starter ===\n")
#cor_starter_same %>%
#  group_by(starter) %>%
#  summarise(
#    mean_rho = mean(rho, na.rm = TRUE),
#    sd_rho = sd(rho, na.rm = TRUE),
#    n_stories = n()
#  ) %>%
#  print()

# Test difference
#t_test_starter <- t.test(rho ~ starter, data = cor_starter_same)
#cat("\nT-test (starter effect on same-turn rho):\n")
#print(t_test_starter)
```


## 6.2 LLM Type Interaction Effects

### 6.2.1 Same-Turn Alignment by LLM Type
```{r llm-same-turn}
# Mixed model with LLM type interaction
df_field_llm <- df_field %>% filter(!is.na(llm_type))

model_llm_same <- lmer(ai_valence ~ user_valence * factor(llm_type) + (1 | conversation_id), 
                       data = df_field_llm)

cat("=== Same-Turn Alignment: LLM Type Interaction ===\n")
print(round(summary(model_llm_same)$coefficients, 4))

p_llm_same <- df_field_llm %>%
  mutate(llm_type = as.factor(llm_type)) %>%
  ggplot(aes(x = user_valence, y = ai_valence, color = llm_type)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  labs(
    title = "Ai Valence Same Turn by LLM (User -> LLM)",
    x = "User Valence",
    y = "Ai Valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_llm_same)

ggsave(file.path(GRAPHS_DIR, "alignment_llm_same.png"),
       p_llm_same, width = 12, height = 6, dpi = 300)

```

### 6.2.2 Previous-Turn Alignment by LLM Type
```{r llm-prev-turn}
# Mixed model with LLM type interaction
df_field_llm_lag <- df_field %>% 
  filter(!is.na(llm_type), !is.na(ai_valence_lag1))

model_llm_prev <- lmer(user_valence ~ ai_valence_lag1 * factor(llm_type) + (1 | conversation_id), 
                       data = df_field_llm_lag)

cat("=== Previous-Turn Alignment: LLM Type Interaction ===\n")
print(round(summary(model_llm_prev)$coefficients, 4))

p_llm_prev <- df_field_llm_lag %>%
  mutate(llm_type = as.factor(llm_type)) %>%
  ggplot(aes(x = ai_valence_lag1, y = user_valence, color = llm_type)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange= TRUE, size = 1.2) +
  labs(
    title = "User Valence by Prev AI (Ai -> User)",
    x = "Previous AI Valence",
    y = "User Valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_llm_prev)

ggsave(file.path(GRAPHS_DIR, "alignment_llm_prev.png"),
       p_llm_prev, width = 12, height = 6, dpi = 300)
```

### 6.2.3 Per-Story Correlations by LLM Type
```{r llm-per-story}
# Same-turn correlations stratified by LLM type
cor_llm_same <- df_field %>%
  filter(!is.na(llm_type)) %>%
  group_by(conversation_id, llm_type) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    .groups = "drop"
  )

cat("=== Same-Turn Correlations by LLM Type ===\n")
cor_llm_same %>%
  group_by(llm_type) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

# ANOVA test
anova_llm <- aov(rho ~ factor(llm_type), data = cor_llm_same)
cat("\nANOVA (LLM type effect on same-turn rho):\n")
print(summary(anova_llm))
```


## 6.3 Valence Distribution by Conditions

### 6.3.1 Valence by Starter
```{r valence-starter}
#df_field %>%
#  filter(!is.na(starter)) %>%
#  pivot_longer(cols = c(user_valence, ai_valence),
#               names_to = "agent", values_to = "valence") %>%
#  group_by(starter, agent) %>%
#  summarise(
#    mean_valence = mean(valence, na.rm = TRUE),
#    sd_valence = sd(valence, na.rm = TRUE),
#    .groups = "drop"
#  ) %>%
#  print()
```

### 6.3.2 Valence by LLM Type
```{r valence-llm}
df_field %>%
  filter(!is.na(llm_type)) %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  group_by(llm_type, agent) %>%
  summarise(
    mean_valence = mean(valence, na.rm = TRUE),
    sd_valence = sd(valence, na.rm = TRUE),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

```{r}

```

## 7. Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

**Analysis complete.** All figures saved to `analysis/figures/` and tables to `analysis/tables/`.
