---
title: "Immediate Emotional Adaptation"
subtitle: "Absolute Marginal Shifts and Three-Turn Response Dynamics"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)

pacman::p_load(
  lme4,
  lmerTest,
  emmeans,
  patchwork,
  stringr
)

theme_set(theme_comparison())
```

# Overview

This short analysis treats the current windowed sentiment marginal as an immediate
exchange-level adaptation signal. The raw signal emitted by
`scripts/04_compute_sentiment.py` is signed:

$$m_i = \operatorname{proj}(E(\text{previous context} + \text{current turn})) -
\operatorname{proj}(E(\text{previous context}))$$

Signed means cancel when averaged. The first analysis therefore uses the
absolute point-level magnitude:

$$A_{\text{shift}} = |m_i|$$

This asks whether a condition has more immediate emotional movement, regardless
of whether the move is positive or negative.

The second analysis combines the two adjacent marginal segments around one
exchange:

$$\text{slot2}_{t-1} \rightarrow \text{slot1}_t \rightarrow \text{slot2}_t$$

Because the first segment is already represented by the first turn's marginal
shift, and the second segment is represented by the second turn's marginal
shift, each complete exchange contributes:

$$\text{incoming}_t = m(\text{slot1}_t \mid \text{slot2}_{t-1})$$

$$\text{response}_t = m(\text{slot2}_t \mid \text{slot1}_t)$$

The two three-turn metrics are:

$$\text{adaptation closure}_t = |\text{incoming}_t| - |\text{response}_t|$$

Positive closure means the response segment is smaller than the incoming
segment, consistent with damping or convergence. Negative closure means the
response segment is larger, consistent with amplification.

$$\text{reversal score}_t = -\text{incoming}_t \cdot \text{response}_t$$

Positive reversal means the response moves in the opposite signed direction from
the incoming segment. Negative reversal means the response continues in the same
signed direction.

# Load Data

```{r load-data}
manual_excluded_story_ids <- c(
  "conv_ed575a06c11d42358e3eeb7826d2f959",
  "conv_a63a08273d0a4704a7638e4cd6850225",
  "conv_0bb56093-3033-4615-bb70-ebfa4135589a",
  "conv_72218cb5-e59c-4c93-a4b9-a057fe5dad80"
)

df_combined <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(!conversation_id %in% manual_excluded_story_ids) %>%
  filter(author_2 != "Network error while generating a response.") %>%
  prepare_condition_factor(reference = "human-ai")

raw_shift_cols <- c(
  "author_1_valence_marginal_window",
  "author_2_valence_marginal_window"
)

if (all(raw_shift_cols %in% names(df_combined))) {
  shift_signal_label <- "raw windowed marginal projection"
  df_shift <- df_combined %>%
    mutate(
      author_1_shift = author_1_valence_marginal_window,
      author_2_shift = author_2_valence_marginal_window
    )
} else {
  shift_signal_label <- "promoted author_*_valence fallback"
  df_shift <- df_combined %>%
    mutate(
      author_1_shift = author_1_valence,
      author_2_shift = author_2_valence
    )
}

df_shift <- df_shift %>%
  mutate(
    condition_id = normalize_condition_id(as.character(condition)),
    chronological_first_slot = case_when(
      condition_id == "human-ai" ~ "author_1",
      starter_side == "author_2" ~ "author_2",
      TRUE ~ "author_1"
    ),
    first_shift = if_else(
      chronological_first_slot == "author_1",
      author_1_shift,
      author_2_shift
    ),
    second_shift = if_else(
      chronological_first_slot == "author_1",
      author_2_shift,
      author_1_shift
    ),
    first_speaker_type = if_else(
      chronological_first_slot == "author_1",
      author_1_type,
      author_2_type
    ),
    second_speaker_type = if_else(
      chronological_first_slot == "author_1",
      author_2_type,
      author_1_type
    )
  )

cat("Shift signal used:", shift_signal_label, "\n")

df_shift %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_exchanges = n(),
    min_turn = min(turn_index, na.rm = TRUE),
    max_turn = max(turn_index, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded complete exchange window by condition")
```

# Absolute Marginal Shift

Each speaker contribution becomes one point. The metric is the absolute value of
the signed windowed marginal shift.

```{r absolute-shift-preprocess}
df_abs_point <- bind_rows(
  df_shift %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      chronological_position = "First",
      speaker_type = first_speaker_type,
      signed_shift = first_shift
    ),
  df_shift %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      chronological_position = "Second",
      speaker_type = second_speaker_type,
      signed_shift = second_shift
    )
) %>%
  mutate(
    chronological_position = factor(chronological_position, levels = c("First", "Second")),
    abs_shift = abs(signed_shift)
  ) %>%
  drop_na(abs_shift, turn_index) %>%
  mutate(turn_scaled = as.numeric(scale(turn_index)))

df_abs_story <- df_abs_point %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_points = n(),
    mean_abs_shift = mean(abs_shift, na.rm = TRUE),
    median_abs_shift = median(abs_shift, na.rm = TRUE),
    .groups = "drop"
  )

df_abs_point %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_points = n(),
    mean_abs_shift = mean(abs_shift),
    sd_abs_shift = sd(abs_shift),
    median_abs_shift = median(abs_shift),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Point-level absolute marginal shift by condition"
  )

df_abs_story %>%
  group_by(condition) %>%
  summarise(
    n_stories = n(),
    mean_story_abs_shift = mean(mean_abs_shift),
    sd_story_abs_shift = sd(mean_abs_shift),
    median_story_abs_shift = median(mean_abs_shift),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story mean absolute marginal shift by condition"
  )
```

```{r absolute-shift-plot, fig.width=12, fig.height=5}
p_abs_density <- ggplot(df_abs_point, aes(x = abs_shift, fill = condition)) +
  geom_density(alpha = 0.45) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Absolute Immediate Sentiment Shift",
    subtitle = "Point-level |windowed marginal shift|",
    x = "Absolute marginal shift",
    y = "Density"
  ) +
  theme(legend.position = "bottom")

p_abs_story <- ggplot(df_abs_story, aes(x = condition, y = mean_abs_shift, fill = condition)) +
  geom_violin(alpha = 0.55, trim = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.85, outlier.alpha = 0.15) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 1.3) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Story-Level Adaptation Intensity",
    subtitle = "Mean |windowed marginal shift| per story",
    x = "Condition",
    y = "Mean absolute marginal shift"
  ) +
  theme(legend.position = "none")

p_abs_combined <- p_abs_density + p_abs_story
print(p_abs_combined)

ggsave(
  file.path(GRAPHS_DIR, "immediate_adaptation_absolute_shift.png"),
  p_abs_combined,
  width = 12,
  height = 5,
  dpi = 300
)
```

```{r absolute-shift-tests}
cat("Per-story ANOVA: mean_abs_shift ~ condition\n")
print(summary(aov(mean_abs_shift ~ condition, data = df_abs_story)))

cat("\nPer-story Kruskal-Wallis: mean_abs_shift ~ condition\n")
print(kruskal.test(mean_abs_shift ~ condition, data = df_abs_story))

cat("\nPairwise Wilcoxon on per-story means (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_abs_story$mean_abs_shift,
  df_abs_story$condition,
  p.adjust.method = "BH"
))

bootstrap_condition_means(df_abs_story, "mean_abs_shift", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on condition mean of per-story absolute shift"
  )

effect_sizes_pairwise(df_abs_story, "mean_abs_shift") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on per-story absolute shift"
  )

planned_contrast_HA_vs_same(df_abs_story, "mean_abs_shift") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: Human-AI vs same-type average"
  )
```

```{r absolute-shift-mixed-model}
model_abs_shift <- lmer(
  abs_shift ~ condition + chronological_position + turn_scaled + (1 | conversation_id),
  data = df_abs_point
)

cat("Mixed model: abs_shift ~ condition + chronological_position + turn_scaled + (1 | story)\n")
print(summary(model_abs_shift))

cat("\nEstimated marginal means by condition:\n")
print(emmeans(model_abs_shift, specs = "condition"))

cat("\nPairwise condition contrasts:\n")
print(pairs(emmeans(model_abs_shift, specs = "condition"), adjust = "bonferroni"))
```

# Three-Turn Response Dynamics

Each complete exchange contributes one three-turn point:

`previous second turn -> current first turn -> current second turn`

Operationally, this uses the two adjacent windowed marginal shifts:

- `incoming_shift = first_shift`
- `response_shift = second_shift`

```{r triplet-preprocess}
df_triplet <- df_shift %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    incoming_shift = first_shift,
    response_shift = second_shift,
    incoming_abs = abs(incoming_shift),
    response_abs = abs(response_shift),
    adaptation_closure = incoming_abs - response_abs,
    reversal_score = -incoming_shift * response_shift,
    reversal_flag = case_when(
      incoming_shift == 0 | response_shift == 0 ~ NA,
      TRUE ~ sign(incoming_shift) != sign(response_shift)
    ),
    response_to_incoming_ratio = if_else(
      incoming_abs > 1e-8,
      response_abs / incoming_abs,
      NA_real_
    )
  ) %>%
  ungroup() %>%
  drop_na(incoming_shift, response_shift, turn_index) %>%
  mutate(turn_scaled = as.numeric(scale(turn_index)))

df_triplet_story <- df_triplet %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_triplets = n(),
    mean_adaptation_closure = mean(adaptation_closure, na.rm = TRUE),
    mean_reversal_score = mean(reversal_score, na.rm = TRUE),
    reversal_rate = mean(reversal_flag, na.rm = TRUE),
    median_response_ratio = median(response_to_incoming_ratio, na.rm = TRUE),
    .groups = "drop"
  )

df_triplet %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_triplets = n(),
    mean_incoming_abs = mean(incoming_abs),
    mean_response_abs = mean(response_abs),
    mean_adaptation_closure = mean(adaptation_closure),
    mean_reversal_score = mean(reversal_score),
    reversal_rate = mean(reversal_flag, na.rm = TRUE),
    median_response_ratio = median(response_to_incoming_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Three-turn response metrics by condition"
  )
```

```{r triplet-plots, fig.width=12, fig.height=5}
p_closure <- ggplot(df_triplet, aes(x = condition, y = adaptation_closure, fill = condition)) +
  geom_violin(alpha = 0.55, trim = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.85, outlier.alpha = 0.12) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Adaptation Closure",
    subtitle = "Positive means response shift is smaller than incoming shift",
    x = "Condition",
    y = "|incoming| - |response|"
  ) +
  theme(legend.position = "none")

p_reversal <- ggplot(df_triplet, aes(x = condition, y = reversal_score, fill = condition)) +
  geom_violin(alpha = 0.55, trim = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.85, outlier.alpha = 0.12) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Response Reversal",
    subtitle = "Positive means response reverses the incoming signed movement",
    x = "Condition",
    y = "-incoming * response"
  ) +
  theme(legend.position = "none")

p_triplet_combined <- p_closure + p_reversal
print(p_triplet_combined)

ggsave(
  file.path(GRAPHS_DIR, "immediate_adaptation_triplet_response.png"),
  p_triplet_combined,
  width = 12,
  height = 5,
  dpi = 300
)
```

```{r triplet-story-tests}
cat("Per-story ANOVA: mean_adaptation_closure ~ condition\n")
print(summary(aov(mean_adaptation_closure ~ condition, data = df_triplet_story)))

cat("\nPer-story Kruskal-Wallis: mean_adaptation_closure ~ condition\n")
print(kruskal.test(mean_adaptation_closure ~ condition, data = df_triplet_story))

cat("\nPairwise Wilcoxon on per-story adaptation closure (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_triplet_story$mean_adaptation_closure,
  df_triplet_story$condition,
  p.adjust.method = "BH"
))

cat("\nPer-story ANOVA: mean_reversal_score ~ condition\n")
print(summary(aov(mean_reversal_score ~ condition, data = df_triplet_story)))

cat("\nPer-story Kruskal-Wallis: mean_reversal_score ~ condition\n")
print(kruskal.test(mean_reversal_score ~ condition, data = df_triplet_story))

cat("\nPairwise Wilcoxon on per-story reversal score (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_triplet_story$mean_reversal_score,
  df_triplet_story$condition,
  p.adjust.method = "BH"
))
```

```{r triplet-mixed-models}
model_closure <- lmer(
  adaptation_closure ~ condition + turn_scaled + (1 | conversation_id),
  data = df_triplet
)

model_reversal <- lmer(
  reversal_score ~ condition + turn_scaled + (1 | conversation_id),
  data = df_triplet
)

cat("Mixed model: adaptation_closure ~ condition + turn_scaled + (1 | story)\n")
print(summary(model_closure))

cat("\nCondition contrasts for adaptation closure:\n")
print(pairs(emmeans(model_closure, specs = "condition"), adjust = "bonferroni"))

cat("\nMixed model: reversal_score ~ condition + turn_scaled + (1 | story)\n")
print(summary(model_reversal))

cat("\nCondition contrasts for reversal score:\n")
print(pairs(emmeans(model_reversal, specs = "condition"), adjust = "bonferroni"))
```

# Notes

Use `mean_abs_shift` as the primary condition-level adaptation intensity result:
it is closest to the original concern that signed marginal shifts cancel. Use
`adaptation_closure` and `reversal_score` as mechanism checks for whether the
subsequent response dampens or reverses the immediately preceding emotional
movement.

