---
title: "LLM-Identity Decomposition: Per-Model Asymmetry within HA and AA"
subtitle: "Does dyad asymmetry depend on which LLM is in the conversation?"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 11,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(lme4, lmerTest, patchwork, emmeans)

theme_set(theme_comparison())

# Map HA llm_type integer codes to model names (matches the experiment design in
# the ACL paper). AA stores model names directly in `model_id`.
HA_LLM_LABELS <- c(
  "1" = "GPT-4.1",
  "2" = "Claude-sonnet-4-5",
  "3" = "Llama-3.3-70B",
  "4" = "Qwen2.5-72B"
)

normalize_aa_model <- function(x) {
  out <- as.character(x)
  out <- dplyr::case_when(
    grepl("gpt", out, ignore.case = TRUE) ~ "GPT-4.1",
    grepl("claude", out, ignore.case = TRUE) ~ "Claude-sonnet-4-5",
    grepl("llama", out, ignore.case = TRUE) ~ "Llama-3.3-70B",
    grepl("qwen", out, ignore.case = TRUE) ~ "Qwen2.5-72B",
    TRUE ~ out
  )
  factor(out, levels = c("GPT-4.1", "Claude-sonnet-4-5", "Llama-3.3-70B", "Qwen2.5-72B"))
}

normalize_ha_llm <- function(x) {
  factor(HA_LLM_LABELS[as.character(x)],
         levels = c("GPT-4.1", "Claude-sonnet-4-5", "Llama-3.3-70B", "Qwen2.5-72B"))
}

attach_llm_identity <- function(df, conditions = c("human-ai", "ai-ai")) {
  metadata <- purrr::map_dfr(conditions, function(cond) {
    meta <- load_interaction_metadata(cond)
    if (!is.null(meta)) {
      meta <- meta %>%
        dplyr::select(dplyr::any_of(c("conversation_id", "llm_type", "model_id")))
    }

    novelty_path <- here::here("data", cond, "processed", "novelty_scores.csv")
    if (file.exists(novelty_path)) {
      novelty_meta <- readr::read_csv(novelty_path, show_col_types = FALSE) %>%
        dplyr::select(dplyr::any_of(c("conversation_id", "llm_type", "model_id")))
      meta <- dplyr::bind_rows(meta, novelty_meta)
    }

    if (is.null(meta)) return(tibble::tibble())

    if (!"llm_type" %in% names(meta)) meta$llm_type <- NA_real_
    if (!"model_id" %in% names(meta)) meta$model_id <- NA_character_

    meta %>%
      dplyr::transmute(
        conversation_id = clean_conversation_id(conversation_id),
        llm_type_meta = suppressWarnings(as.numeric(llm_type)),
        model_id_meta = as.character(model_id)
      ) %>%
      dplyr::distinct(conversation_id, .keep_all = TRUE)
  })

  if (!nrow(metadata)) return(df)

  df %>%
    dplyr::mutate(conversation_id = clean_conversation_id(conversation_id)) %>%
    dplyr::left_join(metadata, by = "conversation_id") %>%
    dplyr::mutate(
      llm_type = dplyr::coalesce(
        suppressWarnings(as.numeric(llm_type)),
        llm_type_meta
      ),
      model_id = dplyr::coalesce(as.character(model_id), model_id_meta)
    ) %>%
    dplyr::select(-llm_type_meta, -model_id_meta)
}

llm_palette <- c(
  "GPT-4.1"            = "#1f77b4",
  "Claude-sonnet-4-5"  = "#d62728",
  "Llama-3.3-70B"      = "#2ca02c",
  "Qwen2.5-72B"        = "#9467bd"
)
```

# Overview

The cross-condition notebooks treat HA and AA as monoliths. This notebook
decomposes both by which LLM is in the dyad and asks two questions per
metric:

1. **Per-LLM magnitude.** Does unsigned asymmetry $A_m$ vary across the four
   LLMs within HA? Within AA?
2. **Per-LLM direction.** Does signed slot-asymmetry vary across LLMs (i.e.
   are some models more likely to drive in a particular direction)?

Sample sizes per LLM are small (HA: ~22-29 stories per model; AA: 20 per
model), so reported tests are non-parametric (Kruskal-Wallis + pairwise
Wilcoxon BH-adjusted). Bootstrap CIs are also reported.

Where it makes sense, the same-model rows in HA and AA are placed
side-by-side: does Claude paired with a human exhibit the same asymmetry
profile as Claude paired with itself?

```{r helper-fns}
# Compute per-condition summary + KW + Cliff's d-style pairwise effect sizes
# on a per-story value. Wraps existing helpers and returns a tidy result.
per_llm_summary <- function(df, value_col, group_col = "llm") {
  d <- df %>%
    dplyr::filter(!is.na(.data[[value_col]]), !is.na(.data[[group_col]]))
  if (!nrow(d)) return(list(by_group = tibble::tibble(), kw_p = NA_real_))

  by_group <- d %>%
    dplyr::group_by(.data[[group_col]]) %>%
    dplyr::summarise(
      n = dplyr::n(),
      mean = mean(.data[[value_col]]),
      median = median(.data[[value_col]]),
      sd = sd(.data[[value_col]]),
      .groups = "drop"
    )

  kw_p <- tryCatch(
    kruskal.test(d[[value_col]] ~ as.factor(d[[group_col]]))$p.value,
    error = function(e) NA_real_
  )

  list(by_group = by_group, kw_p = kw_p)
}

# Reshape directional_stability output to use llm as the grouping column.
per_llm_directional <- function(df, value_col, llm_col = "llm") {
  df_in <- df
  # `directional_stability` reads from `condition`, so swap llm in temporarily.
  df_in$condition <- df_in[[llm_col]]
  res <- directional_stability(df_in, value_col)
  if ("condition" %in% names(res$by_condition)) {
    res$by_condition <- res$by_condition %>%
      dplyr::rename(!!llm_col := condition)
  } else {
    res$by_condition <- tibble::tibble(
      !!llm_col := character(),
      n = integer(),
      mean = numeric(),
      sd = numeric(),
      stability = numeric(),
      sign_consistency = numeric(),
      one_sample_t_p = numeric()
    )
  }
  res
}
```

# Build per-metric per-story tables

```{r build-tables}
# ---- Valence: per-story signed Fisher-z difference and unsigned A_valence
df_val <- load_valence_data(c("human-ai", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959",
         conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_val_pairs <- df_val %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(author_2_valence_lag1 = lag(author_2_valence)) %>%
  ungroup() %>%
  drop_na(author_1_valence, author_2_valence, author_2_valence_lag1)

valence_per_story <- df_val_pairs %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_pairs = sum(!is.na(author_1_valence) & !is.na(author_2_valence)),
    r_12 = if (n_pairs >= 3) suppressWarnings(cor(author_1_valence, author_2_valence)) else NA_real_,
    r_21 = if (n_pairs >= 3) suppressWarnings(cor(author_2_valence_lag1, author_1_valence)) else NA_real_,
    llm_type = suppressWarnings(as.numeric(na.omit(unique(llm_type))[1])),
    model_id = as.character(na.omit(unique(model_id))[1]),
    .groups = "drop"
  ) %>%
  mutate(
    r_12 = pmin(pmax(r_12, -0.999999), 0.999999),
    r_21 = pmin(pmax(r_21, -0.999999), 0.999999),
    z_12 = atanh(r_12),
    z_21 = atanh(r_21),
    signed_diff = z_12 - z_21,
    A_valence = abs(signed_diff),
    llm = ifelse(condition == "Human-AI", as.character(normalize_ha_llm(llm_type)),
                 as.character(normalize_aa_model(model_id))),
    llm = factor(llm, levels = names(llm_palette))
  ) %>%
  drop_na(A_valence, llm)

# ---- Novelty: per-story OLS slope difference (resonance ~ surprise per slot)
df_nov <- load_novelty_data(c("human-ai", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959",
         conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_nov_slots <- bind_rows(
  df_nov %>% transmute(condition, conversation_id, turn_index,
                       slot = "author_1",
                       surprise = author_1_surprise,
                       resonance = author_1_surprise - author_1_transience,
                       llm_type, model_id),
  df_nov %>% transmute(condition, conversation_id, turn_index,
                       slot = "author_2",
                       surprise = author_2_surprise,
                       resonance = author_2_surprise - author_2_transience,
                       llm_type, model_id)
) %>%
  drop_na(surprise, resonance) %>%
  mutate(
    condition = factor(condition,
      levels = c("human-ai", "ai-ai"),
      labels = c("Human-AI", "AI-AI")
    ),
    slot = factor(slot, levels = c("author_1", "author_2"),
                  labels = c("Slot 1", "Slot 2"))
  )

novelty_per_story <- df_nov_slots %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(
    n_pts = sum(!is.na(surprise) & !is.na(resonance)),
    slope = if (n_pts >= 4) coef(lm(resonance ~ surprise))[2] else NA_real_,
    llm_type = suppressWarnings(as.numeric(na.omit(unique(llm_type))[1])),
    model_id = as.character(na.omit(unique(model_id))[1]),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = slot, values_from = slope, names_prefix = "slope_",
              id_cols = c(condition, conversation_id, llm_type, model_id)) %>%
  drop_na(`slope_Slot 1`, `slope_Slot 2`) %>%
  mutate(
    signed_diff = `slope_Slot 1` - `slope_Slot 2`,
    A_influence = abs(signed_diff),
    llm = ifelse(condition == "Human-AI", as.character(normalize_ha_llm(llm_type)),
                 as.character(normalize_aa_model(model_id))),
    llm = factor(llm, levels = names(llm_palette))
  ) %>%
  drop_na(A_influence, llm)

# ---- Semantic exploration: per-story AUC asymmetry
df_sem <- load_exploration_data(c("human-ai", "ai-ai")) %>%
  filter(k > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  attach_llm_identity(c("human-ai", "ai-ai")) %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959",
         conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_sem_speaker <- df_sem %>%
  filter(agent != "interleaved") %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(slot = factor(speaker_slot, levels = c("author_1", "author_2"),
                       labels = c("Slot 1", "Slot 2"))) %>%
  drop_na(distance, k)

trap_auc <- function(x, y) {
  if (length(x) < 2) return(NA_real_)
  ord <- order(x); x <- x[ord]; y <- y[ord]
  sum((y[-1] + y[-length(y)]) * diff(x) / 2)
}

semexp_per_story <- df_sem_speaker %>%
  group_by(condition, conversation_id, slot, k) %>%
  summarise(distance = mean(distance, na.rm = TRUE),
            llm_type = suppressWarnings(as.numeric(na.omit(unique(llm_type))[1])),
            model_id = as.character(na.omit(unique(model_id))[1]),
            .groups = "drop") %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(auc = trap_auc(k, distance),
            llm_type = suppressWarnings(as.numeric(na.omit(unique(llm_type))[1])),
            model_id = as.character(na.omit(unique(model_id))[1]),
            .groups = "drop") %>%
  pivot_wider(names_from = slot, values_from = auc, names_prefix = "auc_",
              id_cols = c(condition, conversation_id, llm_type, model_id)) %>%
  drop_na(`auc_Slot 1`, `auc_Slot 2`) %>%
  mutate(
    signed_diff = `auc_Slot 1` - `auc_Slot 2`,
    A_semexp = abs(signed_diff),
    llm = ifelse(condition == "Human-AI", as.character(normalize_ha_llm(llm_type)),
                 as.character(normalize_aa_model(model_id))),
    llm = factor(llm, levels = names(llm_palette))
  ) %>%
  drop_na(A_semexp, llm)

cat(sprintf("Per-story tables built:\n  valence: %d (HA=%d, AA=%d)\n  novelty: %d (HA=%d, AA=%d)\n  semexp:  %d (HA=%d, AA=%d)\n",
  nrow(valence_per_story),
  sum(valence_per_story$condition == "Human-AI"),
  sum(valence_per_story$condition == "AI-AI"),
  nrow(novelty_per_story),
  sum(novelty_per_story$condition == "Human-AI"),
  sum(novelty_per_story$condition == "AI-AI"),
  nrow(semexp_per_story),
  sum(semexp_per_story$condition == "Human-AI"),
  sum(semexp_per_story$condition == "AI-AI")
))
```

# Within-HA: per-LLM Asymmetry

## Magnitude

```{r ha-mag}
ha_mag <- bind_rows(
  per_llm_summary(valence_per_story %>% filter(condition == "Human-AI"), "A_valence")$by_group %>%
    mutate(metric = "A_valence"),
  per_llm_summary(novelty_per_story %>% filter(condition == "Human-AI"), "A_influence")$by_group %>%
    mutate(metric = "A_influence"),
  per_llm_summary(semexp_per_story %>% filter(condition == "Human-AI"), "A_semexp")$by_group %>%
    mutate(metric = "A_semexp")
) %>%
  select(metric, llm, n, mean, median, sd)

ha_mag %>%
  knitr::kable(
    digits = 3,
    caption = "Per-LLM unsigned asymmetry magnitude within HA"
  )

ha_kw <- tibble::tibble(
  metric = c("A_valence", "A_influence", "A_semexp"),
  kw_p = c(
    per_llm_summary(valence_per_story %>% filter(condition == "Human-AI"), "A_valence")$kw_p,
    per_llm_summary(novelty_per_story %>% filter(condition == "Human-AI"), "A_influence")$kw_p,
    per_llm_summary(semexp_per_story %>% filter(condition == "Human-AI"), "A_semexp")$kw_p
  )
)
ha_kw %>%
  knitr::kable(digits = 4, caption = "Within-HA Kruskal-Wallis: A_m ~ llm")
```

```{r ha-mag-plot, fig.width=12, fig.height=4}
plot_one_mag <- function(df, value_col, title) {
  ggplot(df, aes(x = llm, y = .data[[value_col]], fill = llm)) +
    geom_violin(alpha = 0.4, colour = NA) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.85) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 0.8) +
    scale_fill_manual(values = llm_palette) +
    labs(title = title, x = NULL, y = value_col) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 25, hjust = 1))
}

(plot_one_mag(valence_per_story %>% filter(condition == "Human-AI"), "A_valence", "Valence asymmetry") |
 plot_one_mag(novelty_per_story %>% filter(condition == "Human-AI"), "A_influence", "Novelty-resonance asymmetry") |
 plot_one_mag(semexp_per_story %>% filter(condition == "Human-AI"), "A_semexp", "Semexp AUC asymmetry"))
```

## Direction

Per-LLM directional stability of the SIGNED slot difference within HA. If
the human-side identity is what drives directionality, all four LLMs should
look similar (same signed direction). If specific LLMs shape directionality
(e.g., Claude is more deferential than Qwen), the stability index will
differ across models.

```{r ha-dir-tables}
ha_dir_val <- per_llm_directional(
  valence_per_story %>% filter(condition == "Human-AI"),
  "signed_diff", "llm"
)
ha_dir_nov <- per_llm_directional(
  novelty_per_story %>% filter(condition == "Human-AI"),
  "signed_diff", "llm"
)
ha_dir_sem <- per_llm_directional(
  semexp_per_story %>% filter(condition == "Human-AI"),
  "signed_diff", "llm"
)

bind_rows(
  ha_dir_val$by_condition %>% mutate(metric = "valence"),
  ha_dir_nov$by_condition %>% mutate(metric = "novelty"),
  ha_dir_sem$by_condition %>% mutate(metric = "semexp")
) %>%
  select(metric, llm, n, mean, sd, stability, sign_consistency, one_sample_t_p) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-HA directional stability of signed asymmetry per LLM"
  )

tibble::tibble(
  metric = c("valence", "novelty", "semexp"),
  brown_forsythe_p = c(
    ha_dir_val$variance_equality_p,
    ha_dir_nov$variance_equality_p,
    ha_dir_sem$variance_equality_p
  )
) %>%
  knitr::kable(digits = 4, caption = "Variance-equality across LLMs (within HA)")
```

```{r ha-dir-plot, fig.width=12, fig.height=4}
plot_signed <- function(df, value_col, title) {
  ggplot(df, aes(x = llm, y = .data[[value_col]], fill = llm)) +
    geom_violin(alpha = 0.4, colour = NA) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.85) +
    geom_jitter(width = 0.08, alpha = 0.4, size = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = llm_palette) +
    labs(title = title, x = NULL, y = "signed slot difference") +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 25, hjust = 1))
}

(plot_signed(valence_per_story %>% filter(condition == "Human-AI"), "signed_diff", "Valence direction") |
 plot_signed(novelty_per_story %>% filter(condition == "Human-AI"), "signed_diff", "Novelty direction") |
 plot_signed(semexp_per_story %>% filter(condition == "Human-AI"), "signed_diff", "Semexp direction"))
```

# Within-AA: per-Model Asymmetry

The AA dataset only contains same-model self-pairings (no cross-model
pairs). The four levels are therefore pure within-model ai-vs-ai
conversations.

## Magnitude

```{r aa-mag}
aa_mag <- bind_rows(
  per_llm_summary(valence_per_story %>% filter(condition == "AI-AI"), "A_valence")$by_group %>%
    mutate(metric = "A_valence"),
  per_llm_summary(novelty_per_story %>% filter(condition == "AI-AI"), "A_influence")$by_group %>%
    mutate(metric = "A_influence"),
  per_llm_summary(semexp_per_story %>% filter(condition == "AI-AI"), "A_semexp")$by_group %>%
    mutate(metric = "A_semexp")
) %>%
  select(metric, llm, n, mean, median, sd)

aa_mag %>%
  knitr::kable(
    digits = 3,
    caption = "Per-model unsigned asymmetry magnitude within AA"
  )

aa_kw <- tibble::tibble(
  metric = c("A_valence", "A_influence", "A_semexp"),
  kw_p = c(
    per_llm_summary(valence_per_story %>% filter(condition == "AI-AI"), "A_valence")$kw_p,
    per_llm_summary(novelty_per_story %>% filter(condition == "AI-AI"), "A_influence")$kw_p,
    per_llm_summary(semexp_per_story %>% filter(condition == "AI-AI"), "A_semexp")$kw_p
  )
)
aa_kw %>%
  knitr::kable(digits = 4, caption = "Within-AA Kruskal-Wallis: A_m ~ model")
```

```{r aa-mag-plot, fig.width=12, fig.height=4}
(plot_one_mag(valence_per_story %>% filter(condition == "AI-AI"), "A_valence", "Valence asymmetry") |
 plot_one_mag(novelty_per_story %>% filter(condition == "AI-AI"), "A_influence", "Novelty-resonance asymmetry") |
 plot_one_mag(semexp_per_story %>% filter(condition == "AI-AI"), "A_semexp", "Semexp AUC asymmetry"))
```

## Direction

```{r aa-dir-tables}
aa_dir_val <- per_llm_directional(
  valence_per_story %>% filter(condition == "AI-AI"), "signed_diff", "llm"
)
aa_dir_nov <- per_llm_directional(
  novelty_per_story %>% filter(condition == "AI-AI"), "signed_diff", "llm"
)
aa_dir_sem <- per_llm_directional(
  semexp_per_story %>% filter(condition == "AI-AI"), "signed_diff", "llm"
)

bind_rows(
  aa_dir_val$by_condition %>% mutate(metric = "valence"),
  aa_dir_nov$by_condition %>% mutate(metric = "novelty"),
  aa_dir_sem$by_condition %>% mutate(metric = "semexp")
) %>%
  select(metric, llm, n, mean, sd, stability, sign_consistency, one_sample_t_p) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-AA directional stability of signed asymmetry per model"
  )

tibble::tibble(
  metric = c("valence", "novelty", "semexp"),
  brown_forsythe_p = c(
    aa_dir_val$variance_equality_p,
    aa_dir_nov$variance_equality_p,
    aa_dir_sem$variance_equality_p
  )
) %>%
  knitr::kable(digits = 4, caption = "Variance-equality across models (within AA)")
```

# Same Model Across Conditions: HA vs AA per-LLM

For each LLM, place its HA stories (paired with a human) next to its AA
stories (paired with itself). Asks whether a given model's asymmetry
profile depends on partner type.

```{r ha-vs-aa-per-llm-mag, fig.width=12, fig.height=7}
ha_vs_aa_mag_plot_df <- bind_rows(
  valence_per_story %>%
    transmute(condition, llm, metric = "Valence asymmetry", value = A_valence),
  novelty_per_story %>%
    transmute(condition, llm, metric = "Novelty-resonance asymmetry", value = A_influence),
  semexp_per_story %>%
    transmute(condition, llm, metric = "Semexp AUC asymmetry", value = A_semexp)
) %>%
  mutate(
    metric = factor(
      metric,
      levels = c(
        "Valence asymmetry",
        "Novelty-resonance asymmetry",
        "Semexp AUC asymmetry"
      )
    )
  ) %>%
  drop_na(condition, llm, value)

ggplot(ha_vs_aa_mag_plot_df, aes(x = llm, y = value, fill = condition)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.25,
               position = position_dodge(width = 0.75)) +
  facet_wrap(vars(metric), ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("Human-AI" = "#1f77b4", "AI-AI" = "#ff7f0e")) +
  labs(x = NULL, y = "Unsigned asymmetry", fill = "Condition") +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(angle = 25, hjust = 1, size = 9)
  )
```

```{r ha-vs-aa-per-llm-tests}
# For each metric x LLM, Wilcoxon test HA vs AA on the unsigned A
ha_vs_aa_per_llm <- function(df, value_col, metric_label) {
  df %>%
    group_by(llm) %>%
    summarise(
      n_HA = sum(condition == "Human-AI"),
      n_AA = sum(condition == "AI-AI"),
      median_HA = median(.data[[value_col]][condition == "Human-AI"], na.rm = TRUE),
      median_AA = median(.data[[value_col]][condition == "AI-AI"], na.rm = TRUE),
      wilcox_p = tryCatch(
        wilcox.test(.data[[value_col]] ~ condition)$p.value,
        error = function(e) NA_real_
      ),
      cliff_delta = tryCatch({
        a <- .data[[value_col]][condition == "Human-AI"]
        b <- .data[[value_col]][condition == "AI-AI"]
        (sum(outer(a, b, ">")) - sum(outer(a, b, "<"))) / (length(a) * length(b))
      }, error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(metric = metric_label, p_BH = p.adjust(wilcox_p, method = "BH")) %>%
    select(metric, llm, n_HA, n_AA, median_HA, median_AA, cliff_delta, wilcox_p, p_BH)
}

bind_rows(
  ha_vs_aa_per_llm(valence_per_story, "A_valence", "valence"),
  ha_vs_aa_per_llm(novelty_per_story, "A_influence", "novelty"),
  ha_vs_aa_per_llm(semexp_per_story, "A_semexp", "semexp")
) %>%
  knitr::kable(
    digits = 3,
    caption = "HA vs AA per-LLM Wilcoxon (BH-adjusted within metric)"
  )
```

# Discussion-friendly summary tables

Quick-read summary across all metrics: which LLMs (if any) stand out on
magnitude, on direction, and on partner-sensitivity?

```{r summary-quickread}
ha_dir_summary <- bind_rows(
  ha_dir_val$by_condition %>% mutate(metric = "valence"),
  ha_dir_nov$by_condition %>% mutate(metric = "novelty"),
  ha_dir_sem$by_condition %>% mutate(metric = "semexp")
) %>%
  mutate(condition = "Human-AI")

aa_dir_summary <- bind_rows(
  aa_dir_val$by_condition %>% mutate(metric = "valence"),
  aa_dir_nov$by_condition %>% mutate(metric = "novelty"),
  aa_dir_sem$by_condition %>% mutate(metric = "semexp")
) %>%
  mutate(condition = "AI-AI")

bind_rows(ha_dir_summary, aa_dir_summary) %>%
  select(condition, metric, llm, n, stability, sign_consistency, one_sample_t_p) %>%
  arrange(condition, metric, llm) %>%
  knitr::kable(
    digits = 3,
    caption = "Directional stability per LLM, by condition (Human-AI vs AI-AI)"
  )
```

# Summary

What this notebook adds:

- **Within-HA per-LLM**: tests whether the four LLMs differ in unsigned
  asymmetry $A_m$ (KW + per-LLM bootstrap-style CIs in the magnitude
  table) and in directional stability of the signed slot difference. If
  the HA-specific identity-tied directionality reading from the main
  notebook holds, all four LLMs should show high stability and a
  consistent sign on each metric. If it breaks down for one LLM, that
  LLM contributes less to the HA-distinct signature.

- **Within-AA per-model**: tests whether self-pairings of different LLMs
  produce different asymmetry profiles. A pure mathematical-baseline
  reading predicts no difference; a model-specific stylistic reading
  predicts variation (e.g., Claude self-pair more aligning, Qwen
  self-pair more divergent).

- **HA vs AA per-LLM**: holds the LLM constant and varies the partner.
  Same-model HA-vs-AA differences (Wilcoxon + Cliff's $\delta$, BH-adjusted
  per metric) isolate the effect of swapping a human partner for a same-LLM
  partner.

Article-use notes:
- Use the Within-HA per-LLM table to support a sentence in the Discussion
  on whether the HA finding is model-general or model-driven.
- Use Within-AA per-LLM to qualify the structural-null result: if AA's
  rubber-band excess is concentrated in one model (likely Claude or GPT),
  this reads as model-specific homogenization, not "AI-AI dyads in
  general".
- Use HA vs AA per-LLM as a partner-sensitivity panel that complements the
  three primary cross-condition tests.

*Analysis completed: `r Sys.time()`*
