---
title: "Novelty Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Novelty, Transience, and Resonance Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize
)

theme_set(theme_comparison())
```

# Overview

This notebook compares three information-theoretic metrics across conditions:

- **Novelty / surprise**: how unexpected a contribution is given prior context
- **Transience**: how quickly its influence fades
- **Resonance**: novelty minus transience

The notebook now follows the new data structure:

- it loads novelty scores through the shared comparison helpers,
- it restricts the analysis to **complete exchanges** in the aligned turn window,
- it treats `author_1` and `author_2` as **slot 1 / slot 2** rather than fixed identities,
- it uses role labels only where they are structurally valid:
  - **Human / AI** in Human-AI
  - **Starter-side / Non-starter-side** in same-type conditions
- role-labeled plots are therefore **within-condition descriptive views**, not direct cross-condition comparisons

**Primary cross-condition object.** Following the paper's narrative-influence model, the main inferential result is the **within-condition novelty-to-resonance slope interaction** between the two slots, fit as $\text{resonance} \sim \text{surprise} \times \text{slot} + (1 \mid \text{conversation\_id})$ per condition. The cross-condition asymmetry magnitude is
$$A_{\text{influence}}(\text{condition}) \;=\; \bigl|\, \beta_{\text{surprise},\,\text{slot}=1} - \beta_{\text{surprise},\,\text{slot}=2} \,\bigr|,$$
extracted via `emtrends`. Slot mean differences are retained as a supplemental view. Starter-mechanism and Human-AI speaker-type decompositions are deferred.

# Load Data

```{r load-data}
df_raw <- load_novelty_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_raw %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded novelty data by condition")
```

# Preprocessing

```{r preprocess}
df_slots <- bind_rows(
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_1",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_1_type),
        starter_side == "author_1" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_1_surprise,
      transience = author_1_transience,
      resonance = author_1_surprise - author_1_transience
    ),
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_2",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_2_type),
        starter_side == "author_2" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_2_surprise,
      transience = author_2_transience,
      resonance = author_2_surprise - author_2_transience
    )
) %>%
  drop_na(surprise, transience, resonance) %>%
  mutate(
    condition = factor(
      condition,
      levels = c("human-ai", "human-human", "ai-ai"),
      labels = c("Human-AI", "Human-Human", "AI-AI")
    ),
    slot = factor(slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  )

df_long <- df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, levels = c("surprise", "transience", "resonance"))
  ) 



cat("Slot-level rows:", nrow(df_slots), "\n")
cat("Metric rows:", nrow(df_long), "\n")
```

# Primary: Within-Condition Slot Slope-Interaction Asymmetry

```{r primary-slope-interaction}
# Fit resonance ~ surprise * slot + (1|story) within each condition.
# The magnitude of the slope difference between slots is A_influence(condition).
slope_results <- purrr::map_dfr(levels(df_slots$condition), function(cond) {
  df_cond <- df_slots %>% filter(condition == cond)
  if (nrow(df_cond) < 50 || n_distinct(df_cond$conversation_id) < 5) {
    return(tibble())
  }

  # Ensure slot is a 2-level factor with sum contrasts for interpretable interaction
  df_cond$slot <- factor(df_cond$slot, levels = c("Slot 1", "Slot 2"))
  contrasts(df_cond$slot) <- contr.sum(2)

  model <- tryCatch(
    lmer(
      resonance ~ surprise * slot + (1 | conversation_id),
      data = df_cond,
      control = lmerControl(optimizer = "bobyqa")
    ),
    error = function(e) NULL
  )
  if (is.null(model)) return(tibble())

  slopes <- as.data.frame(emtrends(model, specs = "slot", var = "surprise"))
  contrast_df <- as.data.frame(
    contrast(emtrends(model, specs = "slot", var = "surprise"),
             method = "revpairwise")
  )

  tibble(
    condition = cond,
    slope_slot1 = slopes$surprise.trend[slopes$slot == "Slot 1"],
    slope_slot2 = slopes$surprise.trend[slopes$slot == "Slot 2"],
    signed_diff = contrast_df$estimate[1],
    se = contrast_df$SE[1],
    t = contrast_df$t.ratio[1],
    p = contrast_df$p.value[1],
    A_influence = abs(contrast_df$estimate[1])
  )
})

slope_results %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Primary: |slope_1 - slope_2| from resonance ~ surprise * slot per condition"
  )

# Sanity: A_influence is unsigned magnitude
stopifnot(all(slope_results$A_influence >= 0))
```

```{r primary-slope-interaction-plot, fig.width=9, fig.height=5}
if (nrow(slope_results) > 0) {
  slope_results %>%
    mutate(
      ci_lower = A_influence - 1.96 * se,
      ci_upper = A_influence + 1.96 * se
    ) %>%
    ggplot(aes(x = condition, y = A_influence, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.85) +
    geom_errorbar(aes(ymin = pmax(ci_lower, 0), ymax = ci_upper), width = 0.18) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "A_influence by condition",
      subtitle = "|slope(resonance~surprise) diff between slot 1 and slot 2| (LMM, story random intercept)",
      x = "Condition",
      y = expression("|" * Delta * beta * "|")
    ) +
    theme(legend.position = "none")
}
```

```{r primary-per-story-robustness}
# Robustness: per-story OLS slope difference. Each story needs >= 4 points per slot.
per_story_slope_diff <- df_slots %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(
    n_points = sum(!is.na(surprise) & !is.na(resonance)),
    slope = if (n_points >= 4) {
      coef(lm(resonance ~ surprise))[2]
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  select(condition, conversation_id, slot, slope) %>%
  pivot_wider(names_from = slot, values_from = slope, names_prefix = "slope_") %>%
  drop_na(`slope_Slot 1`, `slope_Slot 2`) %>%
  mutate(
    signed_diff = `slope_Slot 1` - `slope_Slot 2`,
    A_influence_story = abs(signed_diff)
  )

per_story_slope_diff %>%
  group_by(condition) %>%
  summarise(
    n_stories = n(),
    median_A = median(A_influence_story),
    mean_A = mean(A_influence_story),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Robustness: per-story |slope_1 - slope_2| (OLS per story, \u2265 4 points per slot)"
  )

if (n_distinct(per_story_slope_diff$condition) >= 2 && nrow(per_story_slope_diff) >= 10) {
  cat("\nKruskal-Wallis on per-story A_influence:\n")
  print(kruskal.test(A_influence_story ~ condition, data = per_story_slope_diff))
}
```

# Supplemental: Slot Mean Comparison

```{r slot-asymmetry}
slot_asymmetry <- purrr::map_dfr(levels(df_long$metric), function(metric_name) {
  df_metric <- df_long %>%
    filter(metric == metric_name)

  model <- lmer(value ~ slot * condition + (1 | conversation_id), data = df_metric)
  emm <- emmeans(model, specs = c("condition", "slot"))
  emm_df <- as.data.frame(emm)
  contrast_df <- as.data.frame(pairs(emm, by = "condition"))

  tibble(
    metric = metric_name,
    condition = contrast_df$condition,
    slot_1_mean = purrr::map_dbl(
      contrast_df$condition,
      ~ emm_df$emmean[emm_df$condition == .x & emm_df$slot == "Slot 1"][1]
    ),
    slot_2_mean = purrr::map_dbl(
      contrast_df$condition,
      ~ emm_df$emmean[emm_df$condition == .x & emm_df$slot == "Slot 2"][1]
    ),
    difference = contrast_df$estimate,
    se = contrast_df$SE,
    t = contrast_df$t.ratio,
    p = contrast_df$p.value
  )
})

slot_asymmetry %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Slot asymmetry by condition for novelty metrics"
  )
```

```{r slot-asymmetry-plot, fig.width=12, fig.height=8}
slot_asymmetry %>%
  ggplot(aes(x = condition, y = difference, fill = condition)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
    width = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Slot Asymmetry by Condition",
    subtitle = "Difference: Slot 1 - Slot 2 (supplemental slot-mean view).",
    x = "Condition",
    y = "Asymmetry"
  ) +
  theme(legend.position = "none")
```

# Supplemental: General Condition-Level Comparison

```{r descriptives}
df_long %>%
  group_by(condition, metric) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_obs = n(),
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Overall novelty metrics by condition"
  )
```

```{r condition-distributions, fig.width=12, fig.height=8}
df_long %>%
  ggplot(aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  labs(
    title = "Novelty Metric Distributions by Condition",
    x = "Metric Value",
    y = "Density"
  ) +
  theme(legend.position = "bottom")
```

```{r condition-models}
for (metric_name in levels(df_long$metric)) {
  cat("\n===", metric_name, "===\n")
  model <- lmer(value ~ condition + (1 | conversation_id), data = filter(df_long, metric == metric_name))
  print(summary(model))
  print(pairs(emmeans(model, specs = "condition"), adjust = "bonferroni"))
}
```

# Supplemental: Pooled Novelty-to-Resonance Relationship

```{r novelty-resonance}
innovation_model <- lmer(
  resonance ~ surprise * condition + (1 | conversation_id),
  data = df_slots
)

cat("Novelty-to-resonance model:\n")
print(summary(innovation_model))

cat("\nCondition-specific novelty-to-resonance slopes:\n")
print(emtrends(innovation_model, specs = "condition", var = "surprise"))
```

```{r novelty-resonance-plot, fig.width=12, fig.height=5}
df_slots %>%
  ggplot(aes(x = surprise, y = resonance, color = condition)) +
  geom_point(alpha = 0.08, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Novelty vs Resonance by Condition",
    subtitle = "One novelty-to-resonance slope per condition",
    x = "Novelty / Surprise",
    y = "Resonance"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: Role-Level View

```{r role-view, fig.width=12, fig.height=7}
df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = role_label, y = value, fill = role_label)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.1) +
  facet_grid(metric ~ condition, scales = "free_y") +
  labs(
    title = "Role-Conditioned Novelty Metrics",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = NULL,
    y = "Metric Value"
  ) +
  theme(legend.position = "none")
```

# Summary

Primary cross-condition result: within each condition, $A_{\text{influence}} = |\,\beta_{\text{surprise},\,\text{slot}=1} - \beta_{\text{surprise},\,\text{slot}=2}\,|$ from `resonance ~ surprise * slot + (1|conversation_id)`. Per-story OLS slope differences are reported as a robustness layer.

Supplemental views retained:

- slot mean differences per metric (LMM `value ~ slot * condition + (1|story)`)
- condition-level marginal distributions and pooled `condition` LMMs
- pooled novelty-to-resonance model with `surprise * condition` interaction
- within-condition role box-plots (Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions)

Deferred to a follow-up pass:

- starter-mechanism analysis (signed, starter-normalized asymmetry across all conditions)
- Human-AI decomposition (`speaker_type \u00d7 speaker_is_starter` within Human-AI only)
- length-controlled variant (`+ nwords`) mirroring the paper

*Analysis completed: `r Sys.time()`*
