---
title: "Novelty Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Novelty, Transience, and Resonance Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares **novelty metrics** across all three experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Metrics (Information-Theoretic Framework):**

- **Novelty (Surprisal)**: How unexpected is a turn given past context? Higher = more surprising/novel
- **Transience**: How quickly does a turn's influence fade? Higher = less lasting impact
- **Resonance**: Novelty minus Transience — net forward influence on the narrative

**Key Questions:**
1. Do humans produce higher novelty than AI?
2. Does novelty lead to greater resonance in human contributions?
3. How do Human-Human dynamics compare to Human-AI?
4. Is the AI-AI baseline more predictable (lower novelty)?

# Load Data

```{r load-data}
# Load novelty data from each condition
load_novelty_data <- function(conditions = c("human-ai", "human-human", "ai-ai")) {
  all_data <- list()
  
  for (cond in conditions) {
    # Try different possible filenames
    filenames <- c("novelty_scores.csv", "novelty_scores_40.csv", "novelty_scores.parquet")
    
    for (fname in filenames) {
      path <- here("data", cond, "processed", fname)
      if (file.exists(path)) {
        df <- if (str_detect(fname, "\\.parquet$")) {
          read_parquet(path)
        } else {
          read_csv(path, show_col_types = FALSE)
        }
        cat("Loaded", cond, "from", fname, "\n")
        all_data[[cond]] <- df %>% mutate(condition = cond)
        break
      }
    }
  }
  
  bind_rows(all_data)
}

# Load data
df_combined <- load_novelty_data()
cat("\nLoaded conditions:", paste(unique(df_combined$condition), collapse = ", "), "\n")
```

# Schema Normalization

```{r normalize}
# Normalize column names across conditions
# Human-AI: user_surprise, ai_surprise, user_transience, ai_transience
# Human-Human: user_surprise, user2_surprise, user_transience, user2_transience
# AI-AI: agent_1_surprise, agent_2_surprise, agent_1_transience, agent_2_transience

normalize_novelty_columns <- function(df) {
  # Detect condition and normalize
  df <- df %>%
    mutate(
      # Author 1 metrics
      author_1_surprise = case_when(
        !is.na(user_surprise) ~ user_surprise,
        "agent_1_surprise" %in% names(df) ~ .data[["agent_1_surprise"]],
        TRUE ~ NA_real_
      ),
      author_1_transience = case_when(
        !is.na(user_transience) ~ user_transience,
        "agent_1_transience" %in% names(df) ~ .data[["agent_1_transience"]],
        TRUE ~ NA_real_
      ),
      # Author 2 metrics
      author_2_surprise = case_when(
        condition == "human-ai" & !is.na(ai_surprise) ~ ai_surprise,
        condition == "human-human" & "user2_surprise" %in% names(df) ~ .data[["user2_surprise"]],
        "agent_2_surprise" %in% names(df) ~ .data[["agent_2_surprise"]],
        TRUE ~ NA_real_
      ),
      author_2_transience = case_when(
        condition == "human-ai" & !is.na(ai_transience) ~ ai_transience,
        condition == "human-human" & "user2_transience" %in% names(df) ~ .data[["user2_transience"]],
        "agent_2_transience" %in% names(df) ~ .data[["agent_2_transience"]],
        TRUE ~ NA_real_
      )
    )
  
  # Compute resonance (novelty - transience)
  df <- df %>%
    mutate(
      author_1_resonance = author_1_surprise - author_1_transience,
      author_2_resonance = author_2_surprise - author_2_transience
    )
  
  df
}

# Check available columns before normalization
cat("Columns in combined data:\n")
print(names(df_combined))

# Normalize if we have the expected columns
if ("user_surprise" %in% names(df_combined) || "agent_1_surprise" %in% names(df_combined)) {
  df_analysis <- normalize_novelty_columns(df_combined)
} else {
  # Fallback: use whatever surprise/transience columns exist
  df_analysis <- df_combined
  cat("Using original column names - manual inspection needed\n")
}

# Prepare condition factor
df_analysis <- df_analysis %>%
  prepare_condition_factor(reference = "human-ai")

# Report data summary
cat("\nAnalysis dataset:\n")
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Descriptive Statistics

```{r descriptives}
# Summary statistics by condition
desc_stats <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    # Author 1 (Human in Human-AI, User1 in Human-Human, Agent1 in AI-AI)
    a1_surprise_mean = mean(author_1_surprise, na.rm = TRUE),
    a1_surprise_sd = sd(author_1_surprise, na.rm = TRUE),
    a1_transience_mean = mean(author_1_transience, na.rm = TRUE),
    a1_resonance_mean = mean(author_1_resonance, na.rm = TRUE),
    # Author 2
    a2_surprise_mean = mean(author_2_surprise, na.rm = TRUE),
    a2_surprise_sd = sd(author_2_surprise, na.rm = TRUE),
    a2_transience_mean = mean(author_2_transience, na.rm = TRUE),
    a2_resonance_mean = mean(author_2_resonance, na.rm = TRUE),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3,
    caption = "Novelty metrics by condition"
  )
```

# Novelty (Surprisal) Comparison

## Distribution Visualization

```{r novelty-distributions, fig.width=12, fig.height=5}
# Pivot to long format for plotting
df_novelty_long <- df_analysis %>%
  select(conversation_id, turn, condition, 
         author_1_surprise, author_2_surprise) %>%
  pivot_longer(
    cols = c(author_1_surprise, author_2_surprise),
    names_to = "author",
    values_to = "surprise"
  ) %>%
  mutate(
    author = if_else(str_detect(author, "author_1"), "Author 1", "Author 2")
  ) %>%
  drop_na(surprise)

# Distribution plots
p1 <- df_novelty_long %>%
  ggplot(aes(x = surprise, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~author) +
  labs(
    title = "Novelty (Surprisal) Distributions",
    x = "Surprisal",
    y = "Density"
  )

p2 <- df_novelty_long %>%
  ggplot(aes(x = condition, y = surprise, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~author) +
  labs(
    title = "Novelty by Condition",
    x = "Condition",
    y = "Surprisal"
  ) +
  theme(legend.position = "none")

p1 / p2
```

## Statistical Comparison

```{r novelty-models}
# Author 1 novelty comparison
model_novelty_a1 <- lmer(
  author_1_surprise ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Author 1 Novelty by Condition ===\n")
print(summary(model_novelty_a1))

# Pairwise contrasts
emm_novelty_a1 <- emmeans(model_novelty_a1, specs = "condition")
cat("\n=== Pairwise Contrasts (Author 1) ===\n")
print(pairs(emm_novelty_a1, adjust = "bonferroni"))

# Author 2 novelty comparison
model_novelty_a2 <- lmer(
  author_2_surprise ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("\n=== Author 2 Novelty by Condition ===\n")
print(summary(model_novelty_a2))

emm_novelty_a2 <- emmeans(model_novelty_a2, specs = "condition")
cat("\n=== Pairwise Contrasts (Author 2) ===\n")
print(pairs(emm_novelty_a2, adjust = "bonferroni"))
```

# Transience Comparison

```{r transience-comparison, fig.width=12, fig.height=5}
# Pivot transience to long format
df_transience_long <- df_analysis %>%
  select(conversation_id, turn, condition, 
         author_1_transience, author_2_transience) %>%
  pivot_longer(
    cols = c(author_1_transience, author_2_transience),
    names_to = "author",
    values_to = "transience"
  ) %>%
  mutate(
    author = if_else(str_detect(author, "author_1"), "Author 1", "Author 2")
  ) %>%
  drop_na(transience)

# Visualization
ggplot(df_transience_long, aes(x = condition, y = transience, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~author) +
  labs(
    title = "Transience by Condition and Author",
    subtitle = "Higher transience = influence fades faster",
    x = "Condition",
    y = "Transience"
  ) +
  theme(legend.position = "none")
```

```{r transience-models}
# Author 1 transience
model_trans_a1 <- lmer(
  author_1_transience ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Author 1 Transience by Condition ===\n")
print(summary(model_trans_a1))
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emmeans(model_trans_a1, "condition"), adjust = "bonferroni"))

# Author 2 transience
model_trans_a2 <- lmer(
  author_2_transience ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("\n=== Author 2 Transience by Condition ===\n")
print(summary(model_trans_a2))
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emmeans(model_trans_a2, "condition"), adjust = "bonferroni"))
```

# Resonance Comparison

Resonance = Novelty - Transience (net forward influence)

```{r resonance-comparison, fig.width=12, fig.height=5}
# Pivot resonance to long format
df_resonance_long <- df_analysis %>%
  select(conversation_id, turn, condition, 
         author_1_resonance, author_2_resonance) %>%
  pivot_longer(
    cols = c(author_1_resonance, author_2_resonance),
    names_to = "author",
    values_to = "resonance"
  ) %>%
  mutate(
    author = if_else(str_detect(author, "author_1"), "Author 1", "Author 2")
  ) %>%
  drop_na(resonance)

# Visualization
ggplot(df_resonance_long, aes(x = condition, y = resonance, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~author) +
  labs(
    title = "Resonance by Condition and Author",
    subtitle = "Positive = lasting forward influence on narrative",
    x = "Condition",
    y = "Resonance (Novelty - Transience)"
  ) +
  theme(legend.position = "none")
```

```{r resonance-models}
# Author 1 resonance
model_res_a1 <- lmer(
  author_1_resonance ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Author 1 Resonance by Condition ===\n")
print(summary(model_res_a1))
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emmeans(model_res_a1, "condition"), adjust = "bonferroni"))

# Author 2 resonance
model_res_a2 <- lmer(
  author_2_resonance ~ condition + (1 | conversation_id),
  data = df_analysis
)

cat("\n=== Author 2 Resonance by Condition ===\n")
print(summary(model_res_a2))
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emmeans(model_res_a2, "condition"), adjust = "bonferroni"))
```

# Novelty-Resonance Relationship

Does higher novelty lead to higher resonance? Is this relationship condition-specific?

```{r novelty-resonance-relationship, fig.width=12, fig.height=5}
# Author 1: Novelty vs Resonance
p1 <- df_analysis %>%
  drop_na(author_1_surprise, author_1_resonance) %>%
  ggplot(aes(x = author_1_surprise, y = author_1_resonance, color = condition)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 1: Novelty → Resonance",
    x = "Surprisal (Novelty)",
    y = "Resonance",
    color = "Condition"
  )

# Author 2: Novelty vs Resonance
p2 <- df_analysis %>%
  drop_na(author_2_surprise, author_2_resonance) %>%
  ggplot(aes(x = author_2_surprise, y = author_2_resonance, color = condition)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 2: Novelty → Resonance",
    x = "Surprisal (Novelty)",
    y = "Resonance",
    color = "Condition"
  )

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r novelty-resonance-interaction}
# Test if novelty-resonance slope differs by condition

# Author 1
model_nr_a1 <- lmer(
  author_1_resonance ~ author_1_surprise * condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Author 1: Novelty × Condition Interaction ===\n")
print(summary(model_nr_a1))

# Slopes by condition
emm_slopes_a1 <- emtrends(model_nr_a1, specs = "condition", var = "author_1_surprise")
cat("\n=== Novelty-Resonance Slopes by Condition ===\n")
print(emm_slopes_a1)
cat("\n=== Pairwise Slope Comparisons ===\n")
print(pairs(emm_slopes_a1, adjust = "bonferroni"))
```

# Summary Visualization

```{r summary-plot, fig.width=14, fig.height=10}
# Compile all metrics for summary comparison
summary_data <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    # Author 1
    a1_novelty = mean(author_1_surprise, na.rm = TRUE),
    a1_novelty_se = sd(author_1_surprise, na.rm = TRUE) / sqrt(n()),
    a1_transience = mean(author_1_transience, na.rm = TRUE),
    a1_transience_se = sd(author_1_transience, na.rm = TRUE) / sqrt(n()),
    a1_resonance = mean(author_1_resonance, na.rm = TRUE),
    a1_resonance_se = sd(author_1_resonance, na.rm = TRUE) / sqrt(n()),
    # Author 2
    a2_novelty = mean(author_2_surprise, na.rm = TRUE),
    a2_novelty_se = sd(author_2_surprise, na.rm = TRUE) / sqrt(n()),
    a2_transience = mean(author_2_transience, na.rm = TRUE),
    a2_transience_se = sd(author_2_transience, na.rm = TRUE) / sqrt(n()),
    a2_resonance = mean(author_2_resonance, na.rm = TRUE),
    a2_resonance_se = sd(author_2_resonance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Pivot to long format for faceted plot
summary_long <- summary_data %>%
  pivot_longer(
    cols = -condition,
    names_to = c("author", "metric", ".value"),
    names_pattern = "(a[12])_(.+)_(mean|se)$"
  )

# Alternative pivot that actually works
summary_long <- summary_data %>%
  pivot_longer(
    cols = starts_with("a1_") | starts_with("a2_"),
    names_to = "var",
    values_to = "value"
  ) %>%
  separate(var, into = c("author", "metric_stat"), sep = "_", extra = "merge") %>%
  separate(metric_stat, into = c("metric", "stat"), sep = "_(?=se$|mean$)", fill = "right") %>%
  mutate(stat = if_else(is.na(stat), "mean", stat)) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  rename(mean = `mean`) %>%
  mutate(
    author = if_else(author == "a1", "Author 1", "Author 2"),
    metric = str_to_title(metric)
  )

# Create summary bar plots
ggplot(summary_long, aes(x = condition, y = mean, fill = condition)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se), 
                width = 0.2, linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_grid(author ~ metric, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Summary: Novelty Metrics by Condition and Author",
    subtitle = "Error bars show 95% CI",
    x = "Condition",
    y = "Value"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Key Findings

Based on the analysis above:

1. **Novelty patterns**: [Interpret which authors/conditions show highest surprisal]

2. **Transience patterns**: [Discuss whether contributions have lasting vs. fleeting influence]

3. **Resonance (net influence)**: [Compare forward narrative influence across conditions]

4. **Innovation bias**: [Discuss novelty-resonance slopes — steeper = innovation more rewarded]

5. **AI-AI baseline**: [Note whether simulated baseline shows characteristic patterns]

---

*Analysis completed: `r Sys.time()`*
