---
title: "Novelty Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Novelty, Transience, and Resonance Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize
)

theme_set(theme_comparison())
```

# Overview

This notebook compares three information-theoretic metrics across conditions:

- **Novelty / surprise**: how unexpected a contribution is given prior context
- **Transience**: how quickly its influence fades
- **Resonance**: novelty minus transience

The notebook now follows the new data structure:

- it loads novelty scores through the shared comparison helpers,
- it restricts the analysis to **complete exchanges** in the aligned turn window,
- it treats `author_1` and `author_2` as **slot 1 / slot 2** rather than fixed identities,
- it uses role labels only where they are structurally valid:
  - **Human / AI** in Human-AI
  - **Starter-side / Non-starter-side** in same-type conditions
- role-labeled plots are therefore **within-condition descriptive views**, not direct cross-condition comparisons

# Load Data

```{r load-data}
df_raw <- load_novelty_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window()

df_raw %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded novelty data by condition")
```

# Preprocessing

```{r preprocess}
df_slots <- bind_rows(
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_1",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_1_type),
        starter_side == "author_1" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_1_surprise,
      transience = author_1_transience,
      resonance = author_1_surprise - author_1_transience
    ),
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_2",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_2_type),
        starter_side == "author_2" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_2_surprise,
      transience = author_2_transience,
      resonance = author_2_surprise - author_2_transience
    )
) %>%
  drop_na(surprise, transience, resonance) %>%
  mutate(
    condition = factor(
      condition,
      levels = c("human-ai", "human-human", "ai-ai"),
      labels = c("Human-AI", "Human-Human", "AI-AI")
    ),
    slot = factor(slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  )

df_long <- df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, levels = c("surprise", "transience", "resonance"))
  ) 



cat("Slot-level rows:", nrow(df_slots), "\n")
cat("Metric rows:", nrow(df_long), "\n")
```

# Asymmetry Comparison

```{r slot-asymmetry}
slot_asymmetry <- purrr::map_dfr(levels(df_long$metric), function(metric_name) {
  df_metric <- df_long %>%
    filter(metric == metric_name)

  model <- lmer(value ~ slot * condition + (1 | conversation_id), data = df_metric)
  emm <- emmeans(model, specs = c("condition", "slot"))
  emm_df <- as.data.frame(emm)
  contrast_df <- as.data.frame(pairs(emm, by = "condition"))

  tibble(
    metric = metric_name,
    condition = contrast_df$condition,
    slot_1_mean = purrr::map_dbl(
      contrast_df$condition,
      ~ emm_df$emmean[emm_df$condition == .x & emm_df$slot == "Slot 1"][1]
    ),
    slot_2_mean = purrr::map_dbl(
      contrast_df$condition,
      ~ emm_df$emmean[emm_df$condition == .x & emm_df$slot == "Slot 2"][1]
    ),
    difference = contrast_df$estimate,
    se = contrast_df$SE,
    t = contrast_df$t.ratio,
    p = contrast_df$p.value
  )
})

slot_asymmetry %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Slot asymmetry by condition for novelty metrics"
  )
```

```{r slot-asymmetry-plot, fig.width=12, fig.height=8}
slot_asymmetry %>%
  ggplot(aes(x = condition, y = difference, fill = condition)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
    width = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Slot Asymmetry by Condition",
    subtitle = "Difference: Slot 1 - Slot 2. This is the primary cross-condition asymmetry comparison.",
    x = "Condition",
    y = "Asymmetry"
  ) +
  theme(legend.position = "none")
```

# General Condition-Level Comparison

```{r descriptives}
df_long %>%
  group_by(condition, metric) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_obs = n(),
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Overall novelty metrics by condition"
  )
```

```{r condition-distributions, fig.width=12, fig.height=8}
df_long %>%
  ggplot(aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  labs(
    title = "Novelty Metric Distributions by Condition",
    x = "Metric Value",
    y = "Density"
  ) +
  theme(legend.position = "bottom")
```

```{r condition-models}
for (metric_name in levels(df_long$metric)) {
  cat("\n===", metric_name, "===\n")
  model <- lmer(value ~ condition + (1 | conversation_id), data = filter(df_long, metric == metric_name))
  print(summary(model))
  print(pairs(emmeans(model, specs = "condition"), adjust = "bonferroni"))
}
```

# Condition-Level Novelty-to-Resonance Relationship

```{r novelty-resonance}
innovation_model <- lmer(
  resonance ~ surprise * condition + (1 | conversation_id),
  data = df_slots
)

cat("Novelty-to-resonance model:\n")
print(summary(innovation_model))

cat("\nCondition-specific novelty-to-resonance slopes:\n")
print(emtrends(innovation_model, specs = "condition", var = "surprise"))
```

```{r novelty-resonance-plot, fig.width=12, fig.height=5}
df_slots %>%
  ggplot(aes(x = surprise, y = resonance, color = condition)) +
  geom_point(alpha = 0.08, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Novelty vs Resonance by Condition",
    subtitle = "One novelty-to-resonance slope per condition",
    x = "Novelty / Surprise",
    y = "Resonance"
  ) +
  theme(legend.position = "bottom")
```

# Role-Level View

```{r role-view, fig.width=12, fig.height=7}
df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = role_label, y = value, fill = role_label)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.1) +
  facet_grid(metric ~ condition, scales = "free_y") +
  labs(
    title = "Role-Conditioned Novelty Metrics",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = NULL,
    y = "Metric Value"
  ) +
  theme(legend.position = "none")
```

# Summary

This notebook is now aligned with the new structure by:

- loading novelty data through the shared helper layer,
- restricting the analysis to **complete aligned exchanges**,
- putting the **cross-condition asymmetry comparison first**,
- using **slot-based asymmetry** instead of assuming stable author identities,
- treating the condition-level metric comparison and novelty-to-resonance slope as the second layer.

*Analysis completed: `r Sys.time()`*
