---
title: "Novelty Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Novelty, Transience, and Resonance Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize
)

theme_set(theme_comparison())
```

# Overview

This notebook compares three information-theoretic metrics across conditions:

- **Novelty / surprise**: how unexpected a contribution is given prior context
- **Transience**: how quickly its influence fades
- **Resonance**: novelty minus transience

The notebook now follows the new data structure:

- it loads novelty scores through the shared comparison helpers,
- it restricts the analysis to **complete exchanges** in the aligned turn window,
- it treats `author_1` and `author_2` as **slot 1 / slot 2** rather than fixed identities,
- it uses role labels only where they are structurally valid:
  - **Human / AI** in Human-AI
  - **Starter-side / Non-starter-side** in same-type conditions

# Load Data

```{r load-data}
df_raw <- load_novelty_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window()

df_raw %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded novelty data by condition")
```

# Preprocessing

```{r preprocess}
df_slots <- bind_rows(
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_1",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_1_type),
        starter_side == "author_1" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_1_surprise,
      transience = author_1_transience,
      resonance = author_1_surprise - author_1_transience
    ),
  df_raw %>%
    transmute(
      condition,
      conversation_id,
      turn_index,
      slot = "author_2",
      role_label = case_when(
        condition == "human-ai" ~ str_to_title(author_2_type),
        starter_side == "author_2" ~ "Starter-side",
        TRUE ~ "Non-starter-side"
      ),
      surprise = author_2_surprise,
      transience = author_2_transience,
      resonance = author_2_surprise - author_2_transience
    )
) %>%
  drop_na(surprise, transience, resonance) %>%
  mutate(
    condition = factor(
      condition,
      levels = c("human-ai", "human-human", "ai-ai"),
      labels = c("Human-AI", "Human-Human", "AI-AI")
    ),
    slot = factor(slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  )

df_long <- df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, levels = c("surprise", "transience", "resonance"))
  )

cat("Slot-level rows:", nrow(df_slots), "\n")
cat("Metric rows:", nrow(df_long), "\n")
```

# Descriptive Statistics

```{r descriptives}
df_long %>%
  group_by(condition, metric) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_obs = n(),
    mean_value = mean(value, na.rm = TRUE),
    sd_value = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Overall novelty metrics by condition"
  )
```

```{r condition-distributions, fig.width=12, fig.height=8}
df_long %>%
  ggplot(aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  labs(
    title = "Novelty Metric Distributions by Condition",
    x = "Metric Value",
    y = "Density"
  ) +
  theme(legend.position = "bottom")
```

# Condition-Level Comparison

```{r condition-models}
condition_results <- purrr::map_dfr(levels(df_long$metric), function(metric_name) {
  df_metric <- filter(df_long, metric == metric_name)

  model <- lmer(value ~ condition + (1 | conversation_id), data = df_metric)
  emm <- emmeans(model, specs = "condition")
  pairs_df <- as.data.frame(pairs(emm, adjust = "bonferroni"))

  tibble(
    metric = metric_name,
    omnibus = paste(capture.output(print(anova(model))), collapse = "\n"),
    contrasts = paste(capture.output(print(pairs_df)), collapse = "\n")
  )
})

for (metric_name in levels(df_long$metric)) {
  cat("\n===", metric_name, "===\n")
  model <- lmer(value ~ condition + (1 | conversation_id), data = filter(df_long, metric == metric_name))
  print(summary(model))
  print(pairs(emmeans(model, specs = "condition"), adjust = "bonferroni"))
}
```

# Within-Condition Slot Asymmetry

```{r slot-asymmetry}
slot_asymmetry <- purrr::map_dfr(levels(df_long$metric), function(metric_name) {
  purrr::map_dfr(levels(df_long$condition), function(cond_name) {
    df_subset <- df_long %>%
      filter(metric == metric_name, condition == cond_name)

    if (nrow(df_subset) < 50) {
      return(tibble())
    }

    model <- lmer(value ~ slot + (1 | conversation_id), data = df_subset)
    emm <- emmeans(model, specs = "slot")
    contrast_df <- as.data.frame(pairs(emm))

    tibble(
      metric = metric_name,
      condition = cond_name,
      slot_1_mean = as.data.frame(emm)$emmean[1],
      slot_2_mean = as.data.frame(emm)$emmean[2],
      difference = contrast_df$estimate[1],
      se = contrast_df$SE[1],
      t = contrast_df$t.ratio[1],
      p = contrast_df$p.value[1]
    )
  })
})

slot_asymmetry %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-condition slot asymmetry for novelty metrics"
  )
```

# Role-Level View

```{r role-view, fig.width=12, fig.height=7}
df_slots %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = role_label, y = value, fill = role_label)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.1) +
  facet_grid(metric ~ condition, scales = "free_y") +
  labs(
    title = "Role-Conditioned Novelty Metrics",
    subtitle = "Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = NULL,
    y = "Metric Value"
  ) +
  theme(legend.position = "none")
```

# Novelty-to-Resonance Relationship

```{r novelty-resonance}
innovation_model <- lmer(
  resonance ~ surprise * condition + (1 | conversation_id),
  data = df_slots
)

cat("Novelty-to-resonance model:\n")
print(summary(innovation_model))
```

```{r novelty-resonance-plot, fig.width=12, fig.height=5}
df_slots %>%
  ggplot(aes(x = surprise, y = resonance, color = condition)) +
  geom_point(alpha = 0.08, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Novelty vs Resonance by Condition",
    x = "Novelty / Surprise",
    y = "Resonance"
  ) +
  theme(legend.position = "bottom")
```

# Summary

This notebook is now aligned with the new structure by:

- loading novelty data through the shared helper layer,
- restricting the analysis to **complete aligned exchanges**,
- using **slot-based asymmetry** instead of assuming stable author identities,
- surfacing role labels only where they are valid under the new schema.

*Analysis completed: `r Sys.time()`*
