---
title: "Novelty Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Novelty, Transience, and Resonance Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares **novelty metrics** across all three experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Metrics (Information-Theoretic Framework):**

- **Novelty (Surprisal)**: How unexpected is a turn given past context? Higher = more surprising/novel
- **Transience**: How quickly does a turn's influence fade? Higher = less lasting impact
- **Resonance**: Novelty minus Transience — net forward influence on the narrative

**Key Questions:**
1. Do humans produce higher novelty than AI?
2. Does novelty lead to greater resonance in human contributions?
3. How do Human-Human dynamics compare to Human-AI?
4. Is the AI-AI baseline more predictable (lower novelty)?

# Load Data

```{r load-data}
# Load novelty data from each condition
load_novelty_data <- function(conditions = c("human-ai", "human-human", "ai-ai")) {
  all_data <- list()
  
  for (cond in conditions) {
    # Try different possible filenames
    filenames <- c("novelty_scores_40.csv", "novelty_scores.csv")
    
    for (fname in filenames) {
      path <- here("data", cond, "processed", fname)
      if (file.exists(path)) {
        df <- if (str_detect(fname, "\\.parquet$")) {
          read_parquet(path)
        } else {
          read_csv(path, show_col_types = FALSE)
        }
        cat("Loaded", cond, "from", fname, "\n")
        all_data[[cond]] <- df %>% mutate(condition = cond)
        break
      }
    }
  }
  
  bind_rows(all_data)
}

# Load data
df_combined <- load_novelty_data()
cat("\nLoaded conditions:", paste(unique(df_combined$condition), collapse = ", "), "\n")
```

# Schema Normalization

```{r normalize}
# Data now uses standardized column names from Python pipeline:
# author_1_surprise, author_2_surprise, author_1_transience, author_2_transience

# Compute resonance (novelty - transience)
df_analysis <- df_combined %>%
  mutate(
    author_1_resonance = author_1_surprise - author_1_transience,
    author_2_resonance = author_2_surprise - author_2_transience
  ) %>%
  filter(interaction_count > 1, interaction_count < 10)

# Prepare condition factor
df_analysis <- df_analysis %>%
  prepare_condition_factor(reference = "human-ai")

# Report data summary
cat("\nAnalysis dataset:\n")
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()

cat("\nAvailable columns:\n")
print(names(df_analysis))
```

## Build Long-Format Dataset with Author Factor

Following the within-condition approach, we stack Author 1 and Author 2 metrics into a long format with `author_role` as a factor. This enables proper testing of asymmetry (whether metrics differ by author role) within and across conditions.

```{r build-long-format}
# Create long-format dataset
df_long <- df_analysis %>%
  pivot_longer(
    cols = c(author_1_surprise, author_2_surprise,
             author_1_transience, author_2_transience,
             author_1_resonance, author_2_resonance),
    names_to = c("author_role", "metric"),
    names_pattern = "(author_[12])_(surprise|transience|resonance)",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  mutate(
    author_role = factor(author_role, levels = c("author_1", "author_2"))
  ) %>%
  drop_na(surprise, transience, resonance)

cat("=== Long-Format Dataset ===\n")
df_long %>%
  group_by(condition, author_role) %>%
  summarise(
    n_obs = n(),
    n_stories = n_distinct(conversation_id),
    mean_surprise = mean(surprise, na.rm = TRUE),
    mean_transience = mean(transience, na.rm = TRUE),
    mean_resonance = mean(resonance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

## Standardize Within Conversation × Author

Z-score metrics within each (conversation_id × author_role) cell to ensure comparable coefficients across author roles and conditions.

```{r standardize}
df_std <- df_long %>%
  group_by(conversation_id, author_role) %>%
  mutate(
    surprise_z = as.numeric(scale(surprise)),
    transience_z = as.numeric(scale(transience)),
    resonance_z = as.numeric(scale(resonance))
  ) %>%
  ungroup() %>%
  drop_na(surprise_z, transience_z, resonance_z)

# Set up sum-coded contrasts for author role
contrasts(df_std$author_role) <- contr.sum(2)

cat("Standardized dataset:", nrow(df_std), "observations\n")
```

# Descriptive Statistics

```{r descriptives}
# Summary statistics by condition and author
desc_stats <- df_long %>%
  group_by(condition, author_role) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_obs = n(),
    surprise_mean = mean(surprise, na.rm = TRUE),
    surprise_sd = sd(surprise, na.rm = TRUE),
    transience_mean = mean(transience, na.rm = TRUE),
    transience_sd = sd(transience, na.rm = TRUE),
    resonance_mean = mean(resonance, na.rm = TRUE),
    resonance_sd = sd(resonance, na.rm = TRUE),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3,
    caption = "Novelty metrics by condition and author role"
  )
```

# Within-Condition Asymmetry Analysis

For each condition, we test whether novelty metrics differ by author role using **metric ~ author_role** models. Significant effects indicate asymmetry (e.g., Author 1 shows higher novelty than Author 2).

## Novelty (Surprisal) Asymmetry

```{r novelty-asymmetry-within}
conditions <- unique(df_std$condition)
novelty_asymmetry_results <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_std, condition == cond)
  
  if (nrow(df_cond) < 100) {
    cat("Skipping", cond, "- insufficient data\n")
    next
  }
  
  # Set up contrasts for this subset
  df_cond$author_role <- factor(df_cond$author_role, levels = c("author_1", "author_2"))
  contrasts(df_cond$author_role) <- contr.sum(2)
  
  # Test if surprise differs by author
  m_surprise <- tryCatch({
    lmer(surprise ~ author_role + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(m_surprise)) {
    emm <- emmeans(m_surprise, specs = "author_role")
    contrast_result <- pairs(emm)
    contrast_df <- as.data.frame(contrast_result)
    
    novelty_asymmetry_results <- bind_rows(novelty_asymmetry_results, tibble(
      condition = cond,
      metric = "Novelty",
      author_1_mean = as.data.frame(emm)$emmean[1],
      author_2_mean = as.data.frame(emm)$emmean[2],
      difference = contrast_df$estimate[1],
      se = contrast_df$SE[1],
      t = contrast_df$t.ratio[1],
      p = contrast_df$p.value[1]
    ))
    
    cat("\n=== ", cond, ": Novelty Asymmetry ===\n", sep = "")
    print(summary(m_surprise))
    cat("\n--- Author Means ---\n")
    print(emm)
    cat("\n--- Asymmetry Test ---\n")
    print(contrast_result)
  }
}

novelty_asymmetry_results %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    interpretation = case_when(
      p < 0.05 & difference > 0 ~ "Author 1 > Author 2",
      p < 0.05 & difference < 0 ~ "Author 2 > Author 1",
      TRUE ~ "No asymmetry"
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Novelty asymmetry by condition (Author 1 - Author 2)"
  )
```

## Transience Asymmetry

```{r transience-asymmetry-within}
transience_asymmetry_results <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_std, condition == cond)
  
  if (nrow(df_cond) < 100) next
  
  df_cond$author_role <- factor(df_cond$author_role, levels = c("author_1", "author_2"))
  contrasts(df_cond$author_role) <- contr.sum(2)
  
  m_transience <- tryCatch({
    lmer(transience ~ author_role + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(m_transience)) {
    emm <- emmeans(m_transience, specs = "author_role")
    contrast_result <- pairs(emm)
    contrast_df <- as.data.frame(contrast_result)
    
    transience_asymmetry_results <- bind_rows(transience_asymmetry_results, tibble(
      condition = cond,
      metric = "Transience",
      author_1_mean = as.data.frame(emm)$emmean[1],
      author_2_mean = as.data.frame(emm)$emmean[2],
      difference = contrast_df$estimate[1],
      se = contrast_df$SE[1],
      t = contrast_df$t.ratio[1],
      p = contrast_df$p.value[1]
    ))
    
    cat("\n=== ", cond, ": Transience Asymmetry ===\n", sep = "")
    print(summary(m_transience))
  }
}

transience_asymmetry_results %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    interpretation = case_when(
      p < 0.05 & difference > 0 ~ "Author 1 > Author 2",
      p < 0.05 & difference < 0 ~ "Author 2 > Author 1",
      TRUE ~ "No asymmetry"
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Transience asymmetry by condition (Author 1 - Author 2)"
  )
```

## Resonance Asymmetry

```{r resonance-asymmetry-within}
resonance_asymmetry_results <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_std, condition == cond)
  
  if (nrow(df_cond) < 100) next
  
  df_cond$author_role <- factor(df_cond$author_role, levels = c("author_1", "author_2"))
  contrasts(df_cond$author_role) <- contr.sum(2)
  
  m_resonance <- tryCatch({
    lmer(resonance ~ author_role + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(m_resonance)) {
    emm <- emmeans(m_resonance, specs = "author_role")
    contrast_result <- pairs(emm)
    contrast_df <- as.data.frame(contrast_result)
    
    resonance_asymmetry_results <- bind_rows(resonance_asymmetry_results, tibble(
      condition = cond,
      metric = "Resonance",
      author_1_mean = as.data.frame(emm)$emmean[1],
      author_2_mean = as.data.frame(emm)$emmean[2],
      difference = contrast_df$estimate[1],
      se = contrast_df$SE[1],
      t = contrast_df$t.ratio[1],
      p = contrast_df$p.value[1]
    ))
    
    cat("\n=== ", cond, ": Resonance Asymmetry ===\n", sep = "")
    print(summary(m_resonance))
  }
}

resonance_asymmetry_results %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    interpretation = case_when(
      p < 0.05 & difference > 0 ~ "Author 1 > Author 2",
      p < 0.05 & difference < 0 ~ "Author 2 > Author 1",
      TRUE ~ "No asymmetry"
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Resonance asymmetry by condition (Author 1 - Author 2)"
  )
```

## Summary of Within-Condition Asymmetry

```{r asymmetry-summary}
# Combine all asymmetry results
all_asymmetry <- bind_rows(
  novelty_asymmetry_results,
  transience_asymmetry_results,
  resonance_asymmetry_results
)

all_asymmetry %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  select(condition, metric, author_1_mean, author_2_mean, difference, se, p, sig) %>%
  knitr::kable(
    digits = 3,
    caption = "Summary: Asymmetry in novelty metrics by condition",
    col.names = c("Condition", "Metric", "Author 1", "Author 2", "Diff", "SE", "p", "")
  )
```

# Cross-Condition Comparison

## Test if Asymmetry Differs Across Conditions

Using **metric ~ author_role × condition** interactions to test whether the pattern of asymmetry (Author 1 vs Author 2 difference) varies across conditions.

```{r cross-condition-asymmetry}
# Prepare condition factor with sum coding
df_std$condition <- factor(df_std$condition)
contrasts(df_std$condition) <- contr.sum(length(levels(df_std$condition)))

# Novelty: author × condition interaction
model_novelty_full <- lmer(
  surprise ~ author_role * condition + (1 | conversation_id),
  data = df_std
)

cat("=== Novelty: Author × Condition Interaction ===\n")
print(summary(model_novelty_full))

cat("\n=== Novelty Means by Condition × Author ===\n")
emm_novelty <- emmeans(model_novelty_full, specs = c("condition", "author_role"))
print(emm_novelty)

cat("\n=== Asymmetry Within Each Condition ===\n")
asymmetry_by_cond <- pairs(emm_novelty, by = "condition")
print(asymmetry_by_cond)

cat("\n=== Does Asymmetry Differ Across Conditions? ===\n")
asymmetry_comparison <- contrast(emm_novelty, interaction = c("pairwise", "pairwise"))
print(asymmetry_comparison)
```

```{r transience-cross-condition}
# Transience: author × condition interaction
model_transience_full <- lmer(
  transience ~ author_role * condition + (1 | conversation_id),
  data = df_std
)

cat("=== Transience: Author × Condition Interaction ===\n")
print(summary(model_transience_full))

cat("\n=== Transience Asymmetry by Condition ===\n")
emm_transience <- emmeans(model_transience_full, specs = c("condition", "author_role"))
print(pairs(emm_transience, by = "condition"))
```

```{r resonance-cross-condition}
# Resonance: author × condition interaction
model_resonance_full <- lmer(
  resonance ~ author_role * condition + (1 | conversation_id),
  data = df_std
)

cat("=== Resonance: Author × Condition Interaction ===\n")
print(summary(model_resonance_full))

cat("\n=== Resonance Asymmetry by Condition ===\n")
emm_resonance <- emmeans(model_resonance_full, specs = c("condition", "author_role"))
print(pairs(emm_resonance, by = "condition"))
```

# Novelty-Resonance Relationship

## Does the Innovation Bias (Novelty → Resonance Slope) Differ by Author and Condition?

```{r innovation-bias-within-condition}
# For each condition, test surprise × author_role interaction on resonance
innovation_bias_results <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_long, condition == cond)
  
  if (nrow(df_cond) < 100) next
  
  df_cond$author_role <- factor(df_cond$author_role, levels = c("author_1", "author_2"))
  
  m_innovation <- tryCatch({
    lmer(resonance ~ surprise * author_role + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(m_innovation)) {
    slopes <- emtrends(m_innovation, specs = "author_role", var = "surprise")
    contrast_result <- pairs(slopes)
    contrast_df <- as.data.frame(contrast_result)
    slopes_df <- as.data.frame(slopes)
    
    innovation_bias_results <- bind_rows(innovation_bias_results, tibble(
      condition = cond,
      author_1_slope = slopes_df$surprise.trend[1],
      author_2_slope = slopes_df$surprise.trend[2],
      slope_diff = contrast_df$estimate[1],
      se = contrast_df$SE[1],
      t = contrast_df$t.ratio[1],
      p = contrast_df$p.value[1]
    ))
    
    cat("\n=== ", cond, ": Innovation Bias (Novelty → Resonance) ===\n", sep = "")
    cat("Slopes by author:\n")
    print(slopes)
    cat("\nSlope difference test:\n")
    print(contrast_result)
  }
}

innovation_bias_results %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    interpretation = case_when(
      p < 0.05 & slope_diff > 0 ~ "Author 1 slope > Author 2",
      p < 0.05 & slope_diff < 0 ~ "Author 2 slope > Author 1",
      TRUE ~ "No slope difference"
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Innovation bias (Novelty→Resonance slope) asymmetry by condition"
  )
```

## Cross-Condition Innovation Bias Comparison

```{r innovation-bias-cross-condition}
# Full three-way interaction: surprise × author × condition
model_innovation_full <- lmer(
  resonance ~ surprise * author_role * condition + (1 | conversation_id),
  data = df_std
)

cat("=== Full Model: Novelty × Author × Condition ===\n")
print(summary(model_innovation_full))

cat("\n=== Innovation Slopes by Condition × Author ===\n")
emm_innovation_slopes <- emtrends(model_innovation_full, 
                                   specs = c("condition", "author_role"),
                                   var = "surprise")
print(emm_innovation_slopes)

cat("\n=== Slope Asymmetry Within Each Condition ===\n")
print(pairs(emm_innovation_slopes, by = "condition"))
```

# Visualization

```{r distribution-plots, fig.width=14, fig.height=10}
# Distribution plots for all three metrics
p1 <- df_long %>%
  ggplot(aes(x = surprise, fill = author_role)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~condition, ncol = 3) +
  scale_fill_manual(values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"),
                    labels = c("Author 1", "Author 2")) +
  labs(title = "Novelty (Surprisal) Distribution", x = "Surprisal", y = "Density", fill = "Author")

p2 <- df_long %>%
  ggplot(aes(x = transience, fill = author_role)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~condition, ncol = 3) +
  scale_fill_manual(values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"),
                    labels = c("Author 1", "Author 2")) +
  labs(title = "Transience Distribution", x = "Transience", y = "Density", fill = "Author")

p3 <- df_long %>%
  ggplot(aes(x = resonance, fill = author_role)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~condition, ncol = 3) +
  scale_fill_manual(values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"),
                    labels = c("Author 1", "Author 2")) +
  labs(title = "Resonance Distribution", x = "Resonance", y = "Density", fill = "Author")

p1 / p2 / p3 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r asymmetry-barplot, fig.width=12, fig.height=8}
# Plot asymmetry across metrics and conditions
if (nrow(all_asymmetry) > 0) {
  all_asymmetry %>%
    mutate(
      metric = factor(metric, levels = c("Novelty", "Transience", "Resonance"))
    ) %>%
    ggplot(aes(x = condition, y = difference, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
                  width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    facet_wrap(~metric, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Novelty Metric Asymmetry by Condition",
      subtitle = "Difference: Author 1 - Author 2. Positive = Author 1 > Author 2",
      y = "Asymmetry (Author 1 - Author 2)",
      x = "Condition"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

```{r innovation-bias-plot, fig.width=14, fig.height=5}
# Innovation bias visualization
p1 <- df_long %>%
  drop_na(surprise, resonance) %>%
  ggplot(aes(x = surprise, y = resonance, color = author_role)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  facet_wrap(~condition, ncol = 3) +
  scale_color_manual(values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"),
                     labels = c("Author 1", "Author 2")) +
  labs(
    title = "Innovation Bias: Novelty → Resonance Relationship",
    subtitle = "Steeper slope = novel contributions have greater forward influence",
    x = "Novelty (Surprisal)",
    y = "Resonance",
    color = "Author"
  ) +
  theme(legend.position = "bottom")

p1
```

```{r innovation-slopes-barplot, fig.width=10, fig.height=5}
# Plot innovation bias slopes by condition and author
if (nrow(innovation_bias_results) > 0) {
  innovation_bias_results %>%
    pivot_longer(cols = c(author_1_slope, author_2_slope),
                 names_to = "author",
                 values_to = "slope") %>%
    mutate(author = if_else(author == "author_1_slope", "Author 1", "Author 2")) %>%
    ggplot(aes(x = condition, y = slope, fill = author)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = c("Author 1" = "#377EB8", "Author 2" = "#FF7F00")) +
    labs(
      title = "Innovation Bias Slopes by Condition and Author",
      subtitle = "Slope of Novelty → Resonance relationship",
      y = "Novelty → Resonance Slope (β)",
      x = "Condition",
      fill = "Author"
    ) +
    theme(legend.position = "bottom")
}
```

```{r summary-metrics-plot, fig.width=14, fig.height=10}
# Summary visualization: All metrics by condition and author (original format)
# This recreates the grid layout with authors as rows and metrics as columns

summary_data <- df_long %>%
  group_by(condition, author_role) %>%
  summarise(
    novelty_mean = mean(surprise, na.rm = TRUE),
    novelty_se = sd(surprise, na.rm = TRUE) / sqrt(n()),
    transience_mean = mean(transience, na.rm = TRUE),
    transience_se = sd(transience, na.rm = TRUE) / sqrt(n()),
    resonance_mean = mean(resonance, na.rm = TRUE),
    resonance_se = sd(resonance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Reshape for plotting using the original pivot approach
summary_long <- summary_data %>%
  pivot_longer(
    cols = ends_with("_mean") | ends_with("_se"),
    names_to = "var",
    values_to = "value"
  ) %>%
  separate(var, into = c("metric", "stat"), sep = "_(?=mean$|se$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(
    author = if_else(author_role == "author_1", "Author 1", "Author 2"),
    metric = str_to_title(metric)
  )

# Create summary bar plots with facet_grid (author rows × metric columns)
ggplot(summary_long, aes(x = condition, y = mean, fill = condition)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_errorbar(aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se), 
                width = 0.2, linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_grid(author ~ metric, scales = "free_y") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Summary: Novelty Metrics by Condition and Author",
    subtitle = "Error bars show 95% CI",
    x = "Condition",
    y = "Value"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Summary

```{r summary-table}
# Combine key results
if (nrow(all_asymmetry) > 0) {
  all_asymmetry %>%
    mutate(
      sig = case_when(
        p < 0.001 ~ "***",
        p < 0.01 ~ "**",
        p < 0.05 ~ "*",
        p < 0.1 ~ ".",
        TRUE ~ "ns"
      ),
      result = paste0(round(difference, 3), " (p=", format.pval(p, digits = 2), ") ", sig)
    ) %>%
    select(condition, metric, result) %>%
    pivot_wider(names_from = metric, values_from = result) %>%
    knitr::kable(
      caption = "Summary: Novelty metric asymmetry (Author 1 - Author 2) by condition"
    )
}
```

# Key Findings

Based on the analysis above:

1. **Within-condition asymmetry**: For each condition, we test whether novelty, transience, and resonance differ between Author 1 and Author 2. Significant asymmetry indicates one author contributes more novel/transient/resonant content.

2. **Cross-condition asymmetry comparison**: The author_role × condition interaction tests whether the *pattern* of asymmetry differs across conditions. 
   - If Human-AI shows Author 1 > Author 2 novelty but Human-Human shows no difference, the interaction is significant.
   - If all conditions show similar asymmetry patterns, the interaction is non-significant.

3. **Innovation bias asymmetry**: The novelty → resonance slope indicates whether novel contributions have lasting forward influence (innovation bias). Testing slope asymmetry reveals whether one author's innovations are more "rewarded" in narrative development.

4. **Interpretation for Human-AI**:
   - If humans (Author 1) show higher novelty than AI (Author 2), this indicates humans introduce more surprising content.
   - If humans also show higher transience, their contributions fade faster.
   - If humans show higher resonance despite higher transience, their net forward influence is still positive.
   - A steeper innovation slope for humans means their novel contributions disproportionately shape the story.

5. **Baseline comparison**: The AI-AI condition serves as a baseline for symmetric collaboration. Deviations in Human-AI or Human-Human patterns highlight condition-specific asymmetries.

---

*Analysis completed: `r Sys.time()`*
