---
title: "Rubber-Band Dynamics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Valence Gap and Re-alignment Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares **rubber-band dynamics** — the tendency for divergent valence to "snap back" toward alignment — across all three experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**The Rubber-Band Hypothesis:**
When there is a valence gap (emotional divergence) between co-authors, the subsequent turn should show a correction toward alignment. This is operationalized as:

- **Valence Gap**: Difference between Author 1 and Author 2 valence at turn t
- **Gap Change**: Change in valence gap from turn t to turn t+1
- **Rubber-Band Effect**: Negative correlation between gap and gap change (divergence predicts re-convergence)

**Key Questions:**
1. Is the rubber-band effect present in all conditions?
2. Is the effect stronger in Human-AI (AI accommodating) vs Human-Human (mutual accommodation)?
3. Does AI-AI show the strongest correction (most "compliant" dynamic)?
4. How does the correction direction differ across conditions?

# Load Data

```{r load-data}
# Load valence data using standardized loader
df_combined <- load_valence_data(conditions = c("human-ai", "human-human", "ai-ai"))
cat("Loaded conditions:", paste(unique(df_combined$condition), collapse = ", "), "\n")

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report data summary
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing: Valence Gap Computation

```{r preprocess}
# Compute valence gaps and lagged changes
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn) %>%
  mutate(
    # Current valence gap (Author 1 - Author 2)
    valence_gap = author_1_valence - author_2_valence,
    
    # Lagged gap values
    valence_gap_lag1 = lag(valence_gap, n = 1),
    
    # Gap change (how much the gap changed from previous turn)
    gap_change = valence_gap - valence_gap_lag1,
    
    # Alternative: Author 2's response to Author 1
    # Does Author 2 move toward Author 1's previous valence?
    author_1_valence_lag1 = lag(author_1_valence, n = 1),
    author_2_valence_lag1 = lag(author_2_valence, n = 1),
    
    # Author 2's adjustment
    author_2_adjustment = author_2_valence - author_1_valence_lag1,
    
    # Author 1's adjustment to Author 2's previous valence
    author_1_adjustment = author_1_valence - author_2_valence_lag1
  ) %>%
  ungroup() %>%
  filter(turn > 1) %>%
  drop_na(valence_gap, valence_gap_lag1)

cat("Analysis dataset:", nrow(df_analysis), "turns across", 
    n_distinct(df_analysis$conversation_id), "stories\n")
```

# Descriptive Statistics

```{r descriptives}
# Valence gap statistics by condition
desc_stats <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_turns = n(),
    mean_gap = mean(valence_gap, na.rm = TRUE),
    sd_gap = sd(valence_gap, na.rm = TRUE),
    mean_gap_change = mean(gap_change, na.rm = TRUE),
    sd_gap_change = sd(gap_change, na.rm = TRUE),
    # Rubber-band correlation (within-condition)
    rubber_band_r = cor(valence_gap_lag1, gap_change, use = "complete.obs"),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3,
    caption = "Valence gap statistics by condition",
    col.names = c("Condition", "N Turns", "Mean Gap", "SD Gap", 
                  "Mean Δ Gap", "SD Δ Gap", "Rubber-Band r")
  )
```

# Valence Gap Distribution

```{r gap-distribution, fig.width=12, fig.height=5}
# Gap distributions
p1 <- df_analysis %>%
  ggplot(aes(x = valence_gap, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Valence Gap Distribution",
    subtitle = "Author 1 - Author 2 valence",
    x = "Valence Gap",
    y = "Density"
  )

p2 <- df_analysis %>%
  ggplot(aes(x = condition, y = valence_gap, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Valence Gap by Condition",
    x = "Condition",
    y = "Valence Gap"
  ) +
  theme(legend.position = "none")

p1 + p2 + plot_layout(guides = "collect")
```

# Rubber-Band Effect Analysis

## Core Test: Gap → Gap Change

The rubber-band hypothesis predicts that **previous gap negatively predicts gap change** — when there's divergence, the system corrects toward alignment.

```{r rubber-band-scatter, fig.width=12, fig.height=5}
# Rubber-band scatter plot
df_analysis %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect by Condition",
    subtitle = "Negative slope = divergence predicts re-convergence",
    x = "Valence Gap (t-1)",
    y = "Gap Change (t)",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```

## Within-Condition Models

```{r within-condition-rubberband}
# Fit rubber-band model for each condition
conditions <- unique(df_analysis$condition)

results_within <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_analysis, condition == cond)
  
  if (nrow(df_cond) < 50) {
    cat("Skipping", cond, "- insufficient data\n")
    next
  }
  
  model <- tryCatch({
    lmer(gap_change ~ valence_gap_lag1 + (1 | conversation_id),
         data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(model)) {
    coefs <- summary(model)$coefficients
    results_within <- bind_rows(results_within, tibble(
      condition = cond,
      estimate = coefs["valence_gap_lag1", "Estimate"],
      se = coefs["valence_gap_lag1", "Std. Error"],
      t = coefs["valence_gap_lag1", "t value"],
      p = coefs["valence_gap_lag1", "Pr(>|t|)"]
    ))
  }
}

# Display results
results_within %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    interpretation = if_else(estimate < 0, "Rubber-band present", "No rubber-band")
  ) %>%
  select(condition, estimate, se, ci_lower, ci_upper, t, p, sig, interpretation) %>%
  knitr::kable(
    digits = 3,
    caption = "Rubber-band effect by condition (Gap_t-1 → ΔGap_t)"
  )
```

## Cross-Condition Comparison

```{r cross-condition-rubberband}
# Full model with condition interaction
model_cross <- lmer(
  gap_change ~ valence_gap_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Cross-Condition Rubber-Band Model ===\n")
print(summary(model_cross))

# Slopes by condition
emm_slopes <- emtrends(model_cross, specs = "condition", var = "valence_gap_lag1")
cat("\n=== Rubber-Band Slopes by Condition ===\n")
print(emm_slopes)

cat("\n=== Pairwise Slope Comparisons ===\n")
print(pairs(emm_slopes, adjust = "bonferroni"))
```

# Directional Analysis: Who Accommodates?

Beyond the symmetric rubber-band test, we can ask: **Who does the accommodating?**

- Does Author 2 adjust toward Author 1's previous valence?
- Does Author 1 adjust toward Author 2's previous valence?

```{r directional-accommodation, fig.width=14, fig.height=5}
# Author 2's accommodation to Author 1
p1 <- df_analysis %>%
  drop_na(author_1_valence_lag1, author_2_valence) %>%
  ggplot(aes(x = author_1_valence_lag1, y = author_2_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 2 Accommodation",
    subtitle = "Does Author 2 match Author 1's previous valence?",
    x = "Author 1 Valence (t-1)",
    y = "Author 2 Valence (t)",
    color = "Condition"
  )

# Author 1's accommodation to Author 2
p2 <- df_analysis %>%
  drop_na(author_2_valence_lag1, author_1_valence) %>%
  ggplot(aes(x = author_2_valence_lag1, y = author_1_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 1 Accommodation",
    subtitle = "Does Author 1 match Author 2's previous valence?",
    x = "Author 2 Valence (t-1)",
    y = "Author 1 Valence (t)",
    color = "Condition"
  )

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r directional-models}
# Author 2 → Author 1's previous valence
model_a2_to_a1 <- lmer(
  author_2_valence ~ author_1_valence_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Author 2 Accommodation (A2 → A1_lag) ===\n")
emm_a2_to_a1 <- emtrends(model_a2_to_a1, "condition", var = "author_1_valence_lag1")
print(emm_a2_to_a1)
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emm_a2_to_a1, adjust = "bonferroni"))

# Author 1 → Author 2's previous valence
model_a1_to_a2 <- lmer(
  author_1_valence ~ author_2_valence_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("\n=== Author 1 Accommodation (A1 → A2_lag) ===\n")
emm_a1_to_a2 <- emtrends(model_a1_to_a2, "condition", var = "author_2_valence_lag1")
print(emm_a1_to_a2)
cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emm_a1_to_a2, adjust = "bonferroni"))
```

# Asymmetry in Accommodation

```{r asymmetry-analysis}
# Compare who accommodates more within each condition
accommodation_results <- tibble()

for (cond in unique(df_analysis$condition)) {
  df_cond <- filter(df_analysis, condition == cond)
  
  # A2 → A1_lag
  m1 <- tryCatch({
    lmer(author_2_valence ~ author_1_valence_lag1 + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  # A1 → A2_lag
  m2 <- tryCatch({
    lmer(author_1_valence ~ author_2_valence_lag1 + (1 | conversation_id), data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(m1) && !is.null(m2)) {
    coef1 <- summary(m1)$coefficients["author_1_valence_lag1", "Estimate"]
    se1 <- summary(m1)$coefficients["author_1_valence_lag1", "Std. Error"]
    coef2 <- summary(m2)$coefficients["author_2_valence_lag1", "Estimate"]
    se2 <- summary(m2)$coefficients["author_2_valence_lag1", "Std. Error"]
    
    accommodation_results <- bind_rows(accommodation_results, tibble(
      condition = cond,
      a2_accommodates = coef1,
      a2_se = se1,
      a1_accommodates = coef2,
      a1_se = se2,
      diff = coef1 - coef2,
      diff_se = sqrt(se1^2 + se2^2),
      z = diff / diff_se,
      p = 2 * pnorm(-abs(z))
    ))
  }
}

accommodation_results %>%
  mutate(
    dominant_accommodator = case_when(
      p < 0.05 & diff > 0 ~ "Author 2 (AI/User2)",
      p < 0.05 & diff < 0 ~ "Author 1 (Human/User1)",
      TRUE ~ "Symmetric"
    )
  ) %>%
  select(condition, a2_accommodates, a1_accommodates, diff, z, p, dominant_accommodator) %>%
  knitr::kable(
    digits = 3,
    caption = "Accommodation asymmetry by condition",
    col.names = c("Condition", "A2 → A1", "A1 → A2", "Diff", "z", "p", "Dominant Accommodator")
  )
```

# Temporal Dynamics

How does the rubber-band effect evolve across the story?

```{r temporal-dynamics, fig.width=12, fig.height=5}
# Bin turns into early/middle/late
df_temporal <- df_analysis %>%
  group_by(conversation_id) %>%
  mutate(
    max_turn = max(turn),
    story_phase = case_when(
      turn <= max_turn / 3 ~ "Early",
      turn <= 2 * max_turn / 3 ~ "Middle",
      TRUE ~ "Late"
    ),
    story_phase = factor(story_phase, levels = c("Early", "Middle", "Late"))
  ) %>%
  ungroup()

# Plot rubber-band by phase
df_temporal %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.03, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~story_phase) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect Across Story Phases",
    x = "Valence Gap (t-1)",
    y = "Gap Change (t)",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```

# Summary Visualization

```{r summary-plot, fig.width=12, fig.height=8}
# Coefficient comparison plot
if (nrow(results_within) > 0) {
  p1 <- results_within %>%
    mutate(
      ci_lower = estimate - 1.96 * se,
      ci_upper = estimate + 1.96 * se
    ) %>%
    ggplot(aes(x = condition, y = estimate, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Rubber-Band Effect Strength",
      subtitle = "More negative = stronger re-alignment tendency",
      y = "Coefficient (β)",
      x = "Condition"
    ) +
    theme(legend.position = "none")
  
  # Accommodation comparison
  if (nrow(accommodation_results) > 0) {
    acc_long <- accommodation_results %>%
      select(condition, a2_accommodates, a1_accommodates, a2_se, a1_se) %>%
      pivot_longer(
        cols = c(a2_accommodates, a1_accommodates),
        names_to = "direction",
        values_to = "coefficient"
      ) %>%
      mutate(
        se = if_else(direction == "a2_accommodates", a2_se, a1_se),
        direction = if_else(direction == "a2_accommodates", 
                            "A2 → A1", "A1 → A2")
      )
    
    p2 <- ggplot(acc_long, aes(x = condition, y = coefficient, fill = direction)) +
      geom_col(position = position_dodge(width = 0.7), alpha = 0.8, width = 0.6) +
      geom_errorbar(
        aes(ymin = coefficient - 1.96 * se, ymax = coefficient + 1.96 * se),
        position = position_dodge(width = 0.7),
        width = 0.2
      ) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
      labs(
        title = "Directional Accommodation",
        subtitle = "Who follows whom?",
        y = "Accommodation Coefficient (β)",
        x = "Condition",
        fill = "Direction"
      ) +
      theme(legend.position = "bottom")
    
    p1 / p2
  } else {
    p1
  }
}
```

# Key Findings

Based on the analysis above:

1. **Rubber-band presence**: [Interpret whether the effect is present in all conditions]

2. **Effect strength comparison**: [Compare magnitude across Human-AI, Human-Human, AI-AI]

3. **Accommodation asymmetry**: [Discuss who does more accommodating in each condition]

4. **Temporal patterns**: [Note if the effect changes across story phases]

5. **AI-AI baseline**: [Compare simulated baseline to empirical conditions]

---

*Analysis completed: `r Sys.time()`*
