---
title: "Rubber-Band Dynamics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Valence Gap and Re-alignment Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares **rubber-band dynamics** — the tendency for divergent valence to "snap back" toward alignment — across all three experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**The Rubber-Band Hypothesis:**
When there is a valence gap (emotional divergence) between co-authors, the subsequent turn should show a correction toward alignment. This is operationalized as:

- **Valence Gap**: Difference between Author 1 and Author 2 valence at turn t
- **Gap Change**: Change in valence gap from turn t to turn t+1
- **Rubber-Band Effect**: Negative correlation between gap and gap change (divergence predicts re-convergence)

**Key Questions:**
1. Is the rubber-band effect present in all conditions?
2. Is the effect stronger in Human-AI (AI accommodating) vs Human-Human (mutual accommodation)?
3. Does AI-AI show the strongest correction (most "compliant" dynamic)?
4. How does the correction direction differ across conditions?
5. Could the rubber-band effect be an inherent mathematical phenomenon, rather than behavioural/psycholinguistical? 

# Load Data

```{r load-data}
# Load valence data using standardized loader
df_combined <- load_valence_data(conditions = c("human-ai", "human-human", "ai-ai"))
cat("Loaded conditions:", paste(unique(df_combined$condition), collapse = ", "), "\n")

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report data summary
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing: Valence Gap Computation

```{r preprocess}
# Compute valence gaps and lagged changes
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn) %>%
  mutate(
    # Current valence gap (Author 1 - Author 2)
    valence_gap = author_1_valence - author_2_valence,
    
    # Lagged gap values
    valence_gap_lag1 = lag(valence_gap, n = 1),
    
    # Gap change (how much the gap changed from previous turn)
    gap_change = valence_gap - valence_gap_lag1,
    
    # Alternative: Author 2's response to Author 1
    # Does Author 2 move toward Author 1's previous valence?
    author_1_valence_lag1 = lag(author_1_valence, n = 1),
    author_2_valence_lag1 = lag(author_2_valence, n = 1),
    
    # Author 2's adjustment
    author_2_adjustment = author_2_valence - author_1_valence_lag1,
    
    # Author 1's adjustment to Author 2's previous valence
    author_1_adjustment = author_1_valence - author_2_valence_lag1
  ) %>%
  ungroup() %>%
  filter(turn > 1) %>%
  drop_na(valence_gap, valence_gap_lag1)

cat("Analysis dataset:", nrow(df_analysis), "turns across", 
    n_distinct(df_analysis$conversation_id), "stories\n")
```

# Descriptive Statistics

```{r descriptives}
# Valence gap statistics by condition
desc_stats <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_turns = n(),
    mean_gap = mean(valence_gap, na.rm = TRUE),
    sd_gap = sd(valence_gap, na.rm = TRUE),
    mean_gap_change = mean(gap_change, na.rm = TRUE),
    sd_gap_change = sd(gap_change, na.rm = TRUE),
    # Rubber-band correlation (within-condition)
    rubber_band_r = cor(valence_gap_lag1, gap_change, use = "complete.obs"),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3,
    caption = "Valence gap statistics by condition",
    col.names = c("Condition", "N Turns", "Mean Gap", "SD Gap", 
                  "Mean Δ Gap", "SD Δ Gap", "Rubber-Band r")
  )
```

# Valence Gap Distribution

```{r gap-distribution, fig.width=12, fig.height=5}
# Gap distributions
p1 <- df_analysis %>%
  ggplot(aes(x = valence_gap, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Valence Gap Distribution",
    subtitle = "Author 1 - Author 2 valence",
    x = "Valence Gap",
    y = "Density"
  )

p2 <- df_analysis %>%
  ggplot(aes(x = condition, y = valence_gap, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Valence Gap by Condition",
    x = "Condition",
    y = "Valence Gap"
  ) +
  theme(legend.position = "none")

p1 + p2 + plot_layout(guides = "collect")
```

# Rubber-Band Effect Analysis

## Core Test: Gap → Gap Change

The rubber-band hypothesis predicts that **previous gap negatively predicts gap change** — when there's divergence, the system corrects toward alignment.

```{r rubber-band-scatter, fig.width=12, fig.height=5}
# Rubber-band scatter plot
df_analysis %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect by Condition",
    subtitle = "Negative slope = divergence predicts re-convergence",
    x = "Valence Gap (t-1)",
    y = "Gap Change (t)",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```

## Within-Condition Models

```{r within-condition-rubberband}
# Fit rubber-band model for each condition
conditions <- unique(df_analysis$condition)

results_within <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_analysis, condition == cond)
  
  if (nrow(df_cond) < 50) {
    cat("Skipping", cond, "- insufficient data\n")
    next
  }
  
  model <- tryCatch({
    lmer(gap_change ~ valence_gap_lag1 + (1 | conversation_id),
         data = df_cond)
  }, error = function(e) NULL)
  
  if (!is.null(model)) {
    coefs <- summary(model)$coefficients
    results_within <- bind_rows(results_within, tibble(
      condition = cond,
      estimate = coefs["valence_gap_lag1", "Estimate"],
      se = coefs["valence_gap_lag1", "Std. Error"],
      t = coefs["valence_gap_lag1", "t value"],
      p = coefs["valence_gap_lag1", "Pr(>|t|)"]
    ))
  }
}

# Display results
results_within %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se,
    interpretation = if_else(estimate < 0, "Rubber-band present", "No rubber-band")
  ) %>%
  select(condition, estimate, se, ci_lower, ci_upper, t, p, sig, interpretation) %>%
  knitr::kable(
    digits = 3,
    caption = "Rubber-band effect by condition (Gap_t-1 → ΔGap_t)"
  )
```

## Cross-Condition Comparison

```{r cross-condition-rubberband}
# Full model with condition interaction
model_cross <- lmer(
  gap_change ~ valence_gap_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Cross-Condition Rubber-Band Model ===\n")
print(summary(model_cross))

# Slopes by condition
emm_slopes <- emtrends(model_cross, specs = "condition", var = "valence_gap_lag1")
cat("\n=== Rubber-Band Slopes by Condition ===\n")
print(emm_slopes)

cat("\n=== Pairwise Slope Comparisons ===\n")
print(pairs(emm_slopes, adjust = "bonferroni"))
```



# Temporal Dynamics

How does the rubber-band effect evolve across the story?

```{r temporal-dynamics, fig.width=12, fig.height=5}
# Bin turns into early/middle/late
df_temporal <- df_analysis %>%
  group_by(conversation_id) %>%
  mutate(
    max_turn = max(turn),
    story_phase = case_when(
      turn <= max_turn / 3 ~ "Early",
      turn <= 2 * max_turn / 3 ~ "Middle",
      TRUE ~ "Late"
    ),
    story_phase = factor(story_phase, levels = c("Early", "Middle", "Late"))
  ) %>%
  ungroup()

# Plot rubber-band by phase
df_temporal %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.03, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~story_phase) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect Across Story Phases",
    x = "Valence Gap (t-1)",
    y = "Gap Change (t)",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```

# Summary Visualization

```{r summary-plot, fig.width=12, fig.height=8}
# Coefficient comparison plot
if (nrow(results_within) > 0) {
  p1 <- results_within %>%
    mutate(
      ci_lower = estimate - 1.96 * se,
      ci_upper = estimate + 1.96 * se
    ) %>%
    ggplot(aes(x = condition, y = estimate, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Rubber-Band Effect Strength",
      subtitle = "More negative = stronger re-alignment tendency",
      y = "Coefficient (β)",
      x = "Condition"
    ) +
    theme(legend.position = "none")
  
    
    p1
  }

```

# Key Findings

Based on the analysis above:

1. **Rubber-band presence**: [Interpret whether the effect is present in all conditions]

2. **Effect strength comparison**: [Compare magnitude across Human-AI, Human-Human, AI-AI]

4. **Temporal patterns**: [Note if the effect changes across story phases]

5. **AI-AI baseline**: [Compare simulated baseline to empirical conditions]

---

*Analysis completed: `r Sys.time()`*
