---
title: "Rubber-Band Dynamics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Valence Gap and Re-Alignment Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize
)

theme_set(theme_comparison())
```

# Overview

This notebook compares **rubber-band dynamics** across conditions using the exchange-aligned data structure.

The main quantity is the **valence gap**:

- `valence_gap = slot_1_valence - slot_2_valence`
- `gap_change = current_gap - previous_gap`

A negative slope in `gap_change ~ previous_gap` indicates a rubber-band effect: when the slots diverge, the next aligned exchange tends to pull them back together.

As in the updated valence notebook:

- `author_1` / `author_2` are treated as **slot 1 / slot 2**
- the analysis is restricted to **complete exchanges**
- the temporal window uses `analysis_turn` / `turn_index`, not the old raw-turn logic

**Primary cross-condition object.** For each story $i$, fit the OLS
$$\text{gap\_change}_{i,t} \;=\; \alpha_i + \beta_i \cdot \text{valence\_gap\_lag1}_{i,t} + \varepsilon_{i,t}$$
and define the per-story reversion strength as $R_i = -\beta_i$, so that larger positive $R_i$ means a stronger pull back together. $R_i$ is already label-invariant (the gap enters squared/reversal-style in the relationship). We test $R_i \sim \text{condition}$ with ANOVA + Kruskal and pairwise BH-adjusted Wilcoxon. The pooled LMM slopes are kept as supplemental.

**Article-use note.** This notebook tests a dyad-level regulation mechanism, not an agent-level asymmetry. If $R_i$ does not differ by condition, the conclusion is that emotional re-convergence is a general co-writing dynamic rather than a Human-AI-specific asymmetry.

# Load Data

```{r load-data}
df_combined <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded valence data for rubber-band analysis")
```

# Preprocessing

```{r preprocess}
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    valence_gap = author_1_valence - author_2_valence,
    valence_gap_lag1 = lag(valence_gap),
    gap_change = valence_gap - valence_gap_lag1,
    max_turn_index = max(turn_index, na.rm = TRUE),
    story_phase = case_when(
      turn_index <= max_turn_index / 3 ~ "Early",
      turn_index <= 2 * max_turn_index / 3 ~ "Middle",
      TRUE ~ "Late"
    ),
    story_phase = factor(story_phase, levels = c("Early", "Middle", "Late"))
  ) %>%
  ungroup() %>%
  filter(turn_index > 1) %>%
  drop_na(valence_gap_lag1, gap_change)

cat("Analysis rows:", nrow(df_analysis), "\n")
cat("Stories:", n_distinct(df_analysis$conversation_id), "\n")
```

# Primary: Per-Story Reversion Strength

```{r primary-reversion}
min_points <- 4

per_story_rb <- df_analysis %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_points = sum(!is.na(valence_gap_lag1) & !is.na(gap_change)),
    beta = if (n_points >= min_points) {
      coef(lm(gap_change ~ valence_gap_lag1))[2]
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(R = -beta) %>%
  drop_na(R)

per_story_rb %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_R = mean(R),
    median_R = median(R),
    sd_R = sd(R),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story reversion strength R_i = -beta by condition"
  )
```

```{r primary-reversion-test}
if (n_distinct(per_story_rb$condition) >= 2 && nrow(per_story_rb) >= 10) {
  cat("One-way ANOVA: R ~ condition\n")
  print(summary(aov(R ~ condition, data = per_story_rb)))

  cat("\nKruskal-Wallis (robust): R ~ condition\n")
  print(kruskal.test(R ~ condition, data = per_story_rb))

  cat("\nPairwise Wilcoxon (BH-adjusted):\n")
  print(pairwise.wilcox.test(
    per_story_rb$R,
    per_story_rb$condition,
    p.adjust.method = "BH"
  ))
}
```

```{r primary-reversion-plot, fig.width=9, fig.height=5}
ggplot(per_story_rb, aes(x = condition, y = R, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Per-story reversion strength by condition",
    subtitle = "R_i = -beta_i from gap_change ~ valence_gap_lag1 per story",
    x = "Condition",
    y = expression(R[i] == -beta[i])
  ) +
  theme(legend.position = "none")
```

# Bootstrap CIs and Effect Sizes on `R_i`

```{r boot-effects-rubber}
bootstrap_condition_means(per_story_rb, "R", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of reversion strength R_i (R = 10,000)"
  )

effect_sizes_pairwise(per_story_rb, "R") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on R_i"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-rubber}
planned_contrast_HA_vs_same(per_story_rb, "R") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: R_i in HA vs the average of HH and AA"
  )
```

# Time-Resolved Reversion Strength

```{r time-resolved-rubber, fig.width=12, fig.height=5}
df_phased <- df_analysis %>%
  drop_na(valence_gap_lag1, gap_change) %>%
  group_by(conversation_id, condition) %>%
  mutate(
    max_t = max(turn_index, na.rm = TRUE),
    phase = case_when(
      turn_index <= max_t / 3 ~ "Early",
      turn_index <= 2 * max_t / 3 ~ "Middle",
      TRUE ~ "Late"
    ),
    phase = factor(phase, levels = c("Early", "Middle", "Late"))
  ) %>%
  ungroup()

per_story_phase_rb <- df_phased %>%
  group_by(condition, conversation_id, phase) %>%
  summarise(
    n_points = sum(!is.na(valence_gap_lag1) & !is.na(gap_change)),
    beta = if (n_points >= 3) coef(lm(gap_change ~ valence_gap_lag1))[2] else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(R_phase = -beta) %>%
  drop_na(R_phase)

per_story_phase_rb %>%
  group_by(condition, phase) %>%
  summarise(
    n = n(),
    mean_R = mean(R_phase),
    sd_R = sd(R_phase),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story reversion strength by condition and story phase"
  )

if (nrow(per_story_phase_rb) >= 30 &&
    n_distinct(per_story_phase_rb$condition) >= 2 &&
    n_distinct(per_story_phase_rb$phase) >= 2) {
  cat("LMM: R_phase ~ condition * phase + (1 | conversation_id)\n")
  phase_model <- lmer(R_phase ~ condition * phase + (1 | conversation_id),
                      data = per_story_phase_rb)
  print(summary(phase_model))
}

ggplot(per_story_phase_rb, aes(x = phase, y = R_phase, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Reversion strength by story phase and condition",
    x = "Story phase",
    y = expression(R[phase])
  )
```

# Structural-Null Permutation Test

Motivation: the observed rubber-band slope is near-identical across HH, HA,
and AA, suggesting a structural / mean-reversion artifact of the
`gap_change ~ gap_lag1` parameterization rather than genuine coordination.
This test pools all turn-level (slot1, slot2) valence pairs across all
stories and conditions, randomly repartitions them into pseudo-stories of 10
turns, and recomputes per-pseudo-story `R = -beta`. If the observed mean R
falls inside the resulting null distribution, the rubber-band effect is
structural; if observed exceeds the null, it reflects genuine dyadic
coordination.

```{r structural-null-rubber, fig.width=10, fig.height=5}
# Pool turn-level valence pairs, ignoring story / condition.
df_pool <- df_combined %>%
  drop_na(author_1_valence, author_2_valence) %>%
  select(author_1_valence, author_2_valence)

null_obj <- null_rubber_band(df_pool, B = 1000, story_size = 10, seed = 42)

null_df <- tibble(null_mean_R = null_obj$null_means)

obs_R <- per_story_rb %>%
  group_by(condition) %>%
  summarise(observed_mean_R = mean(R, na.rm = TRUE), .groups = "drop")

obs_R_overall <- tibble(condition = "Pooled", observed_mean_R = mean(per_story_rb$R, na.rm = TRUE))

obs_table <- bind_rows(obs_R, obs_R_overall) %>%
  rowwise() %>%
  mutate(
    null_mean = mean(null_obj$null_means),
    null_sd = sd(null_obj$null_means),
    p_perm_two_sided = (sum(abs(null_obj$null_means - null_mean) >=
                            abs(observed_mean_R - null_mean)) + 1) /
                       (length(null_obj$null_means) + 1)
  ) %>%
  ungroup()

obs_table %>%
  knitr::kable(
    digits = 4,
    caption = sprintf(
      "Structural-null test: observed mean R per condition vs %d-iter null (pseudo-stories of 10 turns from %d pooled pairs)",
      null_obj$B, null_obj$pool_size
    )
  )

ggplot(null_df, aes(x = null_mean_R)) +
  geom_density(fill = "gray70", alpha = 0.5) +
  geom_vline(
    data = obs_table,
    aes(xintercept = observed_mean_R, color = condition),
    linewidth = 1
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Structural null vs observed reversion strength",
    subtitle = sprintf(
      "Density: B = %d pseudo-stories of 10 turns randomly assembled from pooled turn pairs",
      null_obj$B
    ),
    x = "Mean R across pseudo-stories",
    y = "Density",
    color = "Observed (condition)"
  )
```

# Supplemental: Pooled Condition-Specific Slopes

```{r within-condition-models}
within_results <- list()

for (cond in unique(df_analysis$condition)) {
  df_cond <- filter(df_analysis, condition == cond)

  if (nrow(df_cond) < 50) {
    next
  }

  model <- lmer(
    gap_change ~ valence_gap_lag1 + (1 | conversation_id),
    data = df_cond
  )

  coefs <- summary(model)$coefficients
  within_results[[as.character(cond)]] <- tibble(
    condition = cond,
    estimate = coefs["valence_gap_lag1", "Estimate"],
    se = coefs["valence_gap_lag1", "Std. Error"],
    t = coefs["valence_gap_lag1", "t value"],
    p = coefs["valence_gap_lag1", "Pr(>|t|)"]
  )
}

within_summary <- bind_rows(within_results)

within_summary %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Rubber-band slopes by condition"
  )
```

```{r slope-plot, fig.width=10, fig.height=5}
within_summary %>%
  mutate(
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se
  ) %>%
  ggplot(aes(x = condition, y = estimate, fill = condition)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect by Condition",
    subtitle = "More negative slopes indicate stronger re-convergence after divergence",
    x = "Condition",
    y = "Rubber-Band Slope"
  ) +
  theme(legend.position = "none")
```

## Supplemental: Cross-Condition Interaction Model

```{r cross-condition-model}
model_cross <- lmer(
  gap_change ~ valence_gap_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("Cross-condition model summary:\n")
print(summary(model_cross))

emm_slopes <- emtrends(model_cross, specs = "condition", var = "valence_gap_lag1")

cat("\nCondition-specific slopes:\n")
print(emm_slopes)

cat("\nPairwise slope contrasts:\n")
print(pairs(emm_slopes, adjust = "bonferroni"))
```

# Supplemental: General Gap Structure

```{r rubberband-scatter, fig.width=12, fig.height=5}
df_analysis %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Effect by Condition",
    subtitle = "Negative slope indicates re-convergence after divergence",
    x = "Previous Valence Gap",
    y = "Gap Change"
  ) +
  theme(legend.position = "bottom")
```

```{r descriptives}
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_turns = n(),
    mean_gap = mean(valence_gap, na.rm = TRUE),
    sd_gap = sd(valence_gap, na.rm = TRUE),
    mean_gap_change = mean(gap_change, na.rm = TRUE),
    sd_gap_change = sd(gap_change, na.rm = TRUE),
    rubber_band_r = cor(valence_gap_lag1, gap_change, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Gap and gap-change descriptives by condition"
  )
```

```{r gap-distributions, fig.width=12, fig.height=5}
p_gap_density <- df_analysis %>%
  ggplot(aes(x = valence_gap, fill = condition)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Valence Gap Distribution",
    subtitle = "Slot 1 minus Slot 2",
    x = "Valence Gap",
    y = "Density"
  ) +
  theme(legend.position = "none")

p_gap_box <- df_analysis %>%
  ggplot(aes(x = condition, y = valence_gap, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Valence Gap by Condition",
    x = "Condition",
    y = "Valence Gap"
  ) +
  theme(legend.position = "none")

together <- p_gap_density + p_gap_box

ggsave(
  file.path(GRAPHS_DIR, "valence_gap.png"),
  together,
  width = 12,
  height = 8,
  dpi = 300
)
```

```{r temporal-dynamics, fig.width=12, fig.height=5}
df_analysis %>%
  ggplot(aes(x = valence_gap_lag1, y = gap_change, color = condition)) +
  geom_point(alpha = 0.03, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~story_phase) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Rubber-Band Dynamics Across Story Phases",
    x = "Previous Valence Gap",
    y = "Gap Change"
  ) +
  theme(legend.position = "bottom")
```

# Summary

Primary cross-condition result: per-story reversion strength $R_i = -\beta_i$ from `gap_change ~ valence_gap_lag1`, tested with ANOVA + Kruskal and pairwise BH-adjusted Wilcoxon.

Robustness layers added:

- bootstrap 95% CIs on per-condition mean of $R_i$
- pairwise Cliff's $\delta$ effect sizes
- planned single-df contrast HA vs (HH+AA)/2
- time-resolved view of $R$ across early / middle / late story phases
- **structural-null permutation test**: shuffles all turn-level valence pairs across stories and conditions, repartitions into pseudo-stories, and compares observed mean $R$ to the resulting null. Tests whether the rubber-band effect is a coordination dynamic or a mathematical / mean-reversion artifact

Supplemental views retained:

- pooled within-condition LMM slopes (population-level rubber-band estimate per condition)
- cross-condition interaction LMM (`gap_change ~ valence_gap_lag1 * condition + (1|story)`)
- gap distributions and story-phase splits

The Human-AI rubber-band decomposition lives in [analysis/human-ai/rubber_band_analysis.Rmd](../human-ai/rubber_band_analysis.Rmd) and is not duplicated here.

Deferred to a follow-up pass:

- starter-mechanism analysis of the reversion object

*Analysis completed: `r Sys.time()`*
