---
title: "Semantic Exploration Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Analysis of Lag-Based Semantic Distance"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans
)

theme_set(theme_comparison())
```

# Overview

The semantic exploration outputs now use the **current lag-based structure**:

- `k`: lag
- `distance`: mean semantic distance for that lag
- `agent`: `"interleaved"` for the full exchange sequence, or `author_1` / `author_2` for speaker-specific trajectories

The processed filename is still `semantic_exploration_binned.parquet`, but the current generator computes **lag-based distances**, not the older non-overlapping window-centroid bins.

This notebook therefore no longer assumes old turn-level exploration tables. Instead it analyses:

1. **Interleaved sequence exploration** across conditions
2. **Speaker-specific exploration** using recovered role metadata

Important comparison note:

- **Slot 1 / Slot 2** is the directly comparable cross-condition axis.
- **Human / AI** in Human-AI and **Starter-side / Non-starter-side** in same-type conditions are only used as within-condition descriptive views.

**Primary cross-condition object.** Per story, we compute the trapezoidal area under each slot's mean $\text{distance}(k)$ curve over the intersection of $k$ values present for both slots, and take the unsigned difference:
$$A_{\text{semexp}}(i) \;=\; \bigl|\, \mathrm{AUC}_{\text{slot}=1,\,i} - \mathrm{AUC}_{\text{slot}=2,\,i} \,\bigr|.$$
This is label-invariant and comparable across conditions. The lag-binned LMM slot contrast and the log-distance slope view are retained as supplemental. Starter-mechanism and Human-AI speaker-type decompositions are deferred.

**Article-use note.** The primary claim should come from the per-story AUC asymmetry test. The interleaved-distance model answers a different question: whether the whole story sequence explores semantic space differently by condition, not whether the two contributors are more asymmetric.

# Load Data

```{r load-data}
df_combined <- load_exploration_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter(k > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_combined %>%
  group_by(condition, agent) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    min_k = min(k, na.rm = TRUE),
    max_k = max(k, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded semantic exploration rows by condition and agent view")
```

# Preprocessing

```{r preprocess}
df_interleaved <- df_combined %>%
  filter(agent == "interleaved") %>%
  drop_na(distance, k)

df_speaker <- df_combined %>%
  filter(agent != "interleaved") %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(
    slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  ) %>%
  drop_na(distance, k)

cat("Interleaved rows:", nrow(df_interleaved), "\n")
cat("Speaker rows:", nrow(df_speaker), "\n")
```

# Primary: Per-Story AUC Asymmetry

```{r primary-auc-asymmetry}
# Aggregate to one distance value per (story, slot, k) first
df_slot_k <- df_speaker %>%
  group_by(condition, conversation_id, slot, k) %>%
  summarise(distance = mean(distance, na.rm = TRUE), .groups = "drop") %>%
  drop_na(distance)

# Trapezoidal AUC over the k range that is populated for BOTH slots within each story
trap_auc <- function(x, y) {
  if (length(x) < 2) return(NA_real_)
  ord <- order(x)
  x <- x[ord]; y <- y[ord]
  sum((y[-1] + y[-length(y)]) * diff(x) / 2)
}

df_auc <- df_slot_k %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(k_set = list(k), .groups = "drop") %>%
  pivot_wider(names_from = slot, values_from = k_set) %>%
  rowwise() %>%
  mutate(
    k_common = list(sort(intersect(`Slot 1`, `Slot 2`)))
  ) %>%
  ungroup() %>%
  select(condition, conversation_id, k_common) %>%
  # keep stories with enough shared k to integrate
  filter(lengths(k_common) >= 2)

df_auc_stories <- df_slot_k %>%
  inner_join(df_auc, by = c("condition", "conversation_id")) %>%
  rowwise() %>%
  filter(k %in% k_common) %>%
  ungroup() %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(
    auc = trap_auc(k, distance),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = slot, values_from = auc, names_prefix = "auc_") %>%
  drop_na(`auc_Slot 1`, `auc_Slot 2`) %>%
  mutate(
    signed_diff = `auc_Slot 1` - `auc_Slot 2`,
    A_semexp = abs(signed_diff)
  )

stopifnot(all(df_auc_stories$A_semexp >= 0))

df_auc_stories %>%
  count(condition, name = "n_stories") %>%
  knitr::kable(caption = "Stories with per-story AUC computable in both slots")

df_auc_stories %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_A = mean(A_semexp),
    sd_A = sd(A_semexp),
    median_A = median(A_semexp),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story |AUC(slot 1) - AUC(slot 2)| by condition"
  )
```

```{r primary-auc-test}
if (n_distinct(df_auc_stories$condition) >= 2 && nrow(df_auc_stories) >= 10) {
  cat("One-way ANOVA: A_semexp ~ condition\n")
  print(summary(aov(A_semexp ~ condition, data = df_auc_stories)))

  cat("\nKruskal-Wallis (robust): A_semexp ~ condition\n")
  print(kruskal.test(A_semexp ~ condition, data = df_auc_stories))

  cat("\nPairwise Wilcoxon (BH-adjusted):\n")
  print(pairwise.wilcox.test(
    df_auc_stories$A_semexp,
    df_auc_stories$condition,
    p.adjust.method = "BH"
  ))
}
```

```{r primary-auc-plot, fig.width=9, fig.height=5}
ggplot(df_auc_stories, aes(x = condition, y = A_semexp, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Per-story semantic-exploration AUC asymmetry by condition",
    subtitle = "A_semexp(i) = |AUC_{slot 1}(i) - AUC_{slot 2}(i)|",
    x = "Condition",
    y = expression("|" * Delta * " AUC|")
  ) +
  theme(legend.position = "none")
```

```{r primary-logk-slope-supplement}
# Secondary dynamic object: per-story |slope_log(k), slot 1 - slope_log(k), slot 2|
df_logk <- df_slot_k %>%
  filter(k > 0, distance > 0) %>%
  mutate(log_k = log(k), log_distance = log(distance))

df_logk_slopes <- df_logk %>%
  group_by(condition, conversation_id, slot) %>%
  summarise(
    n_points = n(),
    slope = if (n_points >= 3) coef(lm(log_distance ~ log_k))[2] else NA_real_,
    .groups = "drop"
  ) %>%
  select(condition, conversation_id, slot, slope) %>%
  pivot_wider(names_from = slot, values_from = slope, names_prefix = "slope_") %>%
  drop_na(`slope_Slot 1`, `slope_Slot 2`) %>%
  mutate(A_logk_slope = abs(`slope_Slot 1` - `slope_Slot 2`))

if (nrow(df_logk_slopes) > 0) {
  df_logk_slopes %>%
    group_by(condition) %>%
    summarise(
      n = n(),
      mean_A = mean(A_logk_slope),
      median_A = median(A_logk_slope),
      .groups = "drop"
    ) %>%
    knitr::kable(
      digits = 3,
      caption = "Secondary: per-story |log(distance) ~ log(k) slope diff| by condition"
    )
}
```

# Bootstrap CIs and Effect Sizes on `A_semexp`

```{r boot-effects-semexp}
bootstrap_condition_means(df_auc_stories, "A_semexp", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of A_semexp (R = 10,000)"
  )

effect_sizes_pairwise(df_auc_stories, "A_semexp") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on A_semexp"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-semexp}
planned_contrast_HA_vs_same(df_auc_stories, "A_semexp") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: A_semexp in HA vs the average of HH and AA"
  )
```

# Directional Stability of Signed Semantic-Exploration AUC Asymmetry

Tests whether the *direction* of the slot-1 vs slot-2 AUC difference is
consistent across stories within each condition. The signed quantity is
$\text{signed}_i = \mathrm{AUC}_{\text{slot 1},i} - \mathrm{AUC}_{\text{slot 2},i}$.

```{r directional-stability-semexp}
ds_sem <- directional_stability(df_auc_stories, "signed_diff")
ds_sem$by_condition %>%
  knitr::kable(
    digits = 3,
    caption = "Per-condition directional stability of signed AUC asymmetry"
  )

cat(sprintf(
  "Brown-Forsythe variance-equality p = %.4f\n", ds_sem$variance_equality_p
))
```

```{r directional-stability-semexp-plot, fig.width=9, fig.height=5}
ggplot(df_auc_stories, aes(x = condition, y = signed_diff, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Signed AUC asymmetry by condition",
    subtitle = "AUC(slot 1) - AUC(slot 2) per story",
    x = "Condition",
    y = expression(Delta * AUC)
  ) +
  theme(legend.position = "none")
```

# Supplemental: Speaker-Level Slot Trajectories

```{r speaker-plot, fig.width=12, fig.height=6}
speaker_summary <- df_speaker %>%
  group_by(condition, slot, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(speaker_summary, aes(x = k, y = mean_distance, color = slot)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = slot),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Slot",
    subtitle = "Underlying slot trajectories behind the primary AUC object",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Slot",
    fill = "Slot"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: General Condition-Level Comparison

```{r interleaved-descriptives}
df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Interleaved semantic distance by lag and condition"
  )
```

```{r interleaved-plot, fig.width=12, fig.height=5}
interleaved_summary <- df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(interleaved_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition),
    alpha = 0.18,
    color = NA
  ) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Interleaved Semantic Exploration by Lag",
    subtitle = "Mean distance with 95% confidence intervals",
    x = "Lag Bin (k)",
    y = "Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

```{r interleaved-model}
model_interleaved <- lmer(
  distance ~ k * condition + (1 | conversation_id),
  data = df_interleaved
)

cat("Interleaved model summary:\n")
print(summary(model_interleaved))

emm_interleaved <- emmeans(model_interleaved, specs = "condition")

cat("\nCondition contrasts (interleaved view):\n")
print(pairs(emm_interleaved, adjust = "bonferroni"))
```

# Supplemental: Role-Level View

```{r speaker-descriptives}
df_speaker %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Speaker-level semantic distance by condition, role, and lag"
  )
```

```{r role-plot, fig.width=12, fig.height=6}
role_summary <- df_speaker %>%
  filter(!is.na(condition_role)) %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(role_summary, aes(x = k, y = mean_distance, color = condition_role)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition_role),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Role",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Role",
    fill = "Role"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: Interleaved vs Speaker View

```{r view-comparison, fig.width=12, fig.height=5}
view_summary <- bind_rows(
  df_interleaved %>%
    mutate(view = "Interleaved sequence") %>%
    select(condition, conversation_id, k, distance, view),
  df_speaker %>%
    mutate(view = "Speaker-specific") %>%
    select(condition, conversation_id, k, distance, view)
) %>%
  group_by(condition, view, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(view_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  facet_wrap(~view) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Interleaved vs Speaker-Specific Semantic Distance",
    x = "Lag Bin (k)",
    y = "Mean Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

# Summary

Primary cross-condition result: per-story $A_{\text{semexp}} = |\mathrm{AUC}_{\text{slot 1}} - \mathrm{AUC}_{\text{slot 2}}|$ from trapezoidal integration of the distance(k) curve over the intersected k range. Tested with ANOVA + Kruskal and pairwise BH-adjusted Wilcoxon. A secondary log-distance-slope asymmetry is reported for dynamic-shape continuity with the paper.

Robustness layers added:

- bootstrap 95% CIs on per-condition mean of $A_{\text{semexp}}$
- pairwise Cliff's $\delta$ effect sizes
- planned single-df contrast HA vs (HH+AA)/2

Supplemental views retained:

- speaker-level slot trajectories (descriptive, both slots side-by-side without signed contrast)
- interleaved sequence model and condition descriptives
- within-condition role curves (descriptive only; HA Human/AI inferential view lives in [analysis/human-ai/exploration.Rmd](../human-ai/exploration.Rmd))
- interleaved vs speaker-specific side-by-side view

Deferred to a follow-up pass:

- starter-mechanism analysis (signed, starter-normalized AUC asymmetry across all conditions)

*Analysis completed: `r Sys.time()`*
