---
title: "Semantic Exploration Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Analysis of Lag-Based Semantic Distance"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans
)

theme_set(theme_comparison())
```

# Overview

The semantic exploration outputs now use the **current lag-based structure**:

- `k`: lag
- `distance`: mean semantic distance for that lag
- `agent`: `"interleaved"` for the full exchange sequence, or `author_1` / `author_2` for speaker-specific trajectories

The processed filename is still `semantic_exploration_binned.parquet`, but the current generator computes **lag-based distances**, not the older non-overlapping window-centroid bins.

This notebook therefore no longer assumes old turn-level exploration tables. Instead it analyses:

1. **Interleaved sequence exploration** across conditions
2. **Speaker-specific exploration** using recovered role metadata

Important comparison note:

- **Slot 1 / Slot 2** is the directly comparable cross-condition axis.
- **Human / AI** in Human-AI and **Starter-side / Non-starter-side** in same-type conditions are only used as within-condition descriptive views.

# Load Data

```{r load-data}
df_combined <- load_exploration_data(c("human-ai", "human-human", "ai-ai")) %>%
  prepare_condition_factor(reference = "human-ai")

df_combined %>%
  group_by(condition, agent) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    min_k = min(k, na.rm = TRUE),
    max_k = max(k, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded semantic exploration rows by condition and agent view")
```

# Preprocessing

```{r preprocess}
df_interleaved <- df_combined %>%
  filter(agent == "interleaved") %>%
  drop_na(distance, k)

df_speaker <- df_combined %>%
  filter(agent != "interleaved") %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(
    slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  ) %>%
  drop_na(distance, k)

cat("Interleaved rows:", nrow(df_interleaved), "\n")
cat("Speaker rows:", nrow(df_speaker), "\n")
```

# Asymmetry Comparison

```{r speaker-asymmetry}
speaker_asymmetry_model <- lmer(
  distance ~ k + slot * condition + (1 | conversation_id),
  data = df_speaker
)

cat("Speaker-slot asymmetry model:\n")
print(summary(speaker_asymmetry_model))

speaker_asymmetry_emm <- emmeans(
  speaker_asymmetry_model,
  specs = c("condition", "slot")
)

speaker_asymmetry <- as.data.frame(pairs(speaker_asymmetry_emm, by = "condition")) %>%
  transmute(
    condition,
    difference = estimate,
    se = SE,
    t = t.ratio,
    p = p.value
  )

speaker_asymmetry %>%
  knitr::kable(
    digits = 3,
    caption = "Speaker-slot asymmetry in semantic exploration by condition"
  )
```

```{r speaker-asymmetry-plot, fig.width=10, fig.height=5}
speaker_asymmetry %>%
  ggplot(aes(x = condition, y = difference, fill = condition)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
    width = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Speaker-Slot Asymmetry by Condition",
    subtitle = "Difference: Slot 1 - Slot 2. This is the primary cross-condition asymmetry comparison.",
    x = "Condition",
    y = "Asymmetry in Semantic Distance"
  ) +
  theme(legend.position = "none")
```

```{r speaker-plot, fig.width=12, fig.height=6}
speaker_summary <- df_speaker %>%
  group_by(condition, slot, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(speaker_summary, aes(x = k, y = mean_distance, color = slot)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = slot),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Slot",
    subtitle = "Slot trajectories shown first because asymmetry is the primary comparison target",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Slot",
    fill = "Slot"
  ) +
  theme(legend.position = "bottom")
```

# General Condition-Level Comparison

```{r interleaved-descriptives}
df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Interleaved semantic distance by lag and condition"
  )
```

```{r interleaved-plot, fig.width=12, fig.height=5}
interleaved_summary <- df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(interleaved_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition),
    alpha = 0.18,
    color = NA
  ) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Interleaved Semantic Exploration by Lag",
    subtitle = "Mean distance with 95% confidence intervals",
    x = "Lag Bin (k)",
    y = "Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

```{r interleaved-model}
model_interleaved <- lmer(
  distance ~ k * condition + (1 | conversation_id),
  data = df_interleaved
)

cat("Interleaved model summary:\n")
print(summary(model_interleaved))

emm_interleaved <- emmeans(model_interleaved, specs = "condition")

cat("\nCondition contrasts (interleaved view):\n")
print(pairs(emm_interleaved, adjust = "bonferroni"))
```

# Role-Level View

```{r speaker-descriptives}
df_speaker %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Speaker-level semantic distance by condition, role, and lag"
  )
```

```{r role-plot, fig.width=12, fig.height=6}
role_summary <- df_speaker %>%
  filter(!is.na(condition_role)) %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(role_summary, aes(x = k, y = mean_distance, color = condition_role)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition_role),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Role",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Role",
    fill = "Role"
  ) +
  theme(legend.position = "bottom")
```

# Interleaved vs Speaker View

```{r view-comparison, fig.width=12, fig.height=5}
view_summary <- bind_rows(
  df_interleaved %>%
    mutate(view = "Interleaved sequence") %>%
    select(condition, conversation_id, k, distance, view),
  df_speaker %>%
    mutate(view = "Speaker-specific") %>%
    select(condition, conversation_id, k, distance, view)
) %>%
  group_by(condition, view, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(view_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  facet_wrap(~view) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Interleaved vs Speaker-Specific Semantic Distance",
    x = "Lag Bin (k)",
    y = "Mean Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

# Summary

This notebook is now adapted to the new semantic exploration output by:

- reading the current `semantic_exploration_binned.parquet` files,
- modelling **lag bin `k`** instead of the obsolete raw-turn structure,
- putting the **speaker-slot asymmetry comparison first**,
- separating **interleaved** and **speaker-specific** exploration views,
- using recovered speaker-role metadata where that interpretation is valid.

*Analysis completed: `r Sys.time()`*
