---
title: "Semantic Exploration Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Analysis of Adjacent Semantic Distance"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans
)

theme_set(theme_comparison())
```

# Overview

The semantic exploration outputs contain the **current lag-based structure**:

- `k`: lag
- `distance`: mean semantic distance for that lag
- `agent`: `"interleaved"` for the full exchange sequence, or `author_1` / `author_2` for speaker-specific trajectories

The processed filename is still `semantic_exploration_binned.parquet`, but the current generator computes **lag-based distances**, not the older non-overlapping window-centroid bins.

This notebook uses the raw interaction embeddings to reconstruct the alternating dyadic sequence and analyse:

1. **Adjacent turn-to-turn semantic distance** across conditions as the primary analysis
2. **Story-level mean adjacent distance** as the supporting aggregate
3. **Turn-level adjacent semantic-distance networks** as descriptive 2D maps
4. **Lag-based trajectories** as supplemental descriptive context

Important comparison note:

- The adjacent sequence includes both within-exchange steps (e.g., slot 1$_t$ to slot 2$_t$) and between-exchange steps (e.g., slot 2$_t$ to slot 1$_{t+1}$).
- For Human-AI, the sequence is kept in the standard human-to-AI row order. For same-type dyads, `starter_side` determines which slot speaks first within each exchange.
- **Human / AI** in Human-AI and **Starter-side / Non-starter-side** in same-type conditions are only used as descriptive views.

**Primary cross-condition object.** Each adjacent pair in the reconstructed alternating sequence contributes one semantic distance. The main model compares these adjacent distances by condition while nesting observations within story:
$$d_{ij} = \beta_0 + \beta_1 \mathrm{condition}_{ij} + u_{\mathrm{story}(j)} + \epsilon_{ij}.$$

**Article-use note.** This analysis answers whether adjacent turns move through semantic space differently by condition. It is not a directional slot-asymmetry test.

**Network-map note.** The 2D semantic-network layouts are precomputed by `scripts/09_compute_semantic_network_layouts.py`; this notebook only reads and visualizes those files.
By default, the network script samples the same number of stories and adjacent edges per condition for visualization only, so panels are not visually dominated by sample size.

# Load Data

```{r load-data}
excluded_conversations <- c(
  "conv_ed575a06c11d42358e3eeb7826d2f959",
  "conv_a63a08273d0a4704a7638e4cd6850225"
)

df_combined <- load_exploration_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter(k > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(!conversation_id %in% excluded_conversations)

load_embedding_data <- function(conditions = c("human-ai", "human-human", "ai-ai")) {
  conditions %>%
    map(function(cond) {
      df <- load_condition_data(cond, "story_embeddings_interaction_level.parquet")
      if (is.null(df)) return(NULL)

      df %>%
        normalize_schema(cond) %>%
        attach_interaction_metadata(cond) %>%
        ensure_role_metadata() %>%
        filter_complete_exchange_window() %>%
        drop_provider_error_stories()
    }) %>%
    compact() %>%
    bind_rows() %>%
    filter(!conversation_id %in% excluded_conversations) %>%
    prepare_condition_factor(reference = "human-ai")
}

df_embeddings <- load_embedding_data()

df_combined %>%
  group_by(condition, agent) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    min_k = min(k, na.rm = TRUE),
    max_k = max(k, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded semantic exploration rows by condition and agent view")
```

# Preprocessing

```{r preprocess}
df_interleaved <- df_combined %>%
  filter(agent == "interleaved") %>%
  drop_na(distance, k)

df_speaker <- df_combined %>%
  filter(agent != "interleaved") %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(
    slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  ) %>%
  drop_na(distance, k)

cat("Interleaved rows:", nrow(df_interleaved), "\n")
cat("Speaker rows:", nrow(df_speaker), "\n")
cat("Embedding rows:", nrow(df_embeddings), "\n")
```

# Primary: Adjacent Turn-to-Turn Semantic Distance

```{r adjacent-distance-functions}
parse_embedding_r <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  if (is.character(x)) {
    x <- gsub("^\\[|\\]$", "", x)
    out <- suppressWarnings(as.numeric(strsplit(x, ",")[[1]]))
  } else {
    out <- suppressWarnings(as.numeric(unlist(x, use.names = FALSE)))
  }
  if (length(out) == 0 || any(!is.finite(out))) return(NULL)
  out
}

cosine_distance <- function(a, b) {
  if (is.null(a) || is.null(b) || length(a) != length(b)) return(NA_real_)
  denom <- sqrt(sum(a * a)) * sqrt(sum(b * b))
  if (!is.finite(denom) || denom == 0) return(NA_real_)
  sim <- sum(a * b) / denom
  1 - pmax(pmin(sim, 1), -1)
}

story_slot_order <- function(story_df) {
  condition_value <- as.character(story_df$condition[1])
  starter_values <- na.omit(as.character(story_df$starter_side))
  starter_side <- if (length(starter_values) > 0) starter_values[1] else NA_character_

  if (condition_value == "Human-AI") {
    c("author_1", "author_2")
  } else if (!is.na(starter_side) && starter_side == "author_2") {
    c("author_2", "author_1")
  } else {
    c("author_1", "author_2")
  }
}

build_story_sequence <- function(story_df) {
  story_df <- story_df %>%
    arrange(turn_index)

  slot_order <- story_slot_order(story_df)

  map_dfr(seq_len(nrow(story_df)), function(i) {
    row_i <- story_df[i, ]
    first_slot <- slot_order[1]
    second_slot <- slot_order[2]

    tibble(
      sequence_index = c(2 * i - 1, 2 * i),
      turn_index = c(row_i$turn_index, row_i$turn_index),
      slot = c(first_slot, second_slot),
      embedding = list(
        parse_embedding_r(row_i[[paste0(first_slot, "_embedding")]][[1]]),
        parse_embedding_r(row_i[[paste0(second_slot, "_embedding")]][[1]])
      )
    )
  }) %>%
    filter(map_lgl(embedding, ~ !is.null(.x)))
}

build_story_adjacent_distances <- function(story_df) {
  sequence_df <- build_story_sequence(story_df)

  if (nrow(sequence_df) < 2) {
    return(tibble())
  }

  n_seq <- nrow(sequence_df)

  tibble(
    pair_index = seq_len(n_seq - 1),
    from_turn = sequence_df$turn_index[-n_seq],
    to_turn = sequence_df$turn_index[-1],
    from_slot = sequence_df$slot[-n_seq],
    to_slot = sequence_df$slot[-1],
    pair_type = if_else(from_turn == to_turn, "within_exchange", "between_exchange"),
    adjacent_distance = map2_dbl(
      sequence_df$embedding[-n_seq],
      sequence_df$embedding[-1],
      cosine_distance
    )
  ) %>%
    filter(is.finite(adjacent_distance))
}

df_adjacent <- df_embeddings %>%
  group_by(condition, conversation_id) %>%
  group_modify(~ build_story_adjacent_distances(.x), .keep = TRUE) %>%
  ungroup() %>%
  mutate(
    pair_type = factor(pair_type, levels = c("within_exchange", "between_exchange"))
  )

df_adjacent %>%
  group_by(condition) %>%
  summarise(
    n_adjacent_pairs = n(),
    n_stories = n_distinct(conversation_id),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Adjacent semantic-distance observations by condition")

df_adjacent %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    n_stories = n_distinct(conversation_id),
    mean_distance = mean(adjacent_distance),
    sd_distance = sd(adjacent_distance),
    median_distance = median(adjacent_distance),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Adjacent turn-to-turn semantic distance by condition"
  )
```

```{r adjacent-distance-model}
if (n_distinct(df_adjacent$condition) >= 2 && nrow(df_adjacent) >= 10) {
  model_adjacent <- lmer(
    adjacent_distance ~ condition + (1 | conversation_id),
    data = df_adjacent
  )

  cat("Main mixed model: adjacent_distance ~ condition + (1 | conversation_id)\n")
  print(summary(model_adjacent))

  emm_adjacent <- emmeans(model_adjacent, specs = "condition")

  cat("\nEstimated marginal means by condition:\n")
  print(emm_adjacent)

  cat("\nPairwise condition contrasts (BH-adjusted):\n")
  print(pairs(emm_adjacent, adjust = "BH"))

  cat("\nPlanned contrast: Human-AI vs same-type average:\n")
  print(contrast(
    emm_adjacent,
    method = list("Human-AI - mean(Human-Human, AI-AI)" = c(1, -0.5, -0.5))
  ))
}
```

```{r adjacent-distance-plot, fig.width=9, fig.height=5}
ggplot(df_adjacent, aes(x = condition, y = adjacent_distance, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Adjacent turn-to-turn semantic distance by condition",
    subtitle = "Each point is one adjacent pair in the reconstructed alternating dyadic sequence",
    x = "Condition",
    y = "Adjacent semantic distance"
  ) +
  theme(legend.position = "none")
```

# Supporting: Story-Level Mean Adjacent Distance

```{r story-mean-adjacent}
df_adjacent_story <- df_adjacent %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_pairs = n(),
    mean_adjacent_distance = mean(adjacent_distance, na.rm = TRUE),
    median_adjacent_distance = median(adjacent_distance, na.rm = TRUE),
    .groups = "drop"
  )

df_adjacent_story %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_story_distance = mean(mean_adjacent_distance),
    sd_story_distance = sd(mean_adjacent_distance),
    median_story_distance = median(mean_adjacent_distance),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Story-level mean adjacent semantic distance by condition"
  )
```

```{r story-mean-adjacent-tests}
if (n_distinct(df_adjacent_story$condition) >= 2 && nrow(df_adjacent_story) >= 10) {
  cat("One-way ANOVA: mean_adjacent_distance ~ condition\n")
  print(summary(aov(mean_adjacent_distance ~ condition, data = df_adjacent_story)))

  cat("\nKruskal-Wallis (robust): mean_adjacent_distance ~ condition\n")
  print(kruskal.test(mean_adjacent_distance ~ condition, data = df_adjacent_story))

  cat("\nPairwise Wilcoxon (BH-adjusted):\n")
  print(pairwise.wilcox.test(
    df_adjacent_story$mean_adjacent_distance,
    df_adjacent_story$condition,
    p.adjust.method = "BH"
  ))
}
```

```{r story-mean-adjacent-plot, fig.width=9, fig.height=5}
ggplot(df_adjacent_story, aes(x = condition, y = mean_adjacent_distance, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.45, size = 1.2) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Story-level mean adjacent semantic distance by condition",
    subtitle = "One mean adjacent distance per story",
    x = "Condition",
    y = "Mean adjacent semantic distance"
  ) +
  theme(legend.position = "none")
```

# Bootstrap CIs and Effect Sizes on Story Means

```{r boot-effects-adjacent-story}
bootstrap_condition_means(df_adjacent_story, "mean_adjacent_distance", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of story-level adjacent distance (R = 10,000)"
  )

effect_sizes_pairwise(df_adjacent_story, "mean_adjacent_distance") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on story-level mean adjacent distance"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-adjacent-story}
planned_contrast_HA_vs_same(df_adjacent_story, "mean_adjacent_distance") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: story-level mean adjacent distance in HA vs the average of HH and AA"
  )
```

# Supplemental: Turn-Level Adjacent Semantic-Distance Network

```{r turn-adjacent-network-data}
network_dir <- here::here("analysis", "comparison", "semantic_network_layouts")
network_paths <- list(
  turn_nodes = file.path(network_dir, "turn_nodes.csv"),
  turn_edges = file.path(network_dir, "turn_edges.csv"),
  metadata = file.path(network_dir, "metadata.json")
)

network_available <- all(file.exists(unlist(network_paths)))

if (network_available) {
  df_turn_network <- read_csv(network_paths$turn_nodes, show_col_types = FALSE) %>%
    mutate(condition = factor(condition, levels = names(condition_colors)))
  turn_edges <- read_csv(network_paths$turn_edges, show_col_types = FALSE) %>%
    mutate(
      condition = factor(condition, levels = names(condition_colors)),
      adjacent_distance = as.numeric(adjacent_distance),
      closeness = 1 - scales::rescale(
        adjacent_distance,
        to = c(0, 1),
        from = quantile(adjacent_distance, c(0.05, 0.95), na.rm = TRUE)
      ),
      closeness = pmax(pmin(closeness, 1), 0),
      pair_type = factor(pair_type, levels = c("within_exchange", "between_exchange"))
    )
  network_metadata <- jsonlite::fromJSON(network_paths$metadata)

  turn_edge_labels <- turn_edges %>%
    group_by(condition) %>%
    summarise(
      mean_distance = mean(adjacent_distance, na.rm = TRUE),
      x = quantile(c(x, xend), 0.03, na.rm = TRUE),
      y = quantile(c(y, yend), 0.97, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(label = sprintf("Mean d = %.3f", mean_distance))
} else {
  df_turn_network <- tibble()
  turn_edges <- tibble()
  turn_edge_labels <- tibble()
  network_metadata <- list(turn_layout_method = "not computed")
  cat(
    "Precomputed semantic network files not found. Run:\n",
    "venv/bin/python scripts/09_compute_semantic_network_layouts.py\n"
  )
}
```

```{r turn-adjacent-network-plot, fig.width=13, fig.height=6}
if (network_available) {
  ggplot() +
    geom_segment(
      data = turn_edges,
      aes(
        x = x,
        y = y,
        xend = xend,
        yend = yend,
        color = closeness,
        linewidth = closeness,
        group = conversation_id
      ),
      alpha = 0.68,
      lineend = "round"
    ) +
    geom_point(
      data = df_turn_network,
      aes(x = x, y = y),
      color = "grey15",
      alpha = 0.28,
      size = 0.55
    ) +
    geom_label(
      data = turn_edge_labels,
      aes(x = x, y = y, label = label),
      hjust = 0,
      vjust = 1,
      label.size = 0,
      fill = "white",
      alpha = 0.85,
      size = 3.2
    ) +
    facet_wrap(~condition, nrow = 1) +
    scale_color_gradient(
      low = "grey88",
      high = "grey8",
      name = "Semantic\ncloseness",
      limits = c(0, 1),
      breaks = c(0, 1),
      labels = c("far", "close")
    ) +
    scale_linewidth_continuous(range = c(0.03, 1.35), guide = "none") +
    labs(
      title = "Adjacent turn-to-turn semantic distance networks",
      subtitle = paste0(
        "Nodes are turns from a balanced visual sample; edges connect only adjacent turns. Darker/thicker edges are semantically closer; faint grey edges are larger jumps; layout = ",
        network_metadata$turn_layout_method
      ),
      x = NULL,
      y = NULL
    ) +
    theme(
      legend.position = "right",
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Run venv/bin/python scripts/09_compute_semantic_network_layouts.py to generate this plot."
    ) +
    theme_void()
}
```

# Supplemental: Speaker-Level Slot Trajectories

```{r speaker-plot, fig.width=12, fig.height=6}
speaker_summary <- df_speaker %>%
  group_by(condition, slot, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(speaker_summary, aes(x = k, y = mean_distance, color = slot)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = slot),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Slot",
    subtitle = "Supplemental lag-based slot trajectories",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Slot",
    fill = "Slot"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: General Condition-Level Comparison

```{r interleaved-descriptives}
df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Interleaved semantic distance by lag and condition"
  )
```

```{r interleaved-plot, fig.width=12, fig.height=5}
interleaved_summary <- df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(interleaved_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition),
    alpha = 0.18,
    color = NA
  ) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Interleaved Semantic Exploration by Lag",
    subtitle = "Mean distance with 95% confidence intervals",
    x = "Lag Bin (k)",
    y = "Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

```{r interleaved-model}
model_interleaved <- lmer(
  distance ~ k * condition + (1 | conversation_id),
  data = df_interleaved
)

cat("Interleaved model summary:\n")
print(summary(model_interleaved))

emm_interleaved <- emmeans(model_interleaved, specs = "condition")

cat("\nCondition contrasts (interleaved view):\n")
print(pairs(emm_interleaved, adjust = "bonferroni"))
```

# Supplemental: Role-Level View

```{r speaker-descriptives}
df_speaker %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Speaker-level semantic distance by condition, role, and lag"
  )
```

```{r role-plot, fig.width=12, fig.height=6}
role_summary <- df_speaker %>%
  filter(!is.na(condition_role)) %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(role_summary, aes(x = k, y = mean_distance, color = condition_role)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition_role),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Role",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Role",
    fill = "Role"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: Interleaved vs Speaker View

```{r view-comparison, fig.width=12, fig.height=5}
view_summary <- bind_rows(
  df_interleaved %>%
    mutate(view = "Interleaved sequence") %>%
    select(condition, conversation_id, k, distance, view),
  df_speaker %>%
    mutate(view = "Speaker-specific") %>%
    select(condition, conversation_id, k, distance, view)
) %>%
  group_by(condition, view, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(view_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  facet_wrap(~view) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Interleaved vs Speaker-Specific Semantic Distance",
    x = "Lag Bin (k)",
    y = "Mean Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

# Summary

Primary cross-condition result: adjacent turn-to-turn semantic distance from the reconstructed alternating dyadic sequence, tested with a mixed model that nests adjacent observations within story. This includes both within-exchange distances and the cross-exchange step from the second turn of one exchange to the first turn of the next exchange.

Robustness layers added:

- story-level mean adjacent distance as an aggregate supporting analysis
- bootstrap 95% CIs on per-condition story means
- pairwise Cliff's $\delta$ effect sizes on story means
- planned single-df contrast HA vs (HH+AA)/2 on story means

Supplemental views retained:

- turn-level adjacent semantic-distance network, with edge color/width encoding original-space adjacent distance
- lag-based speaker-level slot trajectories
- lag-based interleaved sequence model and condition descriptives
- within-condition role curves (descriptive only; HA Human/AI inferential view lives in [analysis/human-ai/exploration.Rmd](../human-ai/exploration.Rmd))
- interleaved vs speaker-specific side-by-side view

To refresh the semantic-network layouts, run `venv/bin/python scripts/09_compute_semantic_network_layouts.py` before rendering this notebook.

*Analysis completed: `r Sys.time()`*
