---
title: "Semantic Exploration Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Semantic Space Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans       # Estimated marginal means
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares semantic exploration patterns across experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Metrics:**
- Semantic distance between consecutive turns
- Cumulative semantic exploration (story-level trajectory)
- Local vs global novelty
- Embedding space coverage

**Key Questions:**
1. Do humans introduce more semantic novelty than AI?
2. Is Human-Human exploration more divergent than Human-AI?
3. Does AI-AI show characteristic patterns (e.g., staying "on script")?

# Load Data

```{r load-data}
# Check data availability
cat("Data availability:\n")
avail <- check_data_availability()
print(avail)

# Load exploration data
df_combined <- tryCatch({
  load_exploration_data(conditions = c("human-ai", "human-human", "ai-ai"))
}, error = function(e) {
  cat("Primary load failed:", e$message, "\n")
  cat("Attempting fallback loading...\n")
  
  conditions_available <- c()
  for (cond in c("human-ai", "human-human", "ai-ai")) {
    for (fname in c("semantic_exploration.parquet", "exploration_metrics.parquet", "novelty_scores.parquet")) {
      path <- here("data", cond, "processed", fname)
      if (file.exists(path)) {
        conditions_available <- c(conditions_available, cond)
        break
      }
    }
  }
  
  conditions_available <- unique(conditions_available)
  
  if (length(conditions_available) == 0) {
    stop("No exploration data available for any condition")
  }
  
  load_exploration_data(conditions = conditions_available)
})

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report loaded data
cat("\nLoaded data summary:\n")
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing

```{r preprocess}
# Identify available exploration metrics
potential_metrics <- c(
  # Distance-based
  "semantic_distance", "cosine_distance", "euclidean_distance",
  "embedding_distance", "turn_distance",
  # Novelty
  "novelty", "local_novelty", "global_novelty", "novelty_score",
  # Cumulative
  "cumulative_exploration", "cumulative_distance",
  "trajectory_length", "semantic_trajectory"
)

available_metrics <- intersect(potential_metrics, names(df_combined))
cat("Available exploration metrics:", paste(available_metrics, collapse = ", "), "\n")

if (length(available_metrics) == 0) {
  # Try to find any numeric columns that might be exploration metrics
  numeric_cols <- names(df_combined)[sapply(df_combined, is.numeric)]
  exploration_like <- numeric_cols[str_detect(numeric_cols, "dist|novel|explor|semantic|embed")]
  
  if (length(exploration_like) > 0) {
    available_metrics <- exploration_like
    cat("Found exploration-like metrics:", paste(available_metrics, collapse = ", "), "\n")
  } else {
    stop("No exploration metrics found in the data")
  }
}
```

# Descriptive Statistics

```{r descriptives}
# Compute summary statistics for exploration metrics
desc_stats <- df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    across(all_of(available_metrics[1:min(4, length(available_metrics))]),
           list(
             mean = ~mean(., na.rm = TRUE),
             sd = ~sd(., na.rm = TRUE),
             median = ~median(., na.rm = TRUE)
           ),
           .names = "{.col}_{.fn}"),
    .groups = "drop"
  )

# Display first metric's stats
first_metric <- available_metrics[1]
desc_stats %>%
  select(condition, n_stories, n_turns, starts_with(first_metric)) %>%
  knitr::kable(
    digits = 3,
    caption = paste("Summary statistics for", first_metric)
  )
```

# Exploration Distributions

```{r exploration-distributions, fig.width=12, fig.height=6}
# Plot distributions for main exploration metric
main_metric <- available_metrics[1]

p1 <- df_combined %>%
  ggplot(aes(x = !!sym(main_metric), fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = paste(main_metric, "Distribution"),
    x = main_metric,
    y = "Density"
  )

p2 <- df_combined %>%
  ggplot(aes(x = condition, y = !!sym(main_metric), fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.1) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = paste(main_metric, "by Condition"),
    x = "Condition",
    y = main_metric
  ) +
  theme(legend.position = "none")

p1 + p2 + plot_layout(guides = "collect")
```

# Cross-Condition Comparison

## Primary Exploration Metric

```{r primary-exploration}
main_metric <- available_metrics[1]

model_exploration <- lmer(
  as.formula(paste(main_metric, "~ condition + turn + (1 | conversation_id)")),
  data = df_combined
)

cat("===", main_metric, "by Condition (controlling for turn) ===\n")
print(summary(model_exploration))

# Estimated marginal means
emm_exploration <- emmeans(model_exploration, specs = "condition")
cat("\n=== Estimated Marginal Means ===\n")
print(emm_exploration)

cat("\n=== Pairwise Contrasts ===\n")
print(pairs(emm_exploration, adjust = "bonferroni"))
```

## Turn-by-Turn Evolution

```{r turn-evolution, fig.width=12, fig.height=5}
# How does exploration change across turns?
main_metric <- available_metrics[1]

turn_summary <- df_combined %>%
  group_by(condition, turn) %>%
  summarise(
    mean_exploration = mean(!!sym(main_metric), na.rm = TRUE),
    se_exploration = sd(!!sym(main_metric), na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(turn_summary, aes(x = turn, y = mean_exploration, color = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_exploration - 1.96 * se_exploration,
                  ymax = mean_exploration + 1.96 * se_exploration,
                  fill = condition),
              alpha = 0.2, color = NA) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Semantic Exploration Across Story Turns",
    subtitle = "Shaded area = 95% CI",
    x = "Turn",
    y = main_metric,
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```

## Condition × Turn Interaction

```{r condition-turn-interaction}
main_metric <- available_metrics[1]

model_interaction <- lmer(
  as.formula(paste(main_metric, "~ condition * turn + (1 | conversation_id)")),
  data = df_combined
)

cat("=== Condition × Turn Interaction ===\n")
print(summary(model_interaction))

# Test interaction significance
cat("\n=== ANOVA Table ===\n")
print(anova(model_interaction))
```

# Author-Level Exploration

```{r author-exploration}
# Check for author-specific exploration columns
author_1_cols <- names(df_combined)[str_detect(names(df_combined), "^author_1_")]
author_2_cols <- names(df_combined)[str_detect(names(df_combined), "^author_2_")]

exploration_author_1 <- author_1_cols[str_detect(author_1_cols, "novel|explor|dist")]
exploration_author_2 <- author_2_cols[str_detect(author_2_cols, "novel|explor|dist")]

if (length(exploration_author_1) > 0 && length(exploration_author_2) > 0) {
  cat("Author-level exploration metrics found:\n")
  cat("  Author 1:", paste(exploration_author_1, collapse = ", "), "\n")
  cat("  Author 2:", paste(exploration_author_2, collapse = ", "), "\n")
  
  # Use first matching metric
  a1_metric <- exploration_author_1[1]
  a2_metric <- exploration_author_2[1]
  
  # Compare authors within conditions
  df_author <- df_combined %>%
    select(conversation_id, turn, condition, 
           !!sym(a1_metric), !!sym(a2_metric)) %>%
    pivot_longer(
      cols = c(!!sym(a1_metric), !!sym(a2_metric)),
      names_to = "author",
      values_to = "exploration"
    ) %>%
    mutate(author = if_else(str_detect(author, "author_1"), "Author 1", "Author 2"))
  
  # Plot
  ggplot(df_author, aes(x = author, y = exploration, fill = condition)) +
    geom_boxplot(alpha = 0.7, outlier.alpha = 0.1) +
    facet_wrap(~condition) +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Semantic Exploration by Author and Condition",
      x = "Author",
      y = "Exploration Metric"
    ) +
    theme(legend.position = "none")
  
} else {
  cat("No author-level exploration metrics found.\n")
}
```

# Story-Level Analysis

```{r story-level}
# Aggregate to story level
main_metric <- available_metrics[1]

story_summary <- df_combined %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_turns = n(),
    total_exploration = sum(!!sym(main_metric), na.rm = TRUE),
    mean_exploration = mean(!!sym(main_metric), na.rm = TRUE),
    sd_exploration = sd(!!sym(main_metric), na.rm = TRUE),
    max_exploration = max(!!sym(main_metric), na.rm = TRUE),
    .groups = "drop"
  )

# Story-level comparison
model_story <- lm(total_exploration ~ condition, data = story_summary)
cat("=== Total Story Exploration by Condition ===\n")
print(summary(model_story))

# Visualization
ggplot(story_summary, aes(x = condition, y = total_exploration, fill = condition)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.3) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Total Semantic Exploration per Story",
    x = "Condition",
    y = "Cumulative Exploration"
  ) +
  theme(legend.position = "none")
```

# Summary Visualization

```{r summary-plot, fig.width=12, fig.height=5}
# Effect size comparison plot
if (length(unique(df_combined$condition)) > 1) {
  
  # Compute effect sizes relative to human-ai baseline
  baseline_mean <- df_combined %>%
    filter(condition == "Human-AI") %>%
    pull(!!sym(main_metric)) %>%
    mean(na.rm = TRUE)
  
  baseline_sd <- df_combined %>%
    filter(condition == "Human-AI") %>%
    pull(!!sym(main_metric)) %>%
    sd(na.rm = TRUE)
  
  effect_sizes <- df_combined %>%
    group_by(condition) %>%
    summarise(
      mean = mean(!!sym(main_metric), na.rm = TRUE),
      se = sd(!!sym(main_metric), na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    ) %>%
    mutate(
      cohens_d = (mean - baseline_mean) / baseline_sd,
      d_se = se / baseline_sd
    )
  
  p1 <- ggplot(effect_sizes, aes(x = condition, y = mean, fill = condition)) +
    geom_col(alpha = 0.8, width = 0.6) +
    geom_errorbar(aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se), width = 0.2) +
    geom_hline(yintercept = baseline_mean, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(title = "Mean Exploration by Condition", y = main_metric) +
    theme(legend.position = "none")
  
  p2 <- ggplot(effect_sizes, aes(x = condition, y = cohens_d, fill = condition)) +
    geom_col(alpha = 0.8, width = 0.6) +
    geom_errorbar(aes(ymin = cohens_d - 1.96 * d_se, ymax = cohens_d + 1.96 * d_se), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(title = "Effect Size (Cohen's d) vs Human-AI", y = "Cohen's d") +
    theme(legend.position = "none")
  
  p1 + p2
}
```

# Key Findings

Based on the analysis above:

1. **Exploration patterns**: [Interpret which condition shows greatest/least semantic exploration]

2. **Turn dynamics**: [Discuss how exploration changes across the story arc]

3. **Author contributions**: [Compare who introduces more novelty - Author 1 or Author 2]

4. **AI-AI baseline**: [Note characteristic exploration patterns in simulated data]

---

*Analysis completed: `r Sys.time()`*
