---
title: "Surface Metrics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Text Descriptives on the Exchange-Aligned Structure"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans
)

theme_set(theme_comparison())
```

# Overview

This notebook compares surface-level text metrics across conditions using the long-format, exchange-aligned metric tables.

The underlying processed data are now one row per **speaker contribution**, with recovered role metadata:

- **Human / AI** in Human-AI
- **Starter-side / Non-starter-side** in Human-Human and AI-AI

The notebook therefore no longer assumes old wide `author_1_*` / `author_2_*` columns.

# Load Data

```{r load-data}
df_combined <- load_surface_metrics(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  prepare_condition_factor(reference = "human-ai")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded surface-metric rows by condition")
```

# Metric Selection

```{r metric-selection}
first_present <- function(candidates, df) {
  present <- candidates[candidates %in% names(df)]
  if (length(present) == 0) return(NA_character_)
  present[[1]]
}

selected_metrics <- c(
  word_count = first_present(c("n_words", "n_tokens", "doc_length"), df_combined),
  sentence_length = first_present(c("sentence_length_mean", "avg_sentence_length"), df_combined),
  lexical_diversity = first_present(c("proportion_unique_tokens", "ttr"), df_combined),
  readability = first_present(c("flesch_reading_ease"), df_combined),
  coherence = first_present(c("first_order_coherence"), df_combined)
)

selected_metrics <- selected_metrics[!is.na(selected_metrics)]

knitr::kable(
  tibble(metric_alias = names(selected_metrics), column_name = unname(selected_metrics)),
  caption = "Surface metrics selected for the comparison notebook"
)
```

# Descriptive Statistics

```{r descriptives}
descriptive_table <- purrr::imap_dfr(selected_metrics, function(col_name, metric_alias) {
  df_combined %>%
    group_by(condition) %>%
    summarise(
      metric = metric_alias,
      mean_value = mean(.data[[col_name]], na.rm = TRUE),
      sd_value = sd(.data[[col_name]], na.rm = TRUE),
      .groups = "drop"
    )
})

descriptive_table %>%
  knitr::kable(
    digits = 3,
    caption = "Condition-level descriptives for selected surface metrics"
  )
```

```{r distributions, fig.width=12, fig.height=8}
plot_data <- purrr::imap_dfr(selected_metrics, function(col_name, metric_alias) {
  df_combined %>%
    transmute(condition, metric = metric_alias, value = .data[[col_name]])
})

plot_data %>%
  ggplot(aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  labs(
    title = "Surface Metric Distributions by Condition",
    x = "Metric Value",
    y = "Density"
  ) +
  theme(legend.position = "bottom")
```

# Cross-Condition Models

```{r condition-models}
for (metric_alias in names(selected_metrics)) {
  col_name <- selected_metrics[[metric_alias]]
  model <- lmer(
    as.formula(paste(col_name, "~ condition + (1 | conversation_id)")),
    data = df_combined
  )

  cat("\n===", metric_alias, "(", col_name, ") ===\n")
  print(summary(model))
  print(pairs(emmeans(model, specs = "condition"), adjust = "bonferroni"))
}
```

# Role-Level Comparison

```{r role-comparison, fig.width=12, fig.height=6}
primary_length_metric <- selected_metrics[[names(selected_metrics)[1]]]

df_combined %>%
  filter(!is.na(condition_role)) %>%
  ggplot(aes(x = condition_role, y = .data[[primary_length_metric]], fill = condition_role)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.1) +
  facet_wrap(~condition, scales = "free_x") +
  labs(
    title = paste("Role-Level Comparison for", names(selected_metrics)[1]),
    subtitle = "Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = NULL,
    y = primary_length_metric
  ) +
  theme(legend.position = "none")
```

```{r role-models}
if ("condition_role" %in% names(df_combined) && any(!is.na(df_combined$condition_role))) {
  for (metric_alias in names(selected_metrics)[1:min(3, length(selected_metrics))]) {
    col_name <- selected_metrics[[metric_alias]]

    cat("\n=== Role contrast for", metric_alias, "===\n")

    for (cond_name in levels(df_combined$condition)) {
      df_subset <- df_combined %>%
        filter(condition == cond_name, !is.na(condition_role))

      if (nrow(df_subset) < 40 || n_distinct(df_subset$condition_role) < 2) {
        next
      }

      model <- lmer(
        as.formula(paste(col_name, "~ condition_role + (1 | conversation_id)")),
        data = df_subset
      )

      cat("\n", cond_name, "\n", sep = "")
      print(summary(model))
      print(pairs(emmeans(model, specs = "condition_role")))
    }
  }
}
```

# Summary

The notebook is now adapted to the new structure by:

- loading long-format contribution-level surface metrics,
- using the shared backfilled role metadata,
- restricting the analysis to complete, aligned exchanges,
- removing the old wide-format `author_1_*` / `author_2_*` assumptions.

*Analysis completed: `r Sys.time()`*
