---
title: "Surface Metrics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Text Descriptives on the Exchange-Aligned Structure"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans
)

theme_set(theme_comparison())
```

# Overview

This notebook compares surface-level text metrics across conditions using the long-format, exchange-aligned metric tables.

The underlying processed data are now one row per **speaker contribution**, with recovered role metadata:

- **Human / AI** in Human-AI
- **Starter-side / Non-starter-side** in Human-Human and AI-AI

Those role labels are useful for **within-condition description**, but they are not the primary cross-condition comparison axis. For cross-condition comparison, this notebook uses **Slot 1 / Slot 2 asymmetry**.

The notebook therefore no longer assumes old wide `author_1_*` / `author_2_*` columns.

**Primary cross-condition object.** Per story and per metric $m$,
$$A_m(i) \;=\; \bigl|\, \overline{m}_{\text{slot}=1,\,i} - \overline{m}_{\text{slot}=2,\,i} \,\bigr|.$$
This is the unsigned, label-invariant side-difference magnitude. For each metric we test $A_m \sim \text{condition}$ with ANOVA + Kruskal and apply BH adjustment across the metric panel. The LMM slot-marginal contrasts are retained as supplemental. Starter-mechanism and Human-AI speaker-type decompositions are deferred.

**Article-use note.** The primary result is the per-story unsigned side-difference test. Slot-marginal LMMs and role plots are useful for explaining where the difference comes from, but they should not replace the label-invariant cross-condition result.

# Load Data

```{r load-data}
df_combined <- load_surface_metrics(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded surface-metric rows by condition")
```

# Metric Selection

```{r metric-selection}
first_present <- function(candidates, df) {
  present <- candidates[candidates %in% names(df)]
  if (length(present) == 0) return(NA_character_)
  present[[1]]
}

selected_metrics <- c(
  word_count = first_present(c("n_words", "n_tokens", "doc_length"), df_combined),
  sentence_length = first_present(c("sentence_length_mean", "avg_sentence_length"), df_combined),
  lexical_diversity = first_present(c("proportion_unique_tokens", "ttr"), df_combined),
  readability = first_present(c("flesch_reading_ease"), df_combined),
  coherence = first_present(c("first_order_coherence"), df_combined)
)

selected_metrics <- selected_metrics[!is.na(selected_metrics)]

knitr::kable(
  tibble(metric_alias = names(selected_metrics), column_name = unname(selected_metrics)),
  caption = "Surface metrics selected for the comparison notebook"
)
```

# Primary: Per-Story Side-Mean Difference

```{r primary-surface-asymmetry}
df_slots <- df_combined %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(
    slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  )

per_story_abs_diff <- purrr::imap_dfr(selected_metrics, function(col_name, metric_alias) {
  df_slots %>%
    select(condition, conversation_id, slot, value = all_of(col_name)) %>%
    group_by(condition, conversation_id, slot) %>%
    summarise(side_mean = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = slot, values_from = side_mean) %>%
    drop_na(`Slot 1`, `Slot 2`) %>%
    transmute(
      metric = metric_alias,
      condition,
      conversation_id,
      signed_diff = `Slot 1` - `Slot 2`,
      A = abs(signed_diff)
    )
})

stopifnot(all(per_story_abs_diff$A >= 0))

per_story_abs_diff %>%
  group_by(metric, condition) %>%
  summarise(
    n = n(),
    mean_A = mean(A),
    median_A = median(A),
    sd_A = sd(A),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story |side-mean difference| by condition, per metric"
  )
```

```{r primary-surface-tests}
primary_tests <- per_story_abs_diff %>%
  group_by(metric) %>%
  summarise(
    anova_p = {
      fit <- tryCatch(aov(A ~ condition), error = function(e) NULL)
      if (is.null(fit)) NA_real_ else summary(fit)[[1]][["Pr(>F)"]][1]
    },
    kw_p = tryCatch(
      kruskal.test(A ~ condition)$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    anova_p_BH = p.adjust(anova_p, method = "BH"),
    kw_p_BH = p.adjust(kw_p, method = "BH")
  )

primary_tests %>%
  knitr::kable(
    digits = 4,
    caption = "Primary cross-condition tests for A_m ~ condition, per metric (BH-adjusted across metrics)"
  )
```

```{r primary-surface-plot, fig.width=12, fig.height=8}
ggplot(per_story_abs_diff, aes(x = condition, y = A, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Per-story |Slot 1 mean - Slot 2 mean| by condition",
    x = "Condition",
    y = expression("|" * Delta * bar(m) * "|")
  ) +
  theme(legend.position = "none")
```

# Bootstrap CIs and Effect Sizes per Metric

```{r boot-effects-surface}
boot_by_metric <- per_story_abs_diff %>%
  group_by(metric) %>%
  group_modify(~ bootstrap_condition_means(.x, "A", R = 10000)) %>%
  ungroup()

boot_by_metric %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of A per metric (R = 10,000)"
  )

effects_by_metric <- per_story_abs_diff %>%
  group_by(metric) %>%
  group_modify(~ effect_sizes_pairwise(.x, "A")) %>%
  ungroup()

effects_by_metric %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on A per metric"
  )
```

# Planned Contrast: HA vs (HH + AA)/2 per Metric

```{r planned-contrast-surface}
planned_by_metric <- per_story_abs_diff %>%
  group_by(metric) %>%
  group_modify(~ planned_contrast_HA_vs_same(.x, "A")) %>%
  ungroup() %>%
  mutate(p_BH = p.adjust(p, method = "BH"))

planned_by_metric %>%
  knitr::kable(
    digits = 3,
    caption = "Planned HA vs (HH+AA)/2 contrast per metric, BH-adjusted across metrics"
  )
```

# Directional Stability of Signed Side-Mean Differences (per Metric)

Tests whether the *direction* of the slot-1 vs slot-2 side-mean difference is
consistent across stories within each condition, per metric. The signed
quantity is $\text{signed}_i = \overline{m}_{\text{slot 1},i} - \overline{m}_{\text{slot 2},i}$.

```{r directional-stability-surface}
ds_surface <- per_story_abs_diff %>%
  dplyr::group_split(metric) %>%
  purrr::map_dfr(function(df_m) {
    metric_name <- unique(df_m$metric)
    res <- directional_stability(df_m, "signed_diff")
    res$by_condition %>%
      mutate(metric = metric_name, variance_equality_p = res$variance_equality_p)
  })

ds_surface %>%
  select(metric, condition, n, mean, sd, stability, sign_consistency, one_sample_t_p, variance_equality_p) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-metric directional stability of signed side-mean differences"
  )
```

```{r directional-stability-surface-plot, fig.width=12, fig.height=8}
ggplot(per_story_abs_diff, aes(x = condition, y = signed_diff, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Signed Slot 1 - Slot 2 side-mean difference by condition",
    x = "Condition",
    y = expression(Delta * bar(m))
  ) +
  theme(legend.position = "none")
```

# Multivariate Composite over Surface Metrics

Stacks the per-story `A_m` vector across all selected surface metrics into one
matrix and tests whether the joint condition profiles differ via PERMANOVA
(`vegan::adonis2`). A single multivariate answer to "is the cross-metric
surface-asymmetry profile distinct in HA?".

```{r multivariate-surface, fig.width=10, fig.height=6}
A_matrix <- per_story_abs_diff %>%
  pivot_wider(
    id_cols = c(conversation_id, condition),
    names_from = metric,
    values_from = A
  ) %>%
  drop_na()

cat(sprintf(
  "Stories with complete A-vector across all %d metrics: %d\n",
  length(selected_metrics), nrow(A_matrix)
))

if (nrow(A_matrix) >= 30 && requireNamespace("vegan", quietly = TRUE)) {
  X <- A_matrix %>% select(-conversation_id, -condition) %>% as.matrix()
  set.seed(42)
  perm_res <- vegan::adonis2(X ~ condition, data = A_matrix, method = "euclidean", permutations = 9999)
  print(perm_res)

  pca <- prcomp(X, scale. = TRUE)
  pca_df <- as_tibble(pca$x[, 1:2]) %>%
    bind_cols(condition = A_matrix$condition)

  ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(alpha = 0.6, size = 1.8) +
    stat_ellipse(level = 0.95, linewidth = 0.8) +
    scale_color_manual(values = condition_colors) +
    labs(
      title = "PCA of per-story surface-metric asymmetry vector",
      subtitle = "95% data ellipses per condition",
      x = sprintf("PC1 (%.1f%%)", 100 * pca$sdev[1]^2 / sum(pca$sdev^2)),
      y = sprintf("PC2 (%.1f%%)", 100 * pca$sdev[2]^2 / sum(pca$sdev^2))
    )
} else {
  cat("Skipping multivariate test: insufficient stories or vegan package missing.\n")
}
```

# Supplemental: General Condition-Level Comparison

```{r descriptives}
descriptive_table <- purrr::imap_dfr(selected_metrics, function(col_name, metric_alias) {
  df_combined %>%
    group_by(condition) %>%
    summarise(
      metric = metric_alias,
      mean_value = mean(.data[[col_name]], na.rm = TRUE),
      sd_value = sd(.data[[col_name]], na.rm = TRUE),
      .groups = "drop"
    )
})

descriptive_table %>%
  knitr::kable(
    digits = 3,
    caption = "Condition-level descriptives for selected surface metrics"
  )
```

```{r distributions, fig.width=12, fig.height=8}
plot_data <- purrr::imap_dfr(selected_metrics, function(col_name, metric_alias) {
  df_combined %>%
    transmute(condition, metric = metric_alias, value = .data[[col_name]])
})

plot_data %>%
  ggplot(aes(x = value, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  labs(
    title = "Surface Metric Distributions by Condition",
    x = "Metric Value",
    y = "Density"
  ) +
  theme(legend.position = "bottom")
```

```{r condition-models}
for (metric_alias in names(selected_metrics)) {
  col_name <- selected_metrics[[metric_alias]]
  model <- lmer(
    as.formula(paste(col_name, "~ condition + (1 | conversation_id)")),
    data = df_combined
  )

  cat("\n===", metric_alias, "(", col_name, ") ===\n")
  print(summary(model))
  print(pairs(emmeans(model, specs = "condition"), adjust = "bonferroni"))
}
```

# Supplemental: Role-Level Comparison

```{r role-comparison, fig.width=12, fig.height=6}
primary_length_metric <- selected_metrics[[names(selected_metrics)[1]]]

df_combined %>%
  filter(!is.na(condition_role)) %>%
  ggplot(aes(x = condition_role, y = .data[[primary_length_metric]], fill = condition_role)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.1) +
  facet_wrap(~condition, scales = "free_x") +
  labs(
    title = paste("Role-Level Comparison for", names(selected_metrics)[1]),
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = NULL,
    y = primary_length_metric
  ) +
  theme(legend.position = "none")
```

```{r role-models}
# Role-contrast LMMs are run only for HH and AA, where the role labels are
# Starter-side / Non-starter-side and have no counterpart elsewhere. The HA
# Human-vs-AI role LMM lives in
# analysis/human-ai/surface_metrics_analysis.Rmd and is not duplicated here.
if ("condition_role" %in% names(df_combined) && any(!is.na(df_combined$condition_role))) {
  same_type_conditions <- intersect(levels(df_combined$condition),
                                    c("Human-Human", "AI-AI"))

  for (metric_alias in names(selected_metrics)[1:min(3, length(selected_metrics))]) {
    col_name <- selected_metrics[[metric_alias]]

    cat("\n=== Role contrast for", metric_alias, "(HH and AA only) ===\n")

    for (cond_name in same_type_conditions) {
      df_subset <- df_combined %>%
        filter(condition == cond_name, !is.na(condition_role))

      if (nrow(df_subset) < 40 || n_distinct(df_subset$condition_role) < 2) {
        next
      }

      model <- lmer(
        as.formula(paste(col_name, "~ condition_role + (1 | conversation_id)")),
        data = df_subset
      )

      cat("\n", cond_name, "\n", sep = "")
      print(summary(model))
      print(pairs(emmeans(model, specs = "condition_role")))
    }
  }
}
```

# Summary

Primary cross-condition result: per-metric $A_m(i) = |\overline{m}_{\text{slot 1},i} - \overline{m}_{\text{slot 2},i}|$, tested with ANOVA + Kruskal and BH-adjusted across metrics.

Robustness layers added:

- bootstrap 95% CIs on per-condition mean of $A_m$, per metric
- pairwise Cliff's $\delta$ effect sizes per metric
- planned single-df contrast HA vs (HH+AA)/2 per metric, BH-adjusted across metrics
- multivariate composite over the surface-metric panel via PERMANOVA + PCA biplot

Supplemental views retained:

- condition-level distributions and marginal `condition` LMMs
- within-condition role comparisons restricted to HH / AA (Starter-side / Non-starter-side); HA Human-vs-AI role LMM lives in [analysis/human-ai/surface_metrics_analysis.Rmd](../human-ai/surface_metrics_analysis.Rmd) and is not duplicated here

Deferred to a follow-up pass:

- starter-mechanism analysis (signed, starter-normalized side differences across all conditions)

*Analysis completed: `r Sys.time()`*
