---
title: "Surface Metrics Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Text Descriptives Analysis"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans       # Estimated marginal means
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook compares surface-level text features across experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Metrics:**
- Word count per turn
- Sentence length
- Lexical diversity (type-token ratio)
- Readability scores
- Punctuation patterns

**Key Questions:**
1. Do AI contributions differ in length/complexity from human contributions?
2. Are Human-Human stories more varied than Human-AI or AI-AI?
3. Does the AI-AI baseline show characteristic "AI-like" text patterns?

# Load Data

```{r load-data}
# Check data availability
cat("Data availability:\n")
avail <- check_data_availability()
print(avail)

# Load surface metrics
df_combined <- tryCatch({
  load_surface_metrics(conditions = c("human-ai", "human-human", "ai-ai"))
}, error = function(e) {
  cat("Error loading surface metrics:", e$message, "\n")
  cat("Attempting to load from individual conditions...\n")
  
  conditions_available <- c()
  for (cond in c("human-ai", "human-human", "ai-ai")) {
    path <- here("data", cond, "processed", "surface_metrics.parquet")
    if (file.exists(path)) {
      conditions_available <- c(conditions_available, cond)
    }
  }
  
  if (length(conditions_available) == 0) {
    stop("No surface metrics data available")
  }
  
  load_surface_metrics(conditions = conditions_available)
})

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report loaded data
cat("\nLoaded data summary:\n")
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing

```{r preprocess}
# Identify available metrics
# Common surface metrics from textdescriptives
potential_metrics <- c(
  # Length metrics
  "n_tokens", "n_words", "n_chars", "n_sentences",
  # Complexity
  "avg_word_length", "avg_sentence_length", "n_syllables",
  # Diversity
  "ttr", "mtld", "hdd",
  # Readability
  "flesch_reading_ease", "flesch_kincaid_grade", "gunning_fog",
  "smog", "automated_readability_index", "coleman_liau_index",
  # Other
  "n_unique_tokens", "entropy"
)

# Find which metrics are present
available_metrics <- intersect(potential_metrics, names(df_combined))
cat("Available metrics:", paste(available_metrics, collapse = ", "), "\n")

# Ensure we have at least some metrics
if (length(available_metrics) == 0) {
  stop("No recognized surface metrics found in the data")
}
```

# Descriptive Statistics

```{r descriptives}
# Summary statistics by condition for key metrics
if ("n_tokens" %in% available_metrics || "n_words" %in% available_metrics) {
  word_col <- if ("n_words" %in% available_metrics) "n_words" else "n_tokens"
  
  desc_length <- df_combined %>%
    group_by(condition) %>%
    summarise(
      n_turns = n(),
      mean_words = mean(!!sym(word_col), na.rm = TRUE),
      sd_words = sd(!!sym(word_col), na.rm = TRUE),
      median_words = median(!!sym(word_col), na.rm = TRUE),
      .groups = "drop"
    )
  
  desc_length %>%
    knitr::kable(
      digits = 2,
      caption = "Word count by condition"
    )
}
```

```{r metric-distributions, fig.width=12, fig.height=8}
# Visualize distributions for key metrics
plot_metric_dist <- function(df, metric, title) {
  if (!metric %in% names(df)) return(NULL)
  
  ggplot(df, aes(x = !!sym(metric), fill = condition)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = condition_colors) +
    labs(title = title, x = metric, y = "Density") +
    theme(legend.position = "none")
}

# Create distribution plots for available metrics
plots <- list()

if ("n_words" %in% available_metrics) {
  plots$words <- plot_metric_dist(df_combined, "n_words", "Word Count")
}
if ("n_tokens" %in% available_metrics && !("n_words" %in% available_metrics)) {
  plots$tokens <- plot_metric_dist(df_combined, "n_tokens", "Token Count")
}
if ("avg_sentence_length" %in% available_metrics) {
  plots$sent_len <- plot_metric_dist(df_combined, "avg_sentence_length", "Avg Sentence Length")
}
if ("ttr" %in% available_metrics) {
  plots$ttr <- plot_metric_dist(df_combined, "ttr", "Type-Token Ratio")
}
if ("flesch_reading_ease" %in% available_metrics) {
  plots$flesch <- plot_metric_dist(df_combined, "flesch_reading_ease", "Flesch Reading Ease")
}

if (length(plots) > 0) {
  wrap_plots(plots, ncol = 2) + 
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

# Cross-Condition Comparisons

## Word Count

```{r word-count-analysis}
word_col <- if ("n_words" %in% available_metrics) "n_words" else 
            if ("n_tokens" %in% available_metrics) "n_tokens" else NULL

if (!is.null(word_col)) {
  # Mixed model for word count
  model_words <- lmer(
    as.formula(paste(word_col, "~ condition + (1 | conversation_id)")),
    data = df_combined
  )
  
  cat("=== Word Count by Condition ===\n")
  print(summary(model_words))
  
  # Pairwise contrasts
  emm_words <- emmeans(model_words, specs = "condition")
  cat("\n=== Estimated Marginal Means ===\n")
  print(emm_words)
  
  cat("\n=== Pairwise Contrasts ===\n")
  print(pairs(emm_words, adjust = "bonferroni"))
}
```

## Lexical Diversity

```{r diversity-analysis}
if ("ttr" %in% available_metrics) {
  model_ttr <- lmer(
    ttr ~ condition + (1 | conversation_id),
    data = df_combined
  )
  
  cat("=== Type-Token Ratio by Condition ===\n")
  print(summary(model_ttr))
  
  emm_ttr <- emmeans(model_ttr, specs = "condition")
  cat("\n=== Pairwise Contrasts ===\n")
  print(pairs(emm_ttr, adjust = "bonferroni"))
}
```

## Readability

```{r readability-analysis}
if ("flesch_reading_ease" %in% available_metrics) {
  model_flesch <- lmer(
    flesch_reading_ease ~ condition + (1 | conversation_id),
    data = df_combined
  )
  
  cat("=== Flesch Reading Ease by Condition ===\n")
  print(summary(model_flesch))
  
  emm_flesch <- emmeans(model_flesch, specs = "condition")
  cat("\n=== Pairwise Contrasts ===\n")
  print(pairs(emm_flesch, adjust = "bonferroni"))
}
```

# Author-Level Comparison

Compare Author 1 vs Author 2 contributions within each condition.

```{r author-comparison}
# Check if we have author-specific columns
author_1_cols <- names(df_combined)[str_detect(names(df_combined), "^author_1_")]
author_2_cols <- names(df_combined)[str_detect(names(df_combined), "^author_2_")]

if (length(author_1_cols) > 0 && length(author_2_cols) > 0) {
  cat("Author-level metrics available\n")
  
  # Find matching pairs
  author_1_metrics <- str_replace(author_1_cols, "^author_1_", "")
  author_2_metrics <- str_replace(author_2_cols, "^author_2_", "")
  common_metrics <- intersect(author_1_metrics, author_2_metrics)
  
  if ("n_words" %in% common_metrics || "n_tokens" %in% common_metrics) {
    word_metric <- if ("n_words" %in% common_metrics) "n_words" else "n_tokens"
    
    # Pivot to long format for comparison
    df_author <- df_combined %>%
      select(conversation_id, turn, condition, 
             starts_with("author_1_"), starts_with("author_2_")) %>%
      pivot_longer(
        cols = c(paste0("author_1_", word_metric), paste0("author_2_", word_metric)),
        names_to = "author",
        values_to = "words"
      ) %>%
      mutate(author = if_else(str_detect(author, "author_1"), "Author 1", "Author 2"))
    
    # Plot
    ggplot(df_author, aes(x = author, y = words, fill = condition)) +
      geom_boxplot(alpha = 0.7, outlier.alpha = 0.1) +
      facet_wrap(~condition) +
      scale_fill_manual(values = condition_colors) +
      labs(
        title = "Word Count by Author and Condition",
        x = "Author",
        y = "Words per Turn"
      ) +
      theme(legend.position = "none")
  }
} else {
  cat("No author-level surface metrics found. Data may be at turn level without author separation.\n")
}
```

# Visualization Summary

```{r summary-plot, fig.width=12, fig.height=6}
# Compute summary statistics for available metrics
summary_stats <- df_combined %>%
  group_by(condition) %>%
  summarise(
    across(all_of(available_metrics[1:min(6, length(available_metrics))]),
           list(mean = ~mean(., na.rm = TRUE), se = ~sd(., na.rm = TRUE)/sqrt(n())),
           .names = "{.col}_{.fn}"),
    .groups = "drop"
  )

# Create bar plot for first few metrics
metrics_to_plot <- available_metrics[1:min(4, length(available_metrics))]

plots_summary <- lapply(metrics_to_plot, function(m) {
  mean_col <- paste0(m, "_mean")
  se_col <- paste0(m, "_se")
  
  if (!all(c(mean_col, se_col) %in% names(summary_stats))) return(NULL)
  
  ggplot(summary_stats, aes(x = condition, y = !!sym(mean_col), fill = condition)) +
    geom_col(alpha = 0.8, width = 0.6) +
    geom_errorbar(
      aes(ymin = !!sym(mean_col) - 1.96 * !!sym(se_col),
          ymax = !!sym(mean_col) + 1.96 * !!sym(se_col)),
      width = 0.2
    ) +
    scale_fill_manual(values = condition_colors) +
    labs(title = m, x = "", y = "Mean") +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
})

plots_summary <- compact(plots_summary)
if (length(plots_summary) > 0) {
  wrap_plots(plots_summary, ncol = 2)
}
```

# Key Findings

Based on the analysis above:

1. **Word count patterns**: [Interpret whether AI contributions are longer/shorter than human]

2. **Lexical diversity**: [Compare vocabulary richness across conditions]

3. **Readability**: [Discuss complexity/accessibility differences]

4. **AI-AI characteristics**: [Note any distinctive patterns in simulated baseline]

---

*Analysis completed: `r Sys.time()`*
