---
title: "Time-Resolved Asymmetry: Human-AI vs Human-Human vs AI-AI"
subtitle: "How story-internal asymmetry evolves across early / middle / late phases"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(lme4, lmerTest, patchwork, emmeans)

theme_set(theme_comparison())
```

# Overview

This notebook tracks how within-dyad asymmetry evolves *across the story*.
For each metric (valence, novelty-resonance slope, semantic-exploration AUC),
we compute the per-story signed and unsigned slot-1 vs slot-2 difference
within rolling thirds (Early / Middle / Late) of the story turn window, and
test whether condition $\times$ phase interactions appear.

Two outcomes per metric:

- **Magnitude**: per-story $|A|$ in each phase.
- **Direction**: per-story signed slot-difference in each phase (with the
  identity-tied directionality reading: HA should keep a stable sign across
  phases; HH and AA should show no fixed sign).

# Phase definition

```{r phase-fn}
assign_phase <- function(df, turn_col = "turn_index") {
  df %>%
    group_by(conversation_id) %>%
    mutate(
      max_t = max(.data[[turn_col]], na.rm = TRUE),
      phase = case_when(
        .data[[turn_col]] <= max_t / 3 ~ "Early",
        .data[[turn_col]] <= 2 * max_t / 3 ~ "Middle",
        TRUE ~ "Late"
      ),
      phase = factor(phase, levels = c("Early", "Middle", "Late"))
    ) %>%
    ungroup()
}
```

# Valence Asymmetry by Phase

```{r valence-by-phase, fig.width=12, fig.height=5}
df_val <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  add_chronological_valence() %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959",
         conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225") %>%
  assign_phase()

# Pearson correlation per (story, phase, direction) on signed valence pairs.
df_val_pairs <- df_val %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(second_valence_lag1 = lag(second_valence)) %>%
  ungroup() %>%
  drop_na(first_valence, second_valence, second_valence_lag1)

valence_phase_A <- df_val_pairs %>%
  group_by(conversation_id, condition, phase) %>%
  summarise(
    n_pairs = sum(!is.na(first_valence) & !is.na(second_valence)),
    r_12 = if (n_pairs >= 3) suppressWarnings(cor(first_valence, second_valence)) else NA_real_,
    r_21 = if (n_pairs >= 3) suppressWarnings(cor(second_valence_lag1, first_valence)) else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(
    r_12 = pmin(pmax(r_12, -0.999999), 0.999999),
    r_21 = pmin(pmax(r_21, -0.999999), 0.999999),
    z_12 = atanh(r_12),
    z_21 = atanh(r_21),
    signed_diff = z_12 - z_21,
    A_valence = abs(signed_diff)
  ) %>%
  drop_na(signed_diff)

valence_phase_summary <- valence_phase_A %>%
  group_by(condition, phase) %>%
  summarise(
    n = n(),
    mean_signed = mean(signed_diff),
    sd_signed = sd(signed_diff),
    mean_abs = mean(A_valence),
    .groups = "drop"
  )

valence_phase_summary %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story valence asymmetry by condition and story phase"
  )

p_val_abs <- ggplot(valence_phase_A, aes(x = phase, y = A_valence, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Magnitude: |A_valence|", x = "Story phase", y = expression("|" * Delta * z * "|"))

p_val_signed <- ggplot(valence_phase_A, aes(x = phase, y = signed_diff, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Direction: signed Δz", x = "Story phase", y = expression(Delta * z))

p_val_abs + p_val_signed + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r valence-phase-models}
if (nrow(valence_phase_A) >= 30) {
  cat("LMM: A_valence ~ condition * phase + (1 | conversation_id)\n")
  m_abs <- lmer(A_valence ~ condition * phase + (1 | conversation_id),
                data = valence_phase_A)
  print(summary(m_abs))

  cat("\nLMM: signed_diff ~ condition * phase + (1 | conversation_id)\n")
  m_signed <- lmer(signed_diff ~ condition * phase + (1 | conversation_id),
                   data = valence_phase_A)
  print(summary(m_signed))
}
```

# Novelty-Resonance Slope Asymmetry by Phase

```{r novelty-by-phase, fig.width=12, fig.height=5}
df_nov <- load_novelty_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959",
         conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225")

df_nov_slots <- bind_rows(
  df_nov %>% transmute(condition, conversation_id, turn_index,
                       slot = "author_1",
                       surprise = author_1_surprise,
                       resonance = author_1_surprise - author_1_transience),
  df_nov %>% transmute(condition, conversation_id, turn_index,
                       slot = "author_2",
                       surprise = author_2_surprise,
                       resonance = author_2_surprise - author_2_transience)
) %>%
  drop_na(surprise, resonance) %>%
  mutate(
    condition = factor(condition,
      levels = c("human-ai", "human-human", "ai-ai"),
      labels = c("Human-AI", "Human-Human", "AI-AI")
    ),
    slot = factor(slot, levels = c("author_1", "author_2"),
                  labels = c("Slot 1", "Slot 2"))
  ) %>%
  assign_phase()

novelty_phase_A <- df_nov_slots %>%
  group_by(condition, conversation_id, phase, slot) %>%
  summarise(
    n_pts = sum(!is.na(surprise) & !is.na(resonance)),
    slope = if (n_pts >= 3) coef(lm(resonance ~ surprise))[2] else NA_real_,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = slot, values_from = slope, names_prefix = "slope_") %>%
  drop_na(`slope_Slot 1`, `slope_Slot 2`) %>%
  select(-starts_with("n_pts")) %>%
  mutate(
    signed_diff = `slope_Slot 1` - `slope_Slot 2`,
    A_influence = abs(signed_diff)
  )

novelty_phase_A %>%
  group_by(condition, phase) %>%
  summarise(
    n = n(),
    mean_signed = mean(signed_diff),
    sd_signed = sd(signed_diff),
    mean_abs = mean(A_influence),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story novelty-resonance slope asymmetry by condition and story phase"
  )

p_nov_abs <- ggplot(novelty_phase_A, aes(x = phase, y = A_influence, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Magnitude: |A_influence|", x = "Story phase",
       y = expression("|" * Delta * beta * "|"))

p_nov_signed <- ggplot(novelty_phase_A, aes(x = phase, y = signed_diff, fill = condition)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2, position = position_dodge(width = 0.75)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Direction: signed Δβ", x = "Story phase",
       y = expression(Delta * beta))

p_nov_abs + p_nov_signed + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r novelty-phase-models}
if (nrow(novelty_phase_A) >= 30) {
  cat("LMM: A_influence ~ condition * phase + (1 | conversation_id)\n")
  m_abs <- lmer(A_influence ~ condition * phase + (1 | conversation_id),
                data = novelty_phase_A)
  print(summary(m_abs))

  cat("\nLMM: signed_diff ~ condition * phase + (1 | conversation_id)\n")
  m_signed <- lmer(signed_diff ~ condition * phase + (1 | conversation_id),
                   data = novelty_phase_A)
  print(summary(m_signed))
}
```

# Mean Valence Trajectory by Slot and Condition

A descriptive companion view: how does each slot's mean valence drift across
phase, by condition? Useful for the "story-evolution" narrative without
relying on the asymmetry abstraction.

```{r mean-trajectory, fig.width=12, fig.height=5}
df_val_long <- df_val %>%
  pivot_longer(c(author_1_valence, author_2_valence),
               names_to = "slot", values_to = "valence") %>%
  mutate(slot = factor(slot,
                       levels = c("author_1_valence", "author_2_valence"),
                       labels = c("Slot 1", "Slot 2")))

mean_traj <- df_val_long %>%
  group_by(condition, phase, slot) %>%
  summarise(
    mean_val = mean(valence, na.rm = TRUE),
    se = sd(valence, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(mean_traj, aes(x = phase, y = mean_val, color = slot, group = slot)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_val - 1.96 * se, ymax = mean_val + 1.96 * se),
                width = 0.15) +
  facet_wrap(~ condition) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Mean valence by slot, phase, and condition",
    subtitle = "Shows whether slot-level emotional baselines diverge or converge across the story",
    x = "Story phase", y = "Mean valence", color = "Slot"
  )
```

# Summary

What this notebook adds beyond the cross-section comparisons:

- **Per-phase magnitude $|A|$** for valence and novelty-resonance: tests whether asymmetry strengthens, weakens, or stays flat across the story.
- **Per-phase signed asymmetry**: combined with per-story directional-stability, this is the strongest visual evidence for HA-specific identity-tied directionality. A signed-asymmetry boxplot that stays *consistently above zero* in HA across all three phases, while HH/AA hover around zero, is the publication-quality figure for the reframed RQ1.
- **Mean valence trajectory by slot and condition**: a story-evolution view that does not depend on the asymmetry abstraction and helps illustrate qualitative drift.

Open additions worth considering:

- Add a per-phase semantic-exploration AUC view (the lag $k$ axis is itself an evolution axis, but a phase split could surface fine-grained shifts).
- Replace fixed thirds with overlapping rolling windows for smoother trajectories.

*Analysis completed: `r Sys.time()`*
