---
title: "Valence Alignment Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Contrasts on the Exchange-Aligned Data Model"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize,
  stringr
)

theme_set(theme_comparison())
```


This notebook compares **valence alignment** across the three experimental conditions using the new exchange-aligned structure.

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing together

Important structural note:

- `author_1` / `author_2` are now treated as **slot 1 / slot 2** in the cleaned exchange sequence.
- In **Human-AI**, those slots also map onto human vs AI roles.
- In **Human-Human** and **AI-AI**, the slots are positional sides of the exchange, not stable identity labels.

The analysis window is restricted to **complete exchanges** and the default aligned turn range (`analysis_turn` 1 to 9).

**Primary cross-condition object.** Each story contributes one unsigned, label-invariant dyad asymmetry magnitude:
$$A_{\text{valence}}(i) \;=\; \bigl|\, \operatorname{atanh}(r_{1\to 2,i}) - \operatorname{atanh}(r_{2\to 1,i}) \,\bigr|$$
where $r_{1\to 2,i}$ is the per-story Pearson correlation between same-turn slot 1 and slot 2 valence, and $r_{2\to 1,i}$ is the correlation between slot 2's previous-turn valence and slot 1's current-turn valence. The main question is whether $A_{\text{valence}}$ differs across conditions. The directional-slope LMMs are retained as supplemental views. Starter-mechanism and Human-AI speaker-type decompositions are deferred to a follow-up pass.

**Article-use note.** The primary cross-condition claim should use the unsigned per-story $A_{\text{valence}}$ test below. Signed directional slot models are diagnostic only: in Human-Human and AI-AI, Slot 1 / Slot 2 are positional labels rather than theoretically meaningful agent identities.

# Load Data

```{r load-data}
cat("Data availability:\n")
print(check_data_availability())

df_combined <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225", conversation_id != "conv_0bb56093-3033-4615-bb70-ebfa4135589a", conversation_id != "conv_0f18b30f-7d4b-4681-b98e-a0ff4f2b5256", conversation_id != "conv_72218cb5-e59c-4c93-a4b9-a057fe5dad80", conversation_id != "conv_7c23347e-6172-4c84-9fd0-45ef34290bd5")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    min_turn = min(turn_index, na.rm = TRUE),
    max_turn = max(turn_index, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    caption = "Loaded exchange-aligned valence data by condition"
  )
```

# Preprocessing

```{r preprocess}
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    author_1_valence_lag1 = lag(author_1_valence),
    author_2_valence_lag1 = lag(author_2_valence)
  ) %>%
  ungroup() %>%
  filter(turn_index > 1) %>%
  drop_na(
    author_1_valence, author_2_valence,
    author_1_valence_lag1, author_2_valence_lag1
  )

df_alignment_long <- bind_rows(
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "slot1_to_slot2",
      predictor_valence = author_1_valence,
      response_valence = author_2_valence
    ),
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "slot2_to_slot1",
      predictor_valence = author_2_valence_lag1,
      response_valence = author_1_valence
    )
) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("slot1_to_slot2", "slot2_to_slot1"),
      labels = c("Slot 1 → Slot 2", "Slot 2(t-1) → Slot 1")
    )
  )

df_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z = as.numeric(scale(response_valence))
  ) %>%
  ungroup() %>%
  drop_na(predictor_z, response_z)

contrasts(df_std$alignment_direction) <- contr.sum(2)

cat("Analysis rows:", nrow(df_std), "\n")
cat("Stories:", n_distinct(df_std$conversation_id), "\n")
```

# Descriptive Checks

```{r descriptives}
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    slot_1_mean = mean(author_1_valence, na.rm = TRUE),
    slot_2_mean = mean(author_2_valence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Mean projected valence by condition"
  )
```

```{r valence-distributions, fig.width=12, fig.height=4}
p_slot1 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 1 Valence", x = "Projected Valence", y = "Density") +
  theme(legend.position = "none")

p_slot2 <- df_analysis %>%
  ggplot(aes(x = author_2_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 2 Valence", x = "Projected Valence", y = "Density")

p_slot1 + p_slot2 + plot_layout(guides = "collect")
```

# Primary: Per-Story Asymmetry Magnitude

```{r primary-asymmetry}
# Per-story Pearson correlations under both pairing rules, on non-standardized
# predictor/response pairs (r is scale-invariant).
min_pairs <- 3

df_r12 <- df_alignment_long %>%
  filter(alignment_direction == "Slot 1 \u2192 Slot 2") %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_pairs = sum(!is.na(predictor_valence) & !is.na(response_valence)),
    r = if (n_pairs >= min_pairs) {
      suppressWarnings(cor(predictor_valence, response_valence, use = "complete.obs"))
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  transmute(conversation_id, condition, n_pairs_12 = n_pairs, r_12 = r)

df_r21 <- df_alignment_long %>%
  filter(alignment_direction == "Slot 2(t-1) \u2192 Slot 1") %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_pairs = sum(!is.na(predictor_valence) & !is.na(response_valence)),
    r = if (n_pairs >= min_pairs) {
      suppressWarnings(cor(predictor_valence, response_valence, use = "complete.obs"))
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  transmute(conversation_id, condition, n_pairs_21 = n_pairs, r_21 = r)

df_asym <- df_r12 %>%
  inner_join(df_r21, by = c("conversation_id", "condition")) %>%
  mutate(
    # Clamp |r|<1 before atanh to avoid infinities on degenerate stories
    r_12_c = pmin(pmax(r_12, -0.999999), 0.999999),
    r_21_c = pmin(pmax(r_21, -0.999999), 0.999999),
    z_12 = atanh(r_12_c),
    z_21 = atanh(r_21_c),
    signed_z_diff = z_12 - z_21,
    A_valence = abs(signed_z_diff)
  ) %>%
  filter(!is.na(A_valence))

cat("Stories with both directional correlations available:\n")
df_asym %>%
  count(condition) %>%
  knitr::kable(caption = "Per-story A_valence sample sizes by condition")

# Sanity checks
stopifnot(all(df_asym$A_valence >= 0))
```

```{r primary-asymmetry-descriptives}
df_asym %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_A = mean(A_valence),
    sd_A = sd(A_valence),
    median_A = median(A_valence),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story |Fisher-z asymmetry| by condition"
  )
```

```{r primary-asymmetry-test}
cat("One-way ANOVA: A_valence ~ condition\n")
print(summary(aov(A_valence ~ condition, data = df_asym)))

cat("\nKruskal-Wallis (robust): A_valence ~ condition\n")
print(kruskal.test(A_valence ~ condition, data = df_asym))

cat("\nPairwise Wilcoxon (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_asym$A_valence,
  df_asym$condition,
  p.adjust.method = "BH"
))
```

```{r primary-asymmetry-plot, fig.width=9, fig.height=5}
ggplot(df_asym, aes(x = condition, y = A_valence, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Per-story valence alignment asymmetry by condition",
    subtitle = "A_valence(i) = |atanh(r_{1\u21922,i}) - atanh(r_{2\u21921,i})|",
    x = "Condition",
    y = expression("|" * Delta * z * "|")
  ) +
  theme(legend.position = "none")
```

# Bootstrap CIs and Effect Sizes on `A_valence`

```{r boot-effects-valence}
boot_A <- bootstrap_condition_means(df_asym, "A_valence", R = 10000)
boot_A %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of A_valence (R = 10,000)"
  )

effect_sizes_pairwise(df_asym, "A_valence") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on A_valence"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-valence}
planned_contrast_HA_vs_same(df_asym, "A_valence") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: A_valence in HA vs the average of HH and AA"
  )
```

# Directional Stability of Signed Valence Asymmetry

Tests whether the *direction* of the slot-1 vs slot-2 alignment difference is
consistent across stories within each condition. The signed quantity is
$\text{signed}_i = \operatorname{atanh}(r_{1\to 2,i}) - \operatorname{atanh}(r_{2\to 1,i})$.
HA-specific identity-tied directionality predicts high `stability` and
`sign_consistency` in HA, low in HH and AA, even when unsigned `A_valence`
is similar.

```{r directional-stability-valence}
ds_val <- directional_stability(df_asym, "signed_z_diff")
ds_val$by_condition %>%
  knitr::kable(
    digits = 3,
    caption = "Per-condition directional stability of signed valence asymmetry"
  )

cat(sprintf(
  "Brown-Forsythe variance-equality p = %.4f\n", ds_val$variance_equality_p
))
```

```{r directional-stability-valence-plot, fig.width=9, fig.height=5}
ggplot(df_asym, aes(x = condition, y = signed_z_diff, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Signed valence-asymmetry direction by condition",
    subtitle = "atanh(r_{1→2}) - atanh(r_{2→1}); positive = slot 1 leads",
    x = "Condition",
    y = expression(Delta * z)
  ) +
  theme(legend.position = "none")
```

# Permutation Test on `A_valence ~ condition`

```{r permutation-valence}
permutation_test_A(df_asym, "A_valence", R = 10000) %>%
  knitr::kable(
    digits = 4,
    caption = "Permutation test on A_valence ~ condition (R = 10,000 story-label permutations)"
  )
```

# Summary

Primary cross-condition result: per-story $A_{\text{valence}} = |\operatorname{atanh}(r_{1\to2,i}) - \operatorname{atanh}(r_{2\to1,i})|$, tested with ANOVA + Kruskal and pairwise BH-adjusted Wilcoxon.

Robustness layers added:

- bootstrap 95% CIs on per-condition mean of $A_{\text{valence}}$
- pairwise Cliff's $\delta$ effect sizes
- planned single-df contrast HA vs (HH+AA)/2
- 10k story-label permutation test on $A_{\text{valence}} \sim \text{condition}$

The Human-AI signed Human/AI directional decomposition lives in [analysis/human-ai/valence_alignment_analysis.Rmd](../human-ai/valence_alignment_analysis.Rmd) and is not duplicated here.

Deferred to a follow-up pass:

- starter-mechanism analysis (signed, starter-normalized asymmetry across all conditions)

*Analysis completed: `r Sys.time()`*