---
title: "Valence Alignment Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Statistical Contrasts"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook performs comparative analysis across all three experimental conditions:

-   **Human-AI**: Human co-writing with an LLM
-   **Human-Human**: Two humans co-writing together\
-   **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Questions:** 1. Do alignment patterns differ across the three conditions? 2. Is alignment asymmetric in all conditions or condition-specific? 3. Does the AI-AI baseline show the strongest alignment (as expected)? 4. How do effect sizes compare across conditions?

# Load Data

```{r load-data}
# Check data availability
cat("Data availability:\n")
avail <- check_data_availability()
print(avail)

# Load valence data using standardized loader
df_combined <- tryCatch({
  load_valence_data(conditions = c("human-ai", "human-human", "ai-ai"))
}, error = function(e) {
  # Fallback: try loading whatever is available
  conditions_available <- c()
  
  for (cond in c("human-ai", "human-human", "ai-ai")) {
    path <- here("data", cond, "processed", "dyadic_sentiment_scores.parquet")
    if (file.exists(path)) {
      conditions_available <- c(conditions_available, cond)
    }
  }
  
  if (length(conditions_available) == 0) {
    stop("No valence data available for any condition")
  }
  
  cat("Loading available conditions:", paste(conditions_available, collapse = ", "), "\n")
  load_valence_data(conditions = conditions_available)
})

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report loaded data
cat("\nLoaded data summary:\n")
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing

```{r preprocess}
# Compute lagged valence for alignment analysis
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn) %>%
  mutate(
    author_1_valence_lag1 = lag(author_1_valence, n = 1),
    author_2_valence_lag1 = lag(author_2_valence, n = 1)
  ) %>%
  ungroup() %>%
  filter(turn > 1) %>%
  drop_na(author_1_valence, author_2_valence, 
          author_1_valence_lag1, author_2_valence_lag1)

cat("Analysis dataset:", nrow(df_analysis), "turns across", 
    n_distinct(df_analysis$conversation_id), "stories\n")
```

## Build Long-Format Directional Dataset

Following the within-condition analysis approach: we stack both alignment directions into a single long-format dataset. This enables proper statistical testing of directional asymmetry via predictor × direction interactions.

-   **A1 → A2** (same-turn): Author 2 responds to Author 1's current valence
-   **A2 → A1** (cross-turn): Author 1 responds to Author 2's previous valence

```{r build-long-format}
# Same-turn: A1 → A2 (Author 2 aligns to Author 1's current valence)
df_a1_to_a2 <- df_analysis %>%
  transmute(
    conversation_id,
    condition,
    turn,
    alignment_direction = "A1_to_A2",
    predictor_valence = author_1_valence,
    response_valence = author_2_valence
  )

# Cross-turn: A2 → A1 (Author 1 aligns to Author 2's previous valence)
df_a2_to_a1 <- df_analysis %>%
  transmute(
    conversation_id,
    condition,
    turn,
    alignment_direction = "A2_to_A1",
    predictor_valence = author_2_valence_lag1,
    response_valence = author_1_valence
  )

# Combine into long format
df_alignment_long <- bind_rows(df_a1_to_a2, df_a2_to_a1) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("A1_to_A2", "A2_to_A1")
    )
  )

cat("=== Long-Format Alignment Data ===\n")
df_alignment_long %>%
  group_by(condition, alignment_direction) %>%
  summarise(
    n_rows = n(),
    n_stories = n_distinct(conversation_id),
    mean_predictor = mean(predictor_valence, na.rm = TRUE),
    mean_response = mean(response_valence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

## Standardize Within Conversation × Direction

Z-score predictor and response valence within each (conversation_id × direction) cell. This ensures alignment coefficients are comparable across directions and conditions.

```{r standardize}
df_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z = as.numeric(scale(response_valence))
  ) %>%
  ungroup() %>%
  drop_na(predictor_z, response_z)

# Set up sum-coded contrasts for direction (for interpretable interaction)
df_std$alignment_direction <- factor(
  df_std$alignment_direction,
  levels = c("A1_to_A2", "A2_to_A1")
)
contrasts(df_std$alignment_direction) <- contr.sum(2)

cat("Standardized dataset:", nrow(df_std), "observations\n")
```

# Descriptive Statistics

```{r descriptives}
# Valence descriptives by condition
desc_stats <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    mean_a1_valence = mean(author_1_valence, na.rm = TRUE),
    sd_a1_valence = sd(author_1_valence, na.rm = TRUE),
    mean_a2_valence = mean(author_2_valence, na.rm = TRUE),
    sd_a2_valence = sd(author_2_valence, na.rm = TRUE),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3, 
    caption = "Valence descriptives by condition",
    col.names = c("Condition", "Stories", "Turns", 
                  "Author 1 M", "Author 1 SD", 
                  "Author 2 M", "Author 2 SD")
  )
```

```{r valence-distributions, fig.width=12, fig.height=4}
# Valence distributions by condition
p_dist_a1 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Author 1 Valence Distribution", x = "Valence", y = "Density") +
  theme(legend.position = "none")

p_dist_a2 <- df_analysis %>%
  ggplot(aes(x = author_2_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Author 2 Valence Distribution", x = "Valence", y = "Density")

p_dist_a1 + p_dist_a2 + plot_layout(guides = "collect")
```

# Alignment Analysis

## Within-Condition Asymmetry Models

For each condition, we fit a single mixed model with **predictor × direction** interaction on standardized valence. This properly tests whether alignment strength differs by direction within each condition.

```{r within-condition-asymmetry}
conditions <- unique(df_std$condition)
within_condition_results <- list()

for (cond in conditions) {
  df_cond <- filter(df_std, condition == cond)
  
  if (nrow(df_cond) < 100 || n_distinct(df_cond$conversation_id) < 5) {
    cat("Skipping", cond, "- insufficient data\n")
    next
  }
  
  # Set up contrasts for this subset
  df_cond$alignment_direction <- factor(
    df_cond$alignment_direction,
    levels = c("A1_to_A2", "A2_to_A1")
  )
  contrasts(df_cond$alignment_direction) <- contr.sum(2)
  
  # Fit model with direction interaction
  m <- tryCatch({
    lmer(response_z ~ predictor_z * alignment_direction + 
           (1 + predictor_z | conversation_id),
         data = df_cond,
         control = lmerControl(optimizer = "bobyqa"))
  }, error = function(e) {
    # Fallback to simpler random effects if convergence fails
    tryCatch({
      lmer(response_z ~ predictor_z * alignment_direction + 
             (1 | conversation_id),
           data = df_cond)
    }, error = function(e2) NULL)
  })
  
  if (!is.null(m)) {
    within_condition_results[[cond]] <- list(
      model = m,
      n_obs = nrow(df_cond),
      n_stories = n_distinct(df_cond$conversation_id)
    )
    
    cat("\n=== ", cond, " ===\n", sep = "")
    cat("N =", nrow(df_cond), "observations from", 
        n_distinct(df_cond$conversation_id), "stories\n\n")
    print(summary(m))
    
    # Extract slopes per direction and test difference
    slopes <- emtrends(m, specs = "alignment_direction", var = "predictor_z")
    cat("\n--- Slopes by Direction ---\n")
    print(slopes)
    
    cat("\n--- Asymmetry Test (A1→A2 vs A2→A1) ---\n")
    print(contrast(slopes, method = "revpairwise"))
  }
}
```

## Summary of Within-Condition Asymmetry

```{r within-asymmetry-summary}
# Extract key results into a summary table
asymmetry_summary <- tibble()

for (cond in names(within_condition_results)) {
  m <- within_condition_results[[cond]]$model
  
  # Get slopes
  slopes <- emtrends(m, specs = "alignment_direction", var = "predictor_z")
  slopes_df <- as.data.frame(slopes)
  
  # Get contrast
  contrast_result <- contrast(slopes, method = "revpairwise")
  contrast_df <- as.data.frame(contrast_result)
  
  asymmetry_summary <- bind_rows(asymmetry_summary, tibble(
    condition = cond,
    slope_A1_to_A2 = slopes_df$predictor_z.trend[slopes_df$alignment_direction == "A1_to_A2"],
    slope_A2_to_A1 = slopes_df$predictor_z.trend[slopes_df$alignment_direction == "A2_to_A1"],
    difference = contrast_df$estimate[1],
    se = contrast_df$SE[1],
    t = contrast_df$t.ratio[1],
    p = contrast_df$p.value[1],
    interpretation = case_when(
      contrast_df$p.value[1] < 0.05 & contrast_df$estimate[1] > 0 ~ "A1→A2 stronger",
      contrast_df$p.value[1] < 0.05 & contrast_df$estimate[1] < 0 ~ "A2→A1 stronger",
      TRUE ~ "No asymmetry"
    )
  ))
}

asymmetry_summary %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-Condition Asymmetry: Does alignment differ by direction?",
    col.names = c("Condition", "β (A1→A2)", "β (A2→A1)", "Diff", "SE", "t", "p", "Interpretation", "")
  )
```

# Cross-Condition Comparison

## Full Model: Condition × Direction × Predictor

Test whether asymmetry patterns differ across conditions using a three-way interaction.

```{r cross-condition-model}
# Prepare condition factor with sum coding for interpretable interactions
df_std$condition <- factor(df_std$condition)
contrasts(df_std$condition) <- contr.sum(length(levels(df_std$condition)))

# Full model with three-way interaction
model_full <- tryCatch({
  lmer(response_z ~ predictor_z * alignment_direction * condition + 
         (1 + predictor_z | conversation_id),
       data = df_std,
       control = lmerControl(optimizer = "bobyqa"))
}, error = function(e) {
  # Fallback to simpler random effects
  lmer(response_z ~ predictor_z * alignment_direction * condition + 
         (1 | conversation_id),
       data = df_std)
})

cat("=== Full Model: predictor × direction × condition ===\n")
print(summary(model_full))
```

## Condition Comparison of Asymmetry

```{r condition-asymmetry-comparison}
# Get slopes for each condition × direction combination
slopes_by_cond <- emtrends(model_full, 
                            specs = c("condition", "alignment_direction"),
                            var = "predictor_z")

cat("=== Alignment Slopes by Condition × Direction ===\n")
print(slopes_by_cond)

# Test asymmetry within each condition (A1→A2 vs A2→A1)
cat("\n=== Asymmetry Contrasts Within Each Condition ===\n")
asymmetry_contrasts <- contrast(slopes_by_cond, 
                                 method = "revpairwise",
                                 by = "condition")
print(asymmetry_contrasts)

# Test whether asymmetry differs across conditions
cat("\n=== Does Asymmetry Differ Across Conditions? ===\n")
# Compare the A1→A2 vs A2→A1 difference across conditions
asymmetry_by_cond <- contrast(slopes_by_cond, 
                               interaction = c("revpairwise", "revpairwise"))
print(asymmetry_by_cond)
```

## Pairwise Condition Contrasts for Each Direction

```{r direction-specific-contrasts}
# Compare conditions within each direction
cat("=== Condition Contrasts for A1→A2 Direction ===\n")
a1_to_a2_contrasts <- contrast(slopes_by_cond,
                                method = "pairwise",
                                by = "alignment_direction",
                                adjust = "bonferroni")
print(a1_to_a2_contrasts)
```

# Visualization

```{r alignment-plot, fig.width=14, fig.height=5}
# Directional alignment by condition
p1 <- df_alignment_long %>%
  filter(alignment_direction == "A2_to_A1") %>%
  ggplot(aes(x = predictor_valence, y = response_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "A2 → A1 Alignment",
    subtitle = "Does Author 1 follow Author 2's previous valence?",
    x = "Author 2 Valence (t-1)",
    y = "Author 1 Valence (t)",
    color = "Condition"
  ) +
  theme(legend.position = "none")

p2 <- df_alignment_long %>%
  filter(alignment_direction == "A1_to_A2") %>%
  ggplot(aes(x = predictor_valence, y = response_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "A1 → A2 Alignment",
    subtitle = "Does Author 2 match Author 1's current valence?",
    x = "Author 1 Valence (t)",
    y = "Author 2 Valence (t)",
    color = "Condition"
  )

p1 + p2 + plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

```{r coefficient-plot, fig.width=12, fig.height=6}
# Extract slopes for visualization
slopes_df <- as.data.frame(slopes_by_cond) %>%
  mutate(
    ci_lower = predictor_z.trend - 1.96 * SE,
    ci_upper = predictor_z.trend + 1.96 * SE,
    direction_label = recode(alignment_direction,
      "A1_to_A2" = "A1 → A2 (same-turn)",
      "A2_to_A1" = "A2 → A1 (cross-turn)"
    )
  )

ggplot(slopes_df, aes(x = condition, y = predictor_z.trend, fill = direction_label)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8),
    width = 0.25
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("A1 → A2 (same-turn)" = "#377EB8", 
                                "A2 → A1 (cross-turn)" = "#FF7F00")) +
  labs(
    title = "Alignment Coefficients by Condition and Direction",
    subtitle = "Standardized slopes (β) with 95% CI",
    y = "Alignment Coefficient (β)",
    x = "Condition",
    fill = "Direction"
  ) +
  theme(legend.position = "bottom")
```

```{r asymmetry-plot, fig.width=10, fig.height=5}
# Plot asymmetry (difference between directions) by condition
if (nrow(asymmetry_summary) > 0) {
  asymmetry_summary %>%
    ggplot(aes(x = condition, y = difference, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(
      aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
      width = 0.2
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Directional Asymmetry by Condition",
      subtitle = "Difference: (A1→A2) - (A2→A1). Positive = A1→A2 stronger",
      y = "Asymmetry (Δβ)",
      x = "Condition"
    ) +
    theme(legend.position = "none")
}
```

# Summary

```{r summary-table}
# Final summary combining slopes and asymmetry
if (nrow(asymmetry_summary) > 0) {
  asymmetry_summary %>%
    mutate(
      sig = case_when(
        p < 0.001 ~ "***",
        p < 0.01 ~ "**",
        p < 0.05 ~ "*",
        p < 0.1 ~ ".",
        TRUE ~ "ns"
      ),
      `A1→A2` = round(slope_A1_to_A2, 3),
      `A2→A1` = round(slope_A2_to_A1, 3),
      `Asymmetry (p)` = paste0(round(difference, 3), " (", format.pval(p, digits = 2), ") ", sig)
    ) %>%
    select(condition, `A1→A2`, `A2→A1`, `Asymmetry (p)`, interpretation) %>%
    knitr::kable(
      caption = "Summary: Alignment slopes and asymmetry by condition",
      col.names = c("Condition", "β A1→A2", "β A2→A1", "Asymmetry (p)", "Interpretation")
    )
}
```

# Key Findings

Based on the analysis above:

1.  **Within-condition asymmetry**: Each condition is tested for directional asymmetry using a predictor × direction interaction within a single model. A significant interaction indicates that alignment strength differs between the A1→A2 and A2→A1 directions.

2.  **Cross-condition comparison**: The three-way interaction (predictor × direction × condition) tests whether the *pattern* of asymmetry differs across conditions. A significant three-way interaction means some conditions show asymmetry while others do not, or asymmetry is in opposite directions.

3.  **Interpretation for Human-AI**: In the human-ai condition, if A1→A2 (User→AI) is significantly stronger than A2→A1 (AI→User), this indicates the AI adapts to human valence but humans do not follow AI valence.

4.  **Comparison insights**:

    -   If Human-Human shows no asymmetry but Human-AI does, this suggests asymmetric accommodation is specific to human-AI interaction.
    -   If AI-AI shows strong symmetric alignment, it serves as a baseline for what "full mutual alignment" looks like.

------------------------------------------------------------------------

*Analysis completed: `r Sys.time()`*
