---
title: "Valence Alignment Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Contrasts on the Exchange-Aligned Data Model"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize,
  stringr
)

theme_set(theme_comparison())
```


This notebook compares **valence alignment** across the three experimental conditions using the new exchange-aligned structure.

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing together

Important structural note:

- `author_1` / `author_2` are now treated as **slot 1 / slot 2** in the cleaned exchange sequence.
- In **Human-AI**, those slots also map onto human vs AI roles.
- In **Human-Human** and **AI-AI**, the slots are positional sides of the exchange, not stable identity labels.

The analysis window is restricted to **complete exchanges** and the default aligned turn range (`analysis_turn` 1 to 9).

**Primary alignment object.** Each story contributes one unsigned, label-invariant dyad asymmetry magnitude:
$$A_{\text{valence}}(i) \;=\; \bigl|\, \operatorname{atanh}(r_{\text{first}\to \text{second},i}) - \operatorname{atanh}(r_{\text{second}\to \text{first},i}) \,\bigr|$$
where $r_{\text{first}\to \text{second},i}$ is the per-story Pearson correlation between same-row chronological first and second turns, and $r_{\text{second}\to \text{first},i}$ is the correlation between the previous row's chronological second turn and the current row's chronological first turn. The main question is whether $A_{\text{valence}}$ differs across conditions. The directional-slope LMMs are retained as supplemental views. Starter-mechanism and Human-AI speaker-type decompositions are deferred to a follow-up pass.

**Primary baseline-valence object.** Before the alignment analysis, we also test agent-side baseline valence asymmetry. Each story contributes one unsigned baseline gap:
$$A_{\text{baseline}}(i) \;=\; \bigl|\, \bar{v}_{\text{slot 1},i} - \bar{v}_{\text{slot 2},i} \,\bigr|$$
where $\bar{v}_{\text{slot 1},i}$ and $\bar{v}_{\text{slot 2},i}$ are the mean projected valence scores for the two sides of story $i$. This is a baseline-level difference, not a turn-by-turn adaptation measure.

**Article-use note.** The primary cross-condition claim should use the unsigned per-story $A_{\text{valence}}$ test below. Signed directional models are diagnostic only: in Human-Human and AI-AI, chronological first / second tracks starter-side ordering rather than theoretically meaningful agent identities.

# Load Data

```{r load-data}
cat("Data availability:\n")
print(check_data_availability())

df_combined <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  prepare_condition_factor(reference = "human-ai") %>%
  add_chronological_valence() %>%
  filter(conversation_id != "conv_ed575a06c11d42358e3eeb7826d2f959", conversation_id != "conv_a63a08273d0a4704a7638e4cd6850225", conversation_id != "conv_0bb56093-3033-4615-bb70-ebfa4135589a", conversation_id != "conv_72218cb5-e59c-4c93-a4b9-a057fe5dad80", author_2 != "Network error while generating a response.")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    min_turn = min(turn_index, na.rm = TRUE),
    max_turn = max(turn_index, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    caption = "Loaded exchange-aligned valence data by condition"
  )
```

# Primary Baseline Valence Asymmetry

This section tests whether the two sides of a story differ in their average valence baseline. The signed slot gap is only role-interpretable in Human-AI; for the cross-condition comparison, the unsigned per-story gap $A_{\text{baseline}}$ is the main quantity.

```{r baseline-asymmetry}
df_baseline_story <- df_combined %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_exchanges = sum(!is.na(author_1_valence) & !is.na(author_2_valence)),
    slot_1_baseline = mean(author_1_valence, na.rm = TRUE),
    slot_2_baseline = mean(author_2_valence, na.rm = TRUE),
    story_baseline = mean(c(author_1_valence, author_2_valence), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_gap = slot_1_baseline - slot_2_baseline,
    A_baseline = abs(baseline_gap)
  ) %>%
  drop_na(A_baseline, story_baseline)

df_baseline_story %>%
  group_by(condition) %>%
  summarise(
    n_stories = n(),
    mean_A_baseline = mean(A_baseline),
    sd_A_baseline = sd(A_baseline),
    median_A_baseline = median(A_baseline),
    mean_signed_gap = mean(baseline_gap),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story baseline valence asymmetry by condition"
  )
```

```{r baseline-asymmetry-tests}
cat("One-way ANOVA: A_baseline ~ condition\n")
print(summary(aov(A_baseline ~ condition, data = df_baseline_story)))

cat("\nKruskal-Wallis (robust): A_baseline ~ condition\n")
print(kruskal.test(A_baseline ~ condition, data = df_baseline_story))

cat("\nPairwise Wilcoxon (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_baseline_story$A_baseline,
  df_baseline_story$condition,
  p.adjust.method = "BH"
))
```

```{r baseline-asymmetry-plot, fig.width=9, fig.height=5}
p_baseline_asym <- ggplot(df_baseline_story, aes(x = condition, y = A_baseline, fill = condition)) +
  geom_violin(alpha = 0.55, trim = FALSE) +
  geom_boxplot(width = 0.18, alpha = 0.85, outlier.alpha = 0.15) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 1.4) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Per-story baseline valence asymmetry by condition",
    subtitle = "A_baseline(i) = |mean(slot_1_valence) - mean(slot_2_valence)|",
    x = "Condition",
    y = "Baseline valence asymmetry"
  ) +
  theme(legend.position = "none")

print(p_baseline_asym)
ggsave(
  file.path(GRAPHS_DIR, "valence_baseline_asymmetry_per_story.png"),
  p_baseline_asym,
  width = 9,
  height = 5,
  dpi = 300
)
```

```{r baseline-effects}
bootstrap_condition_means(df_baseline_story, "A_baseline", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of A_baseline (R = 10,000)"
  )

effect_sizes_pairwise(df_baseline_story, "A_baseline") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on A_baseline"
  )

planned_contrast_HA_vs_same(df_baseline_story, "A_baseline") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: A_baseline in HA vs the average of HH and AA"
  )
```

# Condition Baseline Valence

This plot shows overall valence baseline by condition, pooling turn-level valence scores across both slots. It is not an asymmetry measure; it asks whether the conditions differ in overall emotional tone.

```{r condition-baseline-plot, fig.width=9, fig.height=5}
df_condition_valence <- df_combined %>%
  select(condition, conversation_id, turn_index, author_1_valence, author_2_valence) %>%
  pivot_longer(
    cols = c(author_1_valence, author_2_valence),
    names_to = "slot",
    values_to = "valence"
  ) %>%
  drop_na(valence)

p_condition_baseline <- ggplot(df_condition_valence, aes(x = valence, fill = condition)) +
  geom_density(alpha = 0.45) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Condition-level valence baseline",
    subtitle = "Turn-level valence scores pooled across both slots",
    x = "Projected valence",
    y = "Density"
  ) +
  theme(legend.position = "bottom")

print(p_condition_baseline)
ggsave(
  file.path(GRAPHS_DIR, "valence_condition_baseline_density.png"),
  p_condition_baseline,
  width = 9,
  height = 5,
  dpi = 300
)
```

# Preprocessing

```{r preprocess}
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    second_valence_lag1 = lag(second_valence)
  ) %>%
  ungroup() %>%
  filter(turn_index > 1) %>%
  drop_na(
    first_valence, second_valence, second_valence_lag1
  )

df_alignment_long <- bind_rows(
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "first_to_second",
      predictor_valence = first_valence,
      response_valence = second_valence
    ),
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "second_to_first",
      predictor_valence = second_valence_lag1,
      response_valence = first_valence
    )
) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("first_to_second", "second_to_first"),
      labels = c("First → Second", "Second(t-1) → First")
    )
  )

df_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z = as.numeric(scale(response_valence))
  ) %>%
  ungroup() %>%
  drop_na(predictor_z, response_z)

contrasts(df_std$alignment_direction) <- contr.sum(2)

cat("Analysis rows:", nrow(df_std), "\n")
cat("Stories:", n_distinct(df_std$conversation_id), "\n")
```

# Descriptive Checks

```{r descriptives}
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    slot_1_mean = mean(author_1_valence, na.rm = TRUE),
    slot_2_mean = mean(author_2_valence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Mean projected valence by condition"
  )
```

```{r valence-distributions, fig.width=12, fig.height=4}
p_slot1 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 1 Valence", x = "Projected Valence", y = "Density") +
  theme(legend.position = "none")

p_slot2 <- df_analysis %>%
  ggplot(aes(x = author_2_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 2 Valence", x = "Projected Valence", y = "Density")

slot_combined <- p_slot1 + p_slot2 + plot_layout(guides = "collect")
print(slot_combined)
ggsave(
  file.path(GRAPHS_DIR, "valence_slot_distributions.png"),
  slot_combined,
  width = 12,
  height = 4,
  dpi = 300
)
```

# Directional Valence Alignment by Condition

The figure below visualizes responder valence as a function of partner valence under the two exchange-aligned pairing rules. In Human-AI, the directions map onto Human $\rightarrow$ AI and AI $\rightarrow$ Human response patterns. In Human-Human and AI-AI, they should be interpreted as chronological exchange positions rather than stable agent identities.

```{r directional-alignment-three-panel-summary}
df_alignment_long %>%
  group_by(condition, alignment_direction) %>%
  summarise(
    n_pairs = sum(!is.na(predictor_valence) & !is.na(response_valence)),
    slope = if (n_pairs >= 3) {
      coef(lm(response_valence ~ predictor_valence))[2]
    } else {
      NA_real_
    },
    intercept = if (n_pairs >= 3) {
      coef(lm(response_valence ~ predictor_valence))[1]
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Pooled descriptive slopes for responder valence predicted by partner valence"
  )
```

```{r directional-alignment-three-panel-plot, fig.width=12, fig.height=4.8}
p_directional_alignment_comparison <- df_alignment_long %>%
  drop_na(predictor_valence, response_valence) %>%
  ggplot(aes(
    x = predictor_valence,
    y = response_valence,
    color = alignment_direction
  )) +
  geom_point(alpha = 0.18, size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.05, fullrange = TRUE) +
  facet_wrap(~ condition, nrow = 1) +
  scale_color_manual(
    values = c("First \u2192 Second" = "#377EB8", "Second(t-1) \u2192 First" = "#FF7F00"),
    name = "Direction"
  ) +
  labs(
    title = "Directional valence alignment by condition",
    subtitle = "Responder valence predicted by current or previous partner valence",
    x = "Partner valence",
    y = "Responder valence"
  ) +
  theme(legend.position = "bottom")

print(p_directional_alignment_comparison)
ggsave(
  file.path(GRAPHS_DIR, "directional_valence_alignment_comparison_three_panel.png"),
  p_directional_alignment_comparison,
  width = 12,
  height = 4.8,
  dpi = 300
)
```

# Primary: Per-Story Asymmetry Magnitude

The three-panel figure suggests that the two directional slopes may differ most strongly in Human-AI. The mixed model below tests that visual pattern directly by estimating whether the predictor-valence slope differs by direction and whether that directional slope gap changes by condition. This is a pooled turn-level model with story-level random intercepts; the per-story $A_{\text{valence}}$ analysis that follows is the story-level robustness layer.

```{r directional-slope-interaction-model}
model_direction_condition <- lmer(
  response_z ~ predictor_z * alignment_direction * condition + (1 | conversation_id),
  data = df_std
)

cat("=== Direction x condition slope-interaction model ===\n")
print(summary(model_direction_condition)$coefficients)

directional_slopes <- emtrends(
  model_direction_condition,
  specs = c("condition", "alignment_direction"),
  var = "predictor_z"
)

directional_slopes %>%
  as.data.frame() %>%
  knitr::kable(
    digits = 3,
    caption = "Estimated predictor-valence slopes by condition and direction"
  )

direction_gap_by_condition <- contrast(
  directional_slopes,
  method = list("First_to_second_minus_second_to_first" = c(1, -1)),
  by = "condition"
)

direction_gap_by_condition %>%
  as.data.frame() %>%
  knitr::kable(
    digits = 3,
    caption = "Within-condition directional slope gap"
  )
```

```{r directional-slope-gap-condition-contrasts}
direction_gap_slopes <- contrast(
  directional_slopes,
  method = list("First_to_second_minus_second_to_first" = c(1, -1)),
  by = "condition"
)

contrast(
  direction_gap_slopes,
  method = list(
    "Human-AI vs Human-Human" = c(1, -1, 0),
    "Human-AI vs AI-AI" = c(1, 0, -1),
    "Human-AI vs same-type average" = c(1, -0.5, -0.5)
  ),
  by = NULL,
  adjust = "none"
) %>%
  as.data.frame() %>%
  knitr::kable(
    digits = 3,
    caption = "Cross-condition contrasts of the directional slope gap"
  )
```

```{r primary-asymmetry}
# Per-story Pearson correlations under both pairing rules, on non-standardized
# predictor/response pairs (r is scale-invariant).
min_pairs <- 3

df_r12 <- df_alignment_long %>%
  filter(alignment_direction == "First \u2192 Second") %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_pairs = sum(!is.na(predictor_valence) & !is.na(response_valence)),
    r = if (n_pairs >= min_pairs) {
      suppressWarnings(cor(predictor_valence, response_valence, use = "complete.obs"))
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  transmute(conversation_id, condition, n_pairs_12 = n_pairs, r_12 = r)

df_r21 <- df_alignment_long %>%
  filter(alignment_direction == "Second(t-1) \u2192 First") %>%
  group_by(conversation_id, condition) %>%
  summarise(
    n_pairs = sum(!is.na(predictor_valence) & !is.na(response_valence)),
    r = if (n_pairs >= min_pairs) {
      suppressWarnings(cor(predictor_valence, response_valence, use = "complete.obs"))
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  transmute(conversation_id, condition, n_pairs_21 = n_pairs, r_21 = r)

df_asym <- df_r12 %>%
  inner_join(df_r21, by = c("conversation_id", "condition")) %>%
  mutate(
    # Clamp |r|<1 before atanh to avoid infinities on degenerate stories
    r_12_c = pmin(pmax(r_12, -0.999999), 0.999999),
    r_21_c = pmin(pmax(r_21, -0.999999), 0.999999),
    z_12 = atanh(r_12_c),
    z_21 = atanh(r_21_c),
    signed_z_diff = z_12 - z_21,
    A_valence = abs(signed_z_diff)
  ) %>%
  filter(!is.na(A_valence))

cat("Stories with both directional correlations available:\n")
df_asym %>%
  count(condition) %>%
  knitr::kable(caption = "Per-story A_valence sample sizes by condition")

# Sanity checks
stopifnot(all(df_asym$A_valence >= 0))
```

```{r primary-asymmetry-descriptives}
df_asym %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_A = mean(A_valence),
    sd_A = sd(A_valence),
    median_A = median(A_valence),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Per-story |Fisher-z asymmetry| by condition"
  )
```

```{r primary-asymmetry-test}
cat("One-way ANOVA: A_valence ~ condition\n")
print(summary(aov(A_valence ~ condition, data = df_asym)))

cat("\nKruskal-Wallis (robust): A_valence ~ condition\n")
print(kruskal.test(A_valence ~ condition, data = df_asym))

cat("\nPairwise Wilcoxon (BH-adjusted):\n")
print(pairwise.wilcox.test(
  df_asym$A_valence,
  df_asym$condition,
  p.adjust.method = "BH"
))
```

```{r primary-asymmetry-plot, fig.width=9, fig.height=5}
p_asym <- ggplot(df_asym, aes(x = condition, y = A_valence, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Per-story valence alignment asymmetry by condition",
    subtitle = "A_valence(i) = |atanh(r_first_to_second) - atanh(r_second_to_first)|",
    x = "Condition",
    y = expression("|" * Delta * z * "|")
  ) +
  theme(legend.position = "none")

print(p_asym)
ggsave(
  file.path(GRAPHS_DIR, "valence_asymmetry_per_story.png"),
  p_asym,
  width = 9,
  height = 5,
  dpi = 300
)
```

# Bootstrap CIs and Effect Sizes on `A_valence`

```{r boot-effects-valence}
boot_A <- bootstrap_condition_means(df_asym, "A_valence", R = 10000)
boot_A %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of A_valence (R = 10,000)"
  )

effect_sizes_pairwise(df_asym, "A_valence") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on A_valence"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-valence}
planned_contrast_HA_vs_same(df_asym, "A_valence") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: A_valence in HA vs the average of HH and AA"
  )
```

# Directional Stability of Signed Valence Asymmetry

Tests whether the *direction* of the chronological first vs second alignment difference is
consistent across stories within each condition. The signed quantity is
$\text{signed}_i = \operatorname{atanh}(r_{\text{first}\to \text{second},i}) - \operatorname{atanh}(r_{\text{second}\to \text{first},i})$.
HA-specific identity-tied directionality predicts high `stability` and
`sign_consistency` in HA, low in HH and AA, even when unsigned `A_valence`
is similar.

```{r directional-stability-valence}
ds_val <- directional_stability(df_asym, "signed_z_diff")
ds_val$by_condition %>%
  knitr::kable(
    digits = 3,
    caption = "Per-condition directional stability of signed valence asymmetry"
  )

cat(sprintf(
  "Brown-Forsythe variance-equality p = %.4f\n", ds_val$variance_equality_p
))
```

```{r directional-stability-valence-plot, fig.width=9, fig.height=5}
p_signed <- ggplot(df_asym, aes(x = condition, y = signed_z_diff, fill = condition)) +
  geom_violin(alpha = 0.4, colour = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.35, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Signed valence-asymmetry direction by condition",
    subtitle = "atanh(r_first_to_second) - atanh(r_second_to_first)",
    x = "Condition",
    y = expression(Delta * z)
  ) +
  theme(legend.position = "none")

print(p_signed)
ggsave(
  file.path(GRAPHS_DIR, "valence_signed_asymmetry_direction.png"),
  p_signed,
  width = 9,
  height = 5,
  dpi = 300
)
```

# Permutation Test on `A_valence ~ condition`

```{r permutation-valence}
permutation_test_A(df_asym, "A_valence", R = 10000) %>%
  knitr::kable(
    digits = 4,
    caption = "Permutation test on A_valence ~ condition (R = 10,000 story-label permutations)"
  )
```

# Summary

Primary cross-condition result: per-story $A_{\text{valence}} = |\operatorname{atanh}(r_{\text{first}\to\text{second},i}) - \operatorname{atanh}(r_{\text{second}\to\text{first},i})|$, tested with ANOVA + Kruskal and pairwise BH-adjusted Wilcoxon.

Robustness layers added:

- bootstrap 95% CIs on per-condition mean of $A_{\text{valence}}$
- pairwise Cliff's $\delta$ effect sizes
- planned single-df contrast HA vs (HH+AA)/2
- 10k story-label permutation test on $A_{\text{valence}} \sim \text{condition}$

The Human-AI signed Human/AI directional decomposition lives in [analysis/human-ai/valence_alignment_analysis.Rmd](../human-ai/valence_alignment_analysis.Rmd) and is not duplicated here.

Deferred to a follow-up pass:

- starter-mechanism analysis (signed, starter-normalized asymmetry across all conditions)

*Analysis completed: `r Sys.time()`*
