---
title: "Valence Alignment Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Statistical Contrasts"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load comparison utilities
source(here::here("analysis/comparison/_comparison_utils.R"))

# Additional packages
pacman::p_load(
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  patchwork,    # Combine plots
  emmeans,      # Estimated marginal means
  effectsize    # Effect size calculations
)

# Set global theme
theme_set(theme_comparison())
```

# Overview

This notebook performs comparative analysis across all three experimental conditions:

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together  
- **AI-AI**: Two AI agents co-writing (simulated baseline)

**Key Questions:**
1. Do alignment patterns differ across the three conditions?
2. Is alignment asymmetric in all conditions or condition-specific?
3. Does the AI-AI baseline show the strongest alignment (as expected)?
4. How do effect sizes compare across conditions?

# Load Data

```{r load-data}
# Check data availability
cat("Data availability:\n")
avail <- check_data_availability()
print(avail)

# Load valence data using standardized loader
df_combined <- tryCatch({
  load_valence_data(conditions = c("human-ai", "human-human", "ai-ai"))
}, error = function(e) {
  # Fallback: try loading whatever is available
  conditions_available <- c()
  
  for (cond in c("human-ai", "human-human", "ai-ai")) {
    path <- here("data", cond, "processed", "dyadic_sentiment_scores.parquet")
    if (file.exists(path)) {
      conditions_available <- c(conditions_available, cond)
    }
  }
  
  if (length(conditions_available) == 0) {
    stop("No valence data available for any condition")
  }
  
  cat("Loading available conditions:", paste(conditions_available, collapse = ", "), "\n")
  load_valence_data(conditions = conditions_available)
})

# Prepare condition factor
df_combined <- df_combined %>%
  prepare_condition_factor(reference = "human-ai")

# Report loaded data
cat("\nLoaded data summary:\n")
df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

# Preprocessing

```{r preprocess}
# Compute lagged valence for alignment analysis
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn) %>%
  mutate(
    author_1_valence_lag1 = lag(author_1_valence, n = 1),
    author_2_valence_lag1 = lag(author_2_valence, n = 1)
  ) %>%
  ungroup() %>%
  filter(turn > 1) %>%
  drop_na(author_1_valence, author_2_valence, 
          author_1_valence_lag1, author_2_valence_lag1)

cat("Analysis dataset:", nrow(df_analysis), "turns across", 
    n_distinct(df_analysis$conversation_id), "stories\n")
```

# Descriptive Statistics

```{r descriptives}
# Valence descriptives by condition
desc_stats <- df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    mean_a1_valence = mean(author_1_valence, na.rm = TRUE),
    sd_a1_valence = sd(author_1_valence, na.rm = TRUE),
    mean_a2_valence = mean(author_2_valence, na.rm = TRUE),
    sd_a2_valence = sd(author_2_valence, na.rm = TRUE),
    .groups = "drop"
  )

desc_stats %>%
  knitr::kable(
    digits = 3, 
    caption = "Valence descriptives by condition",
    col.names = c("Condition", "Stories", "Turns", 
                  "Author 1 M", "Author 1 SD", 
                  "Author 2 M", "Author 2 SD")
  )
```

```{r valence-distributions, fig.width=12, fig.height=4}
# Valence distributions by condition
p_dist_a1 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Author 1 Valence Distribution", x = "Valence", y = "Density") +
  theme(legend.position = "none")

p_dist_a2 <- df_analysis %>%
  ggplot(aes(x = author_2_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Author 2 Valence Distribution", x = "Valence", y = "Density")

p_dist_a1 + p_dist_a2 + plot_layout(guides = "collect")
```

# Alignment Analysis

## Within-Condition Alignment

For each condition, we fit mixed models testing bidirectional alignment:

- **Author 2 → Author 1**: Does Author 1 follow Author 2's previous valence?
- **Author 1 → Author 2**: Does Author 2 match Author 1's current valence?

```{r within-condition-models}
# Fit models for each condition and direction
conditions <- unique(df_analysis$condition)

results_within <- tibble()

for (cond in conditions) {
  df_cond <- filter(df_analysis, condition == cond)
  
  # Skip if insufficient data
  if (nrow(df_cond) < 50) {
    cat("Skipping", cond, "- insufficient data\n")
    next
  }
  
  # A2 → A1 model
  m_a2_to_a1 <- tryCatch({
    lmer(author_1_valence ~ author_2_valence_lag1 + (1 | conversation_id),
         data = df_cond)
  }, error = function(e) NULL)
  
  # A1 → A2 model
  m_a1_to_a2 <- tryCatch({
    lmer(author_2_valence ~ author_1_valence + (1 | conversation_id),
         data = df_cond)
  }, error = function(e) NULL)
  
  # Extract results
  if (!is.null(m_a2_to_a1)) {
    coefs <- summary(m_a2_to_a1)$coefficients
    results_within <- bind_rows(results_within, tibble(
      condition = cond,
      direction = "A2 → A1",
      estimate = coefs["author_2_valence_lag1", "Estimate"],
      se = coefs["author_2_valence_lag1", "Std. Error"],
      t = coefs["author_2_valence_lag1", "t value"],
      p = coefs["author_2_valence_lag1", "Pr(>|t|)"]
    ))
  }
  
  if (!is.null(m_a1_to_a2)) {
    coefs <- summary(m_a1_to_a2)$coefficients
    results_within <- bind_rows(results_within, tibble(
      condition = cond,
      direction = "A1 → A2",
      estimate = coefs["author_1_valence", "Estimate"],
      se = coefs["author_1_valence", "Std. Error"],
      t = coefs["author_1_valence", "t value"],
      p = coefs["author_1_valence", "Pr(>|t|)"]
    ))
  }
}

# Display results
results_within %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    ),
    ci_lower = estimate - 1.96 * se,
    ci_upper = estimate + 1.96 * se
  ) %>%
  select(condition, direction, estimate, se, ci_lower, ci_upper, t, p, sig) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-condition alignment coefficients"
  )
```

## Cross-Condition Comparison

Test whether alignment effects differ significantly across conditions using condition × predictor interactions.

```{r cross-condition-models}
# A2 → A1 with condition interaction
model_cross_a2_a1 <- lmer(
  author_1_valence ~ author_2_valence_lag1 * condition + (1 | conversation_id),
  data = df_analysis
)

cat("=== Cross-Condition: Author 2 → Author 1 ===\n")
print(summary(model_cross_a2_a1))

# A1 → A2 with condition interaction
model_cross_a1_a2 <- lmer(
  author_2_valence ~ author_1_valence * condition + (1 | conversation_id),
  data = df_analysis
)

cat("\n=== Cross-Condition: Author 1 → Author 2 ===\n")
print(summary(model_cross_a1_a2))
```

## Pairwise Condition Contrasts

```{r pairwise-contrasts}
# Estimated marginal means and pairwise contrasts
if (length(unique(df_analysis$condition)) > 1) {
  
  # A2 → A1 slopes by condition
  emm_a2_a1 <- emtrends(model_cross_a2_a1, specs = "condition", 
                         var = "author_2_valence_lag1")
  
  cat("=== A2→A1 Alignment Slopes by Condition ===\n")
  print(emm_a2_a1)
  
  cat("\n=== Pairwise Contrasts (A2→A1) ===\n")
  print(pairs(emm_a2_a1, adjust = "bonferroni"))
  
  # A1 → A2 slopes by condition
  emm_a1_a2 <- emtrends(model_cross_a1_a2, specs = "condition",
                         var = "author_1_valence")
  
  cat("\n=== A1→A2 Alignment Slopes by Condition ===\n")
  print(emm_a1_a2)
  
  cat("\n=== Pairwise Contrasts (A1→A2) ===\n")
  print(pairs(emm_a1_a2, adjust = "bonferroni"))
}
```

# Visualization

```{r alignment-plot, fig.width=14, fig.height=5}
# Alignment slopes by condition
p1 <- df_analysis %>%
  ggplot(aes(x = author_2_valence_lag1, y = author_1_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 2 → Author 1 Alignment",
    subtitle = "Does Author 1 follow Author 2's previous valence?",
    x = "Author 2 Valence (t-1)",
    y = "Author 1 Valence (t)",
    color = "Condition"
  )

p2 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, y = author_2_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Author 1 → Author 2 Alignment",
    subtitle = "Does Author 2 match Author 1's current valence?",
    x = "Author 1 Valence (t)",
    y = "Author 2 Valence (t)",
    color = "Condition"
  )

p1 + p2 + plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

```{r coefficient-plot, fig.width=10, fig.height=5}
# Coefficient comparison plot
if (nrow(results_within) > 0) {
  results_within %>%
    mutate(
      ci_lower = estimate - 1.96 * se,
      ci_upper = estimate + 1.96 * se
    ) %>%
    ggplot(aes(x = condition, y = estimate, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    facet_wrap(~direction, scales = "free_x") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Alignment Coefficients by Condition",
      subtitle = "Error bars show 95% CI",
      y = "Alignment Coefficient (β)",
      x = "Condition"
    ) +
    theme(legend.position = "none")
}
```

# Asymmetry Analysis

Test whether alignment is asymmetric (stronger in one direction) within each condition.

```{r asymmetry-analysis}
# Compare A2→A1 vs A1→A2 coefficients within each condition
asymmetry_results <- results_within %>%
  select(condition, direction, estimate, se) %>%
  pivot_wider(
    names_from = direction,
    values_from = c(estimate, se),
    names_sep = "_"
  ) %>%
  mutate(
    diff = `estimate_A1 → A2` - `estimate_A2 → A1`,
    diff_se = sqrt(`se_A1 → A2`^2 + `se_A2 → A1`^2),
    z = diff / diff_se,
    p = 2 * pnorm(-abs(z)),
    interpretation = case_when(
      p < 0.05 & diff > 0 ~ "A1→A2 stronger",
      p < 0.05 & diff < 0 ~ "A2→A1 stronger",
      TRUE ~ "No significant asymmetry"
    )
  )

asymmetry_results %>%
  select(condition, diff, diff_se, z, p, interpretation) %>%
  knitr::kable(
    digits = 3,
    caption = "Alignment asymmetry by condition (A1→A2 minus A2→A1)",
    col.names = c("Condition", "Difference", "SE", "z", "p", "Interpretation")
  )
```

# Summary

```{r summary-table}
# Final summary table
summary_table <- results_within %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ "ns"
    ),
    effect = paste0(round(estimate, 3), " (", round(se, 3), ") ", sig)
  ) %>%
  select(condition, direction, effect) %>%
  pivot_wider(names_from = direction, values_from = effect)

summary_table %>%
  knitr::kable(caption = "Summary: Alignment effects by condition and direction (β (SE) sig)")
```

# Key Findings

Based on the analysis above:

1. **Cross-condition differences**: [Interpret whether alignment patterns differ significantly across Human-AI, Human-Human, and AI-AI conditions]

2. **Asymmetry patterns**: [Discuss whether alignment is bidirectional or asymmetric in each condition]

3. **AI-AI baseline**: [Compare the AI-AI simulation to empirical conditions - is it more aligned as expected?]

4. **Effect sizes**: [Compare the magnitude of alignment effects across conditions]

---

*Analysis completed: `r Sys.time()`*
