---
title: "Valence Alignment Comparison: Human-AI vs Human-Human vs AI-AI"
subtitle: "Cross-Condition Contrasts on the Exchange-Aligned Data Model"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))

pacman::p_load(
  lme4,
  lmerTest,
  patchwork,
  emmeans,
  effectsize,
  stringr
)

theme_set(theme_comparison())
```


```{r}
folder_path <- "~/Documents/PENPAL_analysis/data"
file_names <- list.files(folder_path, pattern = "*interaction_level_stories_filtered.csv", full.names = TRUE, recursive = TRUE)
data_list <- lapply(file_names, read.csv)
selected_data_list <- lapply(data_list, function(df) {
    df %>% select("conversation_id", "author_1", "author_2")
})

combined_data <- bind_rows(selected_data_list)
combined_data <- combined_data[sample(nrow(combined_data)), ]
combined_data$combined_interaction <- str_c(combined_data$author_1, combined_data$author_2, sep = " ")

write.csv(combined_data, "combined_data.csv", row.names = FALSE)
```

This notebook compares **valence alignment** across the three experimental conditions using the new exchange-aligned structure.

- **Human-AI**: Human co-writing with an LLM
- **Human-Human**: Two humans co-writing together
- **AI-AI**: Two AI agents co-writing together

Important structural note:

- `author_1` / `author_2` are now treated as **slot 1 / slot 2** in the cleaned exchange sequence.
- In **Human-AI**, those slots also map onto human vs AI roles.
- In **Human-Human** and **AI-AI**, the slots are positional sides of the exchange, not stable identity labels.

The analysis window is restricted to **complete exchanges** and the default aligned turn range (`analysis_turn` 1 to 9).

# Load Data

```{r load-data}
cat("Data availability:\n")
print(check_data_availability())

df_combined <- load_valence_data(c("human-ai", "human-human", "ai-ai")) %>%
  filter_complete_exchange_window() %>%
  prepare_condition_factor(reference = "human-ai")

df_combined %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    min_turn = min(turn_index, na.rm = TRUE),
    max_turn = max(turn_index, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    caption = "Loaded exchange-aligned valence data by condition"
  )
```

# Preprocessing

```{r preprocess}
df_analysis <- df_combined %>%
  group_by(conversation_id, condition) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    author_1_valence_lag1 = lag(author_1_valence),
    author_2_valence_lag1 = lag(author_2_valence)
  ) %>%
  ungroup() %>%
  filter(turn_index > 1) %>%
  drop_na(
    author_1_valence, author_2_valence,
    author_1_valence_lag1, author_2_valence_lag1
  )

df_alignment_long <- bind_rows(
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "slot1_to_slot2",
      predictor_valence = author_1_valence,
      response_valence = author_2_valence
    ),
  df_analysis %>%
    transmute(
      conversation_id,
      condition,
      turn_index,
      alignment_direction = "slot2_to_slot1",
      predictor_valence = author_2_valence_lag1,
      response_valence = author_1_valence
    )
) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("slot1_to_slot2", "slot2_to_slot1"),
      labels = c("Slot 1 → Slot 2", "Slot 2(t-1) → Slot 1")
    )
  )

df_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z = as.numeric(scale(response_valence))
  ) %>%
  ungroup() %>%
  drop_na(predictor_z, response_z)

contrasts(df_std$alignment_direction) <- contr.sum(2)

cat("Analysis rows:", nrow(df_std), "\n")
cat("Stories:", n_distinct(df_std$conversation_id), "\n")
```

# Descriptive Checks

```{r descriptives}
df_analysis %>%
  group_by(condition) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_turns = n(),
    slot_1_mean = mean(author_1_valence, na.rm = TRUE),
    slot_2_mean = mean(author_2_valence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Mean projected valence by condition"
  )
```

```{r valence-distributions, fig.width=12, fig.height=4}
p_slot1 <- df_analysis %>%
  ggplot(aes(x = author_1_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 1 Valence", x = "Projected Valence", y = "Density") +
  theme(legend.position = "none")

p_slot2 <- df_analysis %>%
  ggplot(aes(x = author_2_valence, fill = condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = condition_colors) +
  labs(title = "Slot 2 Valence", x = "Projected Valence", y = "Density")

p_slot1 + p_slot2 + plot_layout(guides = "collect")
```

# Within-Condition Directional Asymmetry

```{r within-condition-models}
within_condition_results <- list()

for (cond in unique(df_std$condition)) {
  df_cond <- filter(df_std, condition == cond)

  if (nrow(df_cond) < 100 || n_distinct(df_cond$conversation_id) < 5) {
    next
  }

  df_cond$alignment_direction <- factor(
    df_cond$alignment_direction,
    levels = levels(df_std$alignment_direction)
  )
  contrasts(df_cond$alignment_direction) <- contr.sum(2)

  model <- tryCatch({
    lmer(
      response_z ~ predictor_z * alignment_direction + (1 + predictor_z | conversation_id),
      data = df_cond,
      control = lmerControl(optimizer = "bobyqa")
    )
  }, error = function(e) {
    lmer(
      response_z ~ predictor_z * alignment_direction + (1 | conversation_id),
      data = df_cond
    )
  })

  slopes <- emtrends(model, specs = "alignment_direction", var = "predictor_z")
  slope_contrast <- contrast(slopes, method = "revpairwise")

  within_condition_results[[as.character(cond)]] <- list(
    model = model,
    slopes = as.data.frame(slopes),
    contrast = as.data.frame(slope_contrast)
  )
}

within_summary <- purrr::imap_dfr(within_condition_results, function(res, cond) {
  tibble(
    condition = cond,
    slope_slot1_to_slot2 = res$slopes$predictor_z.trend[res$slopes$alignment_direction == "Slot 1 → Slot 2"],
    slope_slot2_to_slot1 = res$slopes$predictor_z.trend[res$slopes$alignment_direction == "Slot 2(t-1) → Slot 1"],
    difference = res$contrast$estimate[1],
    se = res$contrast$SE[1],
    t = res$contrast$t.ratio[1],
    p = res$contrast$p.value[1]
  )
})

within_summary %>%
  mutate(
    sig = case_when(
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      p < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Within-condition alignment asymmetry by slot direction"
  )
```

```{r asymmetry-plot, fig.width=10, fig.height=5}
if (nrow(within_summary) > 0) {
  within_summary %>%
    ggplot(aes(x = condition, y = difference, fill = condition)) +
    geom_col(width = 0.6, alpha = 0.8) +
    geom_errorbar(
      aes(ymin = difference - 1.96 * se, ymax = difference + 1.96 * se),
      width = 0.2
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(values = condition_colors) +
    labs(
      title = "Directional Asymmetry by Condition",
      subtitle = "Difference: (Slot 1 → Slot 2) - (Slot 2(t-1) → Slot 1). Positive = Slot 1 → Slot 2 stronger",
      y = "Asymmetry (Δβ)",
      x = "Condition"
    ) +
    theme(legend.position = "none")
}
```

# Cross-Condition Model

```{r cross-condition-model}
df_std$condition <- factor(df_std$condition)
contrasts(df_std$condition) <- contr.sum(length(levels(df_std$condition)))

model_full <- tryCatch({
  lmer(
    response_z ~ predictor_z * alignment_direction * condition + (1 + predictor_z | conversation_id),
    data = df_std,
    control = lmerControl(optimizer = "bobyqa")
  )
}, error = function(e) {
  lmer(
    response_z ~ predictor_z * alignment_direction * condition + (1 | conversation_id),
    data = df_std
  )
})

cat("Full model summary:\n")
print(summary(model_full))

slopes_by_condition <- emtrends(
  model_full,
  specs = c("condition", "alignment_direction"),
  var = "predictor_z"
)

cat("\nAlignment slopes by condition and direction:\n")
print(slopes_by_condition)

cat("\nWithin-condition direction contrasts:\n")
print(contrast(slopes_by_condition, method = "revpairwise", by = "condition"))
```

# Visualizations

```{r scatterplots, fig.width=13, fig.height=5}
p_cross_turn <- df_alignment_long %>%
  filter(alignment_direction == "Slot 2(t-1) → Slot 1") %>%
  ggplot(aes(x = predictor_valence, y = response_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Cross-Turn Alignment",
    subtitle = "Does Slot 1 follow Slot 2's previous valence?",
    x = "Slot 2 Valence (t-1)",
    y = "Slot 1 Valence (t)"
  ) +
  theme(legend.position = "none")

p_same_turn <- df_alignment_long %>%
  filter(alignment_direction == "Slot 1 → Slot 2") %>%
  ggplot(aes(x = predictor_valence, y = response_valence, color = condition)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Same-Turn Alignment",
    subtitle = "Does Slot 2 match Slot 1 within the same exchange?",
    x = "Slot 1 Valence (t)",
    y = "Slot 2 Valence (t)"
  )

p_cross_turn + p_same_turn + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r coefficient-plot, fig.width=12, fig.height=5}
slopes_df <- as.data.frame(slopes_by_condition) %>%
  mutate(
    ci_lower = predictor_z.trend - 1.96 * SE,
    ci_upper = predictor_z.trend + 1.96 * SE
  )

ggplot(slopes_df, aes(x = condition, y = predictor_z.trend, fill = alignment_direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.85) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8),
    width = 0.25
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_manual(
    values = c("Slot 1 → Slot 2" = "#377EB8", "Slot 2(t-1) → Slot 1" = "#FF7F00")
  ) +
  labs(
    title = "Alignment Coefficients by Condition and Direction",
    subtitle = "Standardized slopes with 95% confidence intervals",
    x = "Condition",
    y = "Alignment Coefficient (β)",
    fill = "Direction"
  ) +
  theme(legend.position = "bottom")
```

# Summary

The notebook now runs on the new structure by:

- restricting the analysis to **complete exchanges** and the aligned turn window,
- treating `author_1` / `author_2` as **structural slots** rather than fixed identities,
- using `turn_index` / `analysis_turn` rather than the old raw-turn assumptions,
- keeping directional alignment tests comparable across conditions while remaining valid for same-type conversations.

*Analysis completed: `r Sys.time()`*
