---
title: "Valence Alignment in Human-AI Story Co-Creation"
subtitle: "Analysis of Field Data and AI-AI Baseline Comparison"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex,          # ANOVA
  here
)

# Set project root
PROJECT_ROOT <- here()

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

## Summary

- Emotional alignment is asymmetric: the model reliably aligns to the user (User → AI), while humans do not mirror the preceding AI tone.
- Field data shows greater variance in valence and weaker overall coordination compared to the simulated baseline.
- Overall, affective dynamics in the field are primarily model-driven rather than mutually convergent.

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load dyadic sentiment scores

df_field <- read_parquet(here("data/TEXT/processed/dyadic_sentiment_scores.parquet"))
df_simulated <- read_parquet(here("data/Berlin/processed/dyadic_sentiment_scores_simulated.parquet"))

# Display structure
cat("Field data dimensions:", dim(df_field), "\n")
cat("Simulated data dimensions:", dim(df_simulated), "\n")

# Show column names
cat("\nField data columns:\n")
print(names(df_field))

cat("\nSimulated data columns:\n")
print(names(df_simulated))
```

## 1.2 Data Preprocessing

```{r preprocess}
# Preprocess field data - using standardized column names
df_field <- df_field %>%
  rename(
    author_1_valence = author_1_sentiment_projection,
    author_2_valence = author_2_sentiment_projection
  ) %>%
  mutate(
    # Token counts
    author_1_token_count = stri_count_words(author_1),
    author_2_token_count = stri_count_words(author_2),
    token_ratio = if_else(author_2_token_count == 0, NA_real_, 
                         author_1_token_count / author_2_token_count),
    
    # Time parsing
    timestamp = ymd_hms(timestamp),
    
    # Data source label
    data_source = "Field"
  ) %>%
  # Ensure conversation_id is a character/factor
  mutate(conversation_id = as.character(conversation_id)) %>%
  filter(turn > 1, turn < 10)


# Preprocess simulated data - using standardized column names
df_simulated <- df_simulated %>%
  rename(
    author_1_valence = author_1_sentiment_score,
    author_2_valence = author_2_sentiment_score
  ) %>%
  mutate(
    # Token counts
    author_1_token_count = stri_count_words(author_1),
    author_2_token_count = stri_count_words(author_2),
    token_ratio = if_else(author_2_token_count == 0, NA_real_, 
                         author_1_token_count / author_2_token_count),
    
    # Time parsing (if available)
    timestamp = if_else(!is.na(timestamp), ymd_hms(timestamp), as.POSIXct(NA)),
    
    # Data source label
    data_source = "Simulated"
  ) %>%
  mutate(conversation_id = as.character(conversation_id))

# Summary statistics
cat("\n=== Field Data Summary ===\n")
df_field %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_author_1_valence = mean(author_1_valence, na.rm = TRUE),
    mean_author_2_valence = mean(author_2_valence, na.rm = TRUE)
  ) %>%
  print()

cat("\n=== Simulated Data Summary ===\n")
df_simulated %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_author_1_valence = mean(author_1_valence, na.rm = TRUE),
    mean_author_2_valence = mean(author_2_valence, na.rm = TRUE)
  ) %>%
  print()
```

## 1.3 Compute Lagged Variables and Cosine

```{r compute-lags}
# Add lagged AI valence for field data
# Note: lag() within group_by ensures no cross-story contamination
# Turn 1 of each story will have NA for lagged values (correct behavior)

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    author_2_valence_lag1 = lag(author_2_valence, n = 1),
    author_1_valence_lag1 = lag(author_1_valence, n = 1),
  ) %>%
  ungroup()


cosine_sim <- function(x, y) {
  if (is.null(x) || is.null(y) || any(is.na(x)) || any(is.na(y))) {
    return(NA_real_)
  }
  return(sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2))))
}

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    author_2_embedding_prev = lag(author_2_embedding),
    author_1_embedding_prev = lag(author_1_embedding)
  ) %>%
  rowwise() %>%
  mutate(
    cosine_sim_same = cosine_sim(author_1_embedding, author_2_embedding),
    cosine_sim_prev = cosine_sim(author_1_embedding, author_2_embedding_prev),
    cosine_sim_author_1_self = cosine_sim(author_1_embedding, author_1_embedding_prev),
    cosine_sim_author_2_self = cosine_sim(author_2_embedding, author_2_embedding_prev)
  ) %>%
    select(-author_1_embedding, -author_2_embedding, -author_2_embedding_prev, -author_1_embedding_prev) %>%
  ungroup()

# Add lagged AI valence for simulated data
df_simulated <- df_simulated %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    author_2_valence_lag1 = lag(author_2_valence, n = 1),
    author_1_valence_lag1 = lag(author_1_valence, n = 1)
  ) %>%
  ungroup()

# Verify no cross-story contamination
cat("=== Lag Variable Verification ===\n")
cat("Field data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_field$turn == 1 & !is.na(df_field$author_2_valence_lag1)), "\n")
cat("Simulated data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_simulated$turn == 1 & !is.na(df_simulated$author_2_valence_lag1)), "\n")

```
```{r}
# === Step 1 & 2: Calculate Deltas and Alignment Flags ===
df_dynamics <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn) %>%
  mutate(
    # Calculate change from previous turn (Velocity)
    delta_author_1 = author_1_valence - author_1_valence_lag1,
    delta_author_2 = author_2_valence - author_2_valence_lag1,
    
    # Handle cases where change is exactly 0 (rare, but possible)
    # We treat 0 change as "no movement", so sign is 0.
    
    # DIRECTION 1: User 1 -> User 2 (Same Turn)
    # Does User 2 move in the same direction the User 1 just moved?
    is_aligned_same = sign(delta_author_1) == sign(delta_author_2) & delta_author_1 != 0 & delta_author_2 != 0,
    
    # DIRECTION 2: User 2 -> User 1 (Next Turn)
    # Does User 1 move in the same direction the User 2 moved in the PREVIOUS turn?
    # We compare current User 1 delta with lagged User 2 delta
    is_aligned_prev = sign(delta_author_1) == sign(lag(delta_author_2)) & delta_author_1 != 0 & lag(delta_author_2) != 0
  ) %>%
  ungroup()

# === Step 3: Extract Run Lengths (Durations) ===
# Function to get lengths of TRUE sequences from a vector
get_run_lengths <- function(x) {
  # rle() computes lengths of consecutive values
  r <- rle(x)
  # We only want the lengths where the value was TRUE (alignment)
  # If there are no TRUEs, return numeric(0)
  if(all(!x, na.rm=TRUE)) return(numeric(0))
  return(r$lengths[r$values == TRUE & !is.na(r$values)])
}

# Apply this to each conversation and stack the results
alignment_durations <- df_dynamics %>%
  group_by(conversation_id) %>%
  summarise(
    # Extract all duration "streaks" for Same Turn alignment
    duration_same = list(get_run_lengths(is_aligned_same)),
    # Extract all duration "streaks" for Previous Turn alignment
    duration_prev = list(get_run_lengths(is_aligned_prev))
  ) %>%
  pivot_longer(cols = c(duration_same, duration_prev), 
               names_to = "direction", 
               values_to = "duration_list") %>%
  # Unnest the list to get one row per alignment streak
  unnest(duration_list) %>%
  mutate(
    direction = recode(direction, 
                       duration_same = "User 1 -> User 2 (Same Turn)",
                       duration_prev = "User 2 -> User 1 (Next Turn)")
  )

# === Step 4: Visualize ===
p_duration <- alignment_durations %>%
  ggplot(aes(x = direction, y = duration_list, fill = direction)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  # Add mean point
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = c("User 1 -> User 2 (Same Turn)" = "#377EB8", 
                               "User 2 -> User 1 (Next Turn)" = "#FF7F00")) +
  labs(
    title = "Duration of Emotional Alignment",
    subtitle = "How many consecutive turns do they move in sync?",
    y = "Consecutive Turns of Alignment",
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  ylim(1,5)

print(p_duration)

# Statistical Test
t.test(duration_list ~ direction, data = alignment_durations)
```

