---
title: "Novelty Analysis"
subtitle: "Analysis of Novelty, Transience, and Resonance"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

## Summary

- Both AI agents show comparable patterns in novelty (surprisal) and transience metrics.
- Both agents display a positive novelty–resonance relationship (innovation bias), with similar slopes.
- This indicates that novel contributions from both agents have lasting forward influence on narrative development.
- Agent 1 and Agent 2 show similar variance and informational impact in this AI-AI baseline condition.
- The symmetric pattern serves as a baseline for comparison with human-AI and human-human conditions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex,          # ANOVA
  here,          # File path management
  emmeans      # Estimated marginal means

)

# Set project root
PROJECT_ROOT <- here()
setwd(PROJECT_ROOT)

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}

# Load Field Data - already using standardized column names
df_res_field <- read_csv(here("data/ai-ai/processed/novelty_scores.csv")) %>%
  filter(!is.na(author_1_surprise) & !is.na(author_2_surprise)) %>%
  mutate(
    author_1_resonance = author_1_surprise - author_1_transience,
    author_2_resonance = author_2_surprise - author_2_transience,
    datatype = "AI-AI"
  ) %>%
  drop_na(author_1, author_2) %>%
  filter(interaction_count > 1, interaction_count < 10, author_1_surprise > -4, author_2_surprise > -4)


```

## 1.2 Reshape Data

```{r reshape-data}
# Compute word counts for each agent's turn
df_res_field <- df_res_field %>%
  mutate(
    author_1_nwords = str_count(author_1, "\\S+"),
    author_2_nwords = str_count(author_2, "\\S+")
  )

df_res_long <- df_res_field %>%
  pivot_longer(
    cols = c(author_1_surprise, author_1_transience, author_1_nwords,
             author_2_surprise, author_2_transience, author_2_nwords),
    names_to = c("agent", "metric"),
    names_pattern = "(author_1|author_2)_(surprise|transience|nwords)"
  ) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(
    resonance = surprise - transience,
    agent = factor(agent, levels = c("author_1", "author_2"), labels = c("Agent 1", "Agent 2"))
  )
```

## 1.3 Descriptive Statistics

```{r descriptive-stats}
desc_stats <- df_res_long %>%
  group_by(agent) %>%
  summarise(
    n = n(),
    mean_surprise = mean(surprise, na.rm = TRUE),
    sd_surprise = sd(surprise, na.rm = TRUE),
    mean_transience = mean(transience, na.rm = TRUE),
    sd_transience = sd(transience, na.rm = TRUE),
    mean_resonance = mean(resonance, na.rm = TRUE),
    sd_resonance = sd(resonance, na.rm = TRUE)
  )

print(desc_stats)

knitr::kable(desc_stats, caption = "Descriptive Statistics for Novelty Metrics by Agent", digits = 3)
```

```{r}
p_novelty_agent <- ggplot(df_res_long, aes(x = agent, y = surprise, fill = agent)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00"), name = "") +
  scale_color_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00")) +
  labs(title = "Distributions of Novelty", x = "Novelty", y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(p_novelty_agent)
ggsave(
  file.path(GRAPHS_DIR, "Surprise_Distributions_Agent.png"),
  p_novelty_agent,
  width = 10,
  height = 6,
  dpi = 300
)

p_transience_agent <- ggplot(df_res_long, aes(x = transience, fill = agent)) +
  geom_density(alpha = 0.4, color = NA) +
  scale_fill_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00"), name = "") +
  scale_color_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00")) +
  labs(title = "Distributions of Transience", x = "Transience", y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(p_transience_agent)
ggsave(
  file.path(GRAPHS_DIR, "Transience_Distributions_Agent.png"),
  p_transience_agent,
  width = 10,
  height = 6,
  dpi = 300
)

```


# 2. Novelty and Resonance Analysis

## 2.1 Resonance vs Novelty Plot

```{r resonance-plot, fig.width=12, fig.height=6}
# Plot Resonance vs Novelty (Field Data)
p_resonance <- ggplot(df_res_long, aes(x = surprise, y = resonance, color = agent)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, size = 1.2) +
  scale_color_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00")) +
    geom_vline(data = desc_stats,
             aes(xintercept = mean_surprise, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title = "Resonance vs Novelty (AI-AI)",
    x = "Novelty",
    y = "Resonance (Novelty - Transience)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_resonance)

ggsave(file.path(GRAPHS_DIR, "novelty_resonance_vs_novelty.png"),
       p_resonance, width = 12, height = 6, dpi = 300)
```

## 2.2 Transience vs Novelty Plot

```{r transience-plot, fig.width=12, fig.height=6}
p_transience <- ggplot(df_res_long, aes(x = surprise, y = transience, color = agent)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, size = 1.2) +
  scale_color_manual(values = c("Agent 1" = "#377EB8", "Agent 2" = "#FF7F00")) +
      geom_vline(data = desc_stats,
             aes(xintercept = mean_surprise, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title = "Transience vs Novelty (AI-AI)",
    x = "Novelty (Surprisal)",
    y = "Transience"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_transience)

ggsave(file.path(GRAPHS_DIR, "novelty_transience_vs_novelty.png"),
       p_transience, width = 12, height = 6, dpi = 300)
```

# 3. Statistical Models

## 3.1 Resonance Model

```{r resonance-model}
# Model without length control
model_resonance <- lmer(resonance ~ surprise * agent + (1 | conversation_id), data = df_res_long)
cat("=== Resonance Model ===\n")
print(summary(model_resonance)$coefficients)

cat("\n=== Slope Comparison (Resonance) ===\n")
print(emtrends(model_resonance, specs = "agent", var = "surprise"))

# Model WITH length control
model_resonance_len <- lmer(resonance ~ surprise * agent + nwords + (1 | conversation_id), data = df_res_long)
cat("\n=== Resonance Model (controlling for turn length) ===\n")
print(summary(model_resonance_len)$coefficients)

cat("\n=== Slope Comparison (Resonance, length-controlled) ===\n")
print(emtrends(model_resonance_len, specs = "agent", var = "surprise"))
```

## 3.2 Transience Model

```{r transience-model}
# Model without length control
model_transience <- lmer(transience ~ surprise * agent + (1 | conversation_id), data = df_res_long)

cat("=== Transience Model ===\n")
print(summary(model_transience)$coefficients)

cat("\n=== Slope Comparison (Transience) ===\n")
print(emtrends(model_transience, specs = "agent", var = "surprise"))

# Model WITH length control
model_transience_len <- lmer(transience ~ surprise * agent + nwords + (1 | conversation_id), data = df_res_long)

cat("\n=== Transience Model (controlling for turn length) ===\n")
print(summary(model_transience_len)$coefficients)

cat("\n=== Slope Comparison (Transience, length-controlled) ===\n")
print(emtrends(model_transience_len, specs = "agent", var = "surprise"))
```





---

# 5. Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

**Analysis complete.** All figures saved to `analysis/figures/` and tables to `analysis/tables/`.



