---
title: "Valence Alignment in AI-AI Story Co-Creation"
subtitle: "Analysis of AI-AI Baseline Condition"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex,          # ANOVA
  here,
  emmeans
)

# Set project root
PROJECT_ROOT <- here()

# Set global theme
theme_set(theme_minimal(base_size = 14))

format_alignment_label <- function(x) {
  parsed <- dplyr::recode(
    x,
    same_turn = "'(' * A1[t] * ', ' * A2[t] * ')'",
    prev_turn = "'(' * A2[t-1] * ', ' * A1[t] * ')'"
  )
  parse(text = parsed)
}

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

## Summary


# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load dyadic sentiment scores
df_field <- read_parquet(here("data/ai-ai/processed/dyadic_sentiment_scores.parquet"))

# Display structure
cat("Field data dimensions:", dim(df_field), "\n")

# Show column names
cat("\nField data columns:\n")
print(names(df_field))
```

## 1.2 Data Preprocessing

```{r preprocess}
# Preprocess AI-AI data - using standardized column names
df_field <- df_field %>%
  rename(
    author_1_valence = author_1_sentiment_projection,
    author_2_valence = author_2_sentiment_projection
  ) %>%
  mutate(
    # Token counts
    author_1_token_count = stri_count_words(author_1),
    author_2_token_count = stri_count_words(author_2),
    token_ratio = if_else(author_2_token_count == 0, NA_real_, 
                         author_1_token_count / author_2_token_count),
    
    # Time parsing
    timestamp = ymd_hms(timestamp),
    
    # Data source label
    data_source = "AI-AI"
  ) %>%
  # Ensure conversation_id is a character/factor
  mutate(conversation_id = as.character(conversation_id)) %>%
    filter(turn > 1, turn < 10)

# Summary statistics
cat("\n=== AI-AI Data Summary ===\n")
df_field %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_author_1_valence = mean(author_1_valence, na.rm = TRUE),
    mean_author_2_valence = mean(author_2_valence, na.rm = TRUE)
  ) %>%
  print()
```

## 1.3 Compute Lagged Variables and Cosine

```{r compute-lags}
# Add lagged agent valence
# Note: lag() within group_by ensures no cross-story contamination
# Turn 1 of each story will have NA for lagged values (correct behavior)

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    author_2_valence_lag1 = lag(author_2_valence, n = 1),
    author_1_valence_lag1 = lag(author_1_valence, n = 1)
  ) %>%
  ungroup()


cosine_sim <- function(x, y) {
  if (is.null(x) || is.null(y) || any(is.na(x)) || any(is.na(y))) {
    return(NA_real_)
  }
  return(sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2))))
}

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    author_2_embedding_prev = lag(author_2_embedding),
    author_1_embedding_prev = lag(author_1_embedding)
  ) %>%
  rowwise() %>%
  mutate(
    cosine_sim_same = cosine_sim(author_1_embedding, author_2_embedding),
    cosine_sim_prev = cosine_sim(author_1_embedding, author_2_embedding_prev),
    cosine_sim_author_1_self = cosine_sim(author_1_embedding, author_1_embedding_prev),
    cosine_sim_author_2_self = cosine_sim(author_2_embedding, author_2_embedding_prev)
  ) %>%
  ungroup()

# Verify no cross-story contamination
cat("=== Lag Variable Verification ===\n")
cat("Field data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_field$turn == 1 & !is.na(df_field$author_2_valence_lag1)), "\n")
```

# 2. Descriptive Statistics

## 2.1 Token Count Statistics

```{r token-stats}
# Field data token statistics
token_stats_field <- df_field %>%
  summarise(
    mean_author_1_tokens = mean(author_1_token_count, na.rm = TRUE),
    sd_author_1_tokens = sd(author_1_token_count, na.rm = TRUE),
    mean_author_2_tokens = mean(author_2_token_count, na.rm = TRUE),
    sd_author_2_tokens = sd(author_2_token_count, na.rm = TRUE),
    mean_token_ratio = mean(token_ratio, na.rm = TRUE)
  )

cat("=== Field Data Token Statistics ===\n")
print(token_stats_field)

```

## 2.2 Valence Distribution Statistics

```{r valence-stats}
# Combined statistics
valence_stats <- bind_rows(
  df_field %>% 
    pivot_longer(cols = c(author_1_valence, author_2_valence),
                 names_to = "agent", values_to = "valence") %>%
    mutate(agent = str_remove(agent, "_valence")) %>%
    group_by(data_source, agent) %>%
    summarise(
      mean_valence = mean(valence, na.rm = TRUE),
      sd_valence = sd(valence, na.rm = TRUE),
      median_valence = median(valence, na.rm = TRUE),
      .groups = "drop"
    )
)

print(valence_stats)

# Save to CSV
write_csv(valence_stats, file.path(TABLES_DIR, "valence_descriptive_stats.csv"))
```

# 3. Valence Distribution Analysis

## 3.1 Valence Distribution Difference Test
```{r test-valence-distribution}
df_field_long <- df_field %>%
  pivot_longer(cols = c(author_1_valence, author_2_valence),
               names_to = "agent", values_to = "valence") %>%
  mutate(agent = recode(agent, 
                       author_1_valence = "author_1", 
                       author_2_valence = "author_2"),
         agent = factor(agent, levels = c("author_1", "author_2")))



df_field_cosine_long <- df_field %>%
  pivot_longer(cols = c(cosine_sim_same, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_same = "Same",
                            cosine_sim_prev = "Prev"),
         direction = factor(direction, levels = c("Same", "Prev"))
         )

# Calculate means for vertical lines
means_field <- df_field_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")



valence_difference_mixed_model <- lmer(valence ~ agent + (1 | conversation_id), data = df_field_long)
cat("=== Valence Difference Model Coefficients ===\n")
print(round(summary(valence_difference_mixed_model)$coefficients, 4))

```


## 3.2 Combined Valence Distribution Plot

```{r valence-distribution, fig.width=12, fig.height=6}
# Prepare data for plotting

# Calculate KS distance and means
ks_test <- ks.test(df_field_long %>% filter(agent == "author_1") %>% pull(valence),
           df_field_long %>% filter(agent == "author_2") %>% pull(valence))

means_field <- df_field_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")

# Overlapping density plot with means and KS distance
p_field <- ggplot(df_field_long, aes(x = valence, fill = agent, color = agent)) +
  geom_density(alpha = 0.4, linewidth = 1.2) +
  geom_vline(data = means_field,
       aes(xintercept = mean_val, color = agent),
       linetype = "dashed", linewidth = 1.2, show.legend = FALSE) +
  scale_fill_manual(
    values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"), 
    name = "Agent",
    labels = c("author_1" = "Agent 1", "author_2" = "Agent 2")
  ) +
  scale_color_manual(
    values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"), 
    name = "Agent",
    labels = c("author_1" = "Agent 1", "author_2" = "Agent 2")
  ) +
  labs(
  title = "Valence Distribution: Field Data",
  subtitle = sprintf("Kolmogorov-Smirnov D = %.3f, p-value = %.4f", 
             ks_test$statistic, ks_test$p.value),
  x = "Valence Score",
  y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
  legend.position = "top",
  plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40")
  )

print(p_field)

ggsave(file.path(GRAPHS_DIR, "valence_distribution_comparison_density_ks.png"), 
     p_field, width = 10, height = 6, dpi = 300)

# Print KS test results
cat("\n=== Kolmogorov-Smirnov Test: Agent 1 vs Agent 2 Valence Distribution ===\n")
print(ks_test)
```

# 4. Valence Alignment Analysis

## 4.1 Per-Story Correlation (Same Turn)

```{r alignment-same-turn}
# Field data: same-turn correlation
cor_field_same <- df_field %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%  # Require at least 3 turns
  summarise(
    rho = cor(author_1_valence, author_2_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )



# Combine
cor_same_combined <- bind_rows(cor_field_same) %>%
  mutate(alignment_type = "same_turn")

# Summary
cat("=== Same-Turn Alignment Summary ===\n")
cor_same_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

```

## 4.2 Per-Story Correlation (Previous Turn)

```{r alignment-prev-turn}
# Field data: previous-turn correlation (author_1 response to previous author_2)
cor_field_prev <- df_field %>%
  group_by(conversation_id) %>%
  filter(!is.na(author_2_valence_lag1)) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(author_1_valence, author_2_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )

# Combine
cor_prev_combined <- bind_rows(cor_field_prev) %>%
  mutate(alignment_type = "prev_turn")

# Summary
cat("=== Previous-Turn Alignment Summary ===\n")
cor_prev_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```

## 4.3.1 All turns story correlation (same turn)
```{r alignment across all turns}
#Violates assumptions since turns are not independent (nested in stories)
cor_field_same_all_turns <- df_field %>%
  summarise(
    rho = cor(author_1_valence, author_2_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field"
  )



cor_same_combined_all_turns <- bind_rows(cor_field_same_all_turns) %>%
  mutate(alignment_type = "Same Turn All Turns")

cat("=== Same-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```
## 4.3.2 All turns story correlation (Previous Turn)

```{r previous-turn-alignment-all-turns}
# Violates independence assumption (turns nested in stories)
# Note: use="complete.obs" automatically excludes turn 1 (where author_2_valence_lag1 is NA)
# This ensures author_1_t1 is never paired with author_2 from previous story

# Verify exclusion of turn 1
cat("=== Previous-Turn Aggregate Correlation ===\n")
cat("Field: Turn 1 observations (excluded):", sum(df_field$turn == 1), "\n")
cat("Field: Valid lag observations:", sum(!is.na(df_field$author_2_valence_lag1)), "\n")

cor_field_same_all_turns_prev <- df_field %>%
  summarise(
    rho = cor(author_1_valence, author_2_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    n_valid = sum(!is.na(author_2_valence_lag1)),
    data_source = "Field"
  )

cor_same_combined_all_turns_prev <- bind_rows(cor_field_same_all_turns_prev) %>%
  mutate(alignment_type = "Previous Turn All Turns")

cat("=== Previous-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns_prev %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_total = sum(n_turns),
    n_valid = sum(n_valid)
  ) %>%
  print()



```
## 4.3.3 Probe Sentiment Narrative Development

```{r alignment-probe-example, fig.width=12, fig.height=6}


p_probe_sentiment <- df_field %>%
  mutate(time = ymd_hms(timestamp)) %>%
  group_by(conversation_id) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(turn_index = row_number()) %>%
  ungroup() %>%
  filter(conversation_id %in% unique(conversation_id)[26:28]) %>%
  pivot_longer(cols = c(author_1_valence, author_2_valence),
               names_to = "source", values_to = "valence") %>%
  mutate(
    source = recode(source, "author_1_valence" = "author_1", "author_2_valence" = "author_2")
  ) %>%
  group_by(conversation_id, source) %>%
  arrange(turn_index, .by_group = TRUE) %>%
  mutate(
    valence_smoothed = rollmean(valence, k = 1,
                                fill = NA,
                                align = "center")
  ) %>%
  ungroup() %>%
  ggplot(aes(x = turn_index, y = valence_smoothed, color = source)) +
  geom_line(size = 1.2) +
  geom_point(aes(y = valence), size = 1.2, alpha = 0.5) +
  scale_color_manual(
    values = c("author_1" = "#377EB8", "author_2" = "#FF7F00"),
    labels = c("author_1" = "Agent 1", "author_2" = "Agent 2")
  ) +
  facet_wrap(~ conversation_id, scales = "free_x") +
  labs(
    title = "Smoothed valence over turns (rolling mean, k = 1)",
    x = "Turn index",
    y = "Valence",
    color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0)
  )

print(p_probe_sentiment)
ggsave(file.path(GRAPHS_DIR, "valence_probe_temporal.png"),
       p_probe_sentiment, width = 12, height = 6, dpi = 300)

```

## 4.4.1 Alignment Comparison Plot

```{r alignment-comparison-plot, fig.width=12, fig.height=6}
# Combine all correlations
cor_all <- bind_rows(cor_same_combined, cor_prev_combined) %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("same_turn", "prev_turn"))
  )

# Boxplot with jittered points
p_alignment <- cor_all %>%
  filter(data_source == "Field") %>%
  ggplot(aes(x = alignment_type, y = rho, fill = alignment_type)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.4) +
  scale_fill_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  scale_color_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    name = "Alignment Type",
    labels = format_alignment_label
  ) +
  scale_x_discrete(labels = format_alignment_label) +
  labs(
    title = "Valence Alignment: Field U_t,U2_t vs U2_t-1,Ut",
    x = "Alignment Type",
    y = "Pearson ρ (per story)"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_alignment)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison.png"),
       p_alignment, width = 12, height = 6, dpi = 300)
```

## 4.4.2 Comparison: Per-Story vs Aggregate Correlations

```{r alignment-comparison-aggregate, fig.width=12, fig.height=8}
# Prepare aggregate correlation data for overlay
agg_same <- cor_same_combined_all_turns %>%
  mutate(alignment_type = "same_turn")

agg_prev <- cor_same_combined_all_turns_prev %>%
  mutate(alignment_type = "prev_turn") %>%
  select(data_source, rho, alignment_type)

agg_data <- bind_rows(agg_same, agg_prev) %>%
  filter(data_source == "Field") %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("same_turn", "prev_turn"))
  )

# Create plot with per-story boxplots and aggregate overlay
p_comparison <- cor_all %>%
  filter(data_source == "Field") %>%
  ggplot(aes(x = alignment_type, y = rho, fill = alignment_type)) +
  # Per-story boxplots
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  # Aggregate correlations as large points
  geom_point(data = agg_data, 
             aes(x = alignment_type, y = rho, fill = alignment_type),
             position = position_dodge(width = 0.75),
             size = 6, shape = 23, color = "black", stroke = 1.5) +
  scale_fill_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  scale_color_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    name = "Alignment Type",
    labels = format_alignment_label
  ) +
  scale_x_discrete(labels = format_alignment_label) +
  labs(
    title = "Valence Alignment: Per-Story vs Aggregate Correlations",
    subtitle = "Boxplots = per-story correlations; Diamonds = aggregate (all turns pooled)",
    x = "Alignment Type",
    y = "Pearson ρ"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_comparison)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison_with_aggregate.png"),
       p_comparison, width = 12, height = 8, dpi = 300)

# Print summary comparison
cat("\n=== Per-Story vs Aggregate Correlation Comparison ===\n")
comparison_summary <- bind_rows(
  cor_all %>%
    group_by(data_source, alignment_type) %>%
    summarise(
      mean_rho = mean(rho, na.rm = TRUE),
      method = "Per-Story Mean",
      .groups = "drop"
    ),
  agg_data %>%
    mutate(mean_rho = rho, method = "Aggregate All Turns") %>%
    select(data_source, alignment_type, mean_rho, method)
) %>%
  pivot_wider(names_from = method, values_from = mean_rho)

print(comparison_summary)

```



```{r}
# We assume df_field already has:


# 1) Build long-format dataset with both directions in a common structure

# Same-turn: author_1 → author_2
df_same_dir <- df_field %>%
  transmute(
    conversation_id,
    turn,
    alignment_direction = "author_1_to_author_2",
    predictor_valence = author_1_valence,
    response_valence = author_2_valence
  )

# Previous-turn: author_2(t-1) → author_1(t) (response to previous author_2 valence)
df_prev_dir <- df_field %>%
  filter(!is.na(author_2_valence_lag1)) %>%
  transmute(
    conversation_id,
    turn,
    alignment_direction = "author_2_to_author_1",
    predictor_valence = author_2_valence_lag1,
    response_valence = author_1_valence
  )

# Combine into one long dataset
df_alignment_long <- bind_rows(df_same_dir, df_prev_dir) %>%
  mutate(
    alignment_direction = factor(
      alignment_direction,
      levels = c("author_1_to_author_2", "author_2_to_author_1")
    )
  )

cat("=== Directional Alignment Long-Format Data ===\n")
df_alignment_long %>%
  group_by(alignment_direction) %>%
  summarise(
    n_rows = n(),
    mean_predictor = mean(predictor_valence, na.rm = TRUE),
    mean_response  = mean(response_valence,  na.rm = TRUE),
    n_conversations = n_distinct(conversation_id),
    .groups = "drop"
  ) %>%
  print()
```


```{r}

df_std <- df_alignment_long %>%
  group_by(conversation_id, alignment_direction) %>%   # <-- KEY choice
  mutate(
    predictor_z = as.numeric(scale(predictor_valence)),
    response_z  = as.numeric(scale(response_valence))
  ) %>%
  ungroup()

df_std$alignment_direction <- factor(df_std$alignment_direction,
                                     levels = c("author_1_to_author_2", "author_2_to_author_1"))
contrasts(df_std$alignment_direction) <- contr.sum(2)

m <- lmer(response_z ~ predictor_z * alignment_direction + (1 + predictor_z | conversation_id),
          data = df_std)

summary(m)

# slopes per direction + difference test
slopes <- emtrends(m, specs="alignment_direction", var="predictor_z")
slopes
contrast(slopes, method="revpairwise")
```

```{r}
#general alignment
p_general_alignment <- df_alignment_long %>%
  ggplot(aes(x = predictor_valence, y = response_valence)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.1) +
  labs(
    title = "Valence Alignment (Pooled Across Directions)",
    subtitle = "Single regression line fitted to all predictor–response pairs (direction ignored)",
    x = "Predictor valence",
    y = "Response valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
print(p_general_alignment)

ggsave(
  file.path(GRAPHS_DIR, "directional_valence_alignment_field.png"),
  p_general_alignment,
  width = 10,
  height = 6,
  dpi = 300
)

# 3) Visualise directional alignment with separate regression lines

p_direction_alignment <- df_alignment_long %>%
  ggplot(aes(
    x = predictor_valence,
    y = response_valence,
    color = alignment_direction
  )) +
  geom_point(alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    fullrange = TRUE,
    size = 1.2
  ) +
  scale_color_manual(
    values = c("author_1_to_author_2" = "#377EB8", "author_2_to_author_1" = "#FF7F00"),
    name = "Alignment Direction",
    labels = c(expression((U[t] * ", " * A[t])), expression((A[t-1] * ", " * U[t])))
  ) +
  labs(
    title = "Directional Valence Alignment in Field Data",
    x = "Predictor Valence",
    y = "Response Valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_direction_alignment)

ggsave(
  file.path(GRAPHS_DIR, "directional_valence_alignment_field.png"),
  p_direction_alignment,
  width = 10,
  height = 6,
  dpi = 300
)
```


## 4.6 Statistical Tests: Alignment Differences

```{r alignment-stats}
# Test 1: Same-turn alignment difference
#t_test_same <- t.test(rho ~ data_source, data = cor_same_combined)
#cat("=== T-test: Same-Turn Alignment ===\n")
#print(t_test_same)

# Test 2: Previous-turn alignment difference
#t_test_prev <- t.test(rho ~ data_source, data = cor_prev_combined)
#cat("\n=== T-test: Previous-Turn Alignment ===\n")
#print(t_test_prev)

# Test 3: One-sample t-tests against zero (is alignment significant?)
cat("\n=== One-Sample Tests Against Zero ===\n")
cor_all %>%
  group_by(data_source, alignment_type) %>%
  summarise(
    t_stat = t.test(rho, mu = 0)$statistic,
    p_value = t.test(rho, mu = 0)$p.value,
    mean_rho = mean(rho, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  print()

# Test 4: Paired t-test (Directional Asymmetry within Story)
# We join same and prev correlations by conversation_id to create pairs
cat("\n=== Paired T-Test: Directional Asymmetry (Same vs Prev) ===\n")

# Prepare Field pairs
pairs_field <- inner_join(
  cor_field_same %>% select(conversation_id, rho_same = rho),
  cor_field_prev %>% select(conversation_id, rho_prev = rho),
  by = "conversation_id"
)

# Apply Fisher z-transformation before t-test
t_test_paired_field <- t.test(atanh(pairs_field$rho_same), atanh(pairs_field$rho_prev), paired = TRUE)
cat("Field Data: Paired t-test (Fisher z-transformed Same Turn - Previous Turn)\n")
print(t_test_paired_field)

# Calculate back-transformed means for reporting
mean_z_same <- mean(atanh(pairs_field$rho_same), na.rm = TRUE)
mean_z_prev <- mean(atanh(pairs_field$rho_prev), na.rm = TRUE)

cat("\n--- Interpretation (Back-transformed means) ---\n")
cat("Mean Same-Turn r (author_1->author_2):", round(tanh(mean_z_same), 3), "\n")
cat("Mean Prev-Turn r (author_2->author_1):", round(tanh(mean_z_prev), 3), "\n")

# Save results
alignment_test_results <- tibble(
  test = c("Same-turn", "Previous-turn"),
  #t_statistic = c(t_test_same$statistic, t_test_prev$statistic),
  #p_value = c(t_test_same$p.value, t_test_prev$p.value),
  mean_field = c(mean(cor_field_same$rho, na.rm = TRUE), 
                 mean(cor_field_prev$rho, na.rm = TRUE))
)

write_csv(alignment_test_results, 
          file.path(TABLES_DIR, "alignment_statistical_tests.csv"))
```


## Cosine Distribution Difference Test

```{r cosine-sim-distribution}


df_field_cosine_long <- df_field %>%
  pivot_longer(cols = c(cosine_sim_same, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_same = "same_turn",
                            cosine_sim_prev = "prev_turn"),
         direction = factor(direction, levels = c("same_turn", "prev_turn"))
         )
df_field_cosine_long_prev_self <- df_field %>%
  pivot_longer(cols = c(cosine_sim_author_1_self, cosine_sim_prev),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_author_1_self = "Self",
                            cosine_sim_prev = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor"))
         )
df_field_cosine_long_prev_self_author_2 <- df_field %>%
  pivot_longer(cols = c(cosine_sim_author_2_self, cosine_sim_same),
               names_to = "direction", values_to = "cosine_sim") %>%
  mutate(direction = recode(direction,
                            cosine_sim_author_2_self = "Self",
                            cosine_sim_same = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor")))


# Calculate means for vertical lines
means_field <- df_field_cosine_long %>%
  group_by(direction) %>%
  summarise(mean_val = mean(cosine_sim, na.rm = TRUE), .groups = "drop")




cosine_difference_mixed_model_self_author_1 <- lmer(cosine_sim ~ direction + (direction | conversation_id), data=df_field_cosine_long_prev_self)
cat("\n=== Cosine Sim: Agent 1 (Self vs Interlocutor) ===\n")
print(round(summary(cosine_difference_mixed_model_self_author_1)$coefficients, 4))

cosine_difference_mixed_model_self_author_2 <- lmer(cosine_sim ~ direction + (direction | conversation_id), data = df_field_cosine_long_prev_self_author_2)
cat("\n=== Cosine Sim: Agent 2 (Self vs Interlocutor) ===\n")
print(round(summary(cosine_difference_mixed_model_self_author_2)$coefficients, 4))

cosine_difference_mixed_model <- lmer(cosine_sim ~ direction + (direction | conversation_id), data = df_field_cosine_long)
cat("\n=== Cosine Sim: General Direction Comparison ===\n")
print(round(summary(cosine_difference_mixed_model)$coefficients, 4))
```


```{r}
# Create plot with per-story boxplots and aggregate overlay
p_comparison_cosine_violin <- df_field_cosine_long %>%
  ggplot(aes(x = direction, y = cosine_sim, fill = direction)) +
  # Per-story boxplots
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  scale_color_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    name = "direction",
    labels = format_alignment_label
  ) +
  scale_x_discrete(labels = format_alignment_label) +
  labs(
    title = "Semantic Alignment: Cosine Similarity Between Agents",
    x = "Alignment Type",
    y = "Cosine Similarity"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_comparison_cosine_violin)

ggsave(file.path(GRAPHS_DIR, "cosine_sim_violin.png"),
       p_comparison_cosine_violin, width = 12, height = 8, dpi = 300)

p_field <- ggplot(df_field_cosine_long, aes(x = cosine_sim, fill = direction)) +
  geom_density(alpha = 0.4, color = NA) +
  scale_fill_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    name = "",
    labels = format_alignment_label
  ) +
  scale_color_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  labs(title = "Field Data", x = "Cosine Similarity", y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(p_field)
ggsave(
  file.path(GRAPHS_DIR, "directional_cosine_alignment_field.png"),
  p_field,
  width = 10,
  height = 6,
  dpi = 300
)
```
## Combined Alignment Plot

```{r combined-alignment-plot, fig.width=12, fig.height=12}
# Combine Valence Alignment (p_comparison) and Semantic Alignment (p_comparison_cosine_violin)
# Note: p_comparison is defined in chunk 'alignment-comparison-aggregate'
#       p_comparison_cosine_violin is defined in the chunk above

# Remove x-axis title and text from the top plot to avoid redundancy
p_comparison_clean <- p_comparison + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

combined_alignment_plot <- (p_comparison_clean / p_comparison_cosine_violin) +
  plot_annotation(
    title = "Comparison of Valence and Semantic Alignment",
    tag_levels = 'A'
  )

print(combined_alignment_plot)

ggsave(file.path(GRAPHS_DIR, "combined_valence_semantic_alignment.png"),
       combined_alignment_plot, width = 12, height = 12, dpi = 300)
```


```{r cosine-self-vs-interlocutor-plot, fig.width=10, fig.height=6}
# Plot for Agent 1: Self vs Interlocutor Cosine Similarity
p_cosine_author_1_self_vs_interlocutor <- df_field_cosine_long_prev_self %>%
  ggplot(aes(x = direction, y = cosine_sim, fill = direction)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("Self" = "#377EB8", "Interlocutor" = "#FF7F00")) +
  labs(
    title = "Agent 1 Semantic Alignment: Self vs. Interlocutor",
    subtitle = "Cosine Similarity: Agent 1(t) ↔ Agent 1(t-1) vs. Agent 1(t) ↔ Agent 2(t-1)",
    x = "Alignment Target",
    y = "Cosine Similarity"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40")
  )

print(p_cosine_author_1_self_vs_interlocutor)

ggsave(file.path(GRAPHS_DIR, "cosine_similarity_author_1_self_vs_interlocutor.png"),
       p_cosine_author_1_self_vs_interlocutor, width = 10, height = 6, dpi = 300)
```

## Valence Distance Difference Test

```{r valence-distance-distribution}
# Calculate valence distances
df_field_val <- df_field %>%
  mutate(
    val_dist_same = abs(author_1_valence - author_2_valence),
    val_dist_prev = abs(author_1_valence - author_2_valence_lag1),
    val_dist_author_1_self = abs(author_1_valence - author_1_valence_lag1),
    val_dist_author_2_self = abs(author_2_valence - author_2_valence_lag1)
  )

# Author 1: Self vs Partner (Prev)
df_field_val_long_author_1 <- df_field_val %>%
  pivot_longer(cols = c(val_dist_author_1_self, val_dist_prev),
               names_to = "direction", values_to = "val_dist") %>%
  mutate(direction = recode(direction,
                            val_dist_author_1_self = "Self",
                            val_dist_prev = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor")))

# Author 2: Self vs Partner (Same)
df_field_val_long_author_2 <- df_field_val %>%
  pivot_longer(cols = c(val_dist_author_2_self, val_dist_same),
               names_to = "direction", values_to = "val_dist") %>%
  mutate(direction = recode(direction,
                            val_dist_author_2_self = "Self",
                            val_dist_same = "Interlocutor"),
         direction = factor(direction, levels = c("Self", "Interlocutor")))

cat("\n=== Valence Distance Model: Agent 1 (Self vs Interlocutor) ===\n")
val_model_author_1 <- lmer(val_dist ~ direction + (direction | conversation_id), 
                       data = df_field_val_long_author_1)
print(round(summary(val_model_author_1)$coefficients, 4))

cat("\n=== Valence Distance Model: Agent 2 (Self vs Interlocutor) ===\n")
val_model_author_2 <- lmer(val_dist ~ direction + (direction | conversation_id), 
                      data = df_field_val_long_author_2)
print(round(summary(val_model_author_2)$coefficients, 4))
```

```{r valence-distance-plot, fig.width=12, fig.height=6}
# Combine for plotting
df_val_plot <- bind_rows(
  df_field_val_long_author_1 %>% mutate(agent = "author_1"),
  df_field_val_long_author_2 %>% mutate(agent = "author_2")
)

p_val_dist <- df_val_plot %>%
  ggplot(aes(x = direction, y = val_dist, fill = direction)) +
  geom_violin(width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  facet_wrap(~ agent) +
  scale_fill_manual(values = c("Self" = "#377EB8", "Interlocutor" = "#FF7F00")) +
  labs(
    title = "Valence Distance: Self vs Interlocutor Alignment",
    subtitle = "Lower distance = Higher alignment",
    x = "Alignment Target",
    y = "Absolute Valence Distance"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    strip.text = element_text(size = 14, face = "bold")
  )

print(p_val_dist)

ggsave(file.path(GRAPHS_DIR, "valence_distance_comparison.png"),
       p_val_dist, width = 12, height = 6, dpi = 300)
```

## 4.7 Alignment Duration and Stability

```{r alignment-duration-dynamics}
# 1. Calculate Velocity and Alignment Flags
df_dynamics <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn) %>%
  mutate(
    # Calculate change from previous turn (Velocity)
    delta_author_1 = author_1_valence - lag(author_1_valence),
    delta_author_2 = author_2_valence - lag(author_2_valence),
    
    # Define Alignment: Signs of change match
    # Exclude cases where change is exactly 0 (no movement)
    
    # DIRECTION 1: author_1 → author_2 (Same Turn Response)
    # Does author_2 move in same direction author_1 just moved?
    is_aligned_same = sign(delta_author_1) == sign(delta_author_2) & delta_author_1 != 0 & delta_author_2 != 0,
    
    # DIRECTION 2: author_2(t-1) → author_1(t) (Next Turn Response)
    # Does author_1 move in same direction author_2 moved in previous turn?
    # Compare author_1(t) delta with author_2(t-1) delta
    is_aligned_prev = sign(delta_author_1) == sign(lag(delta_author_2)) & delta_author_1 != 0 & lag(delta_author_2) != 0
  ) %>%
  ungroup()

# 2. Extract Run Lengths (Streak Durations)
get_run_lengths <- function(x) {
  # Ensure input is a standard atomic vector (removes attributes)
  x <- as.vector(x)
  
  # Return empty if null or all NA
  if(length(x) == 0 || all(is.na(x))) return(numeric(0))
  
  # Filter NAs explicitly (cleaner than na.omit for rle)
  x <- x[!is.na(x)]
  
  # Double check length after filtering
  if(length(x) == 0) return(numeric(0))
  
  r <- rle(x)
  # Return lengths where value is TRUE
  return(r$lengths[r$values == TRUE])
}

alignment_durations <- df_dynamics %>%
  group_by(conversation_id) %>%
  summarise(
    duration_same = list(get_run_lengths(is_aligned_same)),
    duration_prev = list(get_run_lengths(is_aligned_prev))
  ) %>%
  pivot_longer(cols = c(duration_same, duration_prev), 
               names_to = "direction", 
               values_to = "duration") %>%
  unnest(duration) %>%
  mutate(
    direction_label = recode(direction, 
                             duration_same = "same_turn",
                             duration_prev = "prev_turn")
  )

# 3. Calculate General Alignment Frequency (Hypothesis: ai1->AI is more frequent)
alignment_frequency <- df_dynamics %>%
  pivot_longer(cols = c(is_aligned_same, is_aligned_prev),
               names_to = "direction",
               values_to = "is_aligned") %>%
  filter(!is.na(is_aligned)) %>%
  group_by(direction) %>%
  summarise(
    total_turns = n(),
    aligned_turns = sum(is_aligned),
    prop_aligned = aligned_turns / total_turns
  ) %>%
  mutate(
    direction_label = recode(direction, 
                             is_aligned_same = "same_turn",
                             is_aligned_prev = "prev_turn")
  )
```

### 4.7.1 Visualization: Alignment Frequency (Stability)

```{r plot-alignment-frequency, fig.width=8, fig.height=5}
# Plot Overall Frequency
p_freq <- ggplot(alignment_frequency, aes(x = direction_label, y = prop_aligned, fill = direction_label)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_text(aes(label = scales::percent(prop_aligned, accuracy = 0.1)), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.8)) + # Scale approx
  scale_fill_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  scale_x_discrete(labels = format_alignment_label) +
  labs(
    title = "Frequency of Directional Alignment",
    subtitle = "Proportion of turns where responder moves in same direction as initiator",
    x = NULL,
    y = "Alignment Probability"
  ) +
  theme(legend.position = "none")

print(p_freq)
ggsave(file.path(GRAPHS_DIR, "alignment_frequency_bar.png"), p_freq, width = 8, height = 5)

# Statistical Test (Proportions Test)
prop_test <- prop.test(
  x = alignment_frequency$aligned_turns,
  n = alignment_frequency$total_turns
)
cat("\n=== Proportions Test: Alignment Frequency ===\n")
print(prop_test)
```

### 4.7.2 Visualization: Duration Distribution (Survival Analysis)

```{r plot-alignment-duration, fig.width=10, fig.height=6}
# Calculate summary stats for durations
duration_stats <- alignment_durations %>%
  group_by(direction_label) %>%
  summarise(
    mean_duration = mean(duration),
    median_duration = median(duration),
    max_duration = max(duration),
    n_streaks = n()
  )
print(duration_stats)

# Plot: Survival Curve style (Retention of Alignment)
# showing % of streaks that last at least X turns
duration_counts <- alignment_durations %>%
  group_by(direction_label, duration) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(direction_label) %>%
  arrange(desc(duration)) %>%
  mutate(
    cum_count = cumsum(count),
    total_streaks = max(cum_count), # max cumsum is total streaks
    prop_remaining = cum_count / total_streaks
  )

p_duration_curve <- ggplot(duration_counts, aes(x = duration, y = prop_remaining, color = direction_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(
    values = c("same_turn" = "#377EB8", "prev_turn" = "#FF7F00"),
    labels = format_alignment_label
  ) +
  labs(
    title = "Persistence of Emotional Alignment",
    subtitle = "Percentage of alignment streaks lasting at least X turns",
    x = "Duration (Consecutive Turns)",
    y = "Proportion of Streaks Remaining",
    color = "Direction"
  ) +
  theme(legend.position = "top")

print(p_duration_curve)
ggsave(file.path(GRAPHS_DIR, "alignment_duration_survival.png"), p_duration_curve, width = 10, height = 6)

# Statistical Test: Wilcoxon Rank Sum Test for Durations
# (Are streaks in one direction statistically longer?)
wilcox_test <- wilcox.test(duration ~ direction_label, data = alignment_durations)

cat("\n=== Wilcoxon Rank Sum Test: Alignment Duration ===\n")
print(wilcox_test)
```

# 6. Summary and Conclusions

## 6.1 Key Findings Summary

```{r summary-table}
# Create summary table
slopes_summary <- as.data.frame(slopes)
beta_author_1_to_author_2 <- slopes_summary[slopes_summary$alignment_direction == "author_1_to_author_2", "predictor_z.trend"]
beta_author_2_to_author_1 <- slopes_summary[slopes_summary$alignment_direction == "author_2_to_author_1", "predictor_z.trend"]

summary_results <- tribble(
  ~Metric, ~author_1, ~author_2,
  "Mean Valence", mean(df_field$author_1_valence, na.rm = TRUE), mean(df_field$author_2_valence, na.rm = TRUE),
  "Sentiment Alignment (β)", beta_author_1_to_author_2, beta_author_2_to_author_1,
  "Alignment Frequency (%)", alignment_frequency$prop_aligned[alignment_frequency$direction == "is_aligned_prev"] * 100, alignment_frequency$prop_aligned[alignment_frequency$direction == "is_aligned_same"] * 100,
  "Mean Alignment Duration (turns)", duration_stats$mean_duration[duration_stats$direction_label == "prev_turn"], duration_stats$mean_duration[duration_stats$direction_label == "same_turn"],
  "Semantic Alignment (Cosine)", mean(df_field_cosine_long_prev_self$cosine_sim[df_field_cosine_long_prev_self$direction == "Interlocutor"], na.rm = TRUE), mean(df_field_cosine_long$cosine_sim[df_field_cosine_long$direction == "same_turn"], na.rm = TRUE)
) %>%
  mutate(
    across(c(author_1, author_2), ~round(., 3))
  )

print(summary_results)

write_csv(summary_results, file.path(TABLES_DIR, "analysis_summary.csv"))
```


# 7. Supplementary: Interaction Effects


## 7.2 LLM Type Interaction Effects

### 7.2.1 Same-Turn Alignment by LLM Type
```{r llm-same-turn}
# Mixed model with LLM type interaction
df_field_llm <- df_field %>% 
  filter(!is.na(model_id)) %>%
  filter(interaction_count > 1) %>%
  mutate(
    model_id_name = recode(model_id, 
                           "1" = "Gpt-4.1", 
                           "2" = "Claude-sonnet-4-5", 
                           "3" = "Llama-3.3-70B-Instruct", 
                           "4" = "Qwen2.5-72B-Instruct")
  )

model_llm_same <- lmer(author_2_valence ~ author_1_valence * factor(model_id) + (1 | conversation_id), 
                       data = df_field_llm)

cat("=== Same-Turn Alignment: LLM Type Interaction ===\n")
print(round(summary(model_llm_same)$coefficients, 4))

p_llm_same <- df_field_llm %>%
  mutate(model_id_name = as.factor(model_id_name)) %>%
  ggplot(aes(x = author_1_valence, y = author_2_valence, color = model_id_name)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  labs(
    title = "Agent 2 Valence Same Turn by LLM (Agent 1 → Agent 2)",
    x = "Agent 1 Valence",
    y = "Agent 2 Valence",
    color = "LLM Model"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_llm_same)

ggsave(file.path(GRAPHS_DIR, "alignment_llm_same.png"),
       p_llm_same, width = 12, height = 6, dpi = 300)

```

### 7.2.2 Previous-Turn Alignment by LLM Type
```{r llm-prev-turn}
# Mixed model with LLM type interaction
df_field_llm_lag <- df_field %>% 
  filter(!is.na(model_id), !is.na(author_2_valence_lag1)) %>%   mutate(
    model_id_name = recode(model_id, 
                           "1" = "Gpt-4.1", 
                           "2" = "Claude-sonnet-4-5", 
                           "3" = "Llama-3.3-70B-Instruct", 
                           "4" = "Qwen2.5-72B-Instruct")
  )

model_llm_prev <- lmer(author_1_valence ~ author_2_valence_lag1 * factor(model_id) + (1 | conversation_id), 
                       data = df_field_llm_lag)

cat("=== Previous-Turn Alignment: LLM Type Interaction ===\n")
print(round(summary(model_llm_prev)$coefficients, 4))

p_llm_prev <- df_field_llm_lag %>%
  mutate(model_id_name = as.factor(model_id_name)) %>%
  ggplot(aes(x = author_2_valence_lag1, y = author_1_valence, color = model_id_name)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange= TRUE, size = 1.2) +
  labs(
    title = "Agent 1 Valence by Previous Agent 2 (Agent 2 → Agent 1)",
    x = "Previous Agent 2 Valence",
    y = "Agent 1 Valence"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_llm_prev)

ggsave(file.path(GRAPHS_DIR, "alignment_llm_prev.png"),
       p_llm_prev, width = 12, height = 6, dpi = 300)
```

### 7.2.3 Per-Story Correlations by LLM Type
```{r llm-per-story}
# Same-turn correlations stratified by LLM type
cor_llm_same <- df_field %>%
  filter(!is.na(model_id)) %>%
  group_by(conversation_id, model_id) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(author_1_valence, author_2_valence, use = "complete.obs"),
    n_turns = n(),
    .groups = "drop"
  )

cat("=== Same-Turn Correlations by LLM Type ===\n")
cor_llm_same %>%
  group_by(model_id) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

# ANOVA test
anova_llm <- aov(rho ~ factor(model_id), data = cor_llm_same)
cat("\nANOVA (LLM type effect on same-turn rho):\n")
print(summary(anova_llm))
```


## 7.3 Valence Distribution by Conditions

### 7.3.2 Valence by LLM Type
```{r valence-llm}
df_field %>%
  filter(!is.na(model_id)) %>%
  pivot_longer(cols = c(author_1_valence, author_2_valence),
               names_to = "agent", values_to = "valence") %>%
  group_by(model_id, agent) %>%
  summarise(
    mean_valence = mean(valence, na.rm = TRUE),
    sd_valence = sd(valence, na.rm = TRUE),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```

## 7.1 Starter Interaction Effects

### 7.1.1 Same-Turn Alignment by Starter (Removed for now)
```{r starter-same-turn}

#lmer_valence_time <- lmer(valence ~ turn + (1 | conversation_id), data = df_field_long)
#print(round(summary(lmer_valence_time)$coefficients, 4))

#Mixed model with starter interaction
#model_starter_same <- lmer(valence ~ starter + (1 | conversation_id), 
#                           data = df_field_long)

#cat("=== Same-Turn Alignment: Starter Interaction ===\n")
#print(round(summary(model_starter_same)$coefficients, 4))
```

### 7.1.2 Previous-Turn Alignment by Starter (Removed for now)
```{r starter-prev-turn}
# Mixed model with starter interaction
#df_field_lag <- df_field %>% filter(!is.na(author_2_valence_lag1))

#model_starter_prev <- lmer(author_1_valence ~ author_2_valence_lag1 * starter + (1 | conversation_id), 
#                           data = df_field_lag)

#cat("=== Previous-Turn Alignment: Starter Interaction ===\n")
#print(round(summary(model_starter_prev)$coefficients, 4))
```

### 7.1.3 Per-Story Correlations by Starter (Removed for now)
```{r starter-per-story}
# Same-turn correlations stratified by starter
#cor_starter_same <- df_field %>%
#  filter(!is.na(starter)) %>%
#  group_by(conversation_id, starter) %>%
#  filter(n() >= 3) %>%
#  summarise(
#    rho = cor(author_1_valence, author_2_valence, use = "complete.obs"),
#    n_turns = n(),
#    .groups = "drop"
#  )

#cat("=== Same-Turn Correlations by Starter ===\n")
#cor_starter_same %>%
#  group_by(starter) %>%
#  summarise(
#    mean_rho = mean(rho, na.rm = TRUE),
#    sd_rho = sd(rho, na.rm = TRUE),
#    n_stories = n()
#  ) %>%
#  print()

# Test difference
#t_test_starter <- t.test(rho ~ starter, data = cor_starter_same)
#cat("\nT-test (starter effect on same-turn rho):\n")
#print(t_test_starter)
```




## 7.3 Valence Distribution by Conditions

### 7.3.1 Valence by Starter
```{r valence-starter}
#df_field %>%
#  filter(!is.na(starter)) %>%
#  pivot_longer(cols = c(author_1_valence, author_2_valence),
#               names_to = "agent", values_to = "valence") %>%
#  group_by(starter, agent) %>%
#  summarise(
#    mean_valence = mean(valence, na.rm = TRUE),
#    sd_valence = sd(valence, na.rm = TRUE),
#    .groups = "drop"
#  ) %>%
#  print()
```

## 8. Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

**Analysis complete.** All figures saved to `analysis/figures/` and tables to `analysis/tables/`.
