---
title: "Rubber_band_analysis"
subtitle: "Analysis of Field Data and AI-AI in Rubber-band dynamics"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex          # ANOVA
)

# Set project root
PROJECT_ROOT <- "/Users/halfidaldal/Documents/PENPAL_analysis"
setwd(PROJECT_ROOT)

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```



# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load dyadic sentiment scores
setwd("/Users/halfidaldal/Documents/PENPAL_analysis")
df_field <- read_parquet("data/TEXT/processed/dyadic_sentiment_scores.parquet")
df_simulated <- read_parquet("data/Berlin/processed/dyadic_sentiment_scores_simulated.parquet")

# Display structure
cat("Field data dimensions:", dim(df_field), "\n")
cat("Simulated data dimensions:", dim(df_simulated), "\n")

# Show column names
cat("\nField data columns:\n")
print(names(df_field))

cat("\nSimulated data columns:\n")
print(names(df_simulated))
```



## 1.2 Data Preprocessing

```{r preprocess}
# Preprocess field data
df_field <- df_field %>%
  rename(
    ai_valence = ai_sentiment_score,
    user_valence = user_sentiment_score
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing
    timestamp = ymd_hms(timestamp),
    
    # Data source label
    data_source = "Field"
  ) %>%
  # Ensure conversation_id is a character/factor
  mutate(conversation_id = as.character(conversation_id))

# Preprocess simulated data
df_simulated <- df_simulated %>%
  rename(
    ai_valence = ai_sentiment_score,
    user_valence = user_sentiment_score
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing (if available)
    timestamp = if_else(!is.na(timestamp), ymd_hms(timestamp), as.POSIXct(NA)),
    
    # Data source label
    data_source = "Simulated"
  ) %>%
  mutate(conversation_id = as.character(conversation_id))

# Summary statistics
cat("\n=== Field Data Summary ===\n")
df_field %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()

cat("\n=== Simulated Data Summary ===\n")
df_simulated %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()
```


## 1.3 Compute Lagged Variables and Cosine

```{r compute-lags}
# Add lagged AI valence for field data
# Note: lag() within group_by ensures no cross-story contamination
# Turn 1 of each story will have NA for lagged values (correct behavior)

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1),
  ) %>%
  ungroup()


cosine_sim <- function(x, y) {
  if (is.null(x) || is.null(y) || any(is.na(x)) || any(is.na(y))) {
    return(NA_real_)
  }
  return(sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2))))
}

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_embedding_prev = lag(ai_embedding)
  ) %>%
  rowwise() %>%
  mutate(
    cosine_sim_same = cosine_sim(user_embedding, ai_embedding),
    cosine_sim_prev = cosine_sim(user_embedding, ai_embedding_prev)
  ) %>%
  ungroup()

# Add lagged AI valence for simulated data
df_simulated <- df_simulated %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1)
  ) %>%
  ungroup()

# Verify no cross-story contamination
cat("=== Lag Variable Verification ===\n")
cat("Field data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_field$turn == 1 & !is.na(df_field$ai_valence_lag1)), "\n")
cat("Simulated data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_simulated$turn == 1 & !is.na(df_simulated$ai_valence_lag1)), "\n")
```


# 2. Temporal Dynamics Analysis

## 2.1 Stage-Based Analysis (Early/Middle/Late)

```{r stage-analysis}
# Function to compute stage-based valence gap
compute_stage_gaps <- function(df, data_source_label) {
  df %>%
    group_by(conversation_id) %>%
    arrange(turn, .by_group = TRUE) %>%
    mutate(
      turn_idx = row_number(),
      n_turns = n()
    ) %>%
    filter(n_turns >= 3) %>%  # Need at least 3 turns
    mutate(
      stage = ntile(turn_idx, 3),
      valence_gap = abs(user_valence - ai_valence)
    ) %>%
    ungroup() %>%
    group_by(conversation_id, stage) %>%
    summarise(
      mean_gap = mean(valence_gap, na.rm = TRUE),
      n_turns = first(n_turns),
      .groups = "drop"
    ) %>%
    mutate(
      data_source = data_source_label,
      stage_label = case_when(
        stage == 1 ~ "Early",
        stage == 2 ~ "Middle",
        stage == 3 ~ "Late"
      ),
      stage_label = factor(stage_label, levels = c("Early", "Middle", "Late")),
      story_length = if_else(n_turns >= median(n_turns), 
                            "High ≥ median", "Low < median")
    )
}

# Compute for both datasets
stage_gaps_field <- compute_stage_gaps(df_field, "Field")
stage_gaps_sim <- compute_stage_gaps(df_simulated, "Simulated")

# Combined
stage_gaps_all <- bind_rows(stage_gaps_field, stage_gaps_sim)

# Summary by stage
cat("=== Mean Valence Gap by Stage ===\n")
stage_gaps_all %>%
  group_by(data_source, stage_label) %>%
  summarise(
    mean_gap = mean(mean_gap, na.rm = TRUE),
    sd_gap = sd(mean_gap, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

## 2.2 Stage Valence Gap Plot

```{r stage-gap-plot, fig.width=12, fig.height=6}
# Field plot
p_stage_field <- ggplot(stage_gaps_field, 
                        aes(x = stage_label, y = mean_gap, 
                            group = conversation_id, color = story_length)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  stat_summary(fun = mean, geom = "line", aes(group = story_length),
               linewidth = 1.2, position = position_dodge(width = 0.2)) +
  stat_summary(fun = mean, geom = "point",
               size = 3, position = position_dodge(width = 0.2)) +
  scale_color_brewer(palette = "Set1", name = "Story Length") +
  labs(title = "Field Data", x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold")
  )

# Simulated plot
p_stage_sim <- ggplot(stage_gaps_sim, 
                      aes(x = stage_label, y = mean_gap, 
                          group = conversation_id, color = story_length)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  stat_summary(fun = mean, geom = "line", aes(group = story_length),
               linewidth = 1.2, position = position_dodge(width = 0.2)) +
  stat_summary(fun = mean, geom = "point",
               size = 3, position = position_dodge(width = 0.2)) +
  scale_color_brewer(palette = "Set2", name = "Story Length") +
  labs(title = "Simulated Data", x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold")
  )

# Combine
combined_stage <- (p_stage_field | p_stage_sim) +
  plot_annotation(
    title = "Mean Valence Gap by Turn Stage",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"))
  )

final_stage <- ggdraw() +
  draw_plot(combined_stage, 0, 0, 1, 1) +
  draw_label("Mean Valence Difference", x = 0.02, y = 0.5, angle = 90, size = 16) +
  draw_label("Turn Stage", x = 0.5, y = 0.02, size = 16)

print(final_stage)

ggsave(file.path(GRAPHS_DIR, "stage_valence_gap_comparison.png"),
       final_stage, width = 12, height = 6, dpi = 300)
```

## 2.3 Turn-by-Turn Valence Gap Trajectory

```{r gap-trajectory-plot, fig.width=12, fig.height=6}

# Compute valence gap for each turn
df_field_gap <- df_field %>%
  mutate(valence_gap = abs(user_valence - ai_valence))

df_sim_gap <- df_simulated %>%
  mutate(valence_gap = abs(user_valence - ai_valence))

# Field data plot
p_gap_field <- ggplot(df_field_gap, aes(x = turn, y = valence_gap, group = conversation_id)) +
  geom_line(alpha = 0.2, color = "#E41A1C") +  # Individual stories (faint)
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "#E41A1C", linewidth = 1.5) +  # Aggregate mean (bold)
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               color = "#E41A1C", size = 3) +
  labs(title = "Field Data", x = NULL, y = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Simulated data plot
p_gap_sim <- ggplot(df_sim_gap, aes(x = turn, y = valence_gap, group = conversation_id)) +
  geom_line(alpha = 0.2, color = "#8DA0CB") +  # Individual stories (faint)
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "#8DA0CB", linewidth = 1.5) +  # Aggregate mean (bold)
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               color = "#8DA0CB", size = 3) +
  labs(title = "Simulated Data", x = NULL, y = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Combine
combined_gap_trajectory <- (p_gap_field | p_gap_sim) +
  plot_annotation(
    title = "Valence Gap Trajectory Across Story Turns",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"))
  )

final_gap_trajectory <- ggdraw() +
  draw_plot(combined_gap_trajectory, 0, 0, 1, 1) +
  draw_label("Mean Valence Difference", x = 0.02, y = 0.5, angle = 90, size = 16) +
  draw_label("Turn Number", x = 0.5, y = 0.02, size = 16)

print(final_gap_trajectory)

ggsave(file.path(GRAPHS_DIR, "gap_trajectory_comparison.png"),
       final_gap_trajectory, width = 12, height = 6, dpi = 300)

```

## 2.4 Rubber-Band Dynamics

```{r rubberband-analysis}
# Function to compute rubber-band dynamics
compute_rubberband <- function(df, data_source_label) {
  df %>%
    group_by(conversation_id) %>%
    arrange(turn, .by_group = TRUE) %>%
    mutate(
      turn_idx = row_number(),
      n_turns = n(),
      valence_gap = abs(user_valence - ai_valence)
    ) %>%
    filter(n_turns >= 3) %>%
    mutate(stage = ntile(turn_idx, 3)) %>%
    ungroup() %>%
    group_by(conversation_id, stage) %>%
    summarise(
      mean_gap = mean(valence_gap, na.rm = TRUE),
      n_turns = first(n_turns),
      .groups = "drop"
    ) %>%
    pivot_wider(
      id_cols = c(conversation_id, n_turns),
      names_from = stage,
      values_from = mean_gap,
      names_prefix = "g"
    ) %>%
    filter(!is.na(g1) & !is.na(g2) & !is.na(g3)) %>%
    mutate(
      delta_12 = g2 - g1,  # Change from Early to Middle
      delta_23 = g3 - g2,  # Change from Middle to Late
      data_source = data_source_label,
      story_length = if_else(n_turns >= median(n_turns), 
                            "High ≥ median", "Low < median")
    )
}

# Compute for both datasets
rb_field <- compute_rubberband(df_field, "Field")
rb_sim <- compute_rubberband(df_simulated, "Simulated")

# Fit linear models
lm_field <- lm(delta_23 ~ delta_12, data = rb_field)
#lm_sim <- lm(delta_23 ~ delta_12 * story_length, data = rb_sim)

cat("=== Rubber-Band Model: Field Data ===\n")
print(summary(lm_field))

cat("\n=== Rubber-Band Model: Simulated Data ===\n")
#print(summary(lm_sim))
```

## 2.5 Turn-by-Turn Rubber-Band Dynamics

```{r rubberband-turn-level}
# Function to compute turn-by-turn rubber-band dynamics
compute_rubberband_turns <- function(df, data_source_label) {
  df %>%
    group_by(conversation_id) %>%
    arrange(turn, .by_group = TRUE) %>%
    mutate(
      valence_gap = abs(user_valence - ai_valence),
      gap_change = valence_gap - lag(valence_gap),  # Change from previous turn
      gap_change_next = lead(valence_gap) - valence_gap  # Change to next turn
    ) %>%
    filter(!is.na(gap_change) & !is.na(gap_change_next)) %>%  # Need both values
    ungroup() %>%
    mutate(
      data_source = data_source_label,
      n_turns = n_distinct(conversation_id)
    )
}

# Compute for both datasets
rb_turns_field <- compute_rubberband_turns(df_field, "Field")
rb_turns_sim <- compute_rubberband_turns(df_simulated, "Simulated")

# Fit linear models for turn-level rubber-band
lm_turns_field <- lm(gap_change_next ~ gap_change, data = rb_turns_field)
lm_turns_sim <- lm(gap_change_next ~ gap_change, data = rb_turns_sim)

cat("=== Turn-Level Rubber-Band Model: Field Data ===\n")
print(summary(lm_turns_field))

cat("\n=== Turn-Level Rubber-Band Model: Simulated Data ===\n")
print(summary(lm_turns_sim))

# Summary statistics
cat("\n=== Turn-Level Rubber-Band Correlation ===\n")
cat("Field correlation (gap_change vs gap_change_next):", 
    cor(rb_turns_field$gap_change, rb_turns_field$gap_change_next, use = "complete.obs"), "\n")
cat("Simulated correlation (gap_change vs gap_change_next):", 
    cor(rb_turns_sim$gap_change, rb_turns_sim$gap_change_next, use = "complete.obs"), "\n")
```

## 2.6 Rubber-Band Dynamics Plot

```{r rubberband-plot, fig.width=12, fig.height=6}
# Field plot
p_rb_field <- ggplot(rb_field, aes(x = delta_12, y = delta_23, color = story_length)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", aes(linetype = story_length), se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "Story Length") +
  scale_linetype_manual(values = c("solid", "dashed"), name = "Story Length") +
  labs(
    title = "Field Data",
    x = expression(Delta[12]~"(Middle − Early)"),
    y = expression(Delta[23]~"(Late − Middle)")
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14, face = "bold")
  )

# Simulated plot
p_rb_sim <- ggplot(rb_sim, aes(x = delta_12, y = delta_23, color = story_length)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", aes(linetype = story_length), se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set2", name = "Story Length") +
  scale_linetype_manual(values = c("solid", "dashed"), name = "Story Length") +
  labs(
    title = "Simulated Data",
    x = expression(Delta[12]~"(Middle − Early)"),
    y = NULL
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold")
  )

# Combine
combined_rb <- (p_rb_field | p_rb_sim) +
  plot_annotation(
    title = "Rubber-Band Dynamics in Valence Gap",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"))
  ) &
  theme(legend.position = "top")

print(combined_rb)

ggsave(file.path(GRAPHS_DIR, "rubberband_dynamics_comparison.png"),
       combined_rb, width = 12, height = 6, dpi = 300)
```

## 2.7 Turn-Level Rubber-Band Dynamics Plot

```{r rubberband-turn-plot, fig.width=12, fig.height=6}
# Field plot
p_rb_turns_field <- ggplot(rb_turns_field, aes(x = gap_change, y = gap_change_next)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.3, size = 2, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C", linewidth = 1.2) +
  labs(
    title = "Field Data",
    x = expression(Delta[t]~"(Gap change from"~t-1~"to"~t*")"),
    y = expression(Delta[t+1]~"(Gap change from"~t~"to"~t+1*")")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Simulated plot
p_rb_turns_sim <- ggplot(rb_turns_sim, aes(x = gap_change, y = gap_change_next)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.3, size = 2, color = "#8DA0CB") +
  geom_smooth(method = "lm", se = TRUE, color = "#8DA0CB", linewidth = 1.2) +
  labs(
    title = "Simulated Data",
    x = expression(Delta[t]~"(Gap change from"~t-1~"to"~t*")"),
    y = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Combine
combined_rb_turns <- (p_rb_turns_field | p_rb_turns_sim) +
  plot_annotation(
    title = "Turn-by-Turn Rubber-Band Dynamics in Valence Gap",
    subtitle = "Negative slope indicates correction: if gap increased, it tends to decrease next turn",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40")
    )
  )

print(combined_rb_turns)

ggsave(file.path(GRAPHS_DIR, "rubberband_turn_level_comparison.png"),
       combined_rb_turns, width = 12, height = 6, dpi = 300)
```

# 3. Permutation tests

## 3.1 Stage-level Permutation Test

```{r permutation-test}
obs_perm_slope <- coef(lm_field)[2]


set.seed(2025)
n_sims <- 1000
perm_slopes <- replicate(n_sims, {
  df_perm <- stage_gaps_field %>%
    group_by(conversation_id) %>%
    mutate(stage_p = sample(stage)) %>%    # shuffle within each id
    ungroup() %>%
    filter(stage_p %in% 1:3) %>%
    pivot_wider(
      id_cols     = conversation_id,
      names_from  = stage_p,
      values_from = mean_gap,
      names_prefix= "g",
      values_fn   = mean,      # collapse duplicates if any
      values_fill = NA
    ) %>%
    filter(!is.na(g1) & !is.na(g2) & !is.na(g3)) %>%
    transmute(
      d12 = g2 - g1,
      d23 = g3 - g2
    )
    coef(lm(d23 ~ d12, data = df_perm))[2]
})

p_stage <- mean(abs(perm_slopes) >= abs(obs_perm_slope))

cat("=== Stage-Level Permutation Test Results ===\n")
cat("Observed slope:", round(obs_perm_slope, 4), "\n")
cat("Null mean:", round(mean(perm_slopes), 4), "\n")
cat("Null SD:", round(sd(perm_slopes), 4), "\n")
cat("Effect size (z):", round((obs_perm_slope - mean(perm_slopes)) / sd(perm_slopes), 2), "\n")
cat("P-value:", format.pval(p_stage, digits = 3), "\n\n")
```

## 3.2 Brutal Stage-Level Permutation
```{r}
set.seed(2025)
n_sims <- 1000

perm_slopes_stage_brutal <- replicate(n_sims, {
  df_perm_brutal <- df_field_gap %>%
    group_by(conversation_id) %>%
    arrange(turn, .by_group = TRUE) %>%
    mutate(
      # shuffle the *sequence* of gaps inside each story
      valence_gap_perm = sample(valence_gap)
    ) %>%
    ungroup() %>%
    group_by(conversation_id) %>%
    mutate(
      turn_idx = row_number(),
      n_turns = n(),
      stage = ntile(turn_idx, 3)
    ) %>%
    group_by(conversation_id, stage) %>%
    summarise(
      mean_gap = mean(valence_gap_perm, na.rm = TRUE),
      n_turns = first(n_turns),
      .groups = "drop"
    ) %>%
    pivot_wider(
      id_cols = c(conversation_id, n_turns),
      names_from = stage,
      values_from = mean_gap,
      names_prefix = "g"
    ) %>%
    filter(!is.na(g1), !is.na(g2), !is.na(g3)) %>%
    mutate(
      delta_12 = g2 - g1,
      delta_23 = g3 - g2
    )

  coef(lm(delta_23 ~ delta_12, data = df_perm_brutal))[2]  # slope only
})

p_stage_brutal <- mean(abs(perm_slopes_stage_brutal) >= abs(obs_perm_slope))

cat("=== Stage-Level Permutation Test Results - Shuffle Turn-Level Valence Gap first ===\n")
cat("Observed slope:", round(obs_perm_slope, 4), "\n")
cat("Null mean:", round(mean(perm_slopes_stage_brutal), 4), "\n")
cat("Null SD:", round(sd(perm_slopes_stage_brutal), 4), "\n")
cat("Effect size (z):", round((obs_perm_slope - mean(perm_slopes_stage_brutal)) / sd(perm_slopes_stage_brutal), 2), "\n")
cat("P-value:", format.pval(p_stage_brutal, digits = 3), "\n\n")

```


## 3.2 Global Shuffle Stage-Level Test
```{r global-shuffle-test}
set.seed(2025)
df_global <- stage_gaps_field%>%
  mutate(stage_g = sample(stage)) %>%      # shuffle across all IDs
  group_by(conversation_id, stage_g) %>%
  summarise(mean_gap = mean(mean_gap), .groups = "drop") %>%  # collapse any duplicates
  pivot_wider(
    id_cols     = conversation_id,
    names_from  = stage_g,
    values_from = mean_gap,
    names_prefix= "g",
    values_fn   = mean,
    values_fill = NA
  ) %>%
  filter(!is.na(g1) & !is.na(g2) & !is.na(g3)) %>%
  transmute(
    d12 = g2 - g1,
    d23 = g3 - g2
  )

lm_global_shuffle <- lm(d23 ~ d12, data = df_global)
print(summary(lm_global_shuffle))

```


## 3.3 Turn-Level Permutation Test
```{r turn-permutation}
# Observed slope
obs_turn_slope <- coef(lm_turns_field)[2]

set.seed(2022)
n_sims <- 1000
perm_slopes_turn <- replicate(n_sims, {
  df_perm <- rb_turns_field %>%
    group_by(conversation_id) %>%
    mutate(gap_change_perm = sample(gap_change)) %>%
    ungroup()
  coef(lm(gap_change_next ~ gap_change_perm, data = df_perm))[2]
})

# Calculate p-value (two-tailed)
p_turn <- mean(abs(perm_slopes_turn) >= abs(obs_turn_slope))

cat("=== Turn-Level Permutation Test Results ===\n")
cat("Observed slope:", round(obs_turn_slope, 4), "\n")
cat("Null mean:", round(mean(perm_slopes_turn), 4), "\n")
cat("Null SD:", round(sd(perm_slopes_turn), 4), "\n")
cat("Effect size (z):", round((obs_turn_slope - mean(perm_slopes_turn)) / sd(perm_slopes_turn), 2), "\n")
cat("P-value:", format.pval(p_turn, digits = 3), "\n\n")

# Tag each turn pair by whether they cross stage boundaries
rb_turns_field %>%
  group_by(conversation_id) %>%
  mutate(
    stage = ntile(turn, 3),
    stage_next = lead(stage)
  ) %>%
  ungroup() %>%
  mutate(
    transition_type = case_when(
      stage == stage_next ~ "Within-Stage",
      stage < stage_next ~ "Stage Boundary",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(transition_type)) %>%
  group_by(transition_type) %>%
  summarise(
    rb_slope = coef(lm(gap_change_next ~ gap_change))[2],
    correlation = cor(gap_change, gap_change_next),
    n = n()
  )


```



# 4. Cosine similarity alignment

## 4.1 Cosine Similarity Distribution 
```{r cosine-distribution, fig.width=12, fig.height=6}

p_cosine_dist <- ggplot(df_field, aes(x = cosine_sim_same)) +
  geom_density(alpha = 0.4, fill = "#66C2A5", color = NA) + 
  geom_vline(aes(xintercept = mean(cosine_sim_same, na.rm = TRUE)),
             color = "#66C2A5", linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Distribution of Cosine Similarity (User vs AI)",
    x = "Cosine Similarity",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_cosine_dist)

ggsave(file.path(GRAPHS_DIR, "cosine_similarity_distribution.png"),
       p_cosine_dist, width = 12, height = 6, dpi = 300)

```

## 4.2 Cosine Similarity Mixed Model
```{r cosine-sim-lm}

df_cosine_sim <- df_field %>%
  filter(turn > 1, turn < 10) 

lm_cosine_turn <- lmer(
  cosine_sim_same ~ turn + (1 | conversation_id),
  data = df_cosine_sim
)
print(summary(lm_cosine_turn))
```

## 4.3 Cosine Similarity Linear Plot
```{r cosine-sim-starter, fig.width=12, fig.height=6}

cosine_sim_lm_plot <- df_field %>%
  filter(turn > 0, turn < 10) %>%
  ggplot(aes(x = turn, y = cosine_sim_same)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.3, size = 2) +  # no color here
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +  # no color here
  xlim(1, 9) +
  ylim(0.5, 0.8) +
  labs(
    title = "Field Data",
    x = "Turn",
    y = "Cosine Similarity Same Turn"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )
print(cosine_sim_lm_plot)

```

## 4.4 Cosine Similarity Over Time
```{r cosine-sim-plot, fig.width=12, fig.height=6}

df_field %>%
  filter(turn > 1, turn < 10) %>%
  group_by(turn) %>%
  summarise(mean_cosine = mean(cosine_sim_same, na.rm = TRUE)) %>%
  ggplot(aes(x = turn, y = mean_cosine)) +
  geom_line() +
  geom_point()


df_field %>%
  filter(turn < 10, turn > 1) %>%
  #group_by(starter) %>%
  ggplot(aes(x = turn, y = cosine_sim_same, group = conversation_id)) +
  geom_line(alpha = 0.2) +  # Individual stories (faint)
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "blue", size = 1.5)  # Aggregate mean (bold)



```

## 4.5 Cosine rubber band
```{r rubber-band}
compute_rubberband_cosine_turns <- function(df, data_source_label) {
  df %>%
    group_by(conversation_id) %>%
    arrange(turn, .by_group = TRUE) %>%
    mutate(
      cosine_dist = 1 - cosine_sim_same,
      dist_change = cosine_dist - lag(cosine_dist),  # C fhange from previous turn
      dist_change_next = lead(cosine_dist) - cosine_dist  # Change to next turn
    ) %>%
    filter(!is.na(dist_change) & !is.na(dist_change_next)) %>%  # Need both values
    ungroup() %>%
    mutate(
      data_source = data_source_label,
      n_turns = n_distinct(conversation_id)
    )
}

rb_cosine <- compute_rubberband_cosine_turns(df_field, "Field")
lm_turns_field_cosine <- lm(dist_change_next ~ dist_change, data = rb_cosine)


cat("=== Turn-Level Rubber-Band Cosine Distance: Field Data ===\n")
print(summary(lm_turns_field_cosine))


```


## 4.6 Turn level Cosine Permutation test
```{r cosine-turn-permutation}
# Observed slope
obs_cosine_slope <- coef(lm_turns_field_cosine)[2]

set.seed(2022)
n_sims <- 1000
perm_slopes_cosine <- replicate(n_sims, {
  df_perm <- rb_cosine %>%
    group_by(conversation_id) %>%
    mutate(dist_change_perm = sample(dist_change)) %>%
    ungroup()
  coef(lm(dist_change_next ~ dist_change_perm, data = df_perm))[2]
})

# Calculate p-value (two-tailed)
p_cosine <- mean(abs(perm_slopes_cosine) >= abs(obs_cosine_slope))

cat("=== Cosine Distance Turn-Level Permutation Test Results ===\n")
cat("Observed slope:", round(obs_cosine_slope, 4), "\n")
cat("Null mean:", round(mean(perm_slopes_cosine), 4), "\n")
cat("Null SD:", round(sd(perm_slopes_cosine), 4), "\n")
cat("Effect size (z):", round((obs_cosine_slope - mean(perm_slopes_cosine)) / sd(perm_slopes_cosine), 2), "\n")
cat("P-value:", format.pval(p_cosine, digits = 3), "\n\n")

# Tag each turn pair by whether they cross stage boundaries
rb_cosine %>%
  group_by(conversation_id) %>%
  mutate(
    stage = ntile(turn, 3),
    stage_next = lead(stage)
  ) %>%
  ungroup() %>%
  mutate(
    transition_type = case_when(
      stage == stage_next ~ "Within-Stage",
      stage < stage_next ~ "Stage Boundary",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(transition_type)) %>%
  group_by(transition_type) %>%
  summarise(
    rb_slope = coef(lm(dist_change_next ~ dist_change))[2],
    correlation = cor(dist_change, dist_change_next),
    n = n()
  )

```
## 4.7 Turn Level Cosine RB Linear plot

```{r, fig.width=12, fig.height=6}
p_rb_turns_field_cosine <- ggplot(rb_cosine, aes(x = dist_change, y = dist_change_next)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.3, size = 2, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C", linewidth = 1.2) +
  labs(
    title = "Field Data",
    x = expression(Delta[t]~"(Gap change from"~t-1~"to"~t*")"),
    y = expression(Delta[t+1]~"(Gap change from"~t~"to"~t+1*")")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )


combined_rb_turns <- (p_rb_turns_field_cosine | p_rb_turns_sim) +
  plot_annotation(
    title = "Turn-by-Turn Rubber-Band Dynamics in Cosine Distance",
    subtitle = "Negative slope indicates correction: if distance increased, it tends to decrease next turn",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40")
    )
  )

print(combined_rb_turns)

ggsave(file.path(GRAPHS_DIR, "rubberband_cosine_turn_level_comparison.png"),
       combined_rb_turns, width = 12, height = 6, dpi = 300)


```


## 4.8 Correlation Between Cosine Distance and Valence Rubber-Band

```{r rb-correlation, fig.width=12, fig.height=6}
# Turn filtering parameters
turn_min <- 2    # Minimum turn to include (set to 1 for no lower bound)
turn_max <- 9  # Maximum turn to include (set to Inf for no upper bound)

# Merge rubber-band data for both metrics
rb_combined <- rb_turns_field %>%
  inner_join(
    rb_cosine %>% select(conversation_id, turn, dist_change, dist_change_next),
    by = c("conversation_id", "turn"),
    suffix = c("_valence", "_cosine")
  ) %>%
  filter(turn >= turn_min, turn <= turn_max)  # Apply turn filtering

# Compute correlation between the two rubber-band patterns
cat("=== Correlation Between Valence and Cosine Distance Rubber-Band ===\n")
cat("Turn range: [", turn_min, ", ", 
    ifelse(is.infinite(turn_max), "Inf", turn_max), "]\n", sep = "")
cat("N observations (turn-level):", nrow(rb_combined), "\n\n")

# 1. Turn-level correlation of changes (Δt)
# This correlates valence and cosine changes at EACH turn
cor_changes <- cor(rb_combined$gap_change, rb_combined$dist_change, 
                   use = "complete.obs")
cat("Turn-level correlation of gap changes (Δt):", round(cor_changes, 3), "\n")
cat("  Interpretation: Do valence and cosine gaps change in the same direction\n")
cat("                  at the same turn across all stories?\n\n")

# 2. Turn-level correlation of next changes (Δt+1)
cor_next_changes <- cor(rb_combined$gap_change_next, rb_combined$dist_change_next, 
                       use = "complete.obs")
cat("Turn-level correlation of next changes (Δt+1):", round(cor_next_changes, 3), "\n")
cat("  Interpretation: Do subsequent valence and cosine corrections correlate?\n\n")

# 3. Per-story rubber-band slopes
# This is the KEY analysis: Do stories with strong valence RB also show strong cosine RB?
rb_slopes <- rb_combined %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%  # Need at least 3 turns for stable regression
  summarise(
    # Valence rubber-band slope: β from (gap_change_next ~ gap_change)
    valence_rb_slope = coef(lm(gap_change_next ~ gap_change))[2],
    # Cosine rubber-band slope: β from (dist_change_next ~ dist_change)
    cosine_rb_slope = coef(lm(dist_change_next ~ dist_change))[2],
    n_turns = n(),
    .groups = "drop"
  )

cat("N stories with >= 3 turns:", nrow(rb_slopes), "\n\n")

cat("N stories with >= 3 turns:", nrow(rb_slopes), "\n\n")

# Correlation between story-level slopes
# This answers: Do stories with strong valence RB also show strong cosine RB?
cor_slopes <- cor(rb_slopes$valence_rb_slope, rb_slopes$cosine_rb_slope, 
                 use = "complete.obs")
cat("STORY-LEVEL correlation of RB slopes:", round(cor_slopes, 3), "\n")
cat("  Interpretation: Across stories, does valence RB strength predict cosine RB strength?\n")
cat("                  Positive r means stories with strong valence correction\n")
cat("                  also show strong cosine distance correction.\n")

# Test significance
slope_test <- cor.test(rb_slopes$valence_rb_slope, rb_slopes$cosine_rb_slope)
cat("  95% CI: [", round(slope_test$conf.int[1], 3), ", ", 
    round(slope_test$conf.int[2], 3), "]\n", sep = "")
cat("  p-value:", format.pval(slope_test$p.value, digits = 3), "\n")
cat("  df:", slope_test$parameter, "\n\n")

# 4. Shared variance (R²)
shared_var <- cor_slopes^2
cat("Shared variance (R²):", round(shared_var, 3), "\n")
cat("Interpretation:", round(shared_var * 100, 1), 
    "% of variation in valence RB slope\n",
    "              is explained by cosine distance RB slope\n\n")

# 5. Scatter plot: Story-level RB slopes
p_rb_correlation <- ggplot(rb_slopes, aes(x = cosine_rb_slope, y = valence_rb_slope)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(alpha = 0.6, size = 2.5, color = "#E41A1C") +
  geom_smooth(method = "lm", se = TRUE, color = "#E41A1C", linewidth = 1.2) +
  annotate("text", x = Inf, y = Inf, 
           label = paste0("r = ", round(cor_slopes, 3), 
                         "\np ", ifelse(slope_test$p.value < 0.001, "< 0.001", 
                                       paste("=", round(slope_test$p.value, 3))),
                         "\nN = ", nrow(rb_slopes)),
           hjust = 1.1, vjust = 1.5, size = 5) +
  labs(
    title = "Relationship Between Valence and Cosine Distance Rubber-Band Dynamics",
    subtitle = paste0("Turn range: [", turn_min, ", ", 
                     ifelse(is.infinite(turn_max), "∞", turn_max), "]"),
    x = "Cosine Distance RB Slope (β per story)",
    y = "Valence Gap RB Slope (β per story)"
  ) +
  xlim(-2,1) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

print(p_rb_correlation)

ggsave(file.path(GRAPHS_DIR, "rubberband_valence_cosine_correlation.png"),
       p_rb_correlation, width = 10, height = 8, dpi = 300)

# 6. Regression: Does cosine RB predict valence RB?
lm_rb_predict <- lm(valence_rb_slope ~ cosine_rb_slope, data = rb_slopes)
cat("\n=== Regression: Cosine RB Predicting Valence RB ===\n")
print(summary(lm_rb_predict))

cat("\n=== VALIDATION CHECK ===\n")
cat("Correlation from cor():", round(cor_slopes, 3), "\n")
cat("Correlation from lm R²:", round(sqrt(summary(lm_rb_predict)$r.squared), 3), "\n")
cat("These should match (they do if correlation is positive).\n\n")
```


# 5. Supplementary: Interaction Effects

## 5.1 Starter Interaction Effects

### 5.1.1 Valence Gap by Starter
```{r starter-valence}
# Turn-level valence gap with starter interaction
lm_starter_valence <- lm(gap_change_next ~ gap_change * starter, data = rb_turns_field)

cat("=== Valence Rubber-Band: Starter Interaction ===\n")
print(summary(lm_starter_valence))
```

### 5.1.2 Cosine Distance by Starter
```{r starter-cosine}
# Turn-level cosine distance with starter interaction
lm_starter_cosine <- lm(dist_change_next ~ dist_change * starter, data = rb_cosine)

cat("=== Cosine Distance Rubber-Band: Starter Interaction ===\n")
print(summary(lm_starter_cosine))
```

### 5.1.3 Alignment (Cosine Similarity) by Starter
```{r starter-alignment}
# Cosine similarity over turns with starter interaction
df_alignment_starter <- df_field %>%
  filter(turn >= 2, turn <= 9, !is.na(cosine_sim_same), !is.na(starter))

lm_alignment_starter <- lm(cosine_sim_same ~ turn * starter, data = df_alignment_starter)

cat("=== Cosine Similarity (Alignment): Starter × Turn Interaction ===\n")
print(summary(lm_alignment_starter))
```


## 5.2 LLM Type Interaction Effects

### 5.2.1 Valence Gap by LLM Type
```{r llm-valence}
# Turn-level valence gap with LLM type interaction
rb_turns_llm <- rb_turns_field %>%
  filter(!is.na(llm_type))

lm_llm_valence <- lm(gap_change_next ~ gap_change * factor(llm_type), data = rb_turns_llm)

cat("=== Valence Rubber-Band: LLM Type Interaction ===\n")
print(summary(lm_llm_valence))
```

### 5.2.2 Cosine Distance by LLM Type
```{r llm-cosine}
# Turn-level cosine distance with LLM type interaction
rb_cosine_llm <- rb_cosine %>%
  filter(!is.na(llm_type))

lm_llm_cosine <- lmer(dist_change_next ~ dist_change * factor(llm_type) + (1 | conversation_id), data = rb_cosine_llm)

cat("=== Cosine Distance Rubber-Band: LLM Type Interaction ===\n")
print(summary(lm_llm_cosine))
```

### 5.2.3 Alignment (Cosine Similarity) by LLM Type
```{r llm-alignment}
# Cosine similarity over turns with LLM type interaction
df_alignment_llm <- df_field %>%
  filter(turn >= 1, turn <= 5, !is.na(cosine_sim_same), !is.na(llm_type))

lm_alignment_llm <- lmer(cosine_sim_same ~ turn * factor(llm_type) + (1 | conversation_id), data = df_alignment_llm)

cat("=== Cosine Similarity (Alignment): LLM Type × Turn Interaction ===\n")
print(summary(lm_alignment_llm))
```
