---
title: "Valence Alignment in Human-AI Story Co-Creation"
subtitle: "Analysis of Field Data and AI-AI Baseline Comparison"
author: "PENPAL Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  lme4,         # Mixed effects models
  lmerTest,     # P-values for mixed models
  arrow,        # Parquet file I/O
  stringi,      # String operations
  lubridate,    # Date/time handling
  zoo,          # Rolling statistics
  patchwork,    # Combine plots
  cowplot,      # Additional plot utilities
  afex          # ANOVA
)

# Set project root
PROJECT_ROOT <- "/Users/halfidaldal/Documents/PENPAL_analysis"
setwd(PROJECT_ROOT)

# Set global theme
theme_set(theme_minimal(base_size = 14))

# Define output directories
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")
TABLES_DIR <- file.path(PROJECT_ROOT, "analysis", "tables")

# Create output directories if they don't exist
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(TABLES_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
# Load dyadic sentiment scores
setwd("/Users/halfidaldal/Documents/PENPAL_analysis")
df_field <- read_parquet("data/TEXT/processed/dyadic_sentiment_scores.parquet")
df_simulated <- read_parquet("data/Berlin/processed/dyadic_sentiment_scores_simulated.parquet")

# Display structure
cat("Field data dimensions:", dim(df_field), "\n")
cat("Simulated data dimensions:", dim(df_simulated), "\n")

# Show column names
cat("\nField data columns:\n")
print(names(df_field))

cat("\nSimulated data columns:\n")
print(names(df_simulated))
```

## 1.2 Data Preprocessing

```{r preprocess}
# Preprocess field data
df_field <- df_field %>%
  rename(
    ai_valence = ai_sentiment_score,
    user_valence = user_sentiment_score
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing
    timestamp = ymd_hms(timestamp),
    
    # Data source label
    data_source = "Field"
  ) %>%
  # Ensure conversation_id is a character/factor
  mutate(conversation_id = as.character(conversation_id))

# Preprocess simulated data
df_simulated <- df_simulated %>%
  rename(
    ai_valence = ai_sentiment_score,
    user_valence = user_sentiment_score
  ) %>%
  mutate(
    # Token counts
    user_token_count = stri_count_words(user),
    ai_token_count = stri_count_words(ai),
    token_ratio = if_else(ai_token_count == 0, NA_real_, 
                         user_token_count / ai_token_count),
    
    # Time parsing (if available)
    timestamp = if_else(!is.na(timestamp), ymd_hms(timestamp), as.POSIXct(NA)),
    
    # Data source label
    data_source = "Simulated"
  ) %>%
  mutate(conversation_id = as.character(conversation_id))

# Summary statistics
cat("\n=== Field Data Summary ===\n")
df_field %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()

cat("\n=== Simulated Data Summary ===\n")
df_simulated %>%
  summarise(
    n_conversations = n_distinct(conversation_id),
    n_turns = n(),
    mean_turns_per_story = n() / n_distinct(conversation_id),
    mean_user_valence = mean(user_valence, na.rm = TRUE),
    mean_ai_valence = mean(ai_valence, na.rm = TRUE)
  ) %>%
  print()
```

## 1.3 Compute Lagged Variables and Cosine

```{r compute-lags}
# Add lagged AI valence for field data
# Note: lag() within group_by ensures no cross-story contamination
# Turn 1 of each story will have NA for lagged values (correct behavior)

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1),
  ) %>%
  ungroup()


cosine_sim <- function(x, y) {
  if (is.null(x) || is.null(y) || any(is.na(x)) || any(is.na(y))) {
    return(NA_real_)
  }
  return(sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2))))
}

df_field <- df_field %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_embedding_prev = lag(ai_embedding)
  ) %>%
  rowwise() %>%
  mutate(
    cosine_sim_same = cosine_sim(user_embedding, ai_embedding),
    cosine_sim_prev = cosine_sim(user_embedding, ai_embedding)
  ) %>%
  ungroup()

# Add lagged AI valence for simulated data
df_simulated <- df_simulated %>%
  group_by(conversation_id) %>%
  arrange(turn, .by_group = TRUE) %>%
  mutate(
    ai_valence_lag1 = lag(ai_valence, n = 1),
    user_valence_lag1 = lag(user_valence, n = 1)
  ) %>%
  ungroup()

# Verify no cross-story contamination
cat("=== Lag Variable Verification ===\n")
cat("Field data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_field$turn == 1 & !is.na(df_field$ai_valence_lag1)), "\n")
cat("Simulated data - Turn 1 with non-NA lag (should be 0):", 
    sum(df_simulated$turn == 1 & !is.na(df_simulated$ai_valence_lag1)), "\n")
```

# 2. Descriptive Statistics

## 2.1 Token Count Statistics

```{r token-stats}
# Field data token statistics
token_stats_field <- df_field %>%
  summarise(
    mean_user_tokens = mean(user_token_count, na.rm = TRUE),
    sd_user_tokens = sd(user_token_count, na.rm = TRUE),
    mean_ai_tokens = mean(ai_token_count, na.rm = TRUE),
    sd_ai_tokens = sd(ai_token_count, na.rm = TRUE),
    mean_token_ratio = mean(token_ratio, na.rm = TRUE)
  )

cat("=== Field Data Token Statistics ===\n")
print(token_stats_field)

# Simulated data token statistics
token_stats_sim <- df_simulated %>%
  summarise(
    mean_user_tokens = mean(user_token_count, na.rm = TRUE),
    sd_user_tokens = sd(user_token_count, na.rm = TRUE),
    mean_ai_tokens = mean(ai_token_count, na.rm = TRUE),
    sd_ai_tokens = sd(ai_token_count, na.rm = TRUE),
    mean_token_ratio = mean(token_ratio, na.rm = TRUE)
  )

cat("\n=== Simulated Data Token Statistics ===\n")
print(token_stats_sim)
```

## 2.2 Valence Distribution Statistics

```{r valence-stats}
# Combined statistics
valence_stats <- bind_rows(
  df_field %>% 
    pivot_longer(cols = c(user_valence, ai_valence),
                 names_to = "agent", values_to = "valence") %>%
    mutate(agent = str_remove(agent, "_valence")) %>%
    group_by(data_source, agent) %>%
    summarise(
      mean_valence = mean(valence, na.rm = TRUE),
      sd_valence = sd(valence, na.rm = TRUE),
      median_valence = median(valence, na.rm = TRUE),
      .groups = "drop"
    ),
  
  df_simulated %>%
    pivot_longer(cols = c(user_valence, ai_valence),
                 names_to = "agent", values_to = "valence") %>%
    mutate(agent = str_remove(agent, "_valence")) %>%
    group_by(data_source, agent) %>%
    summarise(
      mean_valence = mean(valence, na.rm = TRUE),
      sd_valence = sd(valence, na.rm = TRUE),
      median_valence = median(valence, na.rm = TRUE),
      .groups = "drop"
    )
)

print(valence_stats)

# Save to CSV
write_csv(valence_stats, file.path(TABLES_DIR, "valence_descriptive_stats.csv"))
```

# 3. Valence Distribution Analysis

## 3.1 Combined Valence Distribution Plot

```{r valence-distribution, fig.width=12, fig.height=6}
# Prepare data for plotting
df_field_long <- df_field %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  mutate(agent = recode(agent, 
                       user_valence = "User", 
                       ai_valence = "AI"))

df_sim_long <- df_simulated %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  mutate(agent = recode(agent, 
                       user_valence = "AI 1", 
                       ai_valence = "AI 2"))

# Calculate means for vertical lines
means_field <- df_field_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")

means_sim <- df_sim_long %>%
  group_by(agent) %>%
  summarise(mean_val = mean(valence, na.rm = TRUE), .groups = "drop")

# Field data plot
p_field <- ggplot(df_field_long, aes(x = valence, fill = agent)) +
  geom_density(alpha = 0.4, color = NA) +
  geom_vline(data = means_field,
             aes(xintercept = mean_val, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set1", name = "") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Field Data", x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Simulated data plot
p_sim <- ggplot(df_sim_long, aes(x = valence, fill = agent)) +
  geom_density(alpha = 0.4, color = NA) +
  geom_vline(data = means_sim,
             aes(xintercept = mean_val, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2", name = "") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Simulated Data", x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text = element_text(size = 12)
  )

# Combine plots
combined_plot <- (p_field | p_sim) +
  plot_annotation(
    title = "Distribution of Valence Scores: Field vs Simulated Data",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 20, face = "bold"))
  )

# Add axis labels using cowplot
final_valence_dist <- ggdraw() +
  draw_plot(combined_plot, 0, 0, 1, 1) +
  draw_label("Valence Score", x = 0.5, y = 0.02, size = 16) +
  draw_label("Density", x = 0.02, y = 0.5, angle = 90, size = 16)

print(final_valence_dist)

ggsave(file.path(GRAPHS_DIR, "valence_distribution_comparison.png"), 
       final_valence_dist, width = 12, height = 6, dpi = 300)
```

# 4. Valence Alignment Analysis

## 4.1 Per-Story Correlation (Same Turn)

```{r alignment-same-turn}
# Field data: same-turn correlation
cor_field_same <- df_field %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%  # Require at least 3 turns
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )

# Simulated data: same-turn correlation
cor_sim_same <- df_simulated %>%
  group_by(conversation_id) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated",
    .groups = "drop"
  )

# Combine
cor_same_combined <- bind_rows(cor_field_same, cor_sim_same) %>%
  mutate(alignment_type = "Same Turn")

# Summary
cat("=== Same-Turn Alignment Summary ===\n")
cor_same_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

```

## 4.2 Per-Story Correlation (Previous Turn)

```{r alignment-prev-turn}
# Field data: previous-turn correlation (User response to previous AI)
cor_field_prev <- df_field %>%
  group_by(conversation_id) %>%
  filter(!is.na(ai_valence_lag1)) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field",
    .groups = "drop"
  )

# Simulated data: previous-turn correlation
cor_sim_prev <- df_simulated %>%
  group_by(conversation_id) %>%
  filter(!is.na(ai_valence_lag1)) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated",
    .groups = "drop"
  )

# Combine
cor_prev_combined <- bind_rows(cor_field_prev, cor_sim_prev) %>%
  mutate(alignment_type = "Previous Turn")

# Summary
cat("=== Previous-Turn Alignment Summary ===\n")
cor_prev_combined %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```

## 4.3.1 All turns story correlation (same turn)
```{r alignment across all turns}
#Violates assumptions since turns are not independent (nested in stories)
cor_field_same_all_turns <- df_field %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Field"
  )

cor_sim_same_all_turns <- df_simulated %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    data_source = "Simulated"
  )

cor_same_combined_all_turns <- bind_rows(cor_field_same_all_turns, cor_sim_same_all_turns) %>%
  mutate(alignment_tupe = "Same Turn All Turns")

cat("=== Same-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()
```
## 4.3.2 All turns story correlation (Previous Turn)

```{r previous-turn-alignment-all-turns}
# Violates independence assumption (turns nested in stories)
# Note: use="complete.obs" automatically excludes turn 1 (where ai_valence_lag1 is NA)
# This ensures user_t1 is never paired with ai from previous story

# Verify exclusion of turn 1
cat("=== Previous-Turn Aggregate Correlation ===\n")
cat("Field: Turn 1 observations (excluded):", sum(df_field$turn == 1), "\n")
cat("Field: Valid lag observations:", sum(!is.na(df_field$ai_valence_lag1)), "\n")
cat("Simulated: Turn 1 observations (excluded):", sum(df_simulated$turn == 1), "\n")
cat("Simulated: Valid lag observations:", sum(!is.na(df_simulated$ai_valence_lag1)), "\n\n")

cor_field_same_all_turns_prev <- df_field %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    n_valid = sum(!is.na(ai_valence_lag1)),
    data_source = "Field"
  )

cor_sim_same_all_turns_prev <- df_simulated %>%
  summarise(
    rho = cor(user_valence, ai_valence_lag1, use = "complete.obs"),
    n_turns = n(),
    n_valid = sum(!is.na(ai_valence_lag1)),
    data_source = "Simulated"
  )

cor_same_combined_all_turns_prev <- bind_rows(cor_field_same_all_turns_prev, cor_sim_same_all_turns_prev) %>%
  mutate(alignment_type = "Previous Turn All Turns")

cat("=== Previous-Turn All Turns Alignment Summary ===\n")
cor_same_combined_all_turns_prev %>%
  group_by(data_source) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    median_rho = median(rho, na.rm = TRUE),
    n_total = sum(n_turns),
    n_valid = sum(n_valid)
  ) %>%
  print()



```
## 4.3.3 Probe Sentiment Narrative Development

```{r alignment-probe-example, fig.width=12, fig.height=6}


df_field %>%
  mutate(time = ymd_hms(timestamp)) %>%
  
  group_by(conversation_id) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(turn_index = row_number()) %>%
  ungroup() %>%
  filter(conversation_id %in% unique(conversation_id)[20:25]) %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "source", values_to = "valence") %>%
  mutate(valence_smoothed = rollmean(valence, k = 1, fill = NA, align = "center")) %>%
  ggplot(aes(x = turn_index, y = valence_smoothed, color = source)) +
  geom_line() +
  facet_wrap(~conversation_id, scales = "free_x") +
  labs(
    title = "Rolling Average of Valence (k = 8) Human-AI",
    x = "Turn Index",
    y = "Smoothed Valence Score",
    color = "Source"
  ) +
  theme_minimal()

```

## 4.4.1 Alignment Comparison Plot

```{r alignment-comparison-plot, fig.width=12, fig.height=6}
# Combine all correlations
cor_all <- bind_rows(cor_same_combined, cor_prev_combined) %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("Same Turn", "Previous Turn"))
  )

# Boxplot with jittered points
p_alignment <- ggplot(cor_all, aes(x = alignment_type, y = rho, fill = data_source)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.7,
               position = position_dodge(width = 0.75)) +
  geom_jitter(aes(color = data_source),
              position = position_jitterdodge(
                dodge.width = 0.75,
                jitter.width = 0.15
              ),
              size = 1.5, alpha = 0.4) +
  scale_fill_brewer(palette = "Set2", name = "Data Source") +
  scale_color_brewer(palette = "Set2", name = "Data Source") +
  labs(
    title = "Valence Alignment: Field vs Simulated Data",
    x = "Alignment Type",
    y = "Pearson ρ (per story)"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_alignment)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison.png"),
       p_alignment, width = 12, height = 6, dpi = 300)
```

## 4.4.2 Comparison: Per-Story vs Aggregate Correlations

```{r alignment-comparison-aggregate, fig.width=12, fig.height=8}
# Prepare aggregate correlation data for overlay
agg_same <- cor_same_combined_all_turns %>%
  mutate(alignment_type = "Same Turn")

agg_prev <- cor_same_combined_all_turns_prev %>%
  select(data_source, rho, alignment_type)

agg_data <- bind_rows(agg_same, agg_prev) %>%
  mutate(
    alignment_type = factor(alignment_type, 
                           levels = c("Same Turn", "Previous Turn"))
  )

# Create plot with per-story boxplots and aggregate overlay
p_comparison <- ggplot(cor_all, aes(x = alignment_type, y = rho, fill = data_source)) +
  # Per-story boxplots
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.5,
               position = position_dodge(width = 0.75)) +
  # Per-story jittered points
  geom_jitter(aes(color = data_source),
              position = position_jitterdodge(
                dodge.width = 0.75,
                jitter.width = 0.15
              ),
              size = 1.5, alpha = 0.3) +
  # Aggregate correlations as large points
  geom_point(data = agg_data, 
             aes(x = alignment_type, y = rho, fill = data_source),
             position = position_dodge(width = 0.75),
             size = 6, shape = 23, color = "black", stroke = 1.5) +
  scale_fill_brewer(palette = "Set2", name = "Data Source") +
  scale_color_brewer(palette = "Set2", name = "Data Source") +
  labs(
    title = "Valence Alignment: Per-Story vs Aggregate Correlations",
    subtitle = "Boxplots = per-story correlations; Diamonds = aggregate (all turns pooled)",
    x = "Alignment Type",
    y = "Pearson ρ"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "grey40"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12)
  )

print(p_comparison)

ggsave(file.path(GRAPHS_DIR, "valence_alignment_comparison_with_aggregate.png"),
       p_comparison, width = 12, height = 8, dpi = 300)

# Print summary comparison
cat("\n=== Per-Story vs Aggregate Correlation Comparison ===\n")
comparison_summary <- bind_rows(
  cor_all %>%
    group_by(data_source, alignment_type) %>%
    summarise(
      mean_rho = mean(rho, na.rm = TRUE),
      method = "Per-Story Mean",
      .groups = "drop"
    ),
  agg_data %>%
    mutate(mean_rho = rho, method = "Aggregate All Turns") %>%
    select(data_source, alignment_type, mean_rho, method)
) %>%
  pivot_wider(names_from = method, values_from = mean_rho)

print(comparison_summary)

```


## 4.5 Mixed Effects Model: Random Intercepts per Story

```{r mixed-effects-model}
# Same-turn alignment with random intercepts per story
model_same <- lmer(ai_valence ~ user_valence + (1 | conversation_id), data = df_field)
cat("=== Mixed Effects Model: Same-Turn Alignment (Field Data) ===\n")
summary(model_same)

# Previous-turn alignment with random intercepts per story
model_prev <- lmer(user_valence ~ ai_valence_lag1 + (1 | conversation_id), data = df_field)
cat("\n=== Mixed Effects Model: Previous-Turn Alignment (Field Data) ===\n")
summary(model_prev)
```

## 4.6 Statistical Tests: Alignment Differences

```{r alignment-stats}
# Test 1: Same-turn alignment difference
t_test_same <- t.test(rho ~ data_source, data = cor_same_combined)
cat("=== T-test: Same-Turn Alignment ===\n")
print(t_test_same)

# Test 2: Previous-turn alignment difference
t_test_prev <- t.test(rho ~ data_source, data = cor_prev_combined)
cat("\n=== T-test: Previous-Turn Alignment ===\n")
print(t_test_prev)

# Test 3: One-sample t-tests against zero (is alignment significant?)
cat("\n=== One-Sample Tests Against Zero ===\n")
cor_all %>%
  group_by(data_source, alignment_type) %>%
  summarise(
    t_stat = t.test(rho, mu = 0)$statistic,
    p_value = t.test(rho, mu = 0)$p.value,
    mean_rho = mean(rho, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  print()

# Save results
alignment_test_results <- tibble(
  test = c("Same-turn: Field vs Simulated", "Previous-turn: Field vs Simulated"),
  t_statistic = c(t_test_same$statistic, t_test_prev$statistic),
  p_value = c(t_test_same$p.value, t_test_prev$p.value),
  mean_field = c(mean(cor_field_same$rho, na.rm = TRUE), 
                 mean(cor_field_prev$rho, na.rm = TRUE)),
  mean_sim = c(mean(cor_sim_same$rho, na.rm = TRUE), 
               mean(cor_sim_prev$rho, na.rm = TRUE))
)

write_csv(alignment_test_results, 
          file.path(TABLES_DIR, "alignment_statistical_tests.csv"))
```


# 5. Summary and Conclusions

## 5.1 Key Findings Summary

```{r summary-table}
# Create summary table
summary_results <- tribble(
  ~Metric, ~Field_Data, ~Simulated_Data,
  "N Stories", n_distinct(df_field$conversation_id), n_distinct(df_simulated$conversation_id),
  "Mean User Valence", mean(df_field$user_valence, na.rm = TRUE), mean(df_simulated$user_valence, na.rm = TRUE),
  "Mean AI Valence", mean(df_field$ai_valence, na.rm = TRUE), mean(df_simulated$ai_valence, na.rm = TRUE),
  "Same-Turn Alignment (ρ)", mean(cor_field_same$rho, na.rm = TRUE), mean(cor_sim_same$rho, na.rm = TRUE),
  "Previous-Turn Alignment (ρ)", mean(cor_field_prev$rho, na.rm = TRUE), mean(cor_sim_prev$rho, na.rm = TRUE)
) %>%
  mutate(
    Field_Data = round(Field_Data, 3),
    Simulated_Data = round(Simulated_Data, 3)
  )

print(summary_results)

write_csv(summary_results, file.path(TABLES_DIR, "analysis_summary.csv"))
```


# 6. Supplementary: Interaction Effects

## 6.1 Starter Interaction Effects

### 6.1.1 Same-Turn Alignment by Starter
```{r starter-same-turn}
# Mixed model with starter interaction
model_starter_same <- lmer(ai_valence ~ user_valence * starter + (1 | conversation_id), 
                           data = df_field)

cat("=== Same-Turn Alignment: Starter Interaction ===\n")
print(summary(model_starter_same))
```

### 6.1.2 Previous-Turn Alignment by Starter
```{r starter-prev-turn}
# Mixed model with starter interaction
df_field_lag <- df_field %>% filter(!is.na(ai_valence_lag1))

model_starter_prev <- lmer(user_valence ~ ai_valence_lag1 * starter + (1 | conversation_id), 
                           data = df_field_lag)

cat("=== Previous-Turn Alignment: Starter Interaction ===\n")
print(summary(model_starter_prev))
```

### 6.1.3 Per-Story Correlations by Starter
```{r starter-per-story}
# Same-turn correlations stratified by starter
cor_starter_same <- df_field %>%
  filter(!is.na(starter)) %>%
  group_by(conversation_id, starter) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    .groups = "drop"
  )

cat("=== Same-Turn Correlations by Starter ===\n")
cor_starter_same %>%
  group_by(starter) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

# Test difference
t_test_starter <- t.test(rho ~ starter, data = cor_starter_same)
cat("\nT-test (starter effect on same-turn rho):\n")
print(t_test_starter)
```


## 6.2 LLM Type Interaction Effects

### 6.2.1 Same-Turn Alignment by LLM Type
```{r llm-same-turn}
# Mixed model with LLM type interaction
df_field_llm <- df_field %>% filter(!is.na(llm_type))

model_llm_same <- lmer(ai_valence ~ user_valence * factor(llm_type) + (1 | conversation_id), 
                       data = df_field_llm)

cat("=== Same-Turn Alignment: LLM Type Interaction ===\n")
print(summary(model_llm_same))
```

### 6.2.2 Previous-Turn Alignment by LLM Type
```{r llm-prev-turn}
# Mixed model with LLM type interaction
df_field_llm_lag <- df_field %>% 
  filter(!is.na(llm_type), !is.na(ai_valence_lag1))

model_llm_prev <- lmer(user_valence ~ ai_valence_lag1 * factor(llm_type) + (1 | conversation_id), 
                       data = df_field_llm_lag)

cat("=== Previous-Turn Alignment: LLM Type Interaction ===\n")
print(summary(model_llm_prev))
```

### 6.2.3 Per-Story Correlations by LLM Type
```{r llm-per-story}
# Same-turn correlations stratified by LLM type
cor_llm_same <- df_field %>%
  filter(!is.na(llm_type)) %>%
  group_by(conversation_id, llm_type) %>%
  filter(n() >= 3) %>%
  summarise(
    rho = cor(user_valence, ai_valence, use = "complete.obs"),
    n_turns = n(),
    .groups = "drop"
  )

cat("=== Same-Turn Correlations by LLM Type ===\n")
cor_llm_same %>%
  group_by(llm_type) %>%
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho = sd(rho, na.rm = TRUE),
    n_stories = n()
  ) %>%
  print()

# ANOVA test
anova_llm <- aov(rho ~ factor(llm_type), data = cor_llm_same)
cat("\nANOVA (LLM type effect on same-turn rho):\n")
print(summary(anova_llm))
```


## 6.3 Valence Distribution by Conditions

### 6.3.1 Valence by Starter
```{r valence-starter}
df_field %>%
  filter(!is.na(starter)) %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  group_by(starter, agent) %>%
  summarise(
    mean_valence = mean(valence, na.rm = TRUE),
    sd_valence = sd(valence, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

### 6.3.2 Valence by LLM Type
```{r valence-llm}
df_field %>%
  filter(!is.na(llm_type)) %>%
  pivot_longer(cols = c(user_valence, ai_valence),
               names_to = "agent", values_to = "valence") %>%
  group_by(llm_type, agent) %>%
  summarise(
    mean_valence = mean(valence, na.rm = TRUE),
    sd_valence = sd(valence, na.rm = TRUE),
    n_turns = n(),
    .groups = "drop"
  ) %>%
  print()
```


## 7. Session Information

```{r session-info}
sessionInfo()
```

------------------------------------------------------------------------

**Analysis complete.** All figures saved to `analysis/figures/` and tables to `analysis/tables/`.
